### 2016-01-13 doc/manual.tex
Copied DNest3 manual over

% Every Latex document starts with a documentclass command
\documentclass[a4paper, 11pt]{article}

% Load some packages
\usepackage{graphicx} % This allows you to put figures in
\usepackage{natbib}   % This allows for relatively pain-free reference lists
\usepackage[left=2cm,top=3cm,right=2cm]{geometry} % The way I like the margins
\usepackage{framed}

\newcommand{\dnest}{{\tt DNest4}}

% This helps with figure placement
\renewcommand{\topfraction}{0.85}
\renewcommand{\textfraction}{0.1}
\parindent=0cm

% Set values so you can have a title
\title{\dnest~Manual}
\author{Brendon J. Brewer}
\date{\today}

% Document starts here
\begin{document}

% Actually put the title in
\maketitle

\section{Introduction}

\dnest~is a multi-threaded C++ implementation of Diffusive Nested Sampling
(DNS), a powerful Markov Chain Monte Carlo (MCMC) algorithm that is mainly
useful for solving Bayesian Inference problems.
\dnest~is free software,
released under the terms of the GNU General Public Licence, version 3.
For information about how the
algorithm works, please see the paper, freely available from the following
URL:\\

\begin{center}
{\tt http://arxiv.org/abs/0912.238}\\
\end{center}

If you find \dnest~useful, please feel free to cite the paper.
Throughout this manual I will assume that you have read the paper and have a
reasonable understanding of how the algorithm works, as well as some
understanding of Nested Sampling in general, as introduced by John Skilling.\\

\section{Installation}
Please note that I have only ever compiled \dnest~on GNU/Linux,
using the GNU C++ compiler ({\tt g++}). However, I expect it to be
straightforward to compile it on any other Unix-like operating system such as Mac
OS X or FreeBSD. It should be possible, albeit probably more tricky,
to compile \dnest~in Microsoft Windows. If you want to try,
the MinGW compiler is
probably your best bet. I'd be interested to hear from anyone who
has tried this. I would also be interested to hear if anyone has
compiled \dnest~with a different compiler, such as the Intel C++
compiler.\\

\subsection{Dependencies}
\dnest~depends on some other software that you will need to have installed on
your system. These should be quite easy to obtain. I highly
recommend obtaining these programs from your operating system's package manager,
rather than installing them from the source code. Note that some OSs split up
packages into binaries (the library files) and ``development'' packages,
containing header files (part of a library's source code).
In Ubuntu, Linux Mint, and Debian, the development packages have a -dev suffix.
In Fedora, the development packages usually have a -devel suffix (e.g. Fedora).
To compile \dnest, you will need both the libraries and the development packages
for each of the dependencies. Here's a list of dependencies:\\

\begin{itemize}
\item The GNU Scientific Library (GSL)
\item Python 2, along with the NumPy and matplotlib packages
\item The {\tt zlib} library (this is only required if wanting the ability to
output samples to a {\tt gzipped} file, otherwise it can be excluded)
\item The {\tt thread}, {\tt system} and {\tt iostreams} parts of the Boost C++ library.
It may be easier to just install all of Boost.
\item To compile \dnest, you need {\tt cmake}.\\
\end{itemize}

On Ubuntu, I can install these packages with the following commands:
\begin{verbatim}
sudo apt-get install libgsl0-dev
sudo apt-get install python-numpy python-matplotlib
sudo apt-get install zlib1g-dev
sudo apt-get install libboost-thread-dev libboost-system-dev libboost-iostreams-dev
sudo apt-get install cmake
\end{verbatim}

\dnest~uses GSL's random number generator. Python, NumPy and
matplotlib are used for the postprocessing scripts and for
plotting. Boost is used for parallelization (multithreading).
While random number generation and
multithreading is supported
in the recent C++11 standard, I have not ported \dnest~to C++11 and have no
intention of doing so in the near future.\\

It is also possible to compile \dnest~without Boost, however this is not
recommended, and you will not be able to use any of the multi-threaded options
or the {\tt gzip} option.\\

\subsection{Compiling}
Please follow the instructions in README.md to compile \dnest~using cmake.
This is the recommended option.\\

%Go into the \dnest~directory and type {\tt make}. Everything should compile.
To
verify that the compilation was successful,
go into the {\tt Examples/SpikeSlab} directory
and see if an executable file called {\tt main} has been created. If you
execute this file, you should see a lot of output about ``levels'' and
``particles'' (see Figure~\ref{fig:output} for example output).
You will need to manually terminate this process by using
Control+C, or whatever works in your operating system. If you saw this output,
\dnest~was compiled successfully.\\

\section{Running DNest}
As part of the compilation process, several examples will have been compiled.
There is one executable file for each example: \dnest~does not have a global
executable. The
following discussion will involve running the ``SpikeSlab'' example, which is
the example used in the paper.\\

To run the SpikeSlab example, go into the {\tt Examples/Spikeslab}
directory and execute {\tt main}. You should see some output that looks like
that shown in Figure~\ref{fig:output}.
Note that you must terminate the process manually by pressing Ctrl+C
or in another way. If you do not terminate the process yourself, it will run
forever. There is no definite rule for how long you should run \dnest.
It depends on the difficulty of your problem, the number of posterior samples you
want, and the accuracy you want for the marginal likelihood.\\

\begin{figure}[h!]
\begin{framed}
\begin{verbatim}
$ ./main
# Using 1 thread.
# Target compression factor between levels = 2.7182818284590451
# Seeding random number generator with 1402801184.
# Thread 1: Generated 1 particles from the prior.
# Creating level 1 with logL = -51.65058877.
# Creating level 2 with logL = -37.67550971.
# Creating level 3 with logL = -28.69668897.
# Creating level 4 with logL = -21.61338509.
# Creating level 5 with logL = -15.92331209.
# Creating level 6 with logL = -11.291182.
# Saving a particle to disk. N = 1.
# Creating level 7 with logL = -7.301954778.
# Creating level 8 with logL = -4.116771404.
# Creating level 9 with logL = -1.19393518.
# Creating level 10 with logL = 1.620392784.
# Saving a particle to disk. N = 2.
\end{verbatim}
\end{framed}
\caption{\it An example of \dnest~output that is shown on the screen. More
detailed information is saved to the three output files
{\tt sample.txt}, {\tt sample\_info.txt}, and {\tt levels.txt}.
\label{fig:output}}
\end{figure}

The executable {\tt main} is responsible for the exploration part of the
DNS algorithm (i.e. running the MCMC chain, building
levels, and then exploring all the levels). It creates three output text files,
{\tt sample.txt}, {\tt sample\_info.txt}, and {\tt levels.txt}.\\

The first output
file, {\tt sample.txt}, contains a sampling of parameter values that
represents the {\it mixture of constrained priors} (the target distribution
used in DNS), {\bf not} the
posterior distribution. Each line of {\tt sample.txt} represents a point in
parameter space. In the SpikeSlab example, there are 20 parameters, so there
will be 20 columns in {\tt sample.txt}.
Each time a point is saved to {\tt sample.txt}, \dnest~prints
the message ``Saving a particle to disk. N = ...''.\\

The second output file, {\tt sample\_info.txt}, should have the same number of
rows as {\tt sample.txt}, because it contains metadata about the samples in
{\tt sample.txt}. The first
column is the index $j$, which tells us which ``level'' the particle was in
when it was saved. Level 0 represents the prior, and higher levels represent
more constrained versions of the prior.
The second column is the log-likelihood value, and the third column is
the likelihood ``tiebreaker'', which allows Nested Sampling to work when
there is a region in parameter space with nonzero prior probability where the
likelihood is constant. The final column tells us which thread the particle
belonged to: when you use \dnest~in multithreaded mode, each thread
is responsible for evolving one or more walkers.\\

The third output file, {\tt levels.txt}, contains information about the levels
that were built during the run. The first column has estimates of the $\log(X)$
values of the levels, i.e. how compressed they are relative to the prior, in
units of nats. The second column contains the log likelihoods of the levels.
The first level, with a $\log(X)$ value of 0 and a log likelihood of
$-10^{300}$ (basically ``minus infinity''), is simply the prior. The third
column has the ``tiebreaker'' values for the levels, which again are not
particularly useful unless your problem has likelihood plateaus. The fourth
and fifth columns are the number of accepted proposals and the total number
of proposals that have occurred within each level, which are useful for
monitoring the Metropolis acceptance ratio as a function of level.
The final two columns, called ``exceeds'', and ``visits'', are used to refine
the estimates of the level compressions (and hence the $\log(X)$ values of
the levels in column 1), as discussed in Section 3 of the
paper. The visits column counts the number of times a level (level $j$, say)
has been visited, but only starts counting after the next level ($j+1$) has been created. The exceeds column counts the number of times a particle that was
in level $j$ had a likelihood that exceeded that of level $j+1$.
\\

\section{Postprocessing with showresults.py}
After {\tt main} has been running for
a while, you should run the postprocessing script
by executing {\tt python showresults.py}. This script loads the output files,
plots some graphs, and creates two new output files, {\tt weights.txt} and
{\tt posterior\_samples.txt}. It also prints the current estimate of the
(log of the) marginal likelihood or evidence value (traditionally considered
the main goal of Nested Sampling), as well as the information value (the KL
divergence from the prior to the posterior).\\

The plots made by {\tt showresults.py} are very similar to the plots in
the DNS paper. The first plot shows the level of the saved particles vs.
time, the equivalent of Figure 2 in the paper. The second plot produced by
{\tt showresults.py} shows the estimated compression values of the levels,
like in Figure 3 of the paper, along with the Metropolis acceptance fraction
as a function of the level (as you would expect, this usually shows a
decreasing trend). The red horizontal line on the level compression plot is
at -1, and all of the actual compressions should be close to -1 if everything
is working well (unless you've changed the target compression from $e$ to
another value as discussed in Section~\ref{sec:commandline}).

The final plot corresponds to Figure 4 from the paper, and
shows the shape of the likelihood curve along with the posterior weights of
the saved particles.\\

The plots are useful for diagnosing the progress of your run. For example, you
want to ensure that the Metropolis acceptance fraction doesn't drop to near
zero. If it does, the proposal distributions might be too aggressive, or maybe
you just created too many levels (and the high level distributions are just
extremely narrow). The other key plot for diagnostics is the posterior
weights. On most problems, these will peak at some point, and begin to
come back down towards zero (See Figure~\ref{fig3}).
Once this occurs, you probably have enough levels.
It is wasteful to create large amounts of levels extending well beyond the peak.
However, on problems with phase transitions more than one peak may occur.
An example of not having enough levels is given in
Figure~\ref{fig:not_enough_levels}. To change the number of levels, you
can use the OPTIONS file, described in Section~\ref{sec:options}.\\

\begin{figure}
\begin{center}
\includegraphics[scale=0.5]{fig1.pdf}
\caption{\it Level vs. Time from a DNest3 run with 100 levels.\label{fig:fig1}}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[scale=0.5]{fig2.pdf}
\caption{\it Top panel: Estimated compressions between adjacent levels. Ideally, these should equal -1. Bottom panel: Metropolis-Hastings acceptance ratio
as a function of level.\label{fig:fig2}}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[scale=0.5]{fig3.pdf}
\caption{\it Top panel: Likelihood as a function of enclosed prior mass,
the standard Nested Sampling plot. Bottom panel: Posterior weights as a
function of enclosed prior mass. If this peaks at the left of the plot,
i.e. you can't see an entire peak, you need more levels.\label{fig:fig3}}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[scale=0.5]{not_enough_levels.pdf}
\caption{\it Top panel: Example of a DNest run where the number of levels
was too low. The posterior weights peak at the left of the plot, and these
would have continued to increase if we'd gone further.
\label{fig:not_enough_levels}}
\end{center}
\end{figure}

\section{Command line options}\label{sec:commandline}
\dnest~executables allow for several command line options. A couple of these
options are quite important, while others are much less useful.
You can view the available options by executing
{\tt ./main -h}. Here is the list:\\

\begin{figure}
\begin{framed}
\begin{verbatim}
DNest3 Command Line Options:
-h: display this message
-l <filename>: load level structure from the specified file.
-o <filename>: load DNest3 options from the specified file. Default=OPTIONS
-s <seed>: seed the random number generator with the specified value.
           If unspecified, the system time is used.
-d <filename>: Load data from the specified file, if required.
-c <value>: Specify a compression value (between levels) other than e.
-t <num_threads>: run on the specified number of threads. Default=1.
-f <filename>: a custom configuration file for adding problem specific
           options if required.
-z: the output "sample.txt" file will be gzipped and have the suffix ".gz".
\end{verbatim}
\end{framed}
\end{figure}

The most important option is the one about the number of threads. Depending on
your computer, you probably want to run \dnest~with more than one thread.
Each thread will be responsible for evolving its own walker(s), and periodically
the threads will pool their results (for example, information about the
levels). To execute DNest with 8 threads, simply use {\tt ./main -t 8}. You can
replace 8 with however many threads you want. I usually use 8 threads because
that's how many my computer can run simultaneously at full speed.\\

In many Monte Carlo studies, it is useful to be able to set the random number
seed, for debugging and reproducibility reasons. By default \dnest~seeds the
random number generator with the system time, and displays the value of the
seed in the output (you can see this in the example output displayed above).
If you want to set the seed to
something else, use the {\tt -s} option and specify a non-negative integer for
the seed. For example, to execute \dnest~with
8 threads and a seed of 42, use {\tt ./main -t 8 -s 42}.
When running with multiple threads, \dnest~uses a separate generator for each
thread, and ensures that they are seeded with different values.\\

The {\tt -z} option is only available is you have compiled with the {\tt zlib}
library and Boost {\tt iostreams} library installed. \\

If you have any additional configuration information required by the run, e.g.\
for setting values needed for a particular model that you don't want to hardcode,
then they can be added through specifying a configuration file using the {\tt -f}
option. You will have to write your own function to read in this file, so it can
be in whatever format you are comfortable with (e.g.\ an {\tt ini} file, or {\tt
json} file).\\

\section{The OPTIONS file}\label{sec:options}
Here is an example OPTIONS file.
\begin{framed}
\begin{verbatim}
1       # Number of particles
10000   # new level interval
100000  # save interval
200     # threadSteps
100     # maximum number of levels
10      # Backtracking scale length (lambda in the paper)
10      # Strength of effect to force histogram to equal push. (beta in the paper)
0       # Maximum number of saves (0 = infinite)
\end{verbatim}
\end{framed}

The first option is the number of particles {\it per thread}. If you run
\dnest~with 8 threads, there will actually be 8 particles. For most
applications, 1 particle per thread is sufficient. If you use more particles
per thread, the same amount of CPU time will be spent evolving more particles,
so each one will not be evolved as far. On most problems, one particle per
thread is fine. I would only use more particles per thread when working on
a very hard problem where it is difficult for the particles to find the
important parts of parameter space.\\

The new level interval controls how quickly \dnest~creates new levels. In this
example, this is set to 10,000, so a new level will be created once \dnest~has
seen 10,000 likelihood values above the current top level. This is not affected by
the number of threads. It is difficult to give a sensible default for this
quantity because it depends on the complexity of the problem and how good
the Metropolis proposals are. However, 10,000
will work for many problems (I rarely use anything less than 10,000 or greater
than 300,000). Higher values are slower, but more fail-safe.\\

The save interval controls how often \dnest~writes a model to the output
files: this is what is usually called ``thinning''. Saving more frequently
(i.e. a smaller save interval) is usually better, although it depends on your
problem. If your models have a lot of parameters {\tt sample.txt} can get
very big and take forever to process using {\tt showresults.py}. A default
suggestion that is usually okay is to make the save interval the same as the
new level interval.\\

The option ``threadSteps'' controls how frequently the separate threads pool
their information about the levels. I usually use 100 or 200. If
(threadSteps) $\times$ (number of threads) is not much smaller than
saveInterval or newLevelInterval, the latter won't be what you expect, because
\dnest~only creates levels or saves particles when the threads are pooling
information.\\

\section{Implementing models}
\dnest~currently includes four example models: Atoms, SpikeSlab, FitSine, and
HarlemShake. It also contains an empty Template example which you can use
to implement your own models.\\

To use \dnest~on a particular problem, you need to implement a C++ class
which describes the features of the model (e.g. what the parameters are,
what the prior is, and what the likelihood is). Specifically, an object
(instance) of your class represents a point in the parameter space of your
problem. The class must define several methods (functions) which can be
applied to any object. The first method is {\tt fromPrior()}, which when called,
initialises the object with parameter values from the prior distribution.
The second important method is {\tt perturb()}, which is used to propose
a change to the parameters, as in the basic Metropolis algorithm. The third
important method is {\tt logLikelihood()}, which returns (as you might expect)
the log of the likelihood function evaluated at the object's position. Next,
we have the {\tt print(std::ostream\&)} method which defines how the model
parameters are to be written to the file (i.e. which parameters are to be
printed, and in what order). Finally, there is a method {\tt description()}
which allows you to put a header at the top of the output files with information
about what is in each column. To demonstrate how these methods work, we'll go
through the code for {\tt SpikeSlab}.\\

In the SpikeSlab directory,
the relevant class is the {\tt SpikeSlab} class which is defined in the files
{\tt SpikeSlab.h} and {\tt SpikeSlab.cpp}. If you open {\tt SpikeSlab.h},
you'll see that the class has only one variable, which is a C++ vector called
{\tt params}. This contains the 20 parameter values.\\

\begin{figure}
\begin{framed}
\begin{verbatim}
class SpikeSlab:public DNest3::Model
{
    private:
        std::vector<double> params;
\end{verbatim}
\end{framed}
\end{figure}

It's up to you whether you want to have a single vector of parameters, like
in {\tt SpikeSlab}, or if you want to name all your parameters with names that
make sense to you.\\

In {\tt SpikeSlab.cpp}, all of the methods of the {\tt SpikeSlab} class
are defined. First, we have the {\it constructor}, which defines what happens
when we create an object of class {\tt SpikeSlab}:\\

\begin{framed}
\begin{verbatim}
SpikeSlab::SpikeSlab()
:params(20)
{

}
\end{verbatim}
\end{framed}

This constructor does nothing except specify the size of the {\tt params}
vector, which in this example is 20. Next we have {\tt fromPrior()}, the
method which generates the 20 parameters from the prior. In the SpikeSlab
example, the prior is Uniform$(-0.5, 0.5)$, so this method just has a loop
which generates each value from this distribution. Here I'm making use of
a function {\tt randomU()} which generates values from $U(0, 1)$. This function
is defined in {\tt Utils.h} within \dnest, and is really just a function
wrapping a call to GSL's random number generator.\\

\begin{framed}
\begin{verbatim}
void SpikeSlab::fromPrior()
{
    for(size_t i=0; i<params.size(); i++)
        params[i] = -0.5 + randomU();
}
\end{verbatim}
\end{framed}

The next important method is {\tt perturb()}, which defines the Metropolis
proposals that are used to explore. Since \dnest~spends some time sampling
the prior, and some time sampling narrow distributions like the posterior,
it's a good idea to use very heavy tailed proposal distributions. In the
SpikeSlab example, the prior width is 1, so you shouldn't make moves much
bigger than that, but you don't have any real idea how narrow the posterior
distribution is until you've solved the problem. In the SpikeSlab example,
the proposal is done by choosing one of the 20 parameters to move, and making
a move according to a scale mixture of gaussians (equivalent to ``a gaussian
proposal with a random step size''). Finally, to prevent any proposals outside
the prior range, I use modulo to ensure the proposed value is between -0.5
and 0.5. The resulting proposal is:\\

\begin{framed}
\begin{verbatim}
double SpikeSlab::perturb()
{
    int which = randInt(params.size());
    params[which] += pow(10., 1.5-6*randomU())*randn();
    params[which] = mod(params[which] + 0.5, 1.) - 0.5;
    return 0.;
}
\end{verbatim}
\end{framed}

For convenience, I recently defined two useful functions {\tt randh()}
(to generate heavy tailed proposals) and
{\tt wrap(double\& value, double min, double max)} (to simplify the modulo
stuff). An equivalent version of the {\tt perturb} method is then:\\

\begin{framed}
\begin{verbatim}
double SpikeSlab::perturb()
{
    int which = randInt(params.size());
    params[which] += randh();
    wrap(params[which], -0.5, 0.5);
    return 0.;
}
\end{verbatim}
\end{framed}

The last thing we need to pay attention to in the perturb method is the
returned value. \dnest~assumes that your perturb method, if used repeatedly,
would inherently explore the prior distribution in your problem. With
a uniform prior, it's easy to ensure this. However, for a non-uniform prior,
it's not so easy. One way to do it is to work in a coordinate system where
the prior {\it is} uniform, and then transforming the parameter before using
it in the likelihood function. Another way is to return a nonzero value from
the {\tt perturb} method. Any value returned is added to the log of the
acceptance probability. So if you want to explore a non-uniform prior, you
can return the log of the ratio of the prior density at the proposed point
to the value at the original point. For concreteness, suppose you had a
parameter $a$ with a Uniform$(0,10)$ prior, and a parameter $b$ with a
Normal$(20, \sigma=5)$ prior. Then the {\tt perturb} method could be implemented
like this:\newpage

\begin{framed}
\begin{verbatim}
double MyModel::perturb()
{
    int which = randInt(2);
    double logH = 0.;

    if(which == 0)
    {
        a += 10.*randh();
        wrap(a, 0., 10.);
    }
    else
    {
        logH -= -0.5*pow((b - 20.)/5., 2);
        b += 5.*randh();
        logH += -0.5*pow((b - 20.)/5., 2);
    }

    return logH;
}
\end{verbatim}
\end{framed}

This proposal chooses to move either $a$ or $b$, and uses a heavy tailed
proposal distribution (scaled by the width of the prior). Since the prior for
$b$ is non-uniform, it returns the log of the ``Hastings factor'' that would
be needed to explore the normal prior.\\

When you write a new model for \dnest, I highly recommend sampling the prior
and making sure you at least get that part correct. To do this, edit OPTIONS
so the maximum number of levels is 1, and make the save interval frequent.
Then inspect {\tt sample.txt} and make plots to ensure you're sampling the
prior correctly. If you are, that's a necessary but not sufficient condition
for your {\tt perturb} function being okay (your proposals need to satisfy
detailed balance, and it's possible to have proposals that explore the prior
well but do not satisfy detailed balance. Most simple proposals do satisfy
detailed balance).\\

The {\tt logLikelihood} method does what you'd expect: evaluates and returns
the log likelihood based on the current setting of the parameters. In the
SpikeSlab example, the likelihood is an invented mixture of two gaussians.
However, in general, your likelihood will depend on a data set. When I'm using
real data, I usually write a {\tt Data} class, an object of which represents
a data set. Since you need to access the dataset from within another class,
it's easiest to make the C++ equivalent of a ``global variable'': a single
static instance of the dataset which is accessible from everywhere. See
the FitSine example if you want to see a Data class. Note that the FitSine
example has been superseded by my other \dnest~related project, RJObject.\\

Finally, the {\tt print} method defines how your parameters are printed to the
{\tt sample.txt} (and subsequently {\tt posterior\_sample.txt}) files. In
SpikeSlab, the {\tt print} method writes all 20 parameters, separated by
spaces, to a single line of the stream ``out'' which is input to the method:\\

\begin{framed}
\begin{verbatim}
void SpikeSlab::print(std::ostream& out) const
{
    for(size_t i=0; i<params.size(); i++)
        out<<params[i]<<' ';
}
\end{verbatim}
\end{framed}

Your print method needs to output parameters on a single line, in any order
you want. These will be the order in which they'll appear in the output files
{\tt sample.txt} and {\tt posterior\_sample.txt}. As an example, for our
above hypothetical model with parameters $a$ and $b$, the print function
and the description function might look like this:

\begin{framed}
\begin{verbatim}
void MyModel::print(std::ostream& out) const
{
    out<<a<<' '<<b<<' ';
}

string MyModel::description() const
{
    return string("a, b");
}
\end{verbatim}
\end{framed}

\subsection{Compiling and running your model}
Assuming you implemented your model in the Template directory, there's not
much else for you to do. The file {\tt main.cpp} already contains all the code
you need to construct a sampler and execute it. If your model involves a data
set, you'll just need to make sure you load the dataset at the beginning
of the {\tt main()} function, before doing anything else.\\

The Makefile in the Template directory should allow you to compile your model
by issuing the command {\tt make} in a terminal. If everything worked well
then the executable {\tt main} should exist, which you can use to run
\dnest~on your model. As always, the first time you run \dnest~on a new model,
you should pay attention to the output plots and make sure everything is okay,
such as the number of levels etc.

\subsection{Implementing a model using Julia}
Experimental support for models implemented in Julia is provided in the
{\tt Examples/JuliaModel} directory. To use this, follow these steps:

\begin{enumerate}
\item Implement your log likelihood in {\tt julia\_model.jl}, as a function
	called {\tt log\_likelihood} that takes a {\tt Vector::Float64} as an argument
	and returns a {\tt Float64}. You should assume that the priors for the
	parameters are independent Uniform(0, 1) distributions. To use
	a different prior, transform the parameters inside
	your log likelihood function before moving them.\\
\item You may add other functions to {\tt julia\_model.jl}, for example to load
	a dataset to refer to in your log likelihood.\\
\item Edit the constructor for {\tt MyModel} in {\tt MyModel.cpp} so the parameter
	vector has the correct length for your problem. The default given value
	is 20.\\
\item Open the {\tt Makefile} and make sure the {\tt -L} option passed to
	{\tt g++} points to the
	directory on your computer where the Julia shared library (such
	as {\tt libjulia.so}) is stored. You might also have to configure your
	system so that it searches for libraries here at runtime, e.g.
	by running ldconfig as root. The \dnest~library and include files should
	also be somewhere where the compiler will find them (or you can add them
	to the {\tt g++} calls in the {\tt Makefile} using {\tt -I} for header
	files and {\tt -L} for libraries).\\
\item Open {\tt main.cpp} and edit the first argument to the function
	{\tt jl\_init\_with\_image} to be the path to the file {\tt sys.so} associated with
	Julia on your system.\\
\item Compile by invoking '{\tt make}'.\\
\item Run the executable! Note: only a single thread can be used, so you might want
	to use the {\tt OPTIONS} file to have more than one particle. Five is a
	reasonable default.
\end{enumerate}

\end{document}

### 2016-01-13 doc/manual.tex
Working on the manual to make it correct for DNest4

@@
 useful for solving Bayesian Inference problems.
-\dnest~is free software,
-released under the terms of the GNU General Public Licence, version 3.
+\dnest~is free software, released under the terms of the MIT licence.
 For information about how the
-algorithm works, please see the paper, freely available from the following
-URL:\\
+algorithm works, please see the Diffusive Nested Sampling paper,
+available from the following URL:\\
 
@@
 
-If you find \dnest~useful, please feel free to cite the paper.
+If you find \dnest~useful in a research project,
+please feel free to cite the paper.
 Throughout this manual I will assume that you have read the paper and have a
 reasonable understanding of how the algorithm works, as well as some
-understanding of Nested Sampling in general, as introduced by John Skilling.\\
+understanding of general Nested Sampling concepts.\\
 
@@
 \subsection{Dependencies}
-\dnest~depends on some other software that you will need to have installed on
-your system. These should be quite easy to obtain. I highly
-recommend obtaining these programs from your operating system's package manager,
-rather than installing them from the source code. Note that some OSs split up
-packages into binaries (the library files) and ``development'' packages,
-containing header files (part of a library's source code).
-In Ubuntu, Linux Mint, and Debian, the development packages have a -dev suffix.
-In Fedora, the development packages usually have a -devel suffix (e.g. Fedora).
-To compile \dnest, you will need both the libraries and the development packages
-for each of the dependencies. Here's a list of dependencies:\\
+One of the issues with DNest3 which inspired a new version was that many users
+found it annoying to compile on different operating systems because of the
+dependencies. For that reason, \dnest~now has much less dependencies.
+The C++ part of the code is written in pure C++11 and does not require any
+additional libraries --- you just have to have a modern C++ compiler that
+supports the C++11 standard.
 
-\begin{itemize}
-\item The GNU Scientific Library (GSL)
-\item Python 2, along with the NumPy and matplotlib packages
-\item The {\tt zlib} library (this is only required if wanting the ability to
-output samples to a {\tt gzipped} file, otherwise it can be excluded)
-\item The {\tt thread}, {\tt system} and {\tt iostreams} parts of the Boost C++ library.
-It may be easier to just install all of Boost.
-\item To compile \dnest, you need {\tt cmake}.\\
-\end{itemize}
-
-On Ubuntu, I can install these packages with the following commands:
-\begin{verbatim}
-sudo apt-get install libgsl0-dev
-sudo apt-get install python-numpy python-matplotlib
-sudo apt-get install zlib1g-dev
-sudo apt-get install libboost-thread-dev libboost-system-dev libboost-iostreams-dev
-sudo apt-get install cmake
-\end{verbatim}
-
-\dnest~uses GSL's random number generator. Python, NumPy and
-matplotlib are used for the postprocessing scripts and for
-plotting. Boost is used for parallelization (multithreading).
-While random number generation and
-multithreading is supported
-in the recent C++11 standard, I have not ported \dnest~to C++11 and have no
-intention of doing so in the near future.\\
-
-It is also possible to compile \dnest~without Boost, however this is not
-recommended, and you will not be able to use any of the multi-threaded options
-or the {\tt gzip} option.\\
+The Python postprocessing scripts use the Python packages
+{\tt numpy} and {\tt matplotlib}. They work with both Python 2 or 3.
 
 \subsection{Compiling}
-Please follow the instructions in README.md to compile \dnest~using cmake.
-This is the recommended option.\\
-
-%Go into the \dnest~directory and type {\tt make}. Everything should compile.
+Go into the {\tt code} directory and run {\tt make}. Everything should compile.
 To

### 2016-01-14 doc/manual.tex
Finished manual

@@
 # Target compression factor between levels = 2.7182818284590451
-# Seeding random number generator with 1402801184.
-# Thread 1: Generated 1 particles from the prior.
-# Creating level 1 with logL = -51.65058877.
-# Creating level 2 with logL = -37.67550971.
-# Creating level 3 with logL = -28.69668897.
-# Creating level 4 with logL = -21.61338509.
-# Creating level 5 with logL = -15.92331209.
-# Creating level 6 with logL = -11.291182.
-# Saving a particle to disk. N = 1.
-# Creating level 7 with logL = -7.301954778.
-# Creating level 8 with logL = -4.116771404.
-# Creating level 9 with logL = -1.19393518.
-# Creating level 10 with logL = 1.620392784.
-# Saving a particle to disk. N = 2.
+# Seeding random number generators. First seed = 1452723978.
+# Generating 1 particle from the prior...done.
+# Creating level 1 with log likelihood = -44.1995.
+# Creating level 2 with log likelihood = -32.1028.
+# Creating level 3 with log likelihood = -25.5476.
+# Creating level 4 with log likelihood = -19.911.
+# Creating level 5 with log likelihood = -15.5224.
+# Creating level 6 with log likelihood = -11.0712.
+# Creating level 7 with log likelihood = -6.9397.
+# Creating level 8 with log likelihood = -3.30518.
+# Creating level 9 with log likelihood = 0.522437.
 \end{verbatim}
@@
 The plots are useful for diagnosing the progress of your run. For example, you
-want to ensure that the Metropolis acceptance fraction doesn't drop to near
+want to ensure that the Metropolis acc
+eptance fraction doesn't drop to near
 zero. If it does, the proposal distributions might be too aggressive, or maybe
@@
 \includegraphics[scale=0.5]{fig1.pdf}
-\caption{\it Level vs. Time from a DNest3 run with 100 levels.\label{fig:fig1}}
+\caption{\it Level vs. Time from a \dnest~run with 100 levels.\label{fig:fig1}}
 \end{center}
@@
 \begin{verbatim}
-DNest3 Command Line Options:
+DNest4 Command Line Options: 
 -h: display this message
--l <filename>: load level structure from the specified file.
 -o <filename>: load DNest3 options from the specified file. Default=OPTIONS
--s <seed>: seed the random number generator with the specified value.
-           If unspecified, the system time is used.
+-s <seed>: seed the random number generator with the specified value. If unspecified, the system time is used.
 -d <filename>: Load data from the specified file, if required.
@@
 -t <num_threads>: run on the specified number of threads. Default=1.
--f <filename>: a custom configuration file for adding problem specific
-           options if required.
--z: the output "sample.txt" file will be gzipped and have the suffix ".gz".
 \end{verbatim}
@@
 the threads will pool their results (for example, information about the
-levels). To execute DNest with 8 threads, simply use {\tt ./main -t 8}. You can
+levels). To execute \dnest~with 8 threads, simply use {\tt ./main -t 8}. You can
 replace 8 with however many threads you want. I usually use 8 threads because
@@
 thread, and ensures that they are seeded with different values.\\
-
-The {\tt -z} option is only available is you have compiled with the {\tt zlib}
-library and Boost {\tt iostreams} library installed. \\
-
-If you have any additional configuration information required by the run, e.g.\
-for setting values needed for a particular model that you don't want to hardcode,
-then they can be added through specifying a configuration file using the {\tt -f}
-option. You will have to write your own function to read in this file, so it can
-be in whatever format you are comfortable with (e.g.\ an {\tt ini} file, or {\tt
-json} file).\\
 
@@
 \section{Implementing models}
-\dnest~currently includes four example models: Atoms, SpikeSlab, FitSine, and
-HarlemShake. It also contains an empty Template example which you can use
+\dnest~currently includes three example models: SpikeSlab, StraightLine, and
+RJObject_SineWaves. It also contains an empty Template example which you can use
 to implement your own models.\\
@@
 (instance) of your class represents a point in the parameter space of your
-problem. The class must define several methods (functions) which can be
-applied to any object. The first method is {\tt fromPrior()}, which when called,
-initialises the object with parameter values from the prior distribution.
-The second important method is {\tt perturb()}, which is used to propose
+problem. The class must define several member functions which can be
+applied to any object. The first is {\tt from_prior(DNest4::RNG&)},
+which when called,
+initialises the object with parameter values from the prior distribution
+(the argument is a random number generator object, passed by reference).
+The second important member function is {\tt perturb(DNest4::RNG&)}, which is used to propose
 a change to the parameters, as in the basic Metropolis algorithm. The third
-important method is {\tt logLikelihood()}, which returns (as you might expect)
-the log of the likelihood function evaluated at the object's position. Next,
-we have the {\tt print(std::ostream\&)} method which defines how the model
+is {\tt log_likelihood()}, which returns (as you might expect)
+the log of the likelihood function evaluated at the object's current position. Next,
+we have the {\tt print(std::ostream\&)} member function which defines how the model
 parameters are to be written to the file (i.e. which parameters are to be
-printed, and in what order). Finally, there is a method {\tt description()}
+printed, and in what order). Finally, there is {\tt description()},
 which allows you to put a header at the top of the output files with information
-about what is in each column. To demonstrate how these methods work, we'll go
+about what is in each column. To demonstrate how these member functions work, we'll go
 through the code for {\tt SpikeSlab}.\\
@@
 {\tt SpikeSlab.h} and {\tt SpikeSlab.cpp}. If you open {\tt SpikeSlab.h},
-you'll see that the class has only one variable, which is a C++ vector called
+you'll see that the class has only one variable, which is a C++ valarray called
 {\tt params}. This contains the 20 parameter values.\\
@@
 \begin{verbatim}
-class SpikeSlab:public DNest3::Model
+class SpikeSlab
 {
     private:
-        std::vector<double> params;
+        std::valarray<double> params;
 \end{verbatim}
@@
 
-It's up to you whether you want to have a single vector of parameters, like
+It's up to you whether you want to have a single vector or valarray of parameters, like
 in {\tt SpikeSlab}, or if you want to name all your parameters with names that
@@
 
-In {\tt SpikeSlab.cpp}, all of the methods of the {\tt SpikeSlab} class
+In {\tt SpikeSlab.cpp}, all of the member function of the {\tt SpikeSlab} class
 are defined. First, we have the {\it constructor}, which defines what happens
@@
 This constructor does nothing except specify the size of the {\tt params}
-vector, which in this example is 20. Next we have {\tt fromPrior()}, the
-method which generates the 20 parameters from the prior. In the SpikeSlab
-example, the prior is Uniform$(-0.5, 0.5)$, so this method just has a loop
+valarray, which in this example is 20. Next we have {\tt from_prior(DNest4::RNG&)}, the
+function which generates the 20 parameters from the prior. In the SpikeSlab
+example, the prior is Uniform$(-0.5, 0.5)$, so this function just has a loop
 which generates each value from this distribution. Here I'm making use of
-a function {\tt randomU()} which generates values from $U(0, 1)$. This function
-is defined in {\tt Utils.h} within \dnest, and is really just a function
-wrapping a call to GSL's random number generator.\\
+a function {\tt rng.rand()} which generates values from $U(0, 1)$ using the
+random number generator called rng (which is passed in by \dnest).
+The RNG class is defined in {\tt RNG.h} within \dnest, and is really just a
+wrapper of C++11's random number generator, with more convenient function names.
 
@@
 \begin{verbatim}
-void SpikeSlab::fromPrior()
+void SpikeSlab::from_prior(RNG& rng)
 {
 	for(size_t i=0; i<params.size(); i++)
-        params[i] = -0.5 + randomU();
+		params[i] = -0.5 + rng.rand();
 }
@@
 
-The next important method is {\tt perturb()}, which defines the Metropolis
+The next important member function is {\tt perturb(DNest4::RNG&}, which defines the Metropolis
 proposals that are used to explore. Since \dnest~spends some time sampling
@@
 \begin{verbatim}
-double SpikeSlab::perturb()
+double SpikeSlab::perturb(RNG& rng)
 {
-    int which = randInt(params.size());
-    params[which] += pow(10., 1.5-6*randomU())*randn();
+	int which = rng.rand_int(params.size());
+    params[which] += pow(10., 1.5-6*rng.rand())*rng.randn();
     params[which] = mod(params[which] + 0.5, 1.) - 0.5;
@@
 {\tt wrap(double\& value, double min, double max)} (to simplify the modulo
-stuff). An equivalent version of the {\tt perturb} method is then:\\
+stuff). An equivalent version of {\tt perturb} is then:\\
 
@@
 \begin{verbatim}
-double SpikeSlab::perturb()
+double SpikeSlab::perturb(RNG& rng)
 {
-    int which = randInt(params.size());
-    params[which] += randh();
+	int which = rng.rand_int(params.size());
+	params[which] += rng.randh();
 	wrap(params[which], -0.5, 0.5);
@@
 
-The last thing we need to pay attention to in the perturb method is the
-returned value. \dnest~assumes that your perturb method, if used repeatedly,
+The last thing we need to pay attention to in the perturb function is the
+returned value. \dnest~assumes that your perturb function, if used repeatedly,
 would inherently explore the prior distribution in your problem. With
-a uniform prior, it's easy to ensure this. However, for a non-uniform prior,
+a uniform prior, it's easy to ensure this using wrap, as shown above.
+However, for a non-uniform prior,
 it's not so easy. One way to do it is to work in a coordinate system where
@@
 it in the likelihood function. Another way is to return a nonzero value from
-the {\tt perturb} method. Any value returned is added to the log of the
+the {\tt perturb} function. Any value returned is added to the log of the
 acceptance probability. So if you want to explore a non-uniform prior, you
@@
 parameter $a$ with a Uniform$(0,10)$ prior, and a parameter $b$ with a
-Normal$(20, \sigma=5)$ prior. Then the {\tt perturb} method could be implemented
+Normal$(20, \sigma=5)$ prior. Then the {\tt perturb} member function could be implemented
 like this:\newpage
@@
 \begin{verbatim}
-double MyModel::perturb()
+double MyModel::perturb(RNG& rng)
 {
-    int which = randInt(2);
+    int which = rng.rand_int(2);
     double logH = 0.;
@@
     {
-        a += 10.*randh();
+        a += 10*rng.randh();
         wrap(a, 0., 10.);
@@
         logH -= -0.5*pow((b - 20.)/5., 2);
-        b += 5.*randh();
+        b += 5*rng.randh();
         logH += -0.5*pow((b - 20.)/5., 2);
@@
 
-The {\tt logLikelihood} method does what you'd expect: evaluates and returns
+The {\tt log_likelihood} member does what you'd expect: evaluates and returns
 the log likelihood based on the current setting of the parameters. In the
@@
 static instance of the dataset which is accessible from everywhere. See
-the FitSine example if you want to see a Data class. Note that the FitSine
-example has been superseded by my other \dnest~related project, RJObject.\\
+the StraightLine example if you want to see a Data class.\\
 
-Finally, the {\tt print} method defines how your parameters are printed to the
+Finally, the {\tt print} function defines how your parameters are printed to the
 {\tt sample.txt} (and subsequently {\tt posterior\_sample.txt}) files. In
-SpikeSlab, the {\tt print} method writes all 20 parameters, separated by
-spaces, to a single line of the stream ``out'' which is input to the method:\\
+SpikeSlab, {\tt print} writes all 20 parameters, separated by
+spaces, to a single line of the stream ``out'' which is input to the function:\\
 
@@
 
-Your print method needs to output parameters on a single line, in any order
+Your print function needs to output parameters on a single line, in any order
 you want. These will be the order in which they'll appear in the output files
@@
 The Makefile in the Template directory should allow you to compile your model
-by issuing the command {\tt make} in a terminal. If everything worked well
-then the executable {\tt main} should exist, which you can use to run
+by issuing the command {\tt make} in a terminal. The only preparation you need
+to do is add the path to your \dnest~directory to an environment variable called
+DNEST4\_PATH so that the compiler knows where \dnest~is on your system.
+For example, on my laptop, \dnest~is located in /home/brewer/Projects/DNest4.
+So I set DNEST4_PATH to /home/brewer/Projects.
+
+If everything worked well then after running {\tt make},
+the executable {\tt main} should exist, which you can use to run
 \dnest~on your model. As always, the first time you run \dnest~on a new model,
@@
 
-\subsection{Implementing a model using Julia}
-Experimental support for models implemented in Julia is provided in the
-{\tt Examples/JuliaModel} directory. To use this, follow these steps:
+%\subsection{Implementing a model using Julia}
+%Experimental support for models implemented in Julia is provided in the
+%{\tt Examples/JuliaModel} directory. To use this, follow these steps:
 
-\begin{enumerate}
-\item Implement your log likelihood in {\tt julia\_model.jl}, as a function
-	called {\tt log\_likelihood} that takes a {\tt Vector::Float64} as an argument
-	and returns a {\tt Float64}. You should assume that the priors for the
-	parameters are independent Uniform(0, 1) distributions. To use
-	a different prior, transform the parameters inside
-	your log likelihood function before moving them.\\
-\item You may add other functions to {\tt julia\_model.jl}, for example to load
-	a dataset to refer to in your log likelihood.\\
-\item Edit the constructor for {\tt MyModel} in {\tt MyModel.cpp} so the parameter
-	vector has the correct length for your problem. The default given value
-	is 20.\\
-\item Open the {\tt Makefile} and make sure the {\tt -L} option passed to
-	{\tt g++} points to the
-	directory on your computer where the Julia shared library (such
-	as {\tt libjulia.so}) is stored. You might also have to configure your
-	system so that it searches for libraries here at runtime, e.g.
-	by running ldconfig as root. The \dnest~library and include files should
-	also be somewhere where the compiler will find them (or you can add them
-	to the {\tt g++} calls in the {\tt Makefile} using {\tt -I} for header
-	files and {\tt -L} for libraries).\\
-\item Open {\tt main.cpp} and edit the first argument to the function
-	{\tt jl\_init\_with\_image} to be the path to the file {\tt sys.so} associated with
-	Julia on your system.\\
-\item Compile by invoking '{\tt make}'.\\
-\item Run the executable! Note: only a single thread can be used, so you might want
-	to use the {\tt OPTIONS} file to have more than one particle. Five is a
-	reasonable default.
-\end{enumerate}
+%\begin{enumerate}
+%\item Implement your log likelihood in {\tt julia\_model.jl}, as a function
+%	called {\tt log\_likelihood} that takes a {\tt Vector::Float64} as an argument
+%	and returns a {\tt Float64}. You should assume that the priors for the
+%	parameters are independent Uniform(0, 1) distributions. To use
+%	a different prior, transform the parameters inside
+%	your log likelihood function before moving them.\\
+%\item You may add other functions to {\tt julia\_model.jl}, for example to load
+%	a dataset to refer to in your log likelihood.\\
+%\item Edit the constructor for {\tt MyModel} in {\tt MyModel.cpp} so the parameter
+%	vector has the correct length for your problem. The default given value
+%	is 20.\\
+%\item Open the {\tt Makefile} and make sure the {\tt -L} option passed to
+%	{\tt g++} points to the
+%	directory on your computer where the Julia shared library (such
+%	as {\tt libjulia.so}) is stored. You might also have to configure your
+%	system so that it searches for libraries here at runtime, e.g.
+%	by running ldconfig as root. The \dnest~library and include files should
+%	also be somewhere where the compiler will find them (or you can add them
+%	to the {\tt g++} calls in the {\tt Makefile} using {\tt -I} for header
+%	files and {\tt -L} for libraries).\\
+%\item Open {\tt main.cpp} and edit the first argument to the function
+%	{\tt jl\_init\_with\_image} to be the path to the file {\tt sys.so} associated with
+%	Julia on your system.\\
+%\item Compile by invoking '{\tt make}'.\\
+%\item Run the executable! Note: only a single thread can be used, so you might want
+%	to use the {\tt OPTIONS} file to have more than one particle. Five is a
+%	reasonable default.
+%\end{enumerate}
 

### 2016-01-14 doc/manual.tex
Fixed some problems

@@
 weights. On most problems, these will peak at some point, and begin to
-come back down towards zero (See Figure~\ref{fig3}).
+come back down towards zero (See Figure~\ref{fig:fig3}).
 Once this occurs, you probably have enough levels.
@@
 \dnest~currently includes three example models: SpikeSlab, StraightLine, and
-RJObject_SineWaves. It also contains an empty Template example which you can use
+RJObject\_SineWaves. It also contains an empty Template example which you can use
 to implement your own models.\\
@@
 problem. The class must define several member functions which can be
-applied to any object. The first is {\tt from_prior(DNest4::RNG&)},
+applied to any object. The first is {\tt from\_prior(DNest4::RNG\&)},
 which when called,
@@
 (the argument is a random number generator object, passed by reference).
-The second important member function is {\tt perturb(DNest4::RNG&)}, which is used to propose
+The second important member function is {\tt perturb(DNest4::RNG\&)}, which is used to propose
 a change to the parameters, as in the basic Metropolis algorithm. The third
-is {\tt log_likelihood()}, which returns (as you might expect)
+is {\tt log\_likelihood()}, which returns (as you might expect)
 the log of the likelihood function evaluated at the object's current position. Next,
@@
 This constructor does nothing except specify the size of the {\tt params}
-valarray, which in this example is 20. Next we have {\tt from_prior(DNest4::RNG&)}, the
+valarray, which in this example is 20. Next we have {\tt from\_prior(DNest4::RNG\&)}, the
 function which generates the 20 parameters from the prior. In the SpikeSlab
@@
 
-The next important member function is {\tt perturb(DNest4::RNG&}, which defines the Metropolis
+The next important member function is {\tt perturb(DNest4::RNG\&}, which defines the Metropolis
 proposals that are used to explore. Since \dnest~spends some time sampling
@@
 
-The {\tt log_likelihood} member does what you'd expect: evaluates and returns
+The {\tt log\_likelihood} member does what you'd expect: evaluates and returns
 the log likelihood based on the current setting of the parameters. In the
@@
 For example, on my laptop, \dnest~is located in /home/brewer/Projects/DNest4.
-So I set DNEST4_PATH to /home/brewer/Projects.
+So I set DNEST4\_PATH to /home/brewer/Projects.
 

### 2016-01-14 code/RJObject/ConditionalPriors/BasicCircular.cpp
Include header files from ../..

@@ -1,5 +1,5 @@
 #include "BasicCircular.h"
-#include "Utils.h"
+#include "../../Utils.h"
 #include <cmath>
 
 using namespace DNest4;

### 2016-01-14 code/RJObject/ConditionalPriors/BasicCircular.h

@@ -2,8 +2,8 @@
 #define DNest4_BasicCircular
 
 #include "ConditionalPrior.h"
-#include "RNG.h"
-#include "Utils.h"
+#include "../../RNG.h"
+#include "../../Utils.h"
 
 namespace DNest4
 {

### 2016-01-14 code/RJObject/ConditionalPriors/ClassicMassInf.cpp

@@ -1,5 +1,5 @@
 #include "ClassicMassInf.h"
-#include "Utils.h"
+#include "../../Utils.h"
 #include <cmath>
 
 using namespace DNest4;

### 2016-01-14 code/RJObject/ConditionalPriors/ClassicMassInf.h

@@ -2,7 +2,7 @@
 #define DNest4_ClassicMassInf
 
 #include "ConditionalPrior.h"
-#include "RNG.h"
+#include "../../RNG.h"
 
 namespace DNest4
 {

### 2016-01-14 code/RJObject/ConditionalPriors/ClassicMassInf1D.cpp

@@ -1,5 +1,5 @@
 #include "ClassicMassInf1D.h"
-#include "Utils.h"
+#include "../../Utils.h"
 #include <cmath>
 
 using namespace DNest4;

### 2016-01-14 code/RJObject/ConditionalPriors/ConditionalPrior.h

@@ -14,7 +14,7 @@
 
 #include <vector>
 #include <ostream>
-#include "RNG.h"
+#include "../../RNG.h"
 
 namespace DNest4
 {

### 2016-01-14 code/RJObject/ConditionalPriors/Pareto.cpp

@@ -1,5 +1,5 @@
 #include "Pareto.h"
-#include "Utils.h"
+#include "../../Utils.h"
 #include <cmath>
 
 using namespace DNest4;

### 2016-01-14 code/RJObject/ConditionalPriors/Pareto.h

@@ -2,7 +2,7 @@
 #define DNest4_Pareto
 
 #include "ConditionalPrior.h"
-#include "RNG.h"
+#include "../../RNG.h"
 
 namespace DNest4
 {

### 2016-01-14 code/RJObject/RJObject.h
Include path

@@ -6,9 +6,8 @@
 #include <cmath>
 #include <iostream>
 #include <iomanip>
-#include "Utils.h"
-
-#include "RNG.h"
+#include "../Utils.h"
+#include "../RNG.h"
 
 namespace DNest4
 {

### 2016-01-14 doc/manual.tex
DNest3 mentions changed to DNest4
GalaxyField example

@@
 -h: display this message
--o <filename>: load DNest3 options from the specified file. Default=OPTIONS
+-o <filename>: load DNest4 options from the specified file. Default=OPTIONS
 -s <seed>: seed the random number generator with the specified value. If unspecified, the system time is used.
@@
 \section{Implementing models}
-\dnest~currently includes three example models: SpikeSlab, StraightLine, and
-RJObject\_SineWaves. It also contains an empty Template example which you can use
+\dnest~currently includes three example models: SpikeSlab, StraightLine,
+RJObject\_SineWaves, and RJObject\_GalaxyField.
+It also contains an empty Template example which you can use
 to implement your own models.\\

### 2016-01-14 code/Examples/RJObject_GalaxyField/Data.cpp

@@ -1,0 +1,85 @@
+#include "Data.h"
+#include <iostream>
+#include <fstream>
+#include <cmath>
+
+using namespace std;
+
+Data Data::instance;
+
+Data::Data()
+{
+
+}
+
+void Data::load(const char* metadata_file, const char* image_file,
+			const char* sigma_file)
+{
+	/*
+	* First, read in the metadata
+	*/
+	fstream fin(metadata_file, ios::in);
+	if(!fin)
+		cerr<<"# ERROR: couldn't open file "<<metadata_file<<"."<<endl;
+	fin>>ni>>nj;
+	fin>>x_min>>x_max>>y_min>>y_max;
+	fin.close();
+
+	// Make sure maximum > minimum
+	if(x_max <= x_min || y_max <= y_min)
+		cerr<<"# ERROR: strange input in "<<metadata_file<<"."<<endl;
+
+	// Compute pixel widths
+	dx = (x_max - x_min)/nj;
+	dy = (y_max - y_min)/ni;
+
+	// Check that pixels are square
+	if(abs(log(dx/dy)) >= 1E-3)
+		cerr<<"# ERROR: pixels aren't square."<<endl;
+
+	compute_ray_grid();
+
+	/*
+	* Now, load the image
+	*/
+	fin.open(image_file, ios::in);
+	if(!fin)
+		cerr<<"# ERROR: couldn't open file "<<image_file<<"."<<endl;
+	image.assign(ni, vector<double>(nj));
+	for(size_t i=0; i<image.size(); i++)
+		for(size_t j=0; j<image[i].size(); j++)
+			fin>>image[i][j];
+	fin.close();
+
+	/*
+	* Load the sigma map
+	*/
+	fin.open(sigma_file, ios::in);
+	if(!fin)
+		cerr<<"# ERROR: couldn't open file "<<sigma_file<<"."<<endl;
+	sigma.assign(ni, vector<double>(nj));
+	for(size_t i=0; i<sigma.size(); i++)
+		for(size_t j=0; j<sigma[i].size(); j++)
+			fin>>sigma[i][j];
+	fin.close();
+}
+
+void Data::compute_ray_grid()
+{
+	// Make vectors of the correct size
+	x_rays.assign(ni, vector<double>(nj));
+	y_rays.assign(ni, vector<double>(nj));
+
+	// Distance between adjacent rays
+	double L = dx;
+
[…]

### 2016-01-14 code/Examples/RJObject_GalaxyField/Data.h

@@ -1,0 +1,61 @@
+#ifndef DNest4_GalaxyField_Data
+#define DNest4_GalaxyField_Data
+
+#include <vector>
+
+class Data
+{
+	private:
+		// Number of pixels
+		int ni, nj;
+
+		// Coordinates of image boundaries
+		double x_min, x_max, y_min, y_max;
+
+		// Pixel widths
+		double dx, dy;
+
+		// Coordinates of pixel centers
+		std::vector< std::vector<double> > x_rays;
+		std::vector< std::vector<double> > y_rays;
+
+		// The pixels
+		std::vector< std::vector<double> > image;
+
+		// Sigma map
+		std::vector< std::vector<double> > sigma;
+
+		void compute_ray_grid();
+
+	public:
+		Data();
+		void load(const char* metadata_file, const char* image_file,
+				const char* sigma_file);
+
+		// Getters
+		int get_ni() const { return ni; }
+		int get_nj() const { return nj; }
+		double get_x_min() const { return x_min; }
+		double get_x_max() const { return x_max; }
+		double get_y_min() const { return y_min; }
+		double get_y_max() const { return y_max; }
+		double get_dx() const { return dx; }
+		double get_dy() const { return dy; }
+		const std::vector< std::vector<double> >& get_x_rays() const
+			{ return x_rays; }
+		const std::vector< std::vector<double> >& get_y_rays() const
+			{ return y_rays; }
+		const std::vector< std::vector<double> >& get_image() const
+			{ return image; }
+		const std::vector< std::vector<double> >& get_sigma() const
+			{ return sigma; }
+
+	// Singleton
+	private:
+		static Data instance;
+	public:
+		static Data& get_instance() { return instance; }
+};
+
+#endif
+

### 2016-01-14 code/Examples/RJObject_GalaxyField/Frames/make_movie.sh

@@ -1,0 +1,5 @@
+#!/bin/bash
+avconv -r 5 -f image2 -i '%04d.png' -b:v 4096k movie.avi
+
+
+

### 2016-01-14 code/Examples/RJObject_GalaxyField/MyConditionalPrior.cpp

@@ -1,0 +1,154 @@
+#include "MyConditionalPrior.h"
+#include "Utils.h"
+#include <cmath>
+
+using namespace DNest4;
+
+MyConditionalPrior::MyConditionalPrior(double x_min, double x_max,
+					double y_min, double y_max,
+					double fluxlim_min, double fluxlim_max)
+:x_min(x_min)
+,x_max(x_max)
+,y_min(y_min)
+,y_max(y_max)
+,fluxlim_min(fluxlim_min)
+,fluxlim_max(fluxlim_max)
+,radiuslim_min(1E-3*sqrt((x_max - x_min)*(y_max - y_min)))
+,radiuslim_max(sqrt((x_max - x_min)*(y_max - y_min)))
+{
+
+}
+
+void MyConditionalPrior::from_prior(RNG& rng)
+{
+	fluxlim = exp(log(fluxlim_min) + log(fluxlim_max/fluxlim_min)*rng.rand());
+	gamma = rng.rand();
+
+	radiuslim = exp(log(radiuslim_min) + log(radiuslim_max/radiuslim_min)*rng.rand());
+	gamma_radius = rng.rand();
+
+	b1 = rng.rand();
+	a1 = rng.rand()*b1;
+
+	b2 = rng.rand();
+	a2 = rng.rand()*b2;
+}
+
+double MyConditionalPrior::perturb_hyperparameters(RNG& rng)
+{
+	double logH = 0.;
+
+	int which = rng.rand_int(8);
+
+	if(which == 0)
+	{
+		fluxlim = log(fluxlim);
+		fluxlim += log(fluxlim_max/fluxlim_min)*rng.randh();
+		fluxlim = mod(fluxlim - log(fluxlim_min),
+			log(fluxlim_max/fluxlim_min)) + log(fluxlim_min);
+		fluxlim = exp(fluxlim);
+	}
+	else if(which == 1)
+	{
+		gamma += rng.randh();
+		gamma = mod(gamma, 1.);
+	}
+	else if(which == 2)
+	{
+		radiuslim = log(radiuslim);
+		radiuslim += log(radiuslim_max/radiuslim_min)*rng.randh();
+		radiuslim = mod(radiuslim - log(radiuslim_min),
+			log(radiuslim_max/radiuslim_min)) + log(radiuslim_min);
[…]

### 2016-01-14 code/Examples/RJObject_GalaxyField/MyConditionalPrior.h

@@ -1,0 +1,46 @@
+#ifndef DNest4_GalaxyField_MyConditionalPrior
+#define DNest4_GalaxyField_MyConditionalPrior
+
+#include "RJObject/ConditionalPriors/ConditionalPrior.h"
+
+// Hyperparameters setting interim prior for galaxy properties
+class MyConditionalPrior:public DNest4::ConditionalPrior
+{
+	private:
+		// Limits
+		double x_min, x_max, y_min, y_max;
+		double fluxlim_min, fluxlim_max;
+		double radiuslim_min, radiuslim_max;
+
+		// Lower limit and 1/slope for Pareto interim prior
+		// for masses
+		double fluxlim;
+		double gamma;
+
+		// Lower limit and 1/slope for Pareto interim prior
+		// for radii
+		double radiuslim;
+		double gamma_radius;
+
+		// Uniform interim prior for radius ratio and mass ratio
+		double a1, b1;
+		double a2, b2;
+
+		double perturb_hyperparameters(DNest4::RNG& rng);
+
+	public:
+		MyConditionalPrior(double x_min, double x_max,
+					double y_min, double y_max,
+					double fluxlim_min, double fluxlim_max);
+
+		void from_prior(DNest4::RNG& rng);
+
+		double log_pdf(const std::vector<double>& vec) const;
+		void from_uniform(std::vector<double>& vec) const;
+		void to_uniform(std::vector<double>& vec) const;
+
+		void print(std::ostream& out) const;
+};
+
+#endif
+

### 2016-01-14 code/Examples/RJObject_GalaxyField/MyModel.cpp

@@ -1,0 +1,162 @@
+#include "MyModel.h"
+#include "RNG.h"
+#include "Utils.h"
+#include "Data.h"
+#include <cmath>
+
+using namespace std;
+using namespace DNest4;
+
+MyModel::MyModel()
+:objects(8, 100, false, MyConditionalPrior(
+	Data::get_instance().get_x_min(), Data::get_instance().get_x_max(),
+	Data::get_instance().get_y_min(), Data::get_instance().get_y_max(),
+			1E-3, 1E3))
+,image(Data::get_instance().get_ni(),
+	vector<long double>(Data::get_instance().get_nj()))
+,staleness(0)
+{
+
+}
+
+void MyModel::from_prior(RNG& rng)
+{
+	objects.from_prior(rng);
+	calculate_image();
+	sigma = exp(log(1.) + log(1E6)*rng.rand());
+}
+
+void MyModel::calculate_image()
+{
+	// Get coordinate stuff from data
+	const vector< vector<double> >& x = Data::get_instance().get_x_rays();
+	const vector< vector<double> >& y = Data::get_instance().get_y_rays();
+
+	// Components
+	const vector< vector<double> >& components = objects.get_components();
+
+	// Diff
+	const vector< vector<double> >& added = objects.get_added();
+	const vector< vector<double> >& removed = objects.get_removed();
+	bool update = (added.size() + removed.size() < components.size()) &&
+				(staleness < 100);
+	int num = (update)?(added.size() + removed.size()):(components.size());
+
+	if(update)
+		staleness++;
+	else
+	{
+		staleness = 0;
+		// Zero the image
+		image.assign(Data::get_instance().get_ni(),
+			vector<long double>(Data::get_instance().get_nj(), 0.));
+	}
+
+	double xc, yc, M, w, q, theta, cos_theta, sin_theta, wsq;
+	double rinner, rinnersq, frac;
+	double xx, yy, rsq;
+
+	const vector<double>* component;
+	double coeff = 1.; // switches to -1 for removing components
+
+	for(int k=0; k<num; k++)
+	{
+		if(update)
+		{
+			if(k < (int)added.size())
+			{
+				component = &(added[k]);
+				coeff = 1.;
+			}
+			else
+			{
[…]

### 2016-01-14 code/Examples/RJObject_GalaxyField/MyModel.h

@@ -1,0 +1,43 @@
+#ifndef DNest4_GalaxyField_MyModel
+#define DNest4_GalaxyField_MyModel
+
+#include <vector>
+#include "RJObject/RJObject.h"
+#include "MyConditionalPrior.h"
+
+class MyModel
+{
+	private:
+		DNest4::RJObject<MyConditionalPrior> objects;
+
+		// The model image
+		std::vector< std::vector<long double> > image;
+		void calculate_image();
+
+		// How many steps since image was computed from scratch
+		int staleness;
+
+		// Noise standard deviation
+		double sigma;
+
+	public:
+		MyModel();
+
+		// Generate the point from the prior
+		void from_prior(DNest4::RNG& rng);
+
+		// Metropolis-Hastings proposals
+		double perturb(DNest4::RNG& rng);
+
+		// Likelihood function
+		double log_likelihood() const;
+
+		// Print to stream
+		void print(std::ostream& out) const;
+
+		// Return string with column information
+		std::string description() const;
+};
+
+#endif
+

### 2016-01-14 code/Examples/RJObject_GalaxyField/main.cpp

@@ -1,0 +1,19 @@
+#include <iostream>
+#include "Start.h"
+#include "MyModel.h"
+#include "Data.h"
+
+using namespace std;
+using namespace DNest4;
+
+int main(int argc, char** argv)
+{
+	Data::get_instance().load("Data/test_metadata.txt",
+				"Data/test_image.txt", "Data/test_sigma.txt");
+
+	Sampler<MyModel> sampler = setup<MyModel>(argc, argv);
+	sampler.run();
+
+	return 0;
+}
+

### 2016-01-14 code/Examples/RJObject_GalaxyField/make_plots.py

@@ -1,0 +1,36 @@
+from pylab import *
+
+rc("font", size=14, family="serif", serif="Computer Sans")
+rc("text", usetex=True)
+
+# Piecewise linear stretch
+def stretch(x):
+	y = x.copy()
+	y = (y - y.min())/(y.max() - y.min())
+	y[y > 0.1] = 0.1 + 0.05*(y[y > 0.1] - 0.1)
+	return y
+
+data = loadtxt('Data/test_image.txt')
+
+img = stretch(data)
+
+imshow(img, cmap='gray', interpolation='nearest')
+xlabel('$x$')
+ylabel('$y$')
+gca().set_xticks([-0.5, 99.5, 199.5])
+gca().set_yticks([-0.5, 99.5, 199.5])
+gca().set_xticklabels(['-1', '0', '1'])
+gca().set_yticklabels(['1', '0', '-1'])
+title('GalaxyField Data')
+savefig('galaxyfield_data.pdf', bbox_inches='tight')
+show()
+
+posterior_sample = loadtxt('posterior_sample.txt')
+
+hist(posterior_sample[:,40010], 50)
+xlabel('Number of Galaxies $N$', fontsize=14)
+ylabel('Number of posterior samples', fontsize=14)
+xlim([-0.5, 100.5])
+savefig('N_galaxy_result.pdf', bbox_inches='tight')
+show()
+

### 2016-01-14 code/Examples/RJObject_GalaxyField/showresults.py

@@ -1,0 +1,4 @@
+import dnest4
+dnest4.postprocess()
+import display
+

### 2016-01-14 code/Examples/RJObject_GalaxyField/display.py
Display script

@@ -1,0 +1,49 @@
+from pylab import *
+
+import os
+
+# Piecewise linear stretch
+def stretch(x):
+	y = x.copy()
+	y = (y - y.min())/(y.max() - y.min())
+	y[y > 0.1] = 0.1 + 0.05*(y[y > 0.1] - 0.1)
+	return y
+
+saveFrames = False # For making movies
+if saveFrames:
+	os.system('rm Frames/*.png')
+
+posterior_sample = atleast_2d(loadtxt('posterior_sample.txt'))
+data = loadtxt('Data/test_image.txt')
+sig = loadtxt('Data/test_sigma.txt')
+
+ion()
+hold(False)
+for i in range(0, posterior_sample.shape[0]):
+	img = posterior_sample[i, 0:200**2].reshape((200, 200))
+	subplot(1, 2, 1)
+	imshow(-stretch(img), cmap='gray')
+	title('Model {i}'.format(i=i))
+	gca().set_xticks([-0.5, 99.5, 199.5])
+	gca().set_yticks([-0.5, 99.5, 199.5])
+	gca().set_xticklabels(['-1', '0', '1'])
+	gca().set_yticklabels(['1', '0', '-1'])
+
+	subplot(1, 2, 2)
+	sigma = sqrt(sig**2 + posterior_sample[i,-2]**2)
+	imshow(-(img - data)/sigma, cmap='gray')
+	title('Standardised Residuals')
+	gca().set_xticks([-0.5, 99.5, 199.5])
+	gca().set_yticks([-0.5, 99.5, 199.5])
+	gca().set_xticklabels(['-1', '0', '1'])
+	gca().set_yticklabels(['1', '0', '-1'])
+
+	draw()
+
+	if saveFrames:
+		savefig('Frames/' + '%0.4d'%(i+1) + '.png', bbox_inches='tight')
+		print('Frames/' + '%0.4d'%(i+1) + '.png')
+
+ioff()
+show()
+

### 2016-01-14 code/SamplerImpl.h
10 sig figs

@@ -3,6 +3,7 @@
 #include <fstream>
 #include <thread>
 #include <algorithm>
+#include <iomanip>
 #include "Utils.h"
 
 namespace DNest4
@@ -300,6 +301,7 @@ void Sampler<ModelType>::do_bookkeeping()
 		std::sort(all_above.begin(), all_above.end());
 		int index = static_cast<int>((1. - 1./compression)*all_above.size());
 		std::cout<<"# Creating level "<<levels.size()<<" with log likelihood = ";
+		std::cout<<std::setprecision(10);
 		std::cout<<all_above[index].get_value()<<"."<<std::endl;
 
 		levels.push_back(Level(all_above[index]));

### 2016-01-14 doc/manual.tex
Note about directories

@@
 So I set DNEST4\_PATH to /home/brewer/Projects.
+{\bf Note: }The \#include statements in the working examples expect to find
+the \dnest~header files two directories up: i.e. the working examples will only
+successfully compile without modification from within the Examples directory.
+However, the Template should compile from anywhere on your system provided you
+have set DNEST4\_PATH correctly.
 

### 2016-01-18 doc/manual.tex
tabs to spaces
wrap onto next line
Julia instructions in the manual

@@
 -o <filename>: load DNest4 options from the specified file. Default=OPTIONS
--s <seed>: seed the random number generator with the specified value. If unspecified, the system time is used.
+-s <seed>: seed the random number generator with the specified value.
+           If unspecified, the system time is used.
 -d <filename>: Load data from the specified file, if required.
@@
 RJObject\_SineWaves, and RJObject\_GalaxyField.
-It also contains an empty Template example which you can use
+It also contains an empty Template example (in the Templates/BasicModel
+subdirectory) which you can use
 to implement your own models.\\
@@
 \subsection{Compiling and running your model}
-Assuming you implemented your model in the Template directory, there's not
+Assuming you implemented your model in the Template directory
+(Templates/BasicModel), there's not
 much else for you to do. The file {\tt main.cpp} already contains all the code
@@
 
-%\subsection{Implementing a model using Julia}
-%Experimental support for models implemented in Julia is provided in the
-%{\tt Examples/JuliaModel} directory. To use this, follow these steps:
-
-%\begin{enumerate}
-%\item Implement your log likelihood in {\tt julia\_model.jl}, as a function
-%	called {\tt log\_likelihood} that takes a {\tt Vector::Float64} as an argument
-%	and returns a {\tt Float64}. You should assume that the priors for the
-%	parameters are independent Uniform(0, 1) distributions. To use
-%	a different prior, transform the parameters inside
-%	your log likelihood function before moving them.\\
-%\item You may add other functions to {\tt julia\_model.jl}, for example to load
-%	a dataset to refer to in your log likelihood.\\
-%\item Edit the constructor for {\tt MyModel} in {\tt MyModel.cpp} so the parameter
-%	vector has the correct length for your problem. The default given value
-%	is 20.\\
-%\item Open the {\tt Makefile} and make sure the {\tt -L} option passed to
-%	{\tt g++} points to the
-%	directory on your computer where the Julia shared library (such
-%	as {\tt libjulia.so}) is stored. You might also have to configure your
-%	system so that it searches for libraries here at runtime, e.g.
-%	by running ldconfig as root. The \dnest~library and include files should
-%	also be somewhere where the compiler will find them (or you can add them
-%	to the {\tt g++} calls in the {\tt Makefile} using {\tt -I} for header
-%	files and {\tt -L} for libraries).\\
-%\item Open {\tt main.cpp} and edit the first argument to the function
-%	{\tt jl\_init\_with\_image} to be the path to the file {\tt sys.so} associated with
-%	Julia on your system.\\
-%\item Compile by invoking '{\tt make}'.\\
-%\item Run the executable! Note: only a single thread can be used, so you might want
-%	to use the {\tt OPTIONS} file to have more than one particle. Five is a
-%	reasonable default.
-%\end{enumerate}
+\subsection{Implementing a model using Julia}
+Experimental support for models implemented in Julia is provided in the
+Templates/JuliaModel directory. There's a HOWTO file in that directory
+with instructions for how to use it.
 

### 2016-01-19 doc/manual.tex
'code' in directory names

@@
 verify that the compilation was successful,
-go into the {\tt Examples/SpikeSlab} directory
+go into the code/Examples/SpikeSlab directory
 and see if an executable file called {\tt main} has been created. If you
@@
 
-To run the SpikeSlab example, go into the {\tt Examples/Spikeslab}
+To run the SpikeSlab example, go into the code/Examples/Spikeslab
 directory and execute {\tt main}. You should see some output that looks like
@@
 RJObject\_SineWaves, and RJObject\_GalaxyField.
-It also contains an empty Template example (in the Templates/BasicModel
+It also contains an empty Template example (in the code/Templates/BasicModel
 subdirectory) which you can use
@@
 Assuming you implemented your model in the Template directory
-(Templates/BasicModel), there's not
+(code/Templates/BasicModel), there's not
 much else for you to do. The file {\tt main.cpp} already contains all the code
@@
 Experimental support for models implemented in Julia is provided in the
-Templates/JuliaModel directory. There's a HOWTO file in that directory
+code/Templates/JuliaModel directory. There's a HOWTO file in that directory
 with instructions for how to use it.

### 2016-04-06 doc/manual.tex
PYTHON_PATH
Some editing

@@
 If you find \dnest~useful in a research project,
-please feel free to cite the paper.
+please cite the paper.
 Throughout this manual I will assume that you have read the paper and have a
@@
 \section{Installation}
-Please note that I have only ever compiled \dnest~on GNU/Linux,
+I have only ever compiled \dnest~on GNU/Linux,
 using the GNU C++ compiler ({\tt g++}). However, I expect it to be
 straightforward to compile it on any other Unix-like operating system such as Mac
-OS X or FreeBSD. It should be possible, albeit probably more tricky,
+OS X or FreeBSD.\\
+
+It should be possible, but probably more tricky,
 to compile \dnest~in Microsoft Windows. If you want to try,
@@
 probably your best bet. I'd be interested to hear from anyone who
-has tried this. I would also be interested to hear if anyone has
+has tried this, and any problems that arose.
+I would also be interested to hear if anyone has
 compiled \dnest~with a different compiler, such as the Intel C++
@@
 \subsection{Dependencies}
-One of the issues with DNest3 which inspired a new version was that many users
+One of the issues with DNest3 (an older implementation) which inspired a rewrite
+was that many users
 found it annoying to compile on different operating systems because of the
@@
 The Python postprocessing scripts use the Python packages
-{\tt numpy} and {\tt matplotlib}. They work with both Python 2 or 3.
+{\tt NumPy}, {\tt matplotlib}, {\tt Pandas}, and {\tt Cython}.
+They should work with both Python 2 or 3.
 
@@
 verify that the compilation was successful,
-go into the code/Examples/SpikeSlab directory
+go into the {\tt code/Examples/SpikeSlab} directory
 and see if an executable file called {\tt main} has been created. If you
@@
 Control+C, or whatever works in your operating system. If you saw this output,
-\dnest~was compiled successfully.\\
+the C++ parts of \dnest~were compiled successfully.\\
 
-\section{Running DNest}
+Next, you'll need to install the {\tt dnest4} Python package so you can use
+the postprocessing scripts, and other features. Go into the
+{\tt python} directory and run {\tt python setup.py install} or similar.
+
+\section{Running DNest4}
 As part of the compilation process, several examples will have been compiled.
@@
 
-To run the SpikeSlab example, go into the code/Examples/Spikeslab
+To run the SpikeSlab example, go into the {\tt code/Examples/Spikeslab}
 directory and execute {\tt main}. You should see some output that looks like
 that shown in Figure~\ref{fig:output}.
-Note that you must terminate the process manually by pressing Ctrl+C
+You must terminate the process manually by pressing Ctrl+C
 or in another way. If you do not terminate the process yourself, it will run
-forever. There is no definite rule for how long you should run \dnest.
-It depends on the difficulty of your problem, the number of posterior samples you
-want, and the accuracy you want for the marginal likelihood.\\
+forever. There is no definite rule for how long you should run \dnest. Like
+many MCMC methods, it depends on how hard your problem is, the number of
+posterior samples you
+want, and the accuracy you want for the marginal likelihood or other
+quantities you care about.\\
 
@@
 DNS algorithm (i.e. running the MCMC chain, building
-levels, and then exploring all the levels). It creates three output text files,
+levels, and then exploring all the levels). It creates three text output files,
 {\tt sample.txt}, {\tt sample\_info.txt}, and {\tt levels.txt}.\\
@@
 parameter space. In the SpikeSlab example, there are 20 parameters, so there
-will be 20 columns in {\tt sample.txt}.
+are 20 columns in {\tt sample.txt}.
 Each time a point is saved to {\tt sample.txt}, \dnest~prints
@@
 values of the levels, i.e. how compressed they are relative to the prior, in
-units of nats. The second column contains the log likelihoods of the levels.
+units of nats. For example, if a level has $\log(X) = -1.02$, its likelihood
+value encloses $\exp(-1.02) \approx 36.1\%$ of the prior mass.\\
+
+The second column contains the log likelihoods of the levels.
 The first level, with a $\log(X)$ value of 0 and a log likelihood of
-$-10^{300}$ (basically ``minus infinity''), is simply the prior. The third
+$-10^{308}$ (basically ``minus infinity''), is simply the prior. The third
 column has the ``tiebreaker'' values for the levels, which again are not

### 2016-04-06 python/dnest4/builder.py
Binomial

@@ -3,7 +3,7 @@ from enum import Enum
 import numpy as np
 
 __all__ = ["Uniform", "LogUniform", "Normal", "Cauchy", "Poisson",\
-           "Exponential",\
+           "Exponential", "Binomial",\
            "Deterministic", "NodeType", "Node", "Model"]
 
 class Uniform:
@@ -155,11 +155,43 @@ class Poisson:
         self.mu = mu
 
     def log_prob(self):
-        s  = "logp += {x}*log({mu}) - ({mu}) "
-        s += "- lgamma({x} + 1);\n"
+        s  = "if({x} < 0)\n"
+        s += "    logp = -numeric_limits<double>::max();\n"
+        s += "else\n"
+        s += "    logp += {x}*log({mu}) - ({mu}) - lgamma({x} + 1);\n"
+        return s
 
     def insert_parameters(self, s):
         s = s.replace("{mu}", str(self.mu))
+        return s
+
+
+class Binomial:
+    """
+    Binomial distributions.
+    So far only the log_prob has been implemented.
+    """
+    def __init__(self, N, theta):
+        self.N
+        self.theta = theta
+
+    def from_prior(self):
+        pass
+
+    def perturb(self):
+        pass
+
+    def log_prob(self):
+        s  = "if({x} < 0 || {x} > ({N}))\n"
+        s += "    logp = -numeric_limits<double>::max();\n"
+        s += "else\n"
+        s += "    lgamma({N} + 1) - lgamma({x} + 1) - lgamma({N} - ({x}) + 1)"
+        s += "- lgamma({x} + 1);\n"
+        return s
+
+    def insert_parameters(self, s):
+        s = s.replace("{N}", str(self.N))
+        s = s.replace("{theta}", str(self.theta))
         return s
 
 class Deterministic:

### 2016-04-06 python/dnest4/builder.py
Bugfix

@@ -172,7 +172,7 @@ class Binomial:
     So far only the log_prob has been implemented.
     """
     def __init__(self, N, theta):
-        self.N
+        self.N = N
         self.theta = theta
 
     def from_prior(self):

### 2016-04-06 python/dnest4/builder.py
Binomial

@@ -159,7 +159,7 @@ class Poisson:
         s += "    logp = -numeric_limits<double>::max();\n"
         s += "else\n"
         s += "    logp += {x}*log({mu}) - ({mu}) - lgamma({x} + 1);\n"
-        return s
+        return self.insert_parameters(s)
 
     def insert_parameters(self, s):
         s = s.replace("{mu}", str(self.mu))
@@ -185,9 +185,12 @@ class Binomial:
         s  = "if({x} < 0 || {x} > ({N}))\n"
         s += "    logp = -numeric_limits<double>::max();\n"
         s += "else\n"
-        s += "    lgamma({N} + 1) - lgamma({x} + 1) - lgamma({N} - ({x}) + 1)"
+        s += "{\n"
+        s += "    logp += lgamma({N} + 1) - lgamma({x} + 1) - lgamma({N} - ({x}) + 1)"
         s += "- lgamma({x} + 1);\n"
-        return s
+        s += "    logp += {x}*log({theta}) + ({N} - {x})*log(1.0 - ({theta}));\n"
+        s += "}\n"
+        return self.insert_parameters(s)
 
     def insert_parameters(self, s):
         s = s.replace("{N}", str(self.N))

### 2016-05-10 paper/ms.tex
A software paper

\documentclass[a4paper, 12pt]{article}
\usepackage{amsmath}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage[left=2cm, right=2cm, bottom=3cm, top=2cm]{geometry}

\title{DNest4: A C++11 implementation of Diffusive Nested Sampling}
\author{Brendon J. Brewer\\
Dan Foreman-Mackey}
\date{}

\begin{document}
\maketitle

\abstract{\noindent DNest4 is good}

% Need this after the abstract
\setlength{\parindent}{0pt}
\setlength{\parskip}{8pt}

\section{Introduction}

\begin{thebibliography}{999}
\end{thebibliography}

\end{document}

### 2016-05-10 paper/ms.tex
Some sections

@@
 
-\title{DNest4: A C++11 implementation of Diffusive Nested Sampling}
+\title{DNest4: An implementation of Diffusive Nested Sampling
+in C++11 and Python}
 \author{Brendon J. Brewer\\
@@
 
-\section{Introduction}
+\section{Bayesian computation, statistical mechanics, and Nested Sampling}
+
+\subsection{Diffusive Nested Sampling}
+
+\section{Basic usage of DNest4}
+
+\section{Advanced usage}
+
+\subsection{Multithreading}
+
+\section{Implementing models}
+
+\section{Python package}
+
+% Assign to DFM
+
+\section{Examples}
+
+\subsection{}
+
+\subsection{Approximate Bayesian Computation}
+Nested Sampling can be used to solve ``Approximate Bayesian Computation''
+problems elegantly and efficiently.
 

### 2016-05-10 paper/ms.tex
Copied and pasted a bunch of stuff from ABCNS

@@
 \usepackage[left=2cm, right=2cm, bottom=3cm, top=2cm]{geometry}
+
+\newcommand{\params}{\theta}
+\newcommand{\data}{D}
+\newcommand{\dobs}{D_{\rm obs}}
 
@@
 
+\subsection{Implementing models via {\tt generate.py}}
+
 \section{Python package}
@@
 \section{Examples}
+% Reasonably simple examples
 
@@
 
-\subsection{Approximate Bayesian Computation}
+\section{Approximate Bayesian Computation}
 Nested Sampling can be used to solve ``Approximate Bayesian Computation''
 problems elegantly and efficiently.
+
+Approximate Bayesian Computation (ABC), also known as likelihood-free inference,
+is a set of Monte Carlo techniques for approximating a posterior distribution
+without having to evaluate the likelihood function. Instead, the ability to
+generate simulated data sets from the sampling distribution is required.
+Since a sampling distribution must be specified, it is not the assumption of
+a specific likelihood that is avoided (indeed, a sampling distribution and
+a data set imply a particular likelihood function), rather that we cannot
+evaluate it cheaply as a function of $\params$ and $\data$.
+
+All Bayesian updating conditions on the truth of a proposition. We sometimes
+speak and use notation as if we're conditioning on the value of a variable, for
+example by writing the posterior distribution as $p(\params | \data)$. However,
+this is shorthand for $p(\params | \data = \dobs)$. In the prior state of
+knowledge, the proposition $(\data = \dobs)$ could have been either true or
+false, but in the posterior state of knowledge it is definitely true.
+In the case of ``continuous data'' (really a continuous {\it space of
+possibilities for the data before we learned it}) we're conditioning on a
+proposition like $(\data \in [\dobs, \dobs+\epsilon])$ and implicitly
+taking the limit $\epsilon \to 0$.
+
+\begin{table}[ht!]
+\centering
+\small
+\begin{tabular}{lll}
+\hline
+Bayesian computation		&		Statistical mechanics		&		ABC\\
+\hline
+Parameter space	$\Theta$	&		Phase space	$\Omega$ or configuration space $\Gamma$ 			& (Parameter$\times$Data) space $\Theta \times \mathcal{D}$\\		
+Marginal likelihood / evidence	&	Partition function at $T=1$	&\\
+
+\end{tabular}
+\caption{\it The relationship between standard Bayesian computation, statistical
+mechanics, and ABC. In each case Monte Carlo methods are used to calculate
+integrals over a space. In standard Bayesian computation, it is the parameter
+space of a model that is usually of interest, whereas in ABC it is the space
+of possible parameter values {\it and} data sets.
+\label{tab:relation}}
+\end{table}
+
+\begin{figure}[ht!]
+\centering
+\includegraphics[scale=0.5]{figures/joint.pdf}
+\caption{\it The joint prior for parameters $\params$ and data $\data$.
+\label{fig:joint}}
+\end{figure}
+
+The main challenges associated with ABC are:
+\begin{enumerate}
+\item The choice of summary statistics
+\item The choice of distance function
+\item How to choose the value of $\epsilon$, or how to change it as a run
+progresses
+\item How to achieve good mixing.
+\end{enumerate}
+Challenges 1 and 2 are Bayesian in nature, i.e. they relate to the very
+definition of the posterior distribution and are well defined independent of
+the fact that we are going to use a Monte Carlo method to compute the
+results. On the other hand, challenges 3 and 4 are about the computational implementation.
+
+\section{Parameterising the data space}
+In standard MCMC-based ABC, proposal moves are made by first proposing a
+change to the parameters (from $\params$ to $\params'$), and then generating
+a mock dataset $\data'$ from $p(\data | \params')$. As a Metropolis proposal
+for exploring the product space $\Theta \times \mathcal{D}$ this is a very
+aggressive choice, since the proposed value of the
+dataset $\params'$ is likely to be very different from the previous one.
+One of the most elementary pieces of advice we receive when learning the
+Metropolis algorithm is to be careful about how ambitious our proposal
+distributions are, since tiny and huge proposal moves are both very
+inefficient. The performance of MCMC-based ABC can therefore be improved by
+alternative choices of proposal distribution.
+
+In some cases, we might be able to find efficient proposal distributions for exploring $\Theta \times \mathcal{D}$ by having insight into the particular
+problem we're working on. However, there are also techniques that are
+completely general in that they should work on any ABC-type problem.
+
+Consider the process of generating a mock dataset. Whatever else is involved,
+this process will involve calls to an underlying random number generator.
+Perhaps a function {\tt rand()}, generating variables from a Uniform(0, 1)
+distribution, is called $n$ times in the course of generating a dataset.
+Usually, the simulation process has the property that, if $\theta'$ is
+close to $\theta$, then the same sequence of random numbers returned by
+{\tt rand()} would produce a dataset $\data'$ that is very similar to $\data$.
+This suggests an alternative parameterisation of
+the product space $\Theta\times\mathcal{D}$: instead of using the one that
+naturally describes the data, we can use a coordinate system defined by
+$\theta$ together with $\{u_1, ..., u_n\}$, the $n$ uniform random numbers
+used to construct a mock dataset.
+
+In this alternative coordinate system, a proposed move
+to change $\params$ while keeping
+$\{u_i\}$ fixed results in a proposed new dataset that is similar to the
+previous one, assuming the change to $\params$ is small. The standard procedure
+of generating an entirely new mock dataset each step is now viewed as a
+proposal where $\params$ is moved {\it and all of the $u_i$ coordinates are
+re-generated from their iid uniform prior}.
+
+\section*{Acknowledgements}
+It is a pleasure to thank Ewan Cameron (Oxford), Dan
+Foreman-Mackey (Washington), David Hogg (NYU), Daniela Huppenkothen (NYU),
+Scott Sisson (UNSW), and Yanan Fan (NYU) for valuable discussions.
+
+\begin{thebibliography}{99}
+\bibitem[Caticha(2008)]{caticha} Caticha, A.\ 2008.\ Lectures 
+on Probability, Entropy, and Statistical Physics.\ ArXiv e-prints 
+arXiv:0808.0012.
+
+\end{thebibliography}
 

### 2016-05-10 paper/joint.py

@@ -1,0 +1,28 @@
+import numpy as np
+from matplotlib import rc
+import matplotlib.pyplot as plt
+
+"""
+Make a plot of p(theta, D)
+"""
+
+plt.rc("font", size=18, family="serif", serif="Computer Sans")
+plt.rc("text", usetex=True)
+
+# Resolution
+N = 256
+[x, y] = np.meshgrid(np.linspace(0., 5., N), np.linspace(5., 0., N))
+
+f = np.exp(-0.5*(x-2.5)**2/1.**2)*np.exp(-0.5*((y - 5*(x/5)**2)**2)/0.3**2)
+f /= f.sum()
+
+plt.imshow(f, extent=[x.min(), x.max(), y.min(), y.max()], cmap='Blues')
+plt.xlabel(r'$\theta$')
+plt.ylabel(r'$D$')
+plt.axhline(1.3, color='k')
+plt.axhline(1.5, color='k')
+plt.fill_between(x[0, :], 1.3, 1.5, color=[0.2, 0.2, 0.2], alpha=0.2)
+plt.fill_between(x[0, :], 1.2, 1.8, color=[0.6, 0.6, 0.6], alpha=0.2)
+plt.savefig('figures/joint.pdf', bbox_inches='tight')
+#plt.show()
+

### 2016-05-10 paper/ms.tex
Fixed bibliography
minted

@@
 \usepackage{graphicx}
+\usepackage{natbib}
+\usepackage{dsfont}
+\usepackage{minted}
 \usepackage[left=2cm, right=2cm, bottom=3cm, top=2cm]{geometry}
@@
 % Reasonably simple examples
+
+\begin{minted}[mathescape,
+               numbersep=5pt,
+               gobble=0,
+               frame=single,
+               framesep=2mm, fontsize=\small]{c++}
+void StraightLine::from_prior(RNG& rng)
+{
+    // Naive diffuse prior
+    m = 1E3*rng.randn();
+    b = 1E3*rng.randn();
+
+    // Log-uniform prior
+    sigma = exp(-10. + 20.*rng.rand());
+
+    // Compute the model line
+    calculate_mu();
+}
+\end{minted}
+
+\begin{minted}[mathescape,
+               numbersep=5pt,
+               gobble=0,
+               frame=single,
+               framesep=2mm, fontsize=\small]{c++}
+double StraightLine::perturb(RNG& rng)
+{
+    double log_H = 0.;
+
+    // Proposals explore the prior
+    // For normal priors I usually use the hastings factor to do it
+    int which = rng.rand_int(3);
+    if(which == 0)
+    {
+        log_H -= -0.5*pow(m/1E3, 2);
+        m += 1E3*rng.randh();
+        log_H += -0.5*pow(m/1E3, 2);
+    }
+    else if(which == 1)
+    {
+        log_H -= -0.5*pow(b/1E3, 2);
+        b += 1E3*rng.randh();
+        log_H += -0.5*pow(b/1E3, 2);
+    }
+    else
+    {
+        // Usual log-uniform prior trick
+        sigma = log(sigma);
+        sigma += 20.*rng.randh();
+        wrap(sigma, -10., 10.);
+        sigma = exp(sigma);
+    }
+
+    // Calculate mu again if m or b changed
+    if(which == 0 || which == 1)
+        calculate_mu();
+
+    return log_H;
+}
+\end{minted}
 
@@
 \begin{thebibliography}{99}
+\bibitem[\protect\citeauthoryear{Brewer, P{\'a}rtay,
+\& Cs{\'a}nyi}{2011}]{dnest} Brewer B.~J., P{\'a}rtay L.~B., Cs{\'a}nyi G., 2011,
+Statistics and Computing, 21, 4, 649-656. arXiv:0912.2380
+
 \bibitem[Caticha(2008)]{caticha} Caticha, A.\ 2008.\ Lectures 
@@
 
-\begin{thebibliography}{999}
-\end{thebibliography}
-
 \end{document}

### 2016-05-10 paper/ms.tex
JSS stuff

@@
-\documentclass[a4paper, 12pt]{article}
+\documentclass[article]{jss}
 \usepackage{amsmath}
@@
 \usepackage{minted}
-\usepackage[left=2cm, right=2cm, bottom=3cm, top=2cm]{geometry}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% declarations for jss.cls %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%% almost as usual
+\author{Brendon J. Brewer\\The University of Auckland\And 
+        Dan Foreman-Mackey\\Washington}
+\title{\pkg{DNest4}: An implementation of Diffusive Nested Sampling in
+C++11 and Python}
+
+%% for pretty printing and a nice hypersummary also set:
+\Plainauthor{Brendon J. Brewer, Dan Foreman-Mackey} %% comma-separated
+\Plaintitle{A Capitalized Title: Something about a Package foo} %% without formatting
+\Shorttitle{\pkg{foo}: A Capitalized Title} %% a short title (if necessary)
+
+%% an abstract and keywords
+\Abstract{DNest4 is good.}
+
+\Keywords{keywords, comma-separated, not capitalized, \proglang{C++11},
+\proglang{Python}}
+\Plainkeywords{keywords, comma-separated, not capitalized, Java} %% without formatting
+%% at least one keyword must be supplied
+
+%% publication information
+%% NOTE: Typically, this can be left commented and will be filled out by the technical editor
+%% \Volume{50}
+%% \Issue{9}
+%% \Month{June}
+%% \Year{2012}
+%% \Submitdate{2012-06-04}
+%% \Acceptdate{2012-06-04}
+
+%% The address of (at least) one author should be given
+%% in the following format:
+\Address{
+  Brendon J. Brewer\\
+  Department of Statistics\\
+  The University of Auckland\\
+  Private Bag 92019\\
+  Auckland, 1142\\
+  New Zealand\\
+  E-mail: \email{bj.brewer@auckland.ac.nz}\\
+  URL: \url{https://www.stat.auckland.ac.nz/~brewer/}
+}
 
@@
 \newcommand{\dobs}{D_{\rm obs}}
-
-\title{DNest4: An implementation of Diffusive Nested Sampling
-in C++11 and Python}
-\author{Brendon J. Brewer\\
-Dan Foreman-Mackey}
-\date{}
 
@@
 
-\subsection{Diffusive Nested Sampling}
+%\subsection{Diffusive Nested Sampling}
 
-\section{Basic usage of DNest4}
+%\section{Basic usage of DNest4}
 
-\section{Advanced usage}
+%\section{Advanced usage}
 
-\subsection{Multithreading}
+%\subsection{Multithreading}
 
-\section{Implementing models}
+%\section{Implementing models}
 
-\subsection{Implementing models via {\tt generate.py}}
+%%\subsection{generate}
 
-\section{Python package}
+%\section{Python package}
 
-% Assign to DFM
+%% Assign to DFM
 
-\section{Examples}
+%\section{Examples}
 % Reasonably simple examples
@@
 \end{minted}
-
-\subsection{}
 

### 2016-05-10 paper/ms.tex
Keywords and shit

@@
 \title{\pkg{DNest4}: An implementation of Diffusive Nested Sampling in
-C++11 and Python}
+\proglang{C++11} and \proglang{Python}}
 
@@
 \Plainauthor{Brendon J. Brewer, Dan Foreman-Mackey} %% comma-separated
-\Plaintitle{A Capitalized Title: Something about a Package foo} %% without formatting
-\Shorttitle{\pkg{foo}: A Capitalized Title} %% a short title (if necessary)
+\Plaintitle{DNest4: An implementation of Diffusive Nested Sampling in
+C++11 and Python} %% without formatting
+\Shorttitle{\pkg{DNest4}: Diffusive Nested Sampling} %% a short title (if necessary)
 
@@
 
-\Keywords{keywords, comma-separated, not capitalized, \proglang{C++11},
-\proglang{Python}}
-\Plainkeywords{keywords, comma-separated, not capitalized, Java} %% without formatting
+\Keywords{bayesian inference, markov chain monte carlo,
+metropolis algorithm, bayesian computation, nested sampling, \proglang{c++11},
+\proglang{python}}
+\Plainkeywords{bayesian inference, markov chain monte carlo,
+metropolis algorithm, bayesian computation, nested sampling, c++11, python} %% without formatting
 %% at least one keyword must be supplied

### 2016-05-10 paper/ms.tex
Bye bye minted

@@
 \usepackage{dsfont}
-\usepackage{minted}
 
@@
 
-\begin{minted}[mathescape,
-               numbersep=5pt,
-               gobble=0,
-               frame=single,
-               framesep=2mm, fontsize=\small]{c++}
+The member function used to generate straight line parameters from the
+prior is:
+
+\begin{CodeChunk}
+\begin{CodeInput}
 void StraightLine::from_prior(RNG& rng)
@@
    // Log-uniform prior
-    sigma = exp(-10. + 20.*rng.rand());
+   sigma = exp(-10.0 + 20.0*rng.rand());
 
@@
 }
-\end{minted}
-
-\begin{minted}[mathescape,
-               numbersep=5pt,
-               gobble=0,
-               frame=single,
-               framesep=2mm, fontsize=\small]{c++}
-double StraightLine::perturb(RNG& rng)
-{
-    double log_H = 0.;
-
-    // Proposals explore the prior
-    // For normal priors I usually use the hastings factor to do it
-    int which = rng.rand_int(3);
-    if(which == 0)
-    {
-        log_H -= -0.5*pow(m/1E3, 2);
-        m += 1E3*rng.randh();
-        log_H += -0.5*pow(m/1E3, 2);
-    }
-    else if(which == 1)
-    {
-        log_H -= -0.5*pow(b/1E3, 2);
-        b += 1E3*rng.randh();
-        log_H += -0.5*pow(b/1E3, 2);
-    }
-    else
-    {
-        // Usual log-uniform prior trick
-        sigma = log(sigma);
-        sigma += 20.*rng.randh();
-        wrap(sigma, -10., 10.);
-        sigma = exp(sigma);
-    }
-
-    // Calculate mu again if m or b changed
-    if(which == 0 || which == 1)
-        calculate_mu();
-
-    return log_H;
-}
-\end{minted}
+\end{CodeInput}
+\end{CodeChunk}
 

### 2016-05-10 paper/ms.tex
An abstract

@@
 \author{Brendon J. Brewer\\The University of Auckland\And 
-        Dan Foreman-Mackey\\Washington}
-\title{\pkg{DNest4}: An implementation of Diffusive Nested Sampling in
-\proglang{C++11} and \proglang{Python}}
+        Dan Foreman-Mackey\\University of Washington}
+\title{\pkg{DNest4}: Diffusive Nested Sampling in
+\proglang{C++} and \proglang{Python}}
 
 %% for pretty printing and a nice hypersummary also set:
-\Plainauthor{Brendon J. Brewer, Dan Foreman-Mackey} %% comma-separated
+\Plainauthor{Brendon J. Brewer, Daniel Foreman-Mackey} %% comma-separated
 \Plaintitle{DNest4: An implementation of Diffusive Nested Sampling in
@@
 %% an abstract and keywords
-\Abstract{DNest4 is good.}
+\Abstract{In probabilistic (Bayesian) inferences, we typically want to compute
+properties of the posterior distribution, describing knowledge of
+unknown quantities in the context of a particular dataset and the assumed
+prior information. The marginal likelihood (also known as the ``evidence'')
+is a key quantity in Bayesian model averaging. The Diffusive Nested Sampling algorithm, a variant of Nested Sampling, is a powerful tool for generating
+posterior samples and estimating marginal likelihoods. It is effective at
+solving complex problems including many where the posterior distribution is
+multimodal or has strong dependencies between variables. The key requirement
+is that the likelihood function be fast to evaluate.
+\pkg{DNest4} is a multi-threaded implementation of this algorithm in
+modern \proglang{C++} (specifically, \proglang{C++11}),
+along with associated utilities including: i)
+a \pkg{Python} package allowing basic use without \proglang{C++} coding;
+ii) \pkg{RJObject}, a class template for finite mixture models;
+iii) Experimental support for models implemented in \proglang{Julia};
+iv) An experimental tool \proglang{Python} tool to
+generate \pkg{C++} code from a model specification.
+In this paper we demonstrate \pkg{DNest4} usage through examples including
+simple Bayesian data analysis, finite mixture models, and Approximate
+Bayesian Computation (ABC).
+}
 

### 2016-05-10 paper/ms.tex
Metropolis

@@
 \author{Brendon J. Brewer\\The University of Auckland\And 
-        Dan Foreman-Mackey\\University of Washington}
+        Daniel Foreman-Mackey\\University of Washington}
 \title{\pkg{DNest4}: Diffusive Nested Sampling in
@@
 
-\abstract{\noindent DNest4 is good}
-
 % Need this after the abstract
@@
 
-\section{Bayesian computation, statistical mechanics, and Nested Sampling}
+\section{Bayesian computation and Nested Sampling}
+
+\subsection{Markov Chain Monte Carlo}
+In the Metropolis algorithm, the acceptance probability $\alpha$
+is given by
+\begin{align}
+\alpha &= \min\left(1,
+\frac{q(\params'|\params)}{q(\params | \params')}
+\frac{\pi(\params')}{\pi(\params)}\frac{L(\params')}{L(\params)}
+\right)
+\end{align}
+where $q(\theta' | \theta)$ is the proposal distribution used to generate
+a new position $\theta'$ from the current position $\theta$. Often,
+$q$ is {\em symmetric} so the $q$ terms cancel in the acceptance probability.
 

### 2016-05-10 paper/ms.tex
Metropolis for NS

@@
 $q$ is {\em symmetric} so the $q$ terms cancel in the acceptance probability.
+
+In Nested Sampling, the target distribution is not the posterior but rather
+the constrained prior. Hence, the appropriate acceptance probability is
+\begin{align}
+\alpha &= \min\left[1,
+\frac{q(\params'|\params)}{q(\params | \params')}
+\frac{\pi(\params')}{\pi(\params)}
+\mathds{1}\left(L(\params') > L^*\right)
+\right]
+\end{align}
+where $L^*$ is the threshold value of the likelihood.
+
+The \pkg{DNest4} \code{Sampler} class handles the likelihood check,
+while the \code{perturb(DNest4::RNG\&)} member function handles the
+rest of the terms. The value returned from \code{perturb(DNest4::RNG\&)}
+must be
+$\ln\left[q(\params'|\params)/q(\params'|\params)\times\pi(\params')/\pi(\params)\right]$ for your problem.
 

### 2016-05-10 paper/ms.tex
Notation

@@
 
+The posterior distribution
+\begin{align}
+p(\params | \data) &= \frac{p(\params)p(\data | \params)}{p(\data)}\\
+&= \frac{\pi(\params)L(\params)}{Z}
+\end{align}
+
 \subsection{Markov Chain Monte Carlo}
@@
 is a set of Monte Carlo techniques for approximating a posterior distribution
-without having to evaluate the likelihood function. Instead, the ability to
+without having to evaluate the likelihood function $L(\theta)$.
+
+. Instead, the ability to
 generate simulated data sets from the sampling distribution is required.

### 2016-05-10 paper/ms.tex
Write about the code

@@
 
-The posterior distribution
+In very compact notation, the posterior
+posterior distribution for parameters $\params$ given data $\data$, within
+context $I$, is
 \begin{align}
-p(\params | \data) &= \frac{p(\params)p(\data | \params)}{p(\data)}\\
-&= \frac{\pi(\params)L(\params)}{Z}
+p(\params | \data, I) &=
+\frac{p(\params | I)p(\data | \params, I)}{p(\data | I)}
 \end{align}
+or
+\begin{align}
+\textnormal{posterior} &=
+\frac{\textnormal{prior} \times \textnormal{likelihood}}
+     {\textnormal{marginal likelihood}}.
+\end{align}
+When discussing computational matters, the prior is
+often written $\pi(\theta)$, the likelihood $L(\theta)$,
+and the marginal likelihood $Z$.
 
@@
 
-%\section{Basic usage of DNest4}
+\section{Basic usage of DNest4}
+The
 
-%\section{Advanced usage}
+\section{Advanced usage}
 
@@
 
-%\section{Implementing models}
+\section{Implementing models}
 
@@
 \begin{CodeInput}
-void StraightLine::from_prior(RNG& rng)
+void StraightLine::from_prior(DNest4::RNG\& rng)
 {
@@
 \end{CodeChunk}
+
+This generates $m$, $b$, and $\sigma$ from their joint prior. In this case,
+the priors are all independent, so this reduces to generating the parameters
+each from their own prior distribution. For convenience, the \code{RNG} object
+has \code{rand()} and \code{randn()} member functions to generate from
+a Uniform$(0,1)$ and Normal$(0,1)$ distribution respectively.
+
+The \code{perturb} member function is:
+
+When implementing a model class, it is imperative that the
+\code{from\_prior(DNest4::RNG\&)} and \code{perturb(DNest4::RNG\&)}
+member functions be consistent with each other.
 

### 2016-05-10 paper/ms.tex
Writing about proposals

@@
 
-The \code{perturb} member function is:
+After generating the parameters from the prior, another member function
+\code{calculate\_mu()} is called, which computes the model prediction
+$mx_i + b$ for each data point.
+
+The \code{perturb} member function for the straight line model is:
+\begin{CodeChunk}
+\begin{CodeInput}
+double StraightLine::perturb(DNest4::RNG& rng)
+{
+	double log_H = 0.0;
+
+	// Proposals explore the prior
+	// For normal priors I usually use the hastings factor to do it
+	int which = rng.rand_int(3);
+	if(which == 0)
+	{
+		log_H -= -0.5*pow(m/1E3, 2);
+		m += 1E3*rng.randh();
+		log_H += -0.5*pow(m/1E3, 2);
+	}
+	else if(which == 1)
+	{
+		log_H -= -0.5*pow(b/1E3, 2);
+		b += 1E3*rng.randh();
+		log_H += -0.5*pow(b/1E3, 2);
+	}
+	else
+	{
+		// Usual log-uniform prior trick
+		sigma = log(sigma);
+		sigma += 20.0*rng.randh();
+		DNest4::wrap(sigma, -10.0, 10.0);
+		sigma = exp(sigma);
+	}
+
+	// Calculate mu again if m or b changed
+	if(which == 0 || which == 1)
+		calculate_mu();
+
+	return log_H;
+}
+\end{CodeInput}
+\end{CodeChunk}
+
+This first chooses a random integer from $[0, 3)$ using the
+\code{rand\_int(int)} member function of the \code{DNest4::RNG} class.
 
@@
 \code{from\_prior(DNest4::RNG\&)} and \code{perturb(DNest4::RNG\&)}
-member functions be consistent with each other.
+member functions be consistent with each other. One technique
+for verifying this is to sample the prior for a long time
+(by setting the maximum number of levels to 1) and inspect
+{\tt sample.txt} to ensure that each parameter is exploring
+the correct distribution.
 

### 2016-05-10 paper/ms.tex
The model specification

@@
 \section{Implementing models}
+The ``classic'' method of implementing models in \pkg{DNest4} is by
+writing a \proglang{C++} class, an object of which represents a
+point in your parameter space.
+
+To demonstrate, consider a simple linear regression situation where the
+sampling distribution is
+\begin{align}
+y_i | m, b, \sigma &\sim \textnormal{Normal}(mx_i + b, \sigma^2)
+\end{align}
+and the priors are
+\begin{align}
+m &\sim \textnormal{Normal}(0, 1000^2)\\
+b &\sim \textnormal{Normal}(0, 1000^2)\\
+\ln\sigma &\sim \textnormal{Uniform}(-10, 10)
+\end{align}
+These are naive diffuse priors and we do not claim any special status for them.
 
@@
 \code{rand\_int(int)} member function of the \code{DNest4::RNG} class.
+This determines which of the three parameters is modified.
 

### 2016-05-10 paper/ms.tex
DOn't need escape character in code block
List of member functions required

@@
 iii) Experimental support for models implemented in \proglang{Julia};
-iv) An experimental tool \proglang{Python} tool to
+iv) An experimental \proglang{Python} tool to
 generate \pkg{C++} code from a model specification.
@@
 \mathds{1}\left(L(\params') > L^*\right)
-\right]
+\right]\\
+&= \min\left[1,
+e^{\ln(H)}
+\mathds{1}\left(L(\params') > L^*\right)
+\right]\label{eqn:log_hastings}
 \end{align}
@@
 must be
-$\ln\left[q(\params'|\params)/q(\params'|\params)\times\pi(\params')/\pi(\params)\right]$ for your problem.
+$\ln\left[q(\params'|\params)/q(\params'|\params)\times\pi(\params')/\pi(\params)\right]$
+for your problem.
 
@@
 point in your parameter space.
-
 To demonstrate, consider a simple linear regression situation where the
@@
 These are naive diffuse priors and we do not claim any special status for them.
+
+Running \pkg{DNest4} on this problem requires a class, which we call
+{\tt StraightLine}, which has the following member functions:
+\begin{itemize}
+\item \code{void from\_prior(DNest4::RNG\& rng)}, which generates parameter
+        values from the prior;
+\item \code{double perturb(DNest4::RNG\& rng)}, which proposes a
+        change to the parameter values, and returns $\ln(H)$ as defined in
+        Equation~\ref{eqn:log_hastings};
+\item \code{double log\_likelihood() const}, which evaluates the log of
+        the likelihood function;
+\item \code{void print(std::ostream\& out) const}, which prints parameters of
+        interest to the given output stream; and
+\item \code{std::string description() const}, which returns a \proglang{C++}
+        string naming the parameters printed by \code{print(std::ostream\&) const}.
+\end{itemize}
 
@@
 \begin{CodeInput}
-void StraightLine::from_prior(DNest4::RNG\& rng)
+void StraightLine::from_prior(DNest4::RNG& rng)
 {

### 2016-05-10 paper/ms.tex
enumerate with small roman numerals
Table

@@
 \usepackage{natbib}
+\usepackage{enumerate}
 \usepackage{dsfont}
@@
 {\tt StraightLine}, which has the following member functions:
-\begin{itemize}
+\begin{enumerate}[(i)]
 \item \code{void from\_prior(DNest4::RNG\& rng)}, which generates parameter
@@
         string naming the parameters printed by \code{print(std::ostream\&) const}.
-\end{itemize}
+\end{enumerate}
 
@@
    sigma = exp(-10.0 + 20.0*rng.rand());
-
-   // Compute the model line
-   calculate_mu();
 }
@@
 	}
-
-	// Calculate mu again if m or b changed
-	if(which == 0 || which == 1)
-		calculate_mu();
 
@@
 \small
-\begin{tabular}{lll}
+\begin{tabular}{|l|l|l|}
 \hline
@@
 \hline
-Parameter space	$\Theta$	&		Phase space	$\Omega$ or configuration space $\Gamma$ 			& (Parameter$\times$Data) space $\Theta \times \mathcal{D}$\\		
-Marginal likelihood / evidence	&	Partition function at $T=1$	&\\
-
+Parameter space	$\Theta$	&		Phase space	$\Omega$ 			& Product space $\Theta \times \mathcal{D}$\\
+Posterior $p(\params | D,I)$   &  Canonical distribution $p(\boldsymbol{x}; T)$     & \\ 
+Marginal likelihood $P(D|I)$	&	Partition function $Z(T)$	&   Marginal likelihood $p(D\in | I)$\\
+\hline
 \end{tabular}
@@
 The main challenges associated with ABC are:
-\begin{enumerate}
+\begin{enumerate}[(a)]
 \item The choice of summary statistics

### 2016-05-11 paper/ms.tex
Intro words on statmech

@@
 often written $\pi(\theta)$, the likelihood $L(\theta)$,
-and the marginal likelihood $Z$.
+and the marginal likelihood $Z$. A popular alternative name for the marginal
+likelihood, which emphasizes its role in Bayesian model averaging,
+is the ``evidence''.
+
+Nested Sampling \citep[NS;][]{skilling} is a Monte Carlo method whose main
+aim was to calculate $Z$. However, it can also be used to generate samples
+to represent the posterior distribution $\pi(\theta)L(\theta)/Z$, or
+any other distribution proportional to
+$\pi(\theta)\Phi\left[L(\theta)\right]$ where $\Phi$ is any monotonic function.
+This latter property makes NS particularly useful for statistical mechanics
+calculations \citep{2009arXiv0906.3544P, 2015arXiv150303404B}, where the
+``canonical'' family of distributions proportional to
+$\pi(\theta)L(\theta)^\beta$ is of interest (and $L(\theta)$ is identified as
+$\exp(-\textnormal{energy})$). NS is particularly efficient for this, since
+only a single run (which explores regions of higher and higher $L$)
+is required, and different canonical
+distributions can be obtained by re-weighting the output points.
 
@@
 \begin{thebibliography}{99}
+\bibitem[Baldock et al.(2015)]{2015arXiv150303404B} Baldock, R.~J.~N., 
+P{\'a}rtay, L.~v.~B., Bart{\'o}k, A.~P., Payne, M.~C., Cs{\'a}nyi, G.\ 
+2015.\ Determining pressure-temperature phase diagrams of materials.\ ArXiv 
+e-prints arXiv: 1503.03404.
+
 \bibitem[\protect\citeauthoryear{Brewer, P{\'a}rtay,
@@
 
+\bibitem[P{\'a}rtay et al.(2009)]{2009arXiv0906.3544P} P{\'a}rtay, L.~B., 
+Bart{\'o}k, A.~P., Cs{\'a}nyi, G.\ 2009.\ Efficient sampling of atomic 
+configurational spaces.\ ArXiv e-prints arXiv: 0906.3544.
+
+\bibitem[\protect\citeauthoryear{Skilling}{2006}]{skilling} Skilling, J., 2006, Nested Sampling for General Bayesian Computation, Bayesian Analysis 4, pp. 833-860.
+
 \end{thebibliography}

### 2016-05-11 paper/ms.tex
Talking about the class

@@
 
-Running \pkg{DNest4} on this problem requires a class, which we call
-{\tt StraightLine}, which has the following member functions:
+To run \pkg{DNest4} on any particular problem, such as this linear regression
+example, the user needs to define a \proglang{C++} class to specify the
+model. Specifically, an object of the class represents a point in the
+model's parameter space. Member functions are defined which generate
+the object's parameters from the prior, make proposal steps, evaluate the
+likelihood, and so on. The sampler calls these member functions while
+executing a run.
+
+For the simple linear regression example, we will call the class
+{\tt StraightLine}. The member variables representing the
+unknown parameters are defined in the header file {\tt StraightLine.h}:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+class StraightLine
+{
+    private:
+        double m, b, sigma;
+};
+\end{CodeInput}
+\end{CodeChunk}
+
+ which has the following member functions:
 \begin{enumerate}[(i)]

### 2016-05-11 paper/ms.tex
SMall amount of writing
Mixture of constrained priors

@@
 Nested Sampling \citep[NS;][]{skilling} is a Monte Carlo method whose main
-aim was to calculate $Z$. However, it can also be used to generate samples
+aim is to calculate $Z$. However, it can also be used to generate samples
 to represent the posterior distribution $\pi(\theta)L(\theta)/Z$, or
@@
 distributions can be obtained by re-weighting the output points.
+
+A defining feature of NS is that it works with a sequence of
+{\em constrained prior} distributions, proportional to $\pi$ but
+restricted to regions of the parameter space where $L(\theta)$
+is above some threshold $\ell$:
+\begin{align}
+p(\theta; \ell) &=
+\frac{\pi(\theta)\mathds{1}\left[L(\theta) > \ell\right]}{X(\ell)}
+\label{eqn:constrained_prior}
+\end{align}
+where
+\begin{align}
+X(\ell) &= \int \pi(\theta) \mathds{1}\left[L(\theta) > \ell\right] \, d\theta
+\end{align}
+is the amount of prior mass which has likelihood greater than $\ell$.
+In the standard NS framework, the sequence of $\ell$ values is selected
+so that $X(\ell)$ shrinks by a factor $\approx e^{-1/N}$ per iteration, where
+$N$ is the number of particles used.
+Sampling from the constrained priors (Equation~\ref{eqn:constrained_prior})
+is required, and Markov Chain Monte Carlo (MCMC) is a popular method of doing
+this, although alternatives exist \citep[e.g.][]{multinest}.
+
+Diffusive Nested Sampling \citep[DNS][]{dnest} is an alternative to NS for
+problems where MCMC is the only viable sampling method. DNS is based on the
+Metropolis algorithm. DNS evolves a point in the parameter space, along with
+an integer index variable $j$,
+to explore the following joint distribution:
+\begin{align}
+p(\theta, j) &= p(j)p(\theta | j)\\
+&= w_j \times
+\frac{\pi(\theta)\mathds{1}\left[L(\theta) > \ell_j\right]}{X(\ell_j)}.
+\end{align}
+where $\{\ell_j\}$ is a sequence of increasing likelihood thresholds
+or {\em levels}, and
+$\ell_0 = 0$, and $\{w_j\}$ is the marginal distribution for $j$.
+The marginal distribution for $\theta$ is then a {\em mixture} of
+constrained priors:
+\begin{align}
+p(\theta) &=
+\pi(\theta)\sum_{j=0}^{j_{\rm max}}
+\frac{w_j\mathds{1}\left[L(\theta) > \ell_j\right]}{X(\ell_j)}.
+\label{eqn:mixture_of_constrained_priors}
+\end{align}
 
@@
 
- which has the following member functions:
+The class must also define and implement the following member functions:
 \begin{enumerate}[(i)]
@@
         string naming the parameters printed by \code{print(std::ostream\&) const}.
+        This string is printed (after a comment character \#) at the top of
+        the output file.
 \end{enumerate}
@@
 
+\bibitem[\protect\citeauthoryear{Feroz, Hobson,
+\& Bridges}{2009}]{multinest} Feroz F., Hobson M.~P., Bridges M., 2009, MNRAS, 398, 1601
+
 \bibitem[P{\'a}rtay et al.(2009)]{2009arXiv0906.3544P} P{\'a}rtay, L.~B., 

### 2016-05-11 paper/ms.tex
Likelihood and print

@@
 
+The log-likelihood code is:
+\begin{CodeChunk}
+\begin{CodeInput}
+double StraightLine::log_likelihood() const
+{
+    // Grab the dataset
+    const std::vector<double>& x = Data::get_instance().get_x();
+    const std::vector<double>& y = Data::get_instance().get_y();
+
+    // Variance
+    double var = sigma*sigma;
+
+    // Conventional gaussian sampling distribution
+    double log_L = 0.0;
+    double mu;
+    for(size_t i=0; i<y.size(); ++i)
+    {
+        mu = m*x[i] + b;
+        log_L += -0.5*log(2*M_PI*var) - 0.5*pow(y[i] - mu, 2)/var;
+    }
+
+    return log_L;
+}
+\end{CodeInput}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+\begin{CodeInput}
+void StraightLine::print(std::ostream& out) const
+{
+    out<<m<<' '<<b<<' '<<sigma;
+}
+
+std::string StraightLine::description() const
+{
+    return std::string("m, b, sigma");
+}
+
+\end{CodeInput}
+\end{CodeChunk}
+
 \section{Approximate Bayesian Computation}

### 2016-05-11 paper/ms.tex
Install

@@
 \section{Basic usage of DNest4}
-The
+\pkg{DNest4} can be obtained from the \pkg{git} repository at
+{\tt https://github.com/eggplantbren/DNest4/} and is licensed under the
+MIT license. To obtain \pkg{DNest4} and compile and install the software,
+the following steps are sufficient:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+git clone https://github.com/eggplantbren/DNest4
+cd DNest4/code
+make
+cd ../python
+python setup.py install
+\end{CodeInput}
+\end{CodeChunk}
 
@@
 \end{CodeChunk}
+The dataset is assumed to be accessible inside this function. In the
+regression example, this is achieved by having a Data class to represent
+datasets. The dataset using the {\em singleton pattern}.
 

### 2016-05-11 paper/ms.tex
Dependencies

@@
 
-\section{Basic usage of DNest4}
+\section{Dependencies and installation}
 \pkg{DNest4} can be obtained from the \pkg{git} repository at
@@
 MIT license. To obtain \pkg{DNest4} and compile and install the software,
+you require \pkg{git} and a recent version of \pkg{g++}. The \proglang{Python}
+packages \pkg{NumPy}, \pkg{matplotlib},
+\pkg{Pandas}, and \pkg{Cython} are also needed.
+To download and compile \pkg{DNest4}
 the following steps are sufficient:

### 2016-05-11 paper/ms.tex
Singleton pattern

@@
 The dataset is assumed to be accessible inside this function. In the
-regression example, this is achieved by having a Data class to represent
-datasets. The dataset using the {\em singleton pattern}.
+regression example, this is achieved by having a
+\code{Data} class to represent
+datasets. Since there will usually only be one dataset,
+the {\em singleton pattern} (a class of which there is one instance
+accessible from anywhere) is recommended.
 
+The print and description functions are very simple:
 \begin{CodeChunk}

### 2016-05-12 paper/ms.tex
More writing about acceptance probability

@@
 \frac{\pi(\theta)\mathds{1}\left[L(\theta) > \ell_j\right]}{X(\ell_j)}.
+\label{eqn:target_distribution}
 \end{align}
-where $\{\ell_j\}$ is a sequence of increasing likelihood thresholds
+where the $\{\ell_j\}$ are a sequence of increasing likelihood thresholds
 or {\em levels}, and
@@
 
-In Nested Sampling, the target distribution is not the posterior but rather
-the constrained prior. Hence, the appropriate acceptance probability is
+In DNS, the target distribution is not the posterior but rather
+the joint distribution in Equation~\ref{eqn:target_distribution}.
+Moves of $\theta$ are done keeping $j$ fixed, so we only need
+to consider the Metropolis acceptance probability for fixed $j$,
+i.e. with respect to a single constrained prior like
+Equation~\ref{eqn:constrained_prior}.
+Hence, the appropriate acceptance probability
+for a proposed move from $\theta$ to $\theta'$ is
 \begin{align}
@@
 \frac{\pi(\params')}{\pi(\params)}
-\mathds{1}\left(L(\params') > L^*\right)
+\mathds{1}\left(L(\params') > \ell_j\right)
 \right]\\
@@
 e^{\ln(H)}
-\mathds{1}\left(L(\params') > L^*\right)
+\mathds{1}\left(L(\params') > \ell_j\right)
 \right]\label{eqn:log_hastings}
 \end{align}
-where $L^*$ is the threshold value of the likelihood.
+where $\ell_j$ is the likelihood threshold for level $j$.
 

### 2016-05-12 paper/ms.tex
Slight editing

@@
 \section{Dependencies and installation}
-\pkg{DNest4} can be obtained from the \pkg{git} repository at
-{\tt https://github.com/eggplantbren/DNest4/} and is licensed under the
-MIT license. To obtain \pkg{DNest4} and compile and install the software,
+\pkg{DNest4} can be obtained from the \pkg{git} repository located at
+\begin{center}
+{\tt https://github.com/eggplantbren/DNest4/}\\
+\end{center}
+and is licensed under the
+MIT license. To obtain and compile \pkg{DNest4},
 you require \pkg{git} and a recent version of \pkg{g++}. The \proglang{Python}

### 2016-05-13 paper/ms.tex
A bit of intro writing

@@
 
-In very compact notation, the posterior
+Bayesian inference, where probability theory is used to describe degrees of
+logical implication or subjective certainty, provides a powerful general basis
+for data analysis \citep{ohagan, sivia}. The result of such
+an analysis is typically
+posterior probabilities of various hypotheses, or (more specifically)
+a joint posterior probability distribution for the values of unknown
+parameters.
+
+In a very compact (but standard) notation, the posterior
 posterior distribution for parameters $\params$ given data $\data$, within
-context $I$, is
+context $M$, is
 \begin{align}
-p(\params | \data, I) &=
-\frac{p(\params | I)p(\data | \params, I)}{p(\data | I)}
+p(\params | \data, M) &=
+\frac{p(\params | M)p(\data | \params, M)}{p(\data | M)}
 \end{align}
@@
 
+\bibitem[\protect\citeauthoryear{O'Hagan and Forster}{2004}]{ohagan}
+O'Hagan, A., Forster,~J., 2004, Bayesian inference. London: Arnold.
+
+\bibitem[\protect\citeauthoryear{Sivia and Skilling}{2006}]{sivia} Sivia, 
+D.~ S., Skilling, J., 2006, Data Analysis: A Bayesian Tutorial, 2nd 
+Edition, Oxford University Press
+
 \bibitem[P{\'a}rtay et al.(2009)]{2009arXiv0906.3544P} P{\'a}rtay, L.~B., 

### 2016-05-13 paper/ms.tex
open source

@@
 is that the likelihood function be fast to evaluate.
-\pkg{DNest4} is a multi-threaded implementation of this algorithm in
+\pkg{DNest4} is an open source,
+multi-threaded implementation of this algorithm in
 modern \proglang{C++} (specifically, \proglang{C++11}),

### 2016-05-13 paper/ms.tex
model averaging

--- 2016-05-13 paper/ms.tex
@@
 ii) \pkg{RJObject}, a class template for finite mixture models;
-iii) Experimental support for models implemented in \proglang{Julia};
+iii) Experimental support for models implemented in \proglang{Julia}; and
 iv) An experimental \proglang{Python} tool to
@@
 posterior distribution for parameters $\params$ given data $\data$, within
-context $M$, is
+the context of prior information $M$, is
 \begin{align}
@@
 \end{align}
+
+If prior information $I$ (hereby dropped)
+implies a set of possible `models' $\{M_i\}$,
+rather than a single one $M$, the posterior model probabilities are given by
+\begin{align}
+P(M_i | \data) &=
+\frac{P(M_i)p(\data | M_i)}{\sum_j P(M_j)p(\data | M_j)}
+\end{align}
+where
+\begin{align}
+p(\data | M_j) &= \int p(\theta_j | M_j)p(\data | \theta_j, M_j) \, d\theta_j
+\end{align}
+is the marginal likelihood of model $j$. This kind of calculation is often
+called ``model selection'' or (more accurately) ``model averaging''.
+
 When discussing computational matters, the prior is

### 2016-05-13 paper/ms.tex
random edits

@@
 is that the likelihood function be fast to evaluate.
-\pkg{DNest4} is an open source,
+\pkg{DNest4} is an open source (MIT licensed),
 multi-threaded implementation of this algorithm in
@@
 
-When discussing computational matters, the prior is
+When discussing computational matters, the prior distribution for parameters
+is
 often written $\pi(\theta)$, the likelihood $L(\theta)$,
@@
 $N$ is the number of particles used.
+
 Sampling from the constrained priors (Equation~\ref{eqn:constrained_prior})
@@
 \section{Dependencies and installation}
+The following instructions apply to Unix-like operating systems such as
+GNU/Linux, Mac OS X, and FreeBSD.
+
 \pkg{DNest4} can be obtained from the \pkg{git} repository located at
@@
 \section*{Acknowledgements}
-It is a pleasure to thank Ewan Cameron (Oxford), Dan
-Foreman-Mackey (Washington), David Hogg (NYU), Daniela Huppenkothen (NYU),
-Scott Sisson (UNSW), and Yanan Fan (NYU) for valuable discussions.
+It is a pleasure to thank Anna Pancoast (Harvard),
+Ewan Cameron (Oxford), David Hogg (NYU), and Daniela Huppenkothen (NYU)
+for valuable discussions.
 

### 2016-05-13 code/Distributions/Cauchy.cpp
M_PI is non-standard!

@@ -1,6 +1,7 @@
 #include "Cauchy.h"
 #include <stdexcept>
 #include <cmath>
+#include "Utils.h"
 
 namespace DNest4
 {

### 2016-05-13 code/Utils.h

@@ -11,6 +11,12 @@
 namespace DNest4
 {
 
+// This is non-standard, gcc supports it but mingw64 doesn't,
+// so putting it here
+#ifndef M_PI
+    constexpr double M_PI = 3.141592653589793;
+#endif
+
 double mod(double y, double x);
 int mod(int y, int x);
 void wrap(double& x, double min, double max);

### 2016-05-14 paper/ms.tex
Slight edits
Handley reference

@@
 is required, and Markov Chain Monte Carlo (MCMC) is a popular method of doing
-this, although alternatives exist \citep[e.g.][]{multinest}.
+this, although alternatives exist \citep[e.g.][]{multinest, handley}.
 
@@
 GNU/Linux, Mac OS X, and FreeBSD.
-
 \pkg{DNest4} can be obtained from the \pkg{git} repository located at
@@
 MIT license. To obtain and compile \pkg{DNest4},
-you require \pkg{git} and a recent version of \pkg{g++}. The \proglang{Python}
+you require \pkg{git} and a recent version of the GNU
+\proglang{C++} compiler, \pkg{g++}. The \proglang{Python}
 packages \pkg{NumPy}, \pkg{matplotlib},
@@
 the following steps are sufficient:
-
 \begin{CodeChunk}
@@
 
+\bibitem[Handley et al.(2015)]{2015MNRAS.453.4384H} Handley, W.~J., Hobson, M.~P., Lasenby, A.~N.\ 2015.\ POLYCHORD: next-generation nested sampling.\ Monthly Notices of the Royal Astronomical Society 453, 4384-4398.
+
 \bibitem[\protect\citeauthoryear{O'Hagan and Forster}{2004}]{ohagan}

### 2016-05-14 paper/ms.tex
Cleaning up alpha discussion

@@
 \subsection{Markov Chain Monte Carlo}
-In the Metropolis algorithm, the acceptance probability $\alpha$
+DNS is build upon the Metropolis-Hastings algorithm.
+In this algorithm, the acceptance probability $\alpha$
 is given by
@@
 \mathds{1}\left(L(\params') > \ell_j\right)
-\right]\\
-&= \min\left[1,
-e^{\ln(H)}
-\mathds{1}\left(L(\params') > \ell_j\right)
-\right]\label{eqn:log_hastings}
+\right]
+\label{eqn:log_hastings}
 \end{align}
@@
 
-The \pkg{DNest4} \code{Sampler} class handles the likelihood check,
-while the \code{perturb(DNest4::RNG\&)} member function handles the
-rest of the terms. The value returned from \code{perturb(DNest4::RNG\&)}
-must be
-$\ln\left[q(\params'|\params)/q(\params'|\params)\times\pi(\params')/\pi(\params)\right]$
-for your problem.
+For convenience later on, we 
+separate the prior and proposal-related terms from
+the likelihood-related term, and write the former as
+\begin{align}
+H = \frac{q(\params'|\params)}{q(\params | \params')}
+\times \frac{\pi(\params')}{\pi(\params)}
+\end{align}
+so the acceptance probability becomes
+\begin{align}
+\alpha &= \min\left[1,
+H\times
+\mathds{1}\left(L(\params') > \ell_j\right)
+\right]
+\end{align}
+
+%The \pkg{DNest4} \code{Sampler} class handles the likelihood check,
+%while the \code{perturb(DNest4::RNG\&)} member function handles the
+%rest of the terms. The value returned from \code{perturb(DNest4::RNG\&)}
+%must be
+%for your problem.
 
@@
 \section{Dependencies and installation}
+
+\subsection{Unix-like operating systems}
 The following instructions apply to Unix-like operating systems such as
@@
 \end{CodeChunk}
+
+\subsection{Microsoft Windows}
 

### 2016-05-14 paper/ms.tex
Application references

@@
 \end{align}
+
+DNS has been applied several times in astrophysics
+\citep[e.g.][]{2014MNRAS.445.3055P, 2015ApJ...810...66H, 2015MNRAS.448.3206B}
+and was recently used in a biological application \citep{salmonella}.
 
@@
 
+\bibitem[Brewer and Donovan(2015)]{2015MNRAS.448.3206B} Brewer, B.~J., Donovan, C.~P.\ 2015.\ Fast Bayesian inference for exoplanet discovery in radial velocity data.\ Monthly Notices of the Royal Astronomical Society 448, 3206-3214.
+
 \bibitem[Caticha(2008)]{caticha} Caticha, A.\ 2008.\ Lectures 
@@
 arXiv:0808.0012.
+
+\bibitem[Dybowski et al(2015)]{salmonella}
+Dybowski, Richard, et al. Single passage in mouse organs enhances the survival and spread of Salmonella enterica. Journal of The Royal Society Interface 12.113 (2015): 20150702.
 
@@
 
+\bibitem[Huppenkothen et al.(2015)]{2015ApJ...810...66H} Huppenkothen, D., Brewer, B.~J., Hogg, D.~W., Murray, I., Frean, M., Elenbaas, C., Watts, A.~L., Levin, Y., van der Horst, A.~J., Kouveliotou, C.\ 2015.\ Dissecting Magnetar Variability with Bayesian Hierarchical Models.\ The Astrophysical Journal 810, 66.
+
 \bibitem[\protect\citeauthoryear{O'Hagan and Forster}{2004}]{ohagan}
 O'Hagan, A., Forster,~J., 2004, Bayesian inference. London: Arnold.
+
+\bibitem[Pancoast et al.(2014)]{2014MNRAS.445.3055P} Pancoast, A., Brewer, B.~J., Treu, T.\ 2014.\ Modelling reverberation mapping data - I.
+Improved geometric and dynamical models and comparison with cross-correlation results.\ Monthly Notices of the Royal Astronomical Society 445, 3055-3072.
 

### 2016-05-14 paper/ms.tex
A bit more writing

@@
 
-\section{Bayesian computation and Nested Sampling}
-
+\section{Introduction}
 Bayesian inference, where probability theory is used to describe degrees of
@@
 
-\subsection{Markov Chain Monte Carlo}
+There are several ways of using \pkg{DNest4}. After installing the software
+(Section~\ref{sec:installation}), you can implement a model in \proglang{C++}
+(Section~\ref{sec:models}) and compile it to create an executable file
+to run \pkg{DNest4} on that problem. This method offers the flexibility
+
+\section{Markov Chain Monte Carlo}\label{sec:mcmc}
 DNS is build upon the Metropolis-Hastings algorithm.
@@
 
-\section{Dependencies and installation}
+\section{Dependencies and installation}\label{sec:installation}
 
@@
 
-\section{Advanced usage}
-
 %\subsection{Multithreading}
 
-\section{Implementing models}
+\section{Implementing models}\label{sec:models}
 The ``classic'' method of implementing models in \pkg{DNest4} is by
@@
 
+\subsection{Python Bindings}
+
 \section*{Acknowledgements}

### 2016-05-14 paper/ms.tex
Python bindings

@@
 There are several ways of using \pkg{DNest4}. After installing the software
-(Section~\ref{sec:installation}), you can implement a model in \proglang{C++}
+(Section~\ref{sec:installation}), you can implement a model as
+a \proglang{C++} class
 (Section~\ref{sec:models}) and compile it to create an executable file
-to run \pkg{DNest4} on that problem. This method offers the flexibility
+to run \pkg{DNest4} on that problem. This method offers full control over the
+design of your class, and allows the opportunity (for example) to optimise
+performance by preventing a full re-computation of the log-likelihood when
+only a small subset of the parameters were changed.
+
+Similarly, the \proglang{Python}
+bindings (Section~\ref{sec:python_bindings}) allow you
+to specify a model class in \proglang{Python}, and run \pkg{DNest4} entirely
+in the \proglang{Python} interpreter without having to invoke the \proglang{C++}
+compiler.
 
@@
 
-\subsection{Python Bindings}
+\subsection{Python Bindings}\label{sec:python_bindings}
 

### 2016-05-15 code/CommandLineOptions.cpp
DNest3 --> DNest4

@@ -89,7 +89,7 @@ void CommandLineOptions::print_help() const
 {
 	std::cout<<"DNest4 Command Line Options: "<<std::endl;
 	std::cout<<"-h: display this message"<<std::endl;
-	std::cout<<"-o <filename>: load DNest3 options from the specified file. Default=OPTIONS"<<std::endl;
+	std::cout<<"-o <filename>: load DNest4 options from the specified file. Default=OPTIONS"<<std::endl;
 	std::cout<<"-s <seed>: seed the random number generator with the specified value. If unspecified, the system time is used."<<std::endl;
 	std::cout<<"-d <filename>: Load data from the specified file, if required."<<std::endl;
 	std::cout<<"-c <value>: Specify a compression value (between levels) other than e."<<std::endl;

### 2016-05-15 paper/ms.tex
Marsden
Command line options

@@
 
-% Need this after the abstract
-\setlength{\parindent}{0pt}
-\setlength{\parskip}{8pt}
+%% Need this after the abstract
+%\setlength{\parindent}{0pt}
+%\setlength{\parskip}{8pt}
 
@@
 %\subsection{Multithreading}
+
+\section{Running models}
+
+\subsection{Options}
+A plain-text file called OPTIONS resides in the directory from which you
+execute a run.
+
+Additional options are available on the command line. These are:\\
+
+\code{-o <filename>}\\
+This option loads the \pkg{DNest4} options from the specified file, allowing
+an alternative to the default OPTIONS.\\
+
+\code{-s <seed>}\\
+This seeds the random number generator with the specified value. If unspecified, the system time is used.\\
+
+\code{-d <filename>}\\
+Load data from the specified file, if required.\\
+
+\code{-c <value>}\\
+Standard DNS creates levels with a volume ratio of approximately
+$e\approx 2.71828$. To use a different value, such as 10, use this option.\\
+
+\code{-t <num\_threads>}\\
+Run on the specified number of threads. The default is 1.
 
@@
 Ewan Cameron (Oxford), David Hogg (NYU), and Daniela Huppenkothen (NYU)
-for valuable discussions.
+for valuable discussions. BJB was supported by a Marsden Fast-Start grant
+from the Royal Society of New Zealand.
 

### 2016-05-15 code/Examples/StraightLine/display.py
New figure, moved command line options to an appendix

@@ -8,19 +8,17 @@ xmin = data[:,0].min()
 xmax = data[:,0].max()
 
 x = linspace(xmin - 0.2*(xmax - xmin), xmax + 0.2*(xmax - xmin), 1001)
-ion()
 
+# Plot regression lines
 for i in range(posterior_sample.shape[0]):
-	hold(False)
-	plot(data[:,0], data[:,1], 'bo')
-	hold(True)
-	y = posterior_sample[i, 0]*x + posterior_sample[i, 1]
-	plot(x, y, 'k')
-	xlabel('$x$', fontsize=20)
-	ylabel('$y$', fontsize=20)
-	title('Model {k}/{N}'.format(k=(i+1), N=posterior_sample.shape[0]))
-	axis([0, 100, 0, 800])
-	draw()
-ioff()
+    y = posterior_sample[i, 0]*x + posterior_sample[i, 1]
+    plot(x, y, 'k', alpha=0.01)
+
+plot(data[:,0], data[:,1], 'bo')
+xlabel('$x$', fontsize=18)
+ylabel('$y$', fontsize=18)
+axis([0, 100, 0, 800])
+title('Regression Lines', fontsize=16)
+savefig('regression_lines.pdf', bbox_inches='tight')
 show()
 

### 2016-05-15 paper/ms.tex
New figure, moved command line options to an appendix
appendices, mention DNEST4_PATH

@@
 
+We recommend you create an environment variable called \code{DNEST4\_PATH}
+and set it to the directory {\em above} the \pkg{DNest4} directory. Then,
+if the examples or templates from the \pkg{DNest4} repository are copied
+to any other location on your system, their Makefiles will continue to work.
+
 \subsection{Microsoft Windows}
@@
 
+\begin{figure}
+\centering
+\includegraphics[width=0.8\textwidth]{figures/regression_lines.pdf}
+\caption{Regression lines drawn from the posterior.
+\label{fig:regression_lines}}
+\end{figure}
+
 \subsection{Options}
@@
 
-Additional options are available on the command line. These are:\\
-
-\code{-o <filename>}\\
-This option loads the \pkg{DNest4} options from the specified file, allowing
-an alternative to the default OPTIONS.\\
-
-\code{-s <seed>}\\
-This seeds the random number generator with the specified value. If unspecified, the system time is used.\\
-
-\code{-d <filename>}\\
-Load data from the specified file, if required.\\
-
-\code{-c <value>}\\
-Standard DNS creates levels with a volume ratio of approximately
-$e\approx 2.71828$. To use a different value, such as 10, use this option.\\
-
-\code{-t <num\_threads>}\\
-Run on the specified number of threads. The default is 1.
+Additional options are available on the command line. These are
+described in Appendix~\ref{sec:command_line_options}.
 
@@
 
+\appendix
+\section{Command Line Options}\label{sec:command_line_options}
+The following command line options are available.\\
+
+\code{-o <filename>}\\
+This option loads the \pkg{DNest4} options from the specified file, allowing
+an alternative to the default OPTIONS.\\
+
+\code{-s <seed>}\\
+This seeds the random number generator with the specified value. If unspecified, the system time is used.\\
+
+\code{-d <filename>}\\
+Load data from the specified file, if required.\\
+
+\code{-c <value>}\\
+Standard DNS creates levels with a volume ratio of approximately
+$e\approx 2.71828$. To use a different value, such as 10, use this option.
+In our experience, \code{-c 10} tends to work better than the default
+The output units remain in nats.
+This option is incompatible with setting the maximum number of levels to 0
+(automatic) in the OPTIONS file.
+\\
+
+\code{-t <num\_threads>}\\
+Run on the specified number of threads. The default is 1.
+
+\section{Specifying models in Python with dnest4.builder}
+
+\section{Specifying models in Julia}
+
 \end{document}

### 2016-05-15 paper/ms.tex
showresults figures

@@
 \section{Running models}
+To demonstrate \pkg{DNest4},
+we will use a simple linear regression example where the
+sampling distribution is
+\begin{align}
+y_i | m, b, \sigma &\sim \textnormal{Normal}(mx_i + b, \sigma^2)
+\end{align}
+and the priors are
+\begin{align}
+m &\sim \textnormal{Normal}(0, 1000^2)\\
+b &\sim \textnormal{Normal}(0, 1000^2)\\
+\ln\sigma &\sim \textnormal{Uniform}(-10, 10).
+\end{align}
+These are naive diffuse priors and we do not claim any special status for them.
+This model was implemented in \proglang{C++}
+(see Section~\ref{sec:models} for details) and is included in the
+code/Examples/StraightLine subdirectory.
 
-\begin{figure}
-\centering
-\includegraphics[width=0.8\textwidth]{figures/regression_lines.pdf}
+\begin{figure}[ht!]
+\begin{minipage}{0.5\textwidth}
+\includegraphics[width=8cm]{figures/fig1.pdf}
+\caption{\label{fig:fig1}}
+\end{minipage}
+\begin{minipage}{0.5\textwidth}
+\includegraphics[width=8cm]{figures/fig2.pdf}
+\caption{\label{fig:fig2}}
+\end{minipage}
+\end{figure}
+
+\begin{figure}[ht!]
+\begin{minipage}{0.5\textwidth}
+\includegraphics[width=8cm]{figures/fig3.pdf}
+\caption{\label{fig:fig3}}
+\end{minipage}
+\begin{minipage}{0.5\textwidth}
+\includegraphics[width=8cm]{figures/regression_lines.pdf}
 \caption{Regression lines drawn from the posterior.
 \label{fig:regression_lines}}
+\end{minipage}
 \end{figure}
@@
 point in your parameter space.
-To demonstrate, consider a simple linear regression situation where the
-sampling distribution is
-\begin{align}
-y_i | m, b, \sigma &\sim \textnormal{Normal}(mx_i + b, \sigma^2)
-\end{align}
-and the priors are
-\begin{align}
-m &\sim \textnormal{Normal}(0, 1000^2)\\
-b &\sim \textnormal{Normal}(0, 1000^2)\\
-\ln\sigma &\sim \textnormal{Uniform}(-10, 10)
-\end{align}
-These are naive diffuse priors and we do not claim any special status for them.
 

### 2016-05-15 paper/ms.tex
Fixed reference

@@
 is required, and Markov Chain Monte Carlo (MCMC) is a popular method of doing
-this, although alternatives exist \citep[e.g.][]{multinest, handley}.
+this, although alternatives exist \citep[e.g.][]{multinest, 2015MNRAS.453.4384H}.
 

### 2016-05-15 paper/ms.tex
Finished sentence
Figure captions

@@
 This model was implemented in \proglang{C++}
-(see Section~\ref{sec:models} for details) and is included in the
+(see Section~\ref{sec:models} for details).
+The dataset is shown in Figure~\ref{fig:regression_lines}
+and the code is included in the
 code/Examples/StraightLine subdirectory.
@@
 \begin{figure}[ht!]
-\begin{minipage}{0.5\textwidth}
-\includegraphics[width=8cm]{figures/fig1.pdf}
-\caption{\label{fig:fig1}}
-\end{minipage}
-\begin{minipage}{0.5\textwidth}
-\includegraphics[width=8cm]{figures/fig2.pdf}
-\caption{\label{fig:fig2}}
+\begin{minipage}{7.5cm}
+\centering
+\includegraphics[width=7cm]{figures/fig1.pdf}
+\caption{The level $j$ of the saved particles over time.
+Typically, this will trend upwards until all the levels have
+been created, and then diffuse evenly throughout all the levels.
+\label{fig:fig1}}
+\end{minipage}\hspace{0.5cm}
+\begin{minipage}{7.5cm}
+\centering
+\includegraphics[width=7cm]{figures/fig2.pdf}
+\caption{{\bf Top panel}: The estimated compression factor between subsequent
+levels, expressed as $\ln(X_{i+1}/X_{i})$. {\bf Bottom panel}:
+The Metropolis acceptance fraction as a function of level.
+\label{fig:fig2}}
 \end{minipage}
@@
 \begin{figure}[ht!]
-\begin{minipage}{0.5\textwidth}
-\includegraphics[width=8cm]{figures/fig3.pdf}
-\caption{\label{fig:fig3}}
-\end{minipage}
-\begin{minipage}{0.5\textwidth}
-\includegraphics[width=8cm]{figures/regression_lines.pdf}
+\begin{minipage}{7.5cm}
+\includegraphics[width=7cm]{figures/fig3.pdf}
+\caption{{\bf Top panel}: The log-likelihood curve, showing the relationship
+between log-likelihood and the enclosed prior mass.
+{\bf Bottom panel}: Posterior weights of the saved particles.
+For a successful run, there should be a clear peak, and saved particles
+to the left of this plot should have insignificant posterior weight compared
+to those in the peak.
+\label{fig:fig3}}
+\end{minipage}\hspace{0.5cm}
+\begin{minipage}{7.5cm}
+\includegraphics[width=7cm]{figures/regression_lines.pdf}
 \caption{Regression lines drawn from the posterior.
@@
 In our experience, \code{-c 10} tends to work better than the default
+value on problems with a large information.
 The output units remain in nats.

### 2016-05-16 paper/ms.tex
Julia model section

@@
 In our experience, \code{-c 10} tends to work better than the default
-value on problems with a large information.
+value on problems with a large information $\mathcal{H}$.
 The output units remain in nats.
@@
 \section{Specifying models in Julia}
+Experimental support for model specification in the
+\proglang{Julia} language is provided, with an example given in the
+code/Templates/JuliaModel directory.
+Set an enviroment variable {\tt JULIA\_PATH} to the directory on your
+system that contains the libraries {\tt libjulia.so} and
+{\tt sys.so}. Set another environment variable {\tt JULIA\_INCLUDE\_PATH}
+to the directory where the \proglang{Julia} header files are located
+(do not include the final {\tt julia} directory in this path).
+
+The log likelihood is specified in {\tt julia\_model.jl}, as a function
+called {\tt log\_likelihood} that takes a {\tt Vector::Float64} as an argument
+and returns a {\tt Float64}. You should assume that the priors for the
+parameters are independent Uniform(0, 1) distributions (sensible default
+proposals are used). To use a different prior, transform the parameters inside
+your {\tt log\_likelihood} function before using them.
+You may add other functions to {\tt julia\_model.jl}, for example to load
+a dataset to refer to in your {\tt log\_likelihood}. The example in
+the repository is the ``SpikeSlab'' likelihood from \citet{dnest}.
+
+You can compile the model by invoking {\tt make}
+from the {\tt JuliaModel} directory.
+Only a single thread is supported when running \proglang{Julia}
+models.
 

### 2016-05-16 code/Templates/JuliaModel/main.cpp
JULIA_LIB_PATH

@@ -9,7 +9,7 @@ using namespace DNest4;
 
 int main(int argc, char** argv)
 {
-	jl_init_with_image(getenv("JULIA_PATH"), "sys.so");
+	jl_init_with_image(getenv("JULIA_LIB_PATH"), "sys.so");
 	jl_eval_string("include(\"julia_model.jl\")");
 
 	Sampler<MyModel> sampler = setup<MyModel>(argc, argv);

### 2016-05-16 paper/ms.tex
JULIA_LIB_PATH
Copied output file info from manual

@@
 \usepackage{dsfont}
+
+\newcommand{\dnest}{\pkg{DNest4}}
 
@@
 
-\section{Running models}
+\section{Running DNest4}
 To demonstrate \pkg{DNest4},
@@
 \end{align}
-These are naive diffuse priors and we do not claim any special status for them.
+These are naive diffuse priors and do not have any special status.
 This model was implemented in \proglang{C++}
@@
 and the code is included in the
-code/Examples/StraightLine subdirectory.
+{\tt code/Examples/StraightLine} subdirectory. A simplified version of the
+code is explained in Section~\ref{sec:models}.
+
+To execute \pkg{DNest4} on this problem, run the executable {\tt main}
+in the {\tt code/Examples/StraightLine} directory. The output to the screen
+should contain information about levels and ``saving particles to disk''.
+After 5000 particles have been saved to the disk, the run should terminate.
+
+\section{Output files and plots}
+The executable {\tt main} is responsible for the exploration part of the
+DNS algorithm (i.e. running the MCMC chain, building
+levels, and then exploring all the levels). It creates three text output files,
+{\tt sample.txt}, {\tt sample\_info.txt}, and {\tt levels.txt}.
+
+The first output
+file, {\tt sample.txt}, contains a sampling of parameter values that
+represents the {\it mixture of constrained priors} (the target distribution
+used in DNS), {\bf not} the
+posterior distribution. Each line of {\tt sample.txt} represents a point in
+parameter space. In the linear example, there are three parameters
+($m$, $b$, and $\sigma$), so there
+are three columns in {\tt sample.txt}.
+Each time a point is saved to {\tt sample.txt}, \pkg{DNest4} prints
+the message ``Saving a particle to disk. N = ...''.
+
+The second output file, {\tt sample\_info.txt}, should have the same number of
+rows as {\tt sample.txt}, because it contains metadata about the samples in
+{\tt sample.txt}. The first
+column is the index $j$, which tells us which ``level'' the particle was in
+when it was saved. Level 0 represents the prior, and higher levels represent
+more constrained versions of the prior.
+The second column is the log-likelihood value, and the third column is
+the likelihood ``tiebreaker'', which allows Nested Sampling to work when
+there is a region in parameter space with nonzero prior probability where the
+likelihood is constant. The final column tells us which thread the particle
+belonged to: when you use \dnest~in multithreaded mode, each thread
+is responsible for evolving one or more walkers.
+
+The third output file, {\tt levels.txt}, contains information about the levels
+that were built during the run. The first column has estimates of the $\log(X)$
+values of the levels, i.e. how compressed they are relative to the prior, in
+units of nats. For example, if a level has $\log(X) = -1.02$, its likelihood
+value encloses $\exp(-1.02) \approx 36.1\%$ of the prior mass.
+
+The second column contains the log likelihoods of the levels.
+The first level, with a $\log(X)$ value of 0 and a log likelihood of
+$-10^{308}$ (basically ``minus infinity''), is simply the prior. The third
+column has the ``tiebreaker'' values for the levels, which again are not
+particularly useful unless your problem has likelihood plateaus. The fourth
+and fifth columns are the number of accepted proposals and the total number
+of proposals that have occurred within each level, which are useful for
+monitoring the Metropolis acceptance ratio as a function of level.
+The final two columns, called ``exceeds'', and ``visits'', are used to refine
+the estimates of the level compressions (and hence the $\log(X)$ values of
+the levels in column 1), as discussed in Section 3 of the
+paper. The visits column counts the number of times a level (level $j$, say)
+has been visited, but only starts counting after the next level ($j+1$) has been created. The exceeds column counts the number of times a particle that was
+in level $j$ had a likelihood that exceeded that of level $j+1$.
 
@@
 code/Templates/JuliaModel directory.
-Set an enviroment variable {\tt JULIA\_PATH} to the directory on your
+Set an enviroment variable {\tt JULIA\_LIB\_PATH} to the directory on your
 system that contains the libraries {\tt libjulia.so} and

### 2016-05-17 code/Examples/StraightLine/display.py
hold(True)

@@ -13,6 +13,7 @@ x = linspace(xmin - 0.2*(xmax - xmin), xmax + 0.2*(xmax - xmin), 1001)
 for i in range(posterior_sample.shape[0]):
     y = posterior_sample[i, 0]*x + posterior_sample[i, 1]
     plot(x, y, 'k', alpha=0.01)
+    hold(True)
 
 plot(data[:,0], data[:,1], 'bo')
 xlabel('$x$', fontsize=18)

### 2016-05-17 paper/ms.tex
postprocessing

@@
 
-\section{Output files and plots}
+\section{Output files}
 The executable {\tt main} is responsible for the exploration part of the
@@
 in level $j$ had a likelihood that exceeded that of level $j+1$.
+
+\section{Output files}
+The output files themselves aren't particularly useful. The goal of running
+NS is usually to obtain posterior samples (which, again, {\tt sample.txt}
+{\em does not provide}) and the marginal likelihood $Z$. Additional
+post-processing is required. This can be achieved by running the following
+\proglang{Python} function\footnote{Alternatively, the file
+{\tt showresults.py} in the example directory runs this function, and then
+calls the code in {\tt display.py} to create a further plot shown in
+Figure~\ref{fig:regression_lines}.}:
+\begin{CodeChunk}
+\begin{CodeInput}
+import dnest4.classic
+dnest4.classic.postprocess()
+\end{CodeInput}
+\end{CodeChunk}
+This produces the three diagnostic plots in Figures~\ref{fig:fig1},
+\ref{fig:fig2}, and~\ref{fig:fig3}, along with the following output:
+\begin{CodeChunk}
+\begin{CodeOutput}
+log(Z) = -175.528264138 +- 0.0
+Information = 15.1944947225 +- 0.0 nats.
+Effective sample size = 745.819862383
+\end{CodeOutput}
+\end{CodeChunk}
+These are the natural log of the marginal likelihood, the
+information
+\begin{align}
+\mathcal{H} &= \int p(\theta|D, I)
+\ln\left[\frac{p(\theta | D, I)}{p(\theta | I)}\right] \, d\theta
+\end{align}
+and the number of saved particles with significant posterior weight.
+The {\tt postprocess} function also saves a file,
+{\tt posterior\_sample.txt}, containing posterior samples (one per row).
+
+The zero-sized error bars on $\ln(Z)$ and $\mathcal{H}$ should be ignored.
+These are only relevant if advanced options (not covered in this paper)
+are passed to {\tt postprocess}.
 

### 2016-05-17 paper/ms.tex
OPTIONS stuff

@@
 A plain-text file called OPTIONS resides in the directory from which you
-execute a run.
+execute a run. Here is the OPTIONS file used in the linear regression
+example.
+
+\begin{CodeChunk}
+\begin{CodeInput}
+# File containing parameters for DNest4
+# Put comments at the top, or at the end of the line.
+5       # Number of particles
+10000   # new level interval
+10000   # save interval
+100     # threadSteps - how many steps each thread should do independently before pooling information
+0       # maximum number of levels (0 = auto)
+10      # Backtracking scale length (lambda in the paper)
+100     # Strength of effect to force histogram to equal push. Beta in the paper
+5000    # Maximum number of saves (0 = infinite)
+\end{CodeInput}
+\end{CodeChunk}
+
+The first option is the number of particles per thread, here set to one.
+If you use more particles
+per thread, the same amount of CPU time will be spent evolving more particles,
+so each one will not be evolved as far. On most problems, one particle per
+thread is fine. I would only use more particles per thread when working on
+a very hard problem where it is difficult for the particles to find the
+important parts of parameter space.
+
+The new level interval controls how quickly \dnest~creates new levels. In this
+example, this is set to 10,000, so a new level will be created once \dnest~has
+seen 10,000 likelihood values above the current top level. This is not affected by
+the number of threads. It is difficult to give a sensible default for this
+quantity because it depends on the complexity of the problem and how good
+the Metropolis proposals are. However, 10,000
+will work for many problems (I rarely use anything less than 10,000 or greater
+than 300,000). Higher values are slower, but more fail-safe.
+
+The save interval controls how often \dnest~writes a model to the output
+files: this is what is usually called ``thinning''. Saving more frequently
+(i.e. a smaller save interval) is usually better, although it depends on your
+problem. If your models have a lot of parameters {\tt sample.txt} can get
+very big and take forever to process using {\tt showresults.py}. A default
+suggestion that is usually okay is to make the save interval the same as the
+new level interval.
+
+The option ``threadSteps'' controls how frequently the separate threads pool
+their information about the levels. I usually use 100 or 200. If
+(threadSteps) $\times$ (number of threads) is not much smaller than
+saveInterval or newLevelInterval, the latter won't be what you expect, because
+\dnest~only creates levels or saves particles when the threads are pooling
+information.
 

### 2016-05-23 paper/ms.tex
More advantages

@@
 this, although alternatives exist \citep[e.g.][]{multinest, 2015MNRAS.453.4384H}.
-
 Diffusive Nested Sampling \citep[DNS][]{dnest} is an alternative to NS for
@@
 \end{align}
-
+The mixture of constrained priors tends to be easier to sample than the
+posterior, as the prior is always a mixture component, allowing the
+MCMC chain to mix between different modes in some circumstances. The marginal
+likelihood estimate can also be more accurate than standard MCMC-based NS
+\citep{dnest}, as less information is discarded.
 DNS has been applied several times in astrophysics
@@
 
-Similarly, the \proglang{Python}
+Alternatively, the \proglang{Python}
 bindings (Section~\ref{sec:python_bindings}) allow you

### 2016-05-23 paper/ms.tex
some writing

@@
 
-\section{Output files}
+%\section{Output files}
 The executable {\tt main} is responsible for the exploration part of the
@@
 
+%\subsection{sample.txt}
 The first output
@@
 
-The second output file, {\tt sample\_info.txt}, should have the same number of
+%\subsection{sample\_info.txt}
+%The second output file, {\tt sample\_info.txt}, should have the same number of
 rows as {\tt sample.txt}, because it contains metadata about the samples in
@@
 
+%\subsection{levels.txt}
 The third output file, {\tt levels.txt}, contains information about the levels
@@
 
-\section{Output files}
-The output files themselves aren't particularly useful. The goal of running
-NS is usually to obtain posterior samples (which, again, {\tt sample.txt}
-{\em does not provide}) and the marginal likelihood $Z$. Additional
+%\section{Postprocessing}
+The output files themselves are typically not immediately useful.
+The goal of running
+NS is usually to obtain posterior samples and the marginal likelihood $Z$,
+whereas {\tt sample.txt} only contains samples from the mixture of constrained
+priors. Additional
 post-processing is required. This can be achieved by running the following
@@
 Figure~\ref{fig:regression_lines}.}:
+
 \begin{CodeChunk}
@@
 {\tt posterior\_sample.txt}, containing posterior samples (one per row).
-
 The zero-sized error bars on $\ln(Z)$ and $\mathcal{H}$ should be ignored.
@@
 
-\subsection{Options}
-A plain-text file called OPTIONS resides in the directory from which you
-execute a run. Here is the OPTIONS file used in the linear regression
-example.
+%\subsection{Options}
+A plain-text file called {\tt OPTIONS} resides in the directory from which you
+execute a run. This file contains numerical parameters controlling the
+DNS algorithm. Here are the contents of {\tt OPTIONS} for the linear
+regression example:
 
@@
 10000   # save interval
-100     # threadSteps - how many steps each thread should do independently before pooling information
+100     # threadSteps - pooling interval
 0       # maximum number of levels (0 = auto)
@@
 
-The first option is the number of particles per thread, here set to one.
-If you use more particles
-per thread, the same amount of CPU time will be spent evolving more particles,
-so each one will not be evolved as far. On most problems, one particle per
-thread is fine. I would only use more particles per thread when working on
-a very hard problem where it is difficult for the particles to find the
-important parts of parameter space.
+The first option is the number of particles, here set to five.
+If you use more particles, the same amount of CPU time will be spent evolving more particles,
+so each one will not be evolved as far. On most problems, five is a sensible
+default value. On complex problems where the likelihood function has
+a challenging structure, more particles are useful, but it is usually better
+%to run in multi-threaded mode (see Appendix~\ref{sec:command_line_options}).
 

### 2016-05-23 code/SamplerImpl.h
Slightly different termination criterion for auto max_num_levels

@@ -334,8 +334,9 @@ bool Sampler<ModelType>::enough_levels(const std::vector<Level>& l) const
             return false;
 
         // Check level spacing (in terms of log likelihood)
-        // over last 20 levels
-        constexpr int num_levels_to_check = 20;
+        // over last n levels, where n=
+        // geometric mean of {20, l.size()}
+        int num_levels_to_check = static_cast<int>(sqrt(20)*sqrt(0.25*l.size()));
 
         int k = levels.size() - 1;
         for(int i=0; i<num_levels_to_check; ++i)

### 2016-05-23 code/SamplerImpl.h
Another tweak to auto max_num_levels method

@@ -334,9 +334,11 @@ bool Sampler<ModelType>::enough_levels(const std::vector<Level>& l) const
             return false;
 
         // Check level spacing (in terms of log likelihood)
-        // over last n levels, where n=
-        // geometric mean of {20, l.size()}
-        int num_levels_to_check = static_cast<int>(sqrt(20)*sqrt(0.25*l.size()));
+        // over last n levels, where n={ 20 iff l.size() < 80,
+        // geometric mean of {20, l.size()} otherwise}
+        int num_levels_to_check = 20;
+        if(l.size() > 80)
+            num_levels_to_check = static_cast<int>(sqrt(20)*sqrt(0.25*l.size()));
 
         int k = levels.size() - 1;
         for(int i=0; i<num_levels_to_check; ++i)

### 2016-05-23 paper/ms.tex
More writing

@@
 100     # threadSteps - pooling interval
-0       # maximum number of levels (0 = auto)
+0       # maximum number of levels (0 ==> automatic)
 10      # Backtracking scale length (lambda in the paper)
-100     # Strength of effect to force histogram to equal push. Beta in the paper
-5000    # Maximum number of saves (0 = infinite)
+100     # Equal weight enforcement. Beta in the paper
+5000    # Maximum number of saves (0 ==> run forever)
 \end{CodeInput}
@@
 a challenging structure, more particles are useful, but it is usually better
-%to run in multi-threaded mode (see Appendix~\ref{sec:command_line_options}).
+to run in multi-threaded mode (see Appendix~\ref{sec:command_line_options}).
 
 The new level interval controls how quickly \dnest~creates new levels. In this
-example, this is set to 10,000, so a new level will be created once \dnest~has
-seen 10,000 likelihood values above the current top level. This is not affected by
-the number of threads. It is difficult to give a sensible default for this
-quantity because it depends on the complexity of the problem and how good
-the Metropolis proposals are. However, 10,000
-will work for many problems (I rarely use anything less than 10,000 or greater
-than 300,000). Higher values are slower, but more fail-safe.
+example, this is set to 10,000, so a new level will be created once 10,000
+MCMC steps have resulted in
+10,000 likelihood values above the current top level.
+It is difficult to give a sensible default for this
+quantity because it depends on the complexity of the problem (basically,
+how good the Metropolis proposals are at exploring the target distribution).
+However, 10,000 will work for many problems, so we suggest it as a sensible
+default. Higher values are slower, but more fail-safe.
 
 The save interval controls how often \dnest~writes a model to the output
-files: this is what is usually called ``thinning''. Saving more frequently
-(i.e. a smaller save interval) is usually better, although it depends on your
-problem. If your models have a lot of parameters {\tt sample.txt} can get
-very big and take forever to process using {\tt showresults.py}. A default
-suggestion that is usually okay is to make the save interval the same as the
-new level interval.
+files; what is usually called ``thinning''. Saving more frequently
+(i.e. a smaller save interval) is usually better. However, this can result
+in
+big output files if your model prints a lot of parameters to {\tt sample.txt}.
+causing the postprocessing to take a long time and/or a lot of RAM.
+A default suggestion, used in the example, is to set the save interval to the
+same value as the new level interval.
 

### 2016-05-23 paper/ms.tex
Describing the information, and OPTIONS

@@
 \mathcal{H} &= \int p(\theta|D, I)
-\ln\left[\frac{p(\theta | D, I)}{p(\theta | I)}\right] \, d\theta
+\ln\left[\frac{p(\theta | D, I)}{p(\theta | I)}\right] \, d\theta,
 \end{align}
-and the number of saved particles with significant posterior weight.
+which quantifies the degree to which $D$ restricted the
+range of possible $\theta$ values,
+and the effective sample size,
+or number of saved particles with significant posterior weight.
 The {\tt postprocess} function also saves a file,
@@
 
-The option ``threadSteps'' controls how frequently the separate threads pool
-their information about the levels. I usually use 100 or 200. If
-(threadSteps) $\times$ (number of threads) is not much smaller than
-saveInterval or newLevelInterval, the latter won't be what you expect, because
-\dnest~only creates levels or saves particles when the threads are pooling
-information.
+The ``thread steps'' parameter controls how frequently separate threads pool
+their information about the levels (when running in multi-threaded mode,
+see Appendix~\ref{sec:command_line_options}). It should be set to a moderate
+value, but should also be a small fraction of the new level interval and the
+save interval. 100 is a suggested default value.
 

### 2016-05-23 paper/ms.tex
sections

@@
 
-%\subsection{Multithreading}
-
-\section{Running DNest4}
+\section{Running DNest4}\label{sec:running}
 To demonstrate \pkg{DNest4},
@@
 
-%\section{Postprocessing}
+\section{Postprocessing}\label{sec:postprocessing}
 The output files themselves are typically not immediately useful.
@@
 
-%\subsection{Options}
+\subsection{Options}\label{sec:options}
 A plain-text file called {\tt OPTIONS} resides in the directory from which you
@@
 
+\subsection{Number of particles}
 The first option is the number of particles, here set to five.
@@
 
+\subsection{New level interval}
 The new level interval controls how quickly \dnest~creates new levels. In this
@@
 
+\subsection{Save interval}
 The save interval controls how often \dnest~writes a model to the output
@@
 
+\subsection{Thread steps}
 The ``thread steps'' parameter controls how frequently separate threads pool

### 2016-05-23 paper/ms.tex
Updated Figure 2
max num levels

@@
 \centering
-\includegraphics[width=7cm]{figures/fig2.pdf}
+\includegraphics[width=8cm]{figures/fig2.pdf}
 \caption{{\bf Top panel}: The estimated compression factor between subsequent
@@
 
+Additional options are available on the command line. These are
+described in Appendix~\ref{sec:command_line_options}.
+
 \subsection{Number of particles}
@@
 
-Additional options are available on the command line. These are
-described in Appendix~\ref{sec:command_line_options}.
+\subsection{Maximum number of levels}
+As the name suggests, this tells \pkg{DNest4} how many levels to create, and
+therefore controlls the factor by which the parameter space is ultimately
+compressed. An appropriate value for this quantity depends on the specific
+model and dataset at hand.
+
+Figure~\ref{fig:fig3} is the most useful plot for setting the correct
+maximum number of levels. A clear peak should be visible in the posterior
+weights plot in the lower panel.
+
+Alternatively, \pkg{DNest4} can try to determine the maximum number of levels
+automatically, if you set the maximum number of levels to 0.
 

### 2016-05-23 paper/ms.tex
More talk about options

--- 2016-05-23 paper/ms.tex
@@
 
+The {\tt postprocess} function can be called while \pkg{DNest4} is running.
+This is helpful for monitoring the progress of a run, by inspecting the
+output plots.
+
 \begin{figure}[ht!]
@@
 value, but should also be a small fraction of the new level interval and the
-save interval. 100 is a suggested default value.
+save interval. 100 is a suggested default value that should work without
+problems in the vast majority of cases.
 
-Additional options are available on the command line. These are
-described in Appendix~\ref{sec:command_line_options}.
+\subsection{Maximum number of levels}
+As the name suggests, this tells \pkg{DNest4} how many levels to create, and
+therefore controls the factor by which the parameter space is ultimately
+compressed. An appropriate value for this quantity depends on the specific
+model and dataset at hand --- typically, a larger numbers of parameters
+and larger (more informative) datasets will lead to a larger value of
+$\mathcal{H}$, and therefore need more levels.
+
+In the initial phase of DNS when levels are being created, the particles
+move to the left in Figure~\ref{fig:fig3}, increasing likelihood $L$ and
+decreasing prior mass $X$. The posterior distribution is concentrated
+where the rate of increase of $\ln(L)$ and the rate of decrease of
+$\ln(X)$ are approximately equal.
+Figure~\ref{fig:fig3} is the most useful diagnostic
+plot for setting the correct
+maximum number of levels. A clear peak should be visible in the posterior
+weights plot in the lower panel, such that moving further to the left would not
+add any more particles with comparable weight to those in the peak.
+
+As described by \citet{skilling}, there is no guarantee in principle that
+another peak might have appeared had a run continued for longer. These
+{\em phase changes} are more common in statistical mechanics problems than
+in data analysis problems, but can appear in the latter
+\citep[e.g.][]{rjobject, 2015MNRAS.448.3206B}.
+
+Alternatively, \pkg{DNest4} can try to determine the maximum number of levels
+automatically, if you set the maximum number of levels to 0. This works well
+on problems where the MCMC exploration is efficient.
 
@@
 
+\bibitem[Brewer(2014)]{rjobject} Brewer, B.~J.\ 2014.\ Inference for Trans-dimensional Bayesian Models with Diffusive Nested Sampling.\ ArXiv e-prints arXiv:1411.3921.
+
 \bibitem[Brewer and Donovan(2015)]{2015MNRAS.448.3206B} Brewer, B.~J., Donovan, C.~P.\ 2015.\ Fast Bayesian inference for exoplanet discovery in radial velocity data.\ Monthly Notices of the Royal Astronomical Society 448, 3206-3214.

### 2016-05-24 paper/ms.tex
More writing about OPTIONS

@@
 automatically, if you set the maximum number of levels to 0. This works well
-on problems where the MCMC exploration is efficient.
+on problems where the MCMC exploration is efficient. When it fails, this
+is detectable as the posterior weights plot (lower panel of
+Figure~\ref{fig:fig3}) will peak at the left end of its domain. However,
+such a failed run may still be useful as it can be used to suggest an order of
+magnitude for the required number of levels. For example, if the automatic
+setting results in 150 levels and is later seen to fail, it might be worth
+a try to set the maximum number of levels to, say, $1.3 \times 150 = 195$.
 

### 2016-05-24 paper/ms.tex
More writing, this time about lambda

@@
 \end{align}
+The DNS algorithm consists of two main stages. In the first stage,
+the particle(s) are initialized from the prior $\pi(\theta)$ (equivalent
+to the mixture of constrained priors with a single level whose likelihood
+threshold is $-\infty$), and
+the mixture of constrained priors evolves by adding new levels
+(see \citet{dnest} for details). In this stage,
+the mixture weights $\{w_j\}$ are set according to
+\begin{align}
+w_j &\propto \exp(j/\lambda),\label{eqn:weighting}
+\end{align}
+where $\lambda$ is a scale length.
+This enhances the probability that particles stay close to the
+highest likelihood regions seen.
+The second stage sets the mixture weights to be approximately
+uniform ($w_j \propto 1$), with some tweaks described by \citet{dnest}.
+
 The mixture of constrained priors tends to be easier to sample than the
@@
 a try to set the maximum number of levels to, say, $1.3 \times 150 = 195$.
+
+\subsection{Backtracking scale length}
+The backtracking scale length, denoted by $\lambda$, appeared in
+Equation~\ref{eqn:weighting}, and controls the degree to which particles
+are allowed to ``backtrack'' down in level during the first stage of DNS
+when levels are being built. Higher values are more fail-safe, but
+make it take longer to create the levels. The value 10 used in the linear
+regression example is a suitable default value that should work in almost all
+cases. In simple problems where MCMC exploration is easy, lower values
+from 1 to 5 work sufficiently well.
 

### 2016-05-24 paper/ms.tex
Maximum number of saves

@@
 
+\subsection{Equal weight enforcement}
+This value is the parameter $\beta$ described in \citet{dnest4}, and compensates
+for imprecision in the spacing of the levels, so that the desired mixture
+weights $w_j \propto 1$ are achieved during the second stage of the DNS
+algorithm. The value 100 is recommended.
+
+\subsection{Maximum number of saves}
+This controls the length of a \pkg{DNest4} run, in units of saved particles
+in {\tt sample.txt}, which represent the mixture of constrained priors.
+The number of posterior samples is always less than this. In most applications
+5,000 (as in the linear regression example) provides enough posterior
+samples (typically a few hundred) for sufficiently accurate posterior
+summaries. However, if you want to plot smooth-looking posterior histograms,
+you'll need to increase this value.
+
+If you set the maximum number of saves to zero, \pkg{DNest4} will run until
+you terminate it manually.
+
 \section{Implementing models}\label{sec:models}

### 2016-05-24 paper/ms.tex
Talking about randh

@@
 \subsection{Equal weight enforcement}
-This value is the parameter $\beta$ described in \citet{dnest4}, and compensates
+This value is the parameter $\beta$ described in \citet{dnest}, and compensates
 for imprecision in the spacing of the levels, so that the desired mixture
@@
 \code{rand\_int(int)} member function of the \code{DNest4::RNG} class.
-This determines which of the three parameters is modified.
+This determines which of the three parameters
+$(m, b, \sigma)$ is modified.
+
+\subsection{Proposals for single parameters}
+The proposal for $m$ involves adding a perturbation to the current value
+using the line \code{m += 1E3*rng.randh();}.
+
+A challenge using MCMC for standard Nested Sampling is that the target
+distribution is not static --- it gets compressed over time. Similarly, in
+the first stage of DNS the target distribution gets compressed (as levels are
+added), and in the second stage the target distribution is a mixture of
+distributions that have been compressed to varying degrees.
+
+Rather than trying to adapt proposal distributions as a function of level,
+it is much simpler to just use heavy-tailed proposals which have some
+probability of making a jump of appropriate size. This may seem slightly
+wasteful of CPU time, but it saves a lot of human time.
+In simple experiments, we have found that heavy-tailed proposals are
+about as efficient as slice sampling \citep{slice}, but much easier to
+implement. The following procedure generates
+$x$ from a heavy-tailed distribution:
+
+\begin{enumerate}
+\item Generate $c \sim$ Cauchy$(0, 1)$
+\item Generate $n \sim$ Normal$(0, 1)$
+\item Set $x := 10^{1.5 - |c|}n$
+\end{enumerate}
+This generates values with a maximum scale of tens-hundreds, down to
+the smallest positive value in double precision, by virtue
+of the Cauchy distribution's heavy tails.
+
+For convenience, the \code{RNG} class contains a member function \code{randh()}
+to generate values from this distribution.
+
+The factor of \code{1E3} is
+included to account for the prior width.
 
@@
 
+\bibitem[Neal(2003)]{slice}
+Neal, R.M., 2003. Slice sampling. Annals of statistics, pp.705-741.
+
 \bibitem[\protect\citeauthoryear{O'Hagan and Forster}{2004}]{ohagan}

### 2016-05-25 paper/ms.tex
RJObject section

@@
 
-%\section{Output files}
+\section{Output files}
 The executable {\tt main} is responsible for the exploration part of the
@@
 
-%\subsection{sample.txt}
 The first output
@@
 
-%\subsection{sample\_info.txt}
-%The second output file, {\tt sample\_info.txt}, should have the same number of
+The second output file, {\tt sample\_info.txt}, should have the same number of
 rows as {\tt sample.txt}, because it contains metadata about the samples in
@@
 
-%\subsection{levels.txt}
 The third output file, {\tt levels.txt}, contains information about the levels
@@
 
+\section{Finite Mixture Models with RJObject}
+Formerly a separate project \citep{rjobject},
+
 \section{Approximate Bayesian Computation}

### 2016-05-26 paper/ms.tex
Talking about proposals

@@
 the priors are all independent, so this reduces to generating the parameters
-each from their own prior distribution. For convenience, the \code{RNG} object
-has \code{rand()} and \code{randn()} member functions to generate from
-a Uniform$(0,1)$ and Normal$(0,1)$ distribution respectively.
-
-After generating the parameters from the prior, another member function
-\code{calculate\_mu()} is called, which computes the model prediction
-$mx_i + b$ for each data point.
+each from their own prior distribution. For convenience,
+\pkg{DNest4} provides an \code{RNG} class to represent random number
+generators. The \code{RNG} class is just a convenience wrapper for the
+random number generators built into \proglang{C++11}.
+As you might expect, there are
+\code{rand()} and \code{randn()} member functions to generate
+double precision values
+from a Uniform$(0,1)$ and Normal$(0,1)$ distribution respectively.
 
@@
 This determines which of the three parameters
-$(m, b, \sigma)$ is modified.
+$(m, b, \sigma)$ is modified. In this example, there are no proposals
+that modify more than one of the parameters at a time, and all proposals
+are ``random walk'' proposals that add a perturbation
+(drawn from a symmetric distribution) to the current value.
 
 \subsection{Proposals for single parameters}
-The proposal for $m$ involves adding a perturbation to the current value
-using the line \code{m += 1E3*rng.randh();}.
+The proposal for $m$ involves adding a perturbation to the current
+value using the line
+
+\begin{CodeChunk}
+\begin{CodeInput}
+m += 1E3*rng.randh();
+\end{CodeInput}
+\end{CodeChunk}
 
@@
 distributions that have been compressed to varying degrees.
+This makes it difficult to tune step-sizes as you would when using
+the standard Metropolis algorithm to sample the posterior distribution.
 
@@
 \end{enumerate}
-This generates values with a maximum scale of tens-hundreds, down to
+This generates values with a maximum scale of tens--hundreds, down to
 the smallest positive value in double precision, by virtue
 of the Cauchy distribution's heavy tails.
-
 For convenience, the \code{RNG} class contains a member function \code{randh()}
 to generate values from this distribution.
-
-The factor of \code{1E3} is
-included to account for the prior width.
+The factor of \code{1E3} is included because it is a measure of the prior
+width. Therefore, this proposal will attempt moves whose maximum order of
+magnitude is a bit wider than the prior (since it would be very surprising
+if any bigger moves were needed), and will propose moves a few orders of
+magnitude smaller with moderate probability. We recommend using
+this strategy (a measure of prior width, multiplied by \code{randh()}) as
+a default proposal that works well in almost all problems.
 

### 2016-05-26 paper/ms.tex
ABC writing

@@
 \section{Approximate Bayesian Computation}
-Nested Sampling can be used to solve ``Approximate Bayesian Computation''
+Nested Sampling can be used to solve Approximate Bayesian Computation (ABC)
 problems elegantly and efficiently.
-
-Approximate Bayesian Computation (ABC), also known as likelihood-free inference,
-is a set of Monte Carlo techniques for approximating a posterior distribution
-without having to evaluate the likelihood function $L(\theta)$.
-
-. Instead, the ability to
-generate simulated data sets from the sampling distribution is required.
-Since a sampling distribution must be specified, it is not the assumption of
-a specific likelihood that is avoided (indeed, a sampling distribution and
+Also known (misleadingly) as likelihood-free inference,
+ABC
+is a set of Monte Carlo techniques for approximating the posterior distribution
+without having to evaluate the likelihood function $L(\theta)$. Instead, 
+the user must be able to cheaply
+generate simulated data sets from the sampling distribution $p(D|\theta, I)$.
+Since a sampling distribution must be specified, it is not that there is
+no likelihood function (indeed, a sampling distribution and
 a data set imply a particular likelihood function), rather that we cannot
-evaluate it cheaply as a function of $\params$ and $\data$.
+evaluate it cheaply and therefore cannot use MCMC.
 
@@
 speak and use notation as if we're conditioning on the value of a variable, for
-example by writing the posterior distribution as $p(\params | \data)$. However,
-this is shorthand for $p(\params | \data = \dobs)$. In the prior state of
-knowledge, the proposition $(\data = \dobs)$ could have been either true or
-false, but in the posterior state of knowledge it is definitely true.
+example by writing the posterior distribution as
+$p(\theta | D, I)$. However,
+this is shorthand for $p(\theta | D = D_{\rm observed}, I)$, and
+$D = D_{\rm observed}$ is a statement that could have been either
+true or false in the prior state of knowledge, and is known to be true
+in the posterior state of knowledge.
 In the case of ``continuous data'' (really a continuous {\it space of
 possibilities for the data before we learned it}) we're conditioning on a
-proposition like $(\data \in [\dobs, \dobs+\epsilon])$ and implicitly
-taking the limit $\epsilon \to 0$.
+proposition like $(D \in \mathcal{R})$ where $\mathcal{R}$ is a region, and
+then implicitly taking a limit as the size of $\mathcal{R}$ goes to zero.
+
+ABC works by replacing the proposition $D = D_{\rm observed}$ with a logically
+weaker one.
 

### 2016-05-26 paper/ms.tex
ABC writing

@@
 
-ABC works by replacing the proposition $D = D_{\rm observed}$ with a logically
+A simple ``rejection sampling'' version of
+ABC works by sampling the joint prior distribution for the parameters and
+data
+\begin{align}
+p(\theta, D | I) &= p(\theta | I)p(D | \theta, I)
+\end{align}
+and rejecting samples for which $D \neq D_{\rm observed}$, so that the
+samples represent
+\begin{align}
+p(\theta, D | D = D_{\rm observed}, I) &\propto p(\theta | I)p(D | \theta, I)
+\mathds{1}\left[D = D_{\rm observed}\right].
+\end{align}
+The marginal distribution for $\theta$ is also the conditional distribution
+$p(\theta | D = D_{\rm observed}, I)$, that is, the posterior.
+
+This approach is rarely usable in practice because the probability of
+generating a dataset that matches the observed one is extremely low.
+To work around this, the proposition
+So the proposition $D = D_{\rm observed}$ with a logically
 weaker one.

### 2016-05-26 paper/ms.tex
proposal

@@
 The dataset is shown in Figure~\ref{fig:regression_lines}
-and the code is included in the
-{\tt code/Examples/StraightLine} subdirectory. A simplified version of the
+and the code is included in the {\tt code/Examples/StraightLine}
+subdirectory. A simplified version of the
 code is explained in Section~\ref{sec:models}.
-
-To execute \pkg{DNest4} on this problem, run the executable {\tt main}
-in the {\tt code/Examples/StraightLine} directory. The output to the screen
+To execute \pkg{DNest4} on this problem, go to this directory and
+execute {\tt main}. The output to the screen
 should contain information about levels and ``saving particles to disk''.
@@
 
-When implementing a model class, it is imperative that the
-\code{from\_prior(DNest4::RNG\&)} and \code{perturb(DNest4::RNG\&)}
-member functions be consistent with each other. One technique
+Recall that for Nested Sampling, the Metropolis acceptance probability,
+excluding the term for the likelihood, is
+\begin{align}
+H = \frac{q(\params'|\params)}{q(\params | \params')}
+\times \frac{\pi(\params')}{\pi(\params)}.
+\end{align}
+When implementing a model class, the
+\code{perturb(DNest4::RNG\&)}
+member function must return the logarithm of this value.
+Since the prior for $m$ is a normal distribution with
+mean zero and standard deviation $1000$, $H$ is
+\begin{align}
+H &= \frac{\exp\left[-\frac{1}{2}(m'/1000)^2\right]}
+{\exp\left[-\frac{1}{2}(m/1000)^2\right]}.
+\end{align}
+This explains the use of the \code{log\_H} variable.
+
+It is imperative that \code{from_prior} and
+\code{perturb}
+be consistent with each other. One technique
 for verifying this is to sample the prior for a long time
@@
 {\tt sample.txt} to ensure that each parameter is exploring
-the correct distribution.
+the prior correctly.
 

### 2016-05-26 paper/ms.tex
log likelihood and print
Caticha and Giffin

@@
 
-The log-likelihood code is:
+The log-likelihood code is standard:
 \begin{CodeChunk}
@@
 }
-
 \end{CodeInput}
 \end{CodeChunk}
+The print function specifies that the parameters
+$m$, $b$, and $\sigma$ are printed a single line
+(of {\tt sample.txt}) and are separated by spaces (the
+\code{postprocess} function assumes this is the case).
 
@@
 The marginal distribution for $\theta$ is also the conditional distribution
-$p(\theta | D = D_{\rm observed}, I)$, that is, the posterior.
+$p(\theta | D = D_{\rm observed}, I)$, that is,
+the posterior\footnote{Incidentally, this usage of the {\em joint} posterior
+distribution for the parameters {\em and the data} provides a bridge
+connecting Bayesian updating to maximum entropy updating \citep{caticha,giffin}.}.
 
@@
 
-\begin{table}[ht!]
-\centering
-\small
-\begin{tabular}{|l|l|l|}
-\hline
-Bayesian computation		&		Statistical mechanics		&		ABC\\
-\hline
-Parameter space	$\Theta$	&		Phase space	$\Omega$ 			& Product space $\Theta \times \mathcal{D}$\\
-Posterior $p(\params | D,I)$   &  Canonical distribution $p(\boldsymbol{x}; T)$     & \\ 
-Marginal likelihood $P(D|I)$	&	Partition function $Z(T)$	&   Marginal likelihood $p(D\in | I)$\\
-\hline
-\end{tabular}
-\caption{\it The relationship between standard Bayesian computation, statistical
-mechanics, and ABC. In each case Monte Carlo methods are used to calculate
-integrals over a space. In standard Bayesian computation, it is the parameter
-space of a model that is usually of interest, whereas in ABC it is the space
-of possible parameter values {\it and} data sets.
-\label{tab:relation}}
-\end{table}
-
-\begin{figure}[ht!]
-\centering
-\includegraphics[scale=0.5]{figures/joint.pdf}
-\caption{\it The joint prior for parameters $\params$ and data $\data$.
-\label{fig:joint}}
-\end{figure}
-
-The main challenges associated with ABC are:
-\begin{enumerate}[(a)]
-\item The choice of summary statistics
-\item The choice of distance function
-\item How to choose the value of $\epsilon$, or how to change it as a run
-progresses
-\item How to achieve good mixing.
-\end{enumerate}
-Challenges 1 and 2 are Bayesian in nature, i.e. they relate to the very
-definition of the posterior distribution and are well defined independent of
-the fact that we are going to use a Monte Carlo method to compute the
-results. On the other hand, challenges 3 and 4 are about the computational implementation.
-
-\section{Parameterising the data space}
-In standard MCMC-based ABC, proposal moves are made by first proposing a
-change to the parameters (from $\params$ to $\params'$), and then generating
-a mock dataset $\data'$ from $p(\data | \params')$. As a Metropolis proposal
-for exploring the product space $\Theta \times \mathcal{D}$ this is a very
-aggressive choice, since the proposed value of the
-dataset $\params'$ is likely to be very different from the previous one.
-One of the most elementary pieces of advice we receive when learning the
-Metropolis algorithm is to be careful about how ambitious our proposal
-distributions are, since tiny and huge proposal moves are both very
-inefficient. The performance of MCMC-based ABC can therefore be improved by
-alternative choices of proposal distribution.
-
-In some cases, we might be able to find efficient proposal distributions for exploring $\Theta \times \mathcal{D}$ by having insight into the particular
-problem we're working on. However, there are also techniques that are
-completely general in that they should work on any ABC-type problem.
-
-Consider the process of generating a mock dataset. Whatever else is involved,
-this process will involve calls to an underlying random number generator.
-Perhaps a function {\tt rand()}, generating variables from a Uniform(0, 1)
-distribution, is called $n$ times in the course of generating a dataset.
-Usually, the simulation process has the property that, if $\theta'$ is
-close to $\theta$, then the same sequence of random numbers returned by
-{\tt rand()} would produce a dataset $\data'$ that is very similar to $\data$.
-This suggests an alternative parameterisation of
-the product space $\Theta\times\mathcal{D}$: instead of using the one that
-naturally describes the data, we can use a coordinate system defined by
-$\theta$ together with $\{u_1, ..., u_n\}$, the $n$ uniform random numbers
-used to construct a mock dataset.
-
-In this alternative coordinate system, a proposed move
-to change $\params$ while keeping
-$\{u_i\}$ fixed results in a proposed new dataset that is similar to the
-previous one, assuming the change to $\params$ is small. The standard procedure
-of generating an entirely new mock dataset each step is now viewed as a
-proposal where $\params$ is moved {\it and all of the $u_i$ coordinates are
-re-generated from their iid uniform prior}.
+%\begin{table}[ht!]
+%\centering
+%\small
+%\begin{tabular}{|l|l|l|}
+%\hline
+%Bayesian computation		&		Statistical mechanics		&		ABC\\
+%\hline
+%Parameter space	$\Theta$	&		Phase space	$\Omega$ 			& Product space $\Theta \times \mathcal{D}$\\
+%Posterior $p(\params | D,I)$   &  Canonical distribution $p(\boldsymbol{x}; T)$     & \\ 
+%Marginal likelihood $P(D|I)$	&	Partition function $Z(T)$	&   Marginal likelihood $p(D\in | I)$\\
+%\hline
+%\end{tabular}
+%\caption{\it The relationship between standard Bayesian computation, statistical
+%mechanics, and ABC. In each case Monte Carlo methods are used to calculate
+%integrals over a space. In standard Bayesian computation, it is the parameter
+%space of a model that is usually of interest, whereas in ABC it is the space
+%of possible parameter values {\it and} data sets.
+%\label{tab:relation}}
+%\end{table}
+
+%\begin{figure}[ht!]
+%\centering
+%\includegraphics[scale=0.5]{figures/joint.pdf}
+%\caption{\it The joint prior for parameters $\params$ and data $\data$.
+%\label{fig:joint}}
+%\end{figure}
+
+%The main challenges associated with ABC are:
+%\begin{enumerate}[(a)]
+%\item The choice of summary statistics
+%\item The choice of distance function
+%\item How to choose the value of $\epsilon$, or how to change it as a run
+%progresses
+%\item How to achieve good mixing.
+%\end{enumerate}
+%Challenges 1 and 2 are Bayesian in nature, i.e. they relate to the very
+%definition of the posterior distribution and are well defined independent of
+%the fact that we are going to use a Monte Carlo method to compute the
+%results. On the other hand, challenges 3 and 4 are about the computational implementation.
+
+%\section{Parameterising the data space}
+%In standard MCMC-based ABC, proposal moves are made by first proposing a
+%change to the parameters (from $\params$ to $\params'$), and then generating
+%a mock dataset $\data'$ from $p(\data | \params')$. As a Metropolis proposal
+%for exploring the product space $\Theta \times \mathcal{D}$ this is a very
+%aggressive choice, since the proposed value of the
+%dataset $\params'$ is likely to be very different from the previous one.
+%One of the most elementary pieces of advice we receive when learning the
+%Metropolis algorithm is to be careful about how ambitious our proposal
+%distributions are, since tiny and huge proposal moves are both very
+%inefficient. The performance of MCMC-based ABC can therefore be improved by
+%alternative choices of proposal distribution.
+
+%In some cases, we might be able to find efficient proposal distributions for exploring $\Theta \times \mathcal{D}$ by having insight into the particular
+%problem we're working on. However, there are also techniques that are
+%completely general in that they should work on any ABC-type problem.
+
+%Consider the process of generating a mock dataset. Whatever else is involved,
+%this process will involve calls to an underlying random number generator.
+%Perhaps a function {\tt rand()}, generating variables from a Uniform(0, 1)
+%distribution, is called $n$ times in the course of generating a dataset.
+%Usually, the simulation process has the property that, if $\theta'$ is
+%close to $\theta$, then the same sequence of random numbers returned by
+%{\tt rand()} would produce a dataset $\data'$ that is very similar to $\data$.
+%This suggests an alternative parameterisation of
+%the product space $\Theta\times\mathcal{D}$: instead of using the one that
+%naturally describes the data, we can use a coordinate system defined by
+%$\theta$ together with $\{u_1, ..., u_n\}$, the $n$ uniform random numbers
+%used to construct a mock dataset.
+
+%In this alternative coordinate system, a proposed move
+%to change $\params$ while keeping
+%$\{u_i\}$ fixed results in a proposed new dataset that is similar to the
+%previous one, assuming the change to $\params$ is small. The standard procedure
+%of generating an entirely new mock dataset each step is now viewed as a
+%proposal where $\params$ is moved {\it and all of the $u_i$ coordinates are
+%re-generated from their iid uniform prior}.
 
 \subsection{Python Bindings}\label{sec:python_bindings}
+
+{\bf To be written by Dan FM}
 
@@
 
-\bibitem[Caticha(2008)]{caticha} Caticha, A.\ 2008.\ Lectures 
-on Probability, Entropy, and Statistical Physics.\ ArXiv e-prints 
-arXiv:0808.0012.
+\bibitem[Caticha and Giffin(2006)]{caticha} Caticha, A., Giffin, A.\ 2006.\ Updating Probabilities.\ Bayesian Inference and Maximum Entropy Methods In Science and Engineering 872, 31-42.
 
@@
 \& Bridges}{2009}]{multinest} Feroz F., Hobson M.~P., Bridges M., 2009, MNRAS, 398, 1601
+
+\bibitem[Giffin and Caticha(2007)]{giffin} Giffin, A., Caticha, A.\ 2007.\ Updating Probabilities with Data and Moments.\ Bayesian Inference and Maximum Entropy Methods in Science and Engineering 954, 74-84.
 

### 2016-05-26 paper/ms.tex
More subsections

@@
 \end{enumerate}
+These are all described (using the linear regression example) in the
+following subsections.
 
@@
 
+\subsection{Generating from the prior}
 The member function used to generate straight line parameters from the
 prior is:
-
 \begin{CodeChunk}
@@
 
+\subsection{Proposal moves}
 The \code{perturb} member function for the straight line model is:
@@
 
+\subsection{Log-likelihood}
 The log-likelihood code is standard:
@@
 
+\subsection{Parameter output}
 The print and description functions are very simple:
@@
 \code{postprocess} function assumes this is the case).
+
+\subsection{Running the sampler}
+main.cpp
 

### 2016-05-27 paper/ms.tex
main.cpp

@@
 \subsection{Running the sampler}
-main.cpp
+The file {\tt main.cpp} contains the \code{main()} function which
+is executed after compiling and linking. The contents of {\tt main.cpp}
+for the linear regression example are:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+#include <iostream>
+#include "Data.h"
+#include "DNest4/code/DNest4.h"
+#include "StraightLine.h"
+
+using namespace std;
+using namespace DNest4;
+
+int main(int argc, char** argv)
+{
+	Data::get_instance().load("road.txt");
+	start<StraightLine>(argc, argv);
+	return 0;
+}
+\end{CodeInput}
+\end{CodeChunk}
+
+The first line of {\tt main()} loads the data into its global instance
+(so it can be accessed from within the log-likelihood function)
+and the second line uses a \code{start} template function to construct
+and run the sampler.
 

### 2016-05-27 paper/ms.tex
More abc talk

@@
 generating a dataset that matches the observed one is extremely low.
-To work around this, the proposition
-So the proposition $D = D_{\rm observed}$ with a logically
-weaker one.
+To work around this, we replace the proposition
+$D = D_{\rm observed}$ with a logically
+weaker one (i.e. one that is implied by $D = D_{\rm observed}$ but does not
+imply it). This is usually done by defining a few {\em summary statistics}
+$s_1(D), s_2(D), ..., s_n(D)$, which hopefully capture most of the relevant
+information in the data.
+Then, a discrepancy function
+$\rho$ is defined, measuring how different a simulated dataset $D$ is from
+the real one $D_{\rm observed}$, in terms of the summary statistics:
+\begin{align}
+\rho\left(\{s_1(D), ..., s_n(D)\}; \{s_1(D_{\rm observed}), ..., s_n(D_{\rm observed})\}\right)
+\end{align}
+The analysis then proceeds by Monte Carlo sampling the
+joint posterior for $\theta$ and $D$
+conditional on the logically weaker proposition
+\begin{align}
+\rho\left(D; D_{\rm observed}\right) < \epsilon
+\end{align}
+where $\epsilon$ is some small number.
 

### 2016-05-27 paper/ms.tex
RJObject PGM
Extra thanking

@@
 
+\begin{figure}[ht!]
+\begin{center}
+\includegraphics[width=0.4\textwidth]{figures/rjobject_pgm.pdf}
+\caption{\label{fig:rjobject_pgm}}
+\end{center}
+\end{figure}
+
 \section{Approximate Bayesian Computation}
@@
 \section*{Acknowledgements}
+John Veitch and Jorge Alarcon Ochoa provided helpful comments on an early
+version of the manuscript.
 It is a pleasure to thank Anna Pancoast (Harvard),

### 2016-05-28 paper/ms.tex
Removed lots of parentheses as suggested by Veitch

@@
 unknown quantities in the context of a particular dataset and the assumed
-prior information. The marginal likelihood (also known as the ``evidence'')
-is a key quantity in Bayesian model averaging. The Diffusive Nested Sampling algorithm, a variant of Nested Sampling, is a powerful tool for generating
+prior information. The marginal likelihood, also known as the ``evidence'',
+is a key quantity in Bayesian model selection. The Diffusive Nested Sampling algorithm, a variant of Nested Sampling, is a powerful tool for generating
 posterior samples and estimating marginal likelihoods. It is effective at
 solving complex problems including many where the posterior distribution is
-multimodal or has strong dependencies between variables. The key requirement
-is that the likelihood function be fast to evaluate.
+multimodal or has strong dependencies between variables.
 \pkg{DNest4} is an open source (MIT licensed),
@@
 simple Bayesian data analysis, finite mixture models, and Approximate
-Bayesian Computation (ABC).
+Bayesian Computation.
 }
@@
 an analysis is typically
-posterior probabilities of various hypotheses, or (more specifically)
+posterior probabilities of various hypotheses, or
 a joint posterior probability distribution for the values of unknown
@@
 
-In a very compact (but standard) notation, the posterior
+In compact but standard notation, the posterior
 posterior distribution for parameters $\params$ given data $\data$, within
@@
 is the marginal likelihood of model $j$. This kind of calculation is often
-called ``model selection'' or (more accurately) ``model averaging''.
+called ``model selection'' or ``model averaging''.
 
@@
 ``canonical'' family of distributions proportional to
-$\pi(\theta)L(\theta)^\beta$ is of interest (and $L(\theta)$ is identified as
-$\exp(-\textnormal{energy})$). NS is particularly efficient for this, since
-only a single run (which explores regions of higher and higher $L$)
-is required, and different canonical
+$\pi(\theta)L(\theta)^\beta$ is of interest. In such applications,
+$L(\theta)$ is usually equivalent to
+$\exp(-\textnormal{energy})$. NS is particularly efficient for this, since
+only a single run (exploring regions of higher and higher $L$) is needed, and different canonical
 distributions can be obtained by re-weighting the output points.
@@
 The DNS algorithm consists of two main stages. In the first stage,
-the particle(s) are initialized from the prior $\pi(\theta)$ (equivalent
+the particle(s) are initialized from the prior $\pi(\theta)$, equivalent
 to the mixture of constrained priors with a single level whose likelihood
-threshold is $-\infty$), and
-the mixture of constrained priors evolves by adding new levels
+threshold is $-\infty$. The mixture of constrained priors evolves by adding new levels
 (see \citet{dnest} for details). In this stage,

### 2016-05-28 paper/ms.tex
More veitch comments

@@
 multi-threaded implementation of this algorithm in
-modern \proglang{C++} (specifically, \proglang{C++11}),
+\proglang{C++11},
 along with associated utilities including: i)
@@
 
-If prior information $I$ (hereby dropped)
+If prior information $I$ (dropped hereafter)
 implies a set of possible `models' $\{M_i\}$,
@@
 \end{align}
-is the marginal likelihood of model $j$. This kind of calculation is often
+is the marginal likelihood of model $j$, equal to the expected value of the
+likelihood function with respect to the prior distribution.
+This kind of calculation is often
 called ``model selection'' or ``model averaging''.
@@
 \end{align}
-is the amount of prior mass which has likelihood greater than $\ell$.
+is the amount of prior mass which has likelihood greater than $\ell$, and
+$\mathds{1}()$ is the indicator function which takes the value 1 if the
+argument is true and 0 otherwise.
 In the standard NS framework, the sequence of $\ell$ values is selected
 so that $X(\ell)$ shrinks by a factor $\approx e^{-1/N}$ per iteration, where
-$N$ is the number of particles used.
+$N$ is the number of particles used. This geometric compression of the
+parameter space is a defining feature of NS.
 
@@
 problems where MCMC is the only viable sampling method. DNS is based on the
-Metropolis algorithm. DNS evolves a point in the parameter space, along with
-an integer index variable $j$,
+Metropolis algorithm. DNS evolves one or more particles in the parameter space, along with
+an integer index variable $j$ for each particle,
 to explore the following joint distribution:
@@
 to run \pkg{DNest4} on that problem. This method offers full control over the
-design of your class, and allows the opportunity (for example) to optimise
+design of your class, and allows the opportunity of optimizing
 performance by preventing a full re-computation of the log-likelihood when
-only a small subset of the parameters were changed.
+only a small subset of the parameters have been changed.
 

### 2016-05-28 paper/ms.tex
More Veitch comments

@@
 or number of saved particles with significant posterior weight.
-The {\tt postprocess} function also saves a file,
+The \code{postprocess} function also saves a file,
 {\tt posterior\_sample.txt}, containing posterior samples (one per row).
-The zero-sized error bars on $\ln(Z)$ and $\mathcal{H}$ should be ignored.
-These are only relevant if advanced options (not covered in this paper)
-are passed to {\tt postprocess}.
-
-The {\tt postprocess} function can be called while \pkg{DNest4} is running.
+Unfortunately, it is harder to compute justified error bars on $\ln(Z)$
+in DNS than it is in standard NS.
+The \code{postprocess} function does include some options
+for calculating error bars, but they are not used by default,
+and we do not describe them here.
+Hence, the zero-sized error bars on $\ln(Z)$ and $\mathcal{H}$ should
+be ignored.
+
+The \code{postprocess} function can be called while \pkg{DNest4} is running.
 This is helpful for monitoring the progress of a run, by inspecting the
@@
 
-	// Proposals explore the prior
-	// For normal priors I usually use the hastings factor to do it
+    // Proposals must be consistent with the prior
+    // Choose which of the three parameters to move
     int which = rng.rand_int(3);
+
     if(which == 0)
     {
+        // log_H takes care of the prior ratio
+        // i.e. log_H = log(pi(theta')/pi(theta))
         log_H -= -0.5*pow(m/1E3, 2);
+
+        // Take a step
         m += 1E3*rng.randh();
+
+        // log_H takes care of the prior ratio
         log_H += -0.5*pow(m/1E3, 2);
@@
     {
-		// Usual log-uniform prior trick
+        // Proposal for a parameter with a log-uniform
+        // prior takes the log of the parameter,
+        // takes a step with respect to a uniform prior,
+        // then takes the exp of the parameter
         sigma = log(sigma);
         sigma += 20.0*rng.randh();
+
+        // Wrap proposed value back into the
+        // interval allowed by the prior
         DNest4::wrap(sigma, -10.0, 10.0);

### 2016-05-28 code/Distributions/Cauchy.cpp
log_pdf

@@ -26,5 +26,10 @@ double Cauchy::cdf_inverse(double x) const
     return center + width*tan(M_PI*(x - 0.5));
 }
 
+double Cauchy::log_pdf(double x) const
+{
+    return -log(M_PI) - log(width) - log(1.0 + pow((x - center)/width, 2));
+}
+
 } // namespace DNest4
 

### 2016-05-28 code/Distributions/Cauchy.h

@@ -22,6 +22,7 @@ class Cauchy:public ContinuousDistribution
 
         double cdf(double x) const;
         double cdf_inverse(double x) const;
+        double log_pdf(double x) const;
 };
 
 } // namespace DNest4

### 2016-05-28 code/Distributions/ContinuousDistribution.h

@@ -19,6 +19,7 @@ class ContinuousDistribution
 
         virtual double cdf(double x) const = 0;
         virtual double cdf_inverse(double x) const = 0;
+        virtual double log_pdf(double x) const = 0;
 };
 
 } // namespace DNest4

### 2016-05-29 code/RNG.cpp
Factor of three

@@ -29,7 +29,8 @@ double RNG::randn()
 
 double RNG::randh()
 {
-	return pow(10., 1.5 - std::abs(tan(M_PI*(this->rand() - 0.5))))*this->randn();
+	return pow(10.0, 1.5 - 3*std::abs(tan(M_PI*(this->rand() - 0.5))))
+                        *this->randn();
 }
 
 int RNG::rand_int(int N)

### 2016-05-29 code/RNG.cpp
Use t(nu=2) instead of Cauchy

@@ -29,8 +29,10 @@ double RNG::randn()
 
 double RNG::randh()
 {
-	return pow(10.0, 1.5 - 3*std::abs(tan(M_PI*(this->rand() - 0.5))))
-                        *this->randn();
+    // t-distribution with 2 degrees of freedom (less ridiculous than Cauchy)
+    double T = this->randn()/
+                sqrt((pow(this->randn(), 2) + pow(this->randn(), 2))/2);
+	return pow(10.0, 1.5 - 3*std::abs(T))*this->randn();
 }
 
 int RNG::rand_int(int N)

### 2016-05-29 paper/ms.tex
Factor of three
Updated for t distribution i randh

@@
 \begin{enumerate}
-\item Generate $c \sim$ Cauchy$(0, 1)$
+\item Generate $a, b, c \sim$ Normal$(0, 1)$
+\item Define $t := c/\sqrt{(a^2 + b^2)/2}$
 \item Generate $n \sim$ Normal$(0, 1)$
-\item Set $x := 10^{1.5 - |c|}n$
+\item Set $x := 10^{1.5 - 3|t|}n$
 \end{enumerate}
-This generates values with a maximum scale of tens--hundreds, down to
-the smallest positive value in double precision, by virtue
-of the Cauchy distribution's heavy tails.
+The variable $t$ has a student-$t$ distribution with 2 degrees of freedom.
+Overall, this procedure generates values $x$
+with a maximum scale of tens--hundreds, down to
+a minimum scale of about $10^{-30}$ with 99\% probability, by virtue
+of the $t$-distribution's heavy tails.
 For convenience, the \code{RNG} class contains a member function \code{randh()}

### 2016-05-29 paper/ms.tex
Putting ABC talk back in

@@
 \end{align}
-where $\epsilon$ is some small number.
+where $\epsilon$ is some small number. With this proposition, the
+rejection rate will still
+typically be very high, but lower than with $D = D_{\rm observed}$.
+
+\subsection{Why ABC is difficult}
+The main challenges associated with ABC are:
+\begin{enumerate}[(i)]
+\item The choice of summary statistics $s_1(D), ..., s_n(D)$
+\item The choice of distance function $\rho()$
+\item How to choose the value of $\epsilon$, or how to change it as a run
+progresses
+\item How to make algorithms more efficient than rejection sampling.
+\end{enumerate}
+Challenges (i) and (ii) are Bayesian in nature, i.e. they relate to the very
+definition of the posterior distribution itself.
+On the other hand, challenges 3 and 4 are about the computational
+implementation. Most ABC analyses are done using Sequential Monte Carlo
+\citep[SMC][]{delmoral},
 
@@
 
-%The main challenges associated with ABC are:
-%\begin{enumerate}[(a)]
-%\item The choice of summary statistics
-%\item The choice of distance function
-%\item How to choose the value of $\epsilon$, or how to change it as a run
-%progresses
-%\item How to achieve good mixing.
-%\end{enumerate}
-%Challenges 1 and 2 are Bayesian in nature, i.e. they relate to the very
-%definition of the posterior distribution and are well defined independent of
-%the fact that we are going to use a Monte Carlo method to compute the
-%results. On the other hand, challenges 3 and 4 are about the computational implementation.
-
 %\section{Parameterising the data space}
@@
 
+\bibitem[Del Moral et al(2012)]{delmoral}
+Del Moral, P., Doucet, A. and Jasra, A., 2012. An adaptive sequential Monte Carlo method for approximate Bayesian computation. Statistics and Computing, 22(5), pp.1009-1020.
+
 \bibitem[Dybowski et al(2015)]{salmonella}

### 2016-05-29 paper/ms.tex
Consistency in notation, use M for prior info, not I

@@
 the user must be able to cheaply
-generate simulated data sets from the sampling distribution $p(D|\theta, I)$.
+generate simulated data sets from the sampling distribution $p(D|\theta, M)$.
 Since a sampling distribution must be specified, it is not that there is
@@
 example by writing the posterior distribution as
-$p(\theta | D, I)$. However,
-this is shorthand for $p(\theta | D = D_{\rm observed}, I)$, and
-$D = D_{\rm observed}$ is a statement that could have been either
-true or false in the prior state of knowledge, and is known to be true
+$p(\theta | D, M)$. However,
+this is shorthand for $p(\theta | D = D_{\rm observed}, M)$.
+In the prior state of knowledge, the statement
+$D = D_{\rm observed}$ could have been either
+true or false in the prior state of knowledge, but it is known to be true
 in the posterior state of knowledge.
 In the case of ``continuous data'' (really a continuous {\it space of
-possibilities for the data before we learned it}) we're conditioning on a
+possibilities for the data before we learned it}) we condition on a
 proposition like $(D \in \mathcal{R})$ where $\mathcal{R}$ is a region, and
-then implicitly taking a limit as the size of $\mathcal{R}$ goes to zero.
+then implicitly take a limit as the size of $\mathcal{R}$ goes to zero.
 
@@
 \begin{align}
-p(\theta, D | I) &= p(\theta | I)p(D | \theta, I)
+p(\theta, D | M) &= p(\theta | M)p(D | \theta, M)
 \end{align}
@@
 \begin{align}
-p(\theta, D | D = D_{\rm observed}, I) &\propto p(\theta | I)p(D | \theta, I)
+p(\theta, D | D = D_{\rm observed}, M) &\propto p(\theta | M)p(D | \theta, M)
 \mathds{1}\left[D = D_{\rm observed}\right].
@@
 The marginal distribution for $\theta$ is also the conditional distribution
-$p(\theta | D = D_{\rm observed}, I)$, that is,
+$p(\theta | D = D_{\rm observed}, M)$, that is,
 the posterior\footnote{Incidentally, this usage of the {\em joint} posterior
@@
 
-\subsection{Why ABC is difficult}
 The main challenges associated with ABC are:
@@
 implementation. Most ABC analyses are done using Sequential Monte Carlo
-\citep[SMC][]{delmoral},
+\citep[SMC][]{delmoral}.
 

### 2016-05-31 paper/ms.tex
RJObject prior
wrap and log-uniform prior

@@
 
+In the example, the proposal for $\sigma$ is implemented by taking advantage
+of the uniform prior for $\ln(\sigma)$. So $\sigma$ is transformed by
+taking a logarithm, a proposal move is made (that satisfies detailed balance
+with respect to a uniform distribution), and then $\sigma$ is exponentiated
+again. The step for the uniform prior between -10 and +10
+uses the following code:
+\begin{CodeChunk}
+\begin{CodeInput}
+sigma += 20.0*rng.randh();
+DNest4::wrap(sigma, -10.0, 10.0);
+\end{CodeInput}
+\end{CodeChunk}
+The factor of 20 accounts for the prior width, and
+the \code{wrap(double\&, double, double)} function uses the modulo operator
+to construct periodic boundaries. For example, if the perturbation results
+in a value of 10.2, which is outside the prior range, the value is modified to
+$-9.8$.
+The \code{wrap} function has no return value and works by modifying its
+first argument, which is passed by reference.
+
 It is imperative that \code{from_prior} and
@@
 Formerly a separate project \citep{rjobject},
+
+If $N$ is the number of components, $\boldsymbol{x}_i$ is the parameters
+of the $i$th component, and $\alpha$ is the hyperparameters, the prior
+can be factorised as
+\begin{align}
+p\left(N, \alpha, \left\{\boldsymbol{x}_i\right\}_{i=1}^N\right)
+&= p(N)p(\alpha | N)p\left(\left\{\boldsymbol{x}_i\right\}_{i=1}^N | \alpha, N\right).
+\end{align}
+The specific assumptions of \pkg{RJObject} are
+\begin{align}
+p\left(N, \alpha, \left\{\boldsymbol{x}_i\right\}_{i=1}^N\right)
+&= p(N)p(\alpha)p\left(\left\{\boldsymbol{x}_i\right\}_{i=1}^N | \alpha\right).
+\end{align}
+which is depicted in Figure~\ref{fig:rjobject_pgm}.
 

### 2016-05-31 paper/ms.tex
Direct method for log-uniform prior, and why it's bad

@@
 first argument, which is passed by reference.
-
+Alternatively, the log-uniform prior, which has density proportional
+to $1/\sigma$, could have been used directly by not using the log/exponential
+trick and adding $\ln\left[(1/\sigma')/(1/\sigma)\right]$ to the
+return value. However, this isn't recommended for a distribution like this
+which covers several orders of magnitude, because the appropriate scale size
+for the proposal is less clear. This is likely to cause inefficient
+sampling.
+
+\subsection{Consistency of prior and proposal}
 It is imperative that \code{from_prior} and
 \code{perturb}
-be consistent with each other. One technique
-for verifying this is to sample the prior for a long time
+be consistent with each other, and that each implements
+the prior distributions that you want to use. One technique
+for testing this is to sample the prior for a long time
 (by setting the maximum number of levels to 1) and inspect

### 2016-05-31 python/examples/gaussian/gaussian.py
rng sounds like random number generator. Changed to 'width'

@@ -12,23 +12,24 @@ from scipy.special import erf
 
 class Model(object):
 
-    def __init__(self, ndim=5, rng=5.0):
+    def __init__(self, ndim=5, width=10.0):
         self.ndim = ndim
-        self.rng = rng
+        self.width = width
 
     def analytic_log_Z(self):
         return (
-            self.ndim * np.log(erf(self.rng/np.sqrt(2))) -
-            self.ndim * np.log(2*self.rng)
+            self.ndim * np.log(erf(0.5*self.width/np.sqrt(2))) -
+            self.ndim * np.log(self.width)
         )
 
     def from_prior(self):
-        return np.random.uniform(-self.rng, self.rng, size=(self.ndim,))
+        return np.random.uniform(-0.5*self.width, self.width,
+                                    size=(self.ndim,))
 
     def perturb(self, coords):
         i = np.random.randint(self.ndim)
-        coords[i] += 2*self.rng*dnest4.randh()
-        coords[i] = (coords[i] + self.rng) % (2*self.rng) - self.rng
+        coords[i] += self.width*dnest4.randh()
+        coords[i] = (coords[i] + 0.5*self.width) % self.width - 0.5*self.width
         return 0.0
 
     def log_likelihood(self, coords, const=-0.5*np.log(2*np.pi)):

### 2016-05-31 python/dnest4/__init__.py
Wrap function in Python

@@ -13,5 +13,5 @@ if not __DNEST4_SETUP__:
     from . import analysis
     from .sampler import DNest4Sampler
     from .deprecated import postprocess
-    from .randh import randh
-
+    from .utils import randh
+    from .utils import wrap

### 2016-05-31 python/dnest4/utils.py

@@ -1,5 +1,5 @@
 # -*- coding: utf-8 -*-
-__all__ = ["randh"]
+__all__ = ["randh", "wrap"]
 
 import numpy as np
 import numpy.random as rng
@@ -18,4 +18,8 @@ def randh(N=1):
     if len(x) == 1:
         return x[0]
     return x
+
+def wrap(x, a, b):
+    assert b > a
+    return (x - a)%(b - a) + a
 

### 2016-05-31 python/examples/gaussian/gaussian.py

@@ -29,7 +29,8 @@ class Model(object):
     def perturb(self, coords):
         i = np.random.randint(self.ndim)
         coords[i] += self.width*dnest4.randh()
-        coords[i] = (coords[i] + 0.5*self.width) % self.width - 0.5*self.width
+        # Note: use the return value of wrap, unlike in C++
+        coords[i] = dnest4.wrap(coords[i], -0.5*self.width, 0.5*self.width)
         return 0.0
 
     def log_likelihood(self, coords, const=-0.5*np.log(2*np.pi)):

### 2016-05-31 paper/ms.tex
RJobject writing

@@
 \section{Finite Mixture Models with RJObject}
-Formerly a separate project \citep{rjobject},
+Mixture models are a useful way of representing realistic prior information
+in Bayesian data analysis. To reduce the amount of effort needed to
+implement mixture models in DNS, \citet{rjobject} implemented
+a template class called \code{RJObject} to handle the MCMC moves
+required. The RJ in \code{RJObject} stands for Reversible Jump \citep{green}, as
+\code{RJObject} implements birth and death moves for mixture
+models with an unknown number of components. An updated version of
+\code{RJObject} is included in \pkg{DNest4}.
 
@@
 
+\bibitem[\protect\citeauthoryear{Green}{1995}]{green}
+Green, P.~J., 1995, Reversible Jump Markov Chain Monte Carlo Computation and Bayesian Model Determination, Biometrika 82 (4): 711–732.
+
 \bibitem[Handley et al.(2015)]{2015MNRAS.453.4384H} Handley, W.~J., Hobson, M.~P., Lasenby, A.~N.\ 2015.\ POLYCHORD: next-generation nested sampling.\ Monthly Notices of the Royal Astronomical Society 453, 4384-4398.

### 2016-05-31 python/dnest4/deprecated.py
Extra line

@@ -303,8 +303,9 @@ def levels_plot():
 
 	plt.plot(np.log10(np.diff(levels[:,1])))
 	plt.ylim([-1, 4])
-	plt.axhline(0., color='r')
+	plt.axhline(0., color='k', linewidth=2)
 	plt.axhline(np.log10(np.log(10.)), color='g')
+	plt.axhline(np.log10(0.8), color='r', linestyle='--')
 	plt.xlabel('Level')
 	plt.ylabel('Delta log likelihood')
 	plt.show()

### 2016-05-31 paper/ms.tex
Described prior in words

@@
 of the $i$th component, and $\alpha$ is the hyperparameters, the prior
-can be factorised as
+can be factorised via the product rule, giving
 \begin{align}
@@
 \end{align}
-The specific assumptions of \pkg{RJObject} are
+The specific assumptions of \pkg{RJObject} are that this simplifies to
 \begin{align}
 p\left(N, \alpha, \left\{\boldsymbol{x}_i\right\}_{i=1}^N\right)
-&= p(N)p(\alpha)p\left(\left\{\boldsymbol{x}_i\right\}_{i=1}^N | \alpha\right).
-\end{align}
-which is depicted in Figure~\ref{fig:rjobject_pgm}.
+&= p(N)p(\alpha)
+\prod_{i=1}^N
+p\left(\boldsymbol{x}_i | \alpha\right).
+\end{align}
+That is, the prior for the hyperparameters is independent of the number of
+components, and each component is independent given the hyperparameters.
+A probabilistic graphical model (PGM) is shown in Figure~\ref{fig:rjobject_pgm}.
 
@@
 \includegraphics[width=0.4\textwidth]{figures/rjobject_pgm.pdf}
-\caption{\label{fig:rjobject_pgm}}
+\caption{A PGM created using Daft ({\tt http://daft-pgm.org}).\label{fig:rjobject_pgm}}
 \end{center}

### 2016-05-31 paper/ms.tex
Nicer pgm and notation (no bold)

@@
 
-If $N$ is the number of components, $\boldsymbol{x}_i$ is the parameters
-of the $i$th component, and $\alpha$ is the hyperparameters, the prior
+If $N$ is the number of components, $x_i$ denotes the vector of parameters
+of the $i$th component, and $\alpha$ is the vector of hyperparameters, the prior
 can be factorised via the product rule, giving
 \begin{align}
-p\left(N, \alpha, \left\{\boldsymbol{x}_i\right\}_{i=1}^N\right)
-&= p(N)p(\alpha | N)p\left(\left\{\boldsymbol{x}_i\right\}_{i=1}^N | \alpha, N\right).
+p\left(N, \alpha, \left\{x_i\right\}_{i=1}^N\right)
+&= p(N)p(\alpha | N)p\left(\left\{x_i\right\}_{i=1}^N | \alpha, N\right).
 \end{align}
@@
 \begin{align}
-p\left(N, \alpha, \left\{\boldsymbol{x}_i\right\}_{i=1}^N\right)
+p\left(N, \alpha, \left\{x_i\right\}_{i=1}^N\right)
 &= p(N)p(\alpha)
 \prod_{i=1}^N
-p\left(\boldsymbol{x}_i | \alpha\right).
+p\left(x_i | \alpha\right).
 \end{align}
 That is, the prior for the hyperparameters is independent of the number of
-components, and each component is independent given the hyperparameters.
+components, and each component is independent and identically
+distributed given the hyperparameters.
 A probabilistic graphical model (PGM) is shown in Figure~\ref{fig:rjobject_pgm}.
@@
 \begin{center}
-\includegraphics[width=0.4\textwidth]{figures/rjobject_pgm.pdf}
-\caption{A PGM created using Daft ({\tt http://daft-pgm.org}).\label{fig:rjobject_pgm}}
+\hspace*{-2cm}\includegraphics[width=0.4\textwidth]{figures/rjobject_pgm.pdf}
+\caption{A PGM showing the kind of prior information the \code{RJObject}
+template class expresses.
+Figure created using Daft ({\tt http://daft-pgm.org}).\label{fig:rjobject_pgm}}
 \end{center}

### 2016-06-01 code/Examples/RJObject_1DMixture/display.py
Working, plot

@@ -1,12 +1,11 @@
 from pylab import *
 import dnest4.classic as dn4
-import os
+
+rc("font", size=18, family="serif", serif="Computer Sans")
+rc("text", usetex=True)
 
 data = loadtxt('galaxies.txt')
 posterior_sample = atleast_2d(dn4.my_loadtxt('posterior_sample.txt'))
-
-os.system("mkdir Frames")
-os.system("rm Frames/*.png")
 
 x = linspace(0., 50.0, 10001)
 def mixture(x, params):
@@ -22,23 +21,18 @@ def mixture(x, params):
 
     return y
 
-total = zeros(x.shape)
 
-for i in range(0, posterior_sample.shape[0]):
-    hold(False)
-    hist(data, 100, alpha=0.5, normed=True)
+hold(False)
+hist(data, 100, alpha=0.3, normed=True)
+
+for i in range(0, min([posterior_sample.shape[0], 100])):
     hold(True)
     y = mixture(x, posterior_sample[i, :])
-    total += y
-    plot(x, y, 'r', linewidth=2)
-    title("{a} / {b}".format(a=(i+1), b=posterior_sample.shape[0]))
-    savefig('Frames/' + '%0.6d'%(i+1) + '.png', bbox_inches='tight')
-    print('Frames/' + '%0.6d'%(i+1) + '.png')
+    plot(x, y, 'k', linewidth=2, alpha=0.2)
 
+xlabel("Velocity (1000 km/s)")
+ylabel("Density")
+savefig("galaxies.pdf", bbox_inches="tight")
 show()
 
-hist(data, 100, alpha=0.5, normed=True)
-hold(True)
-plot(x, total/posterior_sample.shape[0], 'r', linewidth=2)
-show()
 

### 2016-06-01 paper/ms.tex
Sentence and subsection
Mixture model equation

@@
 
+\citet{rjobject} described the motivation for \code{RJObject} and some details
+about the Metropolis proposals underlying it. Here, we demonstrate how to
+implement a finite mixture model using \code{RJObject}.
+
+\subsection{An example mixture model}
+Consider a sampling distribution
+for data $D=\{D_1, D_2, ..., D_n\}$ which is a mixture of $N$ gaussians
+with means $\{\mu_j\}$, standard deviations $\{\sigma_j\}$, and
+mixture weights $\{w_j\}$. The likelihood for the $i$th data point is
+\begin{align}
+p(D_i | N, \{\mu_j\}, \{\sigma_j\}, \{w_j\}) &=
+\sum_{j=1}^N \frac{w_j}{\sigma_j\sqrt{2\pi}}
+\exp\left[-\frac{1}{2\sigma_j^2}\left(D_i - \mu_j\right)^2\right].
+\end{align}
+
 \section{Approximate Bayesian Computation}

### 2016-06-01 paper/ms.tex
Affiliations

@@
 \section*{Acknowledgements}
-John Veitch and Jorge Alarcon Ochoa provided helpful comments on an early
+John Veitch (Birmingham)
+and Jorge Alarcon Ochoa (Rensselaer Polytechnic Institute)
+provided helpful comments on an early
 version of the manuscript.

### 2016-06-01 paper/ms.tex
galaxies plot
Priors for the mixture model
Removed +- 0

@@
 \begin{CodeOutput}
-log(Z) = -175.528264138 +- 0.0
-Information = 15.1944947225 +- 0.0 nats.
+log(Z) = -175.528264138
+Information = 15.1944947225 nats.
 Effective sample size = 745.819862383
@@
 in DNS than it is in standard NS.
-The \code{postprocess} function does include some options
-for calculating error bars, but they are not used by default,
-and we do not describe them here.
-Hence, the zero-sized error bars on $\ln(Z)$ and $\mathcal{H}$ should
-be ignored.
+%The \code{postprocess} function does include some options
+%for calculating error bars, but they are not used by default,
+%and we do not describe them here.
+%Hence, the zero-sized error bars on $\ln(Z)$ and $\mathcal{H}$ should
+%be ignored.
 
@@
 
+The prior for $N$ was
+\begin{align}
+p(N) \propto \frac{1}{N+1}
+\end{align}
+for $N \in \{1, 2, ..., 100\}$.
+
+We used the following priors for the hyperparameters:
+\begin{align}
+a_\mu &\sim \textnormal{Uniform}(-1000, 1000)\\
+\ln b_\mu &\sim \textnormal{Uniform}(-10, 10)\\
+a_{\ln\sigma} &\sim \textnormal{Uniform}(-10, 10)\\
+b_{\ln\sigma} &\sim \textnormal{Uniform}(0, 5)\\
+b_{\ln W} &\sim \textnormal{Uniform}(0, 5)
+\end{align}
+Location parameters $a$, scale parameters $b$.
+
+The priors for the parameters given the hyperparameters were:
+\begin{align}
+\mu_j &\sim \textnormal{Laplace}(a_\mu, b_\mu)\\
+\ln\sigma_j &\sim \textnormal{Laplace}(a_{\ln\sigma}, b_{\ln\sigma})\\
+\ln W_j &\sim \textnormal{Laplace}(0, b_{\ln W}).
+\end{align}
+
+A Laplace (or bi-exponential)
+distribution with location parameter $a$ and scale parameter
+$b$ has density
+\begin{align}
+p(x | a, b) &= \frac{1}{2}\exp\left(-\frac{1}{b}|x - a|\right).
+\end{align}
+
+The normalized weights are $w_k = W_k/\sum w_j$.
+
+Some illustrative posterior samples are plotted
+in Figure~\ref{fig:galaxies}.
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=0.6\textwidth]{figures/galaxies.pdf}
+\caption{\label{fig:galaxies}}
+\end{center}
+\end{figure}
+
 \section{Approximate Bayesian Computation}

### 2016-06-01 python/dnest4/deprecated.py

@@ -234,8 +234,13 @@ def postprocess(temperature=1., numResampleLogX=1, plot=True, loaded=[], \
 	H_error = np.std(H_estimates)
 	ESS = np.exp(-np.sum(P_samples*np.log(P_samples+1E-300)))
 
-	print("log(Z) = " + str(logz_estimate) + " +- " + str(logz_error))
-	print("Information = " + str(H_estimate) + " +- " + str(H_error) + " nats.")
+	errorbar1 = ""
+	errorbar2 = ""
+	if numResampleLogX > 1:
+		errorbar1 += " +- " + str(logz_error)
+		errorbar2 += " +- " + str(H_error)
+	print("log(Z) = " + str(logz_estimate) + errorbar1)
+	print("Information = " + str(H_estimate) + errorbar2 + " nats.")
 	print("Effective sample size = " + str(ESS))
 
 	# Resample to uniform weight

### 2016-06-02 paper/ms.tex
More RJObject writing

@@
 \prod_{i=1}^N
-p\left(x_i | \alpha\right).
+p\left(x_i | \alpha\right).\label{eqn:rjobject}
 \end{align}
@@
 distributed given the hyperparameters.
-A probabilistic graphical model (PGM) is shown in Figure~\ref{fig:rjobject_pgm}.
+A probabilistic graphical model (PGM) is showing this dependence
+structure is shown in Figure~\ref{fig:rjobject_pgm}. There are no observed
+data nodes here, but if such a structure forms part of a
+Bayesian model, an \code{RJObject} object within your model class
+can encapsulate this part of the model.
 
@@
 implement a finite mixture model using \code{RJObject}.
+This example can be found in the
+\begin{center}
+{\tt code/Examples/RJObject\_1DMixture}
+\end{center}
+directory.
 
@@
 \end{align}
-
-The prior for $N$ was
+Colloquially, one might say the data were 'drawn from' a mixture of
+$N$ normal distributions, and we want to infer $N$ along with the
+properties of those normal distributions.
+
+Making the connection with Equation~\ref{eqn:rjobject} and
+Figure~\ref{fig:rjobject_pgm}, the ``components'' are the gaussians,
+the gaussian parameters are $x_j = \{\mu_j, \sigma_j, w_j\}$
+and hyperparameters $\alpha$ may be used to help specify
+a sensible joint prior for the gaussian parameters.
+To implement this model for \pkg{DNest4}, the model
+class only needs to contain an instance of an
+\code{RJObject}, like so:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+class MyModel
+{
+	private:
+        DNest4::RJObject<MyConditionalPrior> gaussians;
+....
+\end{CodeInput}
+\end{CodeChunk}
+
+The specific prior for $N$ was
 \begin{align}
@@
 
-\subsection{Python Bindings}\label{sec:python_bindings}
+\section{Python Bindings}\label{sec:python_bindings}
 

### 2016-06-02 code/Examples/RJObject_1DMixture/display.py
Posterior for N

@@ -35,4 +35,13 @@ ylabel("Density")
 savefig("galaxies.pdf", bbox_inches="tight")
 show()
 
+width = 0.3
+bins = arange(0, 101) - 0.5*width
+hist(posterior_sample[:,7], bins, width=width, normed=True, alpha=0.3)
+xlim([0, 10])
+ylim([0, 0.5])
+xlabel("Number of gaussians, $N$")
+ylabel("Posterior Probability")
+savefig("galaxies_N.pdf", bbox_inches="tight")
+show()
 

### 2016-06-02 paper/ms.tex
Mention template argument
New figure, posterior for N

@@
 \end{CodeChunk}
+where the template argument \code{<MyConditionalPrior>}
+is a class implementing the prior for the hyperparameters
+$\alpha$ and the form of the conditional prior $p(x_i | \alpha)$.
 
@@
 for $N \in \{1, 2, ..., 100\}$.
+For the conditional prior of the gaussian parameters,
+we used Laplace (biexponential)
+distributions\footnote{A Laplace
+distribution with location parameter $a$ and scale parameter
+$b$ has density $p(x | a, b) = \frac{1}{2}\exp\left(-\frac{1}{b}|x - a|\right)$.}
+
+\begin{align}
+\mu_j &\sim \textnormal{Laplace}(a_\mu, b_\mu)\\
+\ln\sigma_j &\sim \textnormal{Laplace}(a_{\ln\sigma}, b_{\ln\sigma})\\
+\ln W_j &\sim \textnormal{Laplace}(0, b_{\ln W}).
+\end{align}
 
@@
 
-The priors for the parameters given the hyperparameters were:
-\begin{align}
-\mu_j &\sim \textnormal{Laplace}(a_\mu, b_\mu)\\
-\ln\sigma_j &\sim \textnormal{Laplace}(a_{\ln\sigma}, b_{\ln\sigma})\\
-\ln W_j &\sim \textnormal{Laplace}(0, b_{\ln W}).
-\end{align}
-
-A Laplace (or bi-exponential)
-distribution with location parameter $a$ and scale parameter
-$b$ has density
-\begin{align}
-p(x | a, b) &= \frac{1}{2}\exp\left(-\frac{1}{b}|x - a|\right).
-\end{align}
-
 The normalized weights are $w_k = W_k/\sum w_j$.
@@
 \begin{figure}
-\begin{center}
-\includegraphics[width=0.6\textwidth]{figures/galaxies.pdf}
-\caption{\label{fig:galaxies}}
-\end{center}
+\begin{minipage}{0.45\textwidth}
+\centering
+\includegraphics[width=0.95\textwidth]{figures/galaxies.pdf}
+\caption{The `galaxy' data, with some illustrative fits drawn
+from the posterior distribution.\label{fig:galaxies}}
+\end{minipage} \hspace{0.05\textwidth}
+\begin{minipage}{0.45\textwidth}
+\centering
+\includegraphics[width=0.95\textwidth]{figures/galaxies_N.pdf}
+\caption{The posterior distribution for $N$ given the
+galaxy data.\label{fig:galaxies_N}}
+\end{minipage}
 \end{figure}

### 2016-06-02 code/Examples/RJObject_1DMixture/display.py

@@ -38,7 +38,7 @@ show()
 width = 0.3
 bins = arange(0, 101) - 0.5*width
 hist(posterior_sample[:,7], bins, width=width, normed=True, alpha=0.3)
-xlim([0, 10])
+xlim([0, 15])
 ylim([0, 0.5])
 xlabel("Number of gaussians, $N$")
 ylabel("Posterior Probability")

### 2016-06-02 paper/ms.tex
ABC writing

@@
 p(\theta, D | D = D_{\rm observed}, M) &\propto p(\theta | M)p(D | \theta, M)
-\mathds{1}\left[D = D_{\rm observed}\right].
+\mathds{1}\left[D = D_{\rm observed}\right].\label{eqn:abc_posterior}
 \end{align}
@@
 weaker one (i.e. one that is implied by $D = D_{\rm observed}$ but does not
-imply it). This is usually done by defining a few {\em summary statistics}
-$s_1(D), s_2(D), ..., s_n(D)$, which hopefully capture most of the relevant
-information in the data.
-Then, a discrepancy function
-$\rho$ is defined, measuring how different a simulated dataset $D$ is from
-the real one $D_{\rm observed}$, in terms of the summary statistics:
-\begin{align}
-\rho\left(\{s_1(D), ..., s_n(D)\}; \{s_1(D_{\rm observed}), ..., s_n(D_{\rm observed})\}\right)
-\end{align}
+imply it).
+
+The weaker proposition is defined using
+a discrepancy function
+$\rho$, which measures how different a simulated dataset $D$ is from
+the real one $D_{\rm observed}$:
+\begin{align}
+\rho\left(D; D_{\rm observed}\right).
+\end{align}
+The discrepancy function should take the value zero when
+$D = D_{\rm observed}$.
+
 The analysis then proceeds by Monte Carlo sampling the
@@
 typically be very high, but lower than with $D = D_{\rm observed}$.
+Typically, $\rho$ is defined by introducing a few {\em summary statistics}
+$s_1(D), s_2(D), ..., s_n(D)$, which hopefully capture most of the relevant
+information in the data, and then computing the Euclidian distance
+between the summary statistics of $D$ and $D_{\rm observed}$.
+That is,
+\begin{align}
+\rho(D; D_{\rm observed}) &=
+\sqrt{\sum_i \left[s_i(D) - s_i(D_{\rm observed})\right]^2}.
+\end{align}
 
@@
 \begin{enumerate}[(i)]
-\item The choice of summary statistics $s_1(D), ..., s_n(D)$
-\item The choice of distance function $\rho()$
+\item The choice of discrepancy function $\rho(D; D_{\rm observed})$, which
+may involve a choice of summary statistics,
 \item How to choose the value of $\epsilon$, or how to change it as a run
-progresses
+progresses,
 \item How to make algorithms more efficient than rejection sampling.
 \end{enumerate}
-Challenges (i) and (ii) are Bayesian in nature, i.e. they relate to the very
+Challenge (i) is Bayesian in nature, i.e. it relates to the very
 definition of the posterior distribution itself.
-On the other hand, challenges 3 and 4 are about the computational
+On the other hand, challenges (ii) and (iii) are about the computational
 implementation. Most ABC analyses are done using Sequential Monte Carlo
-\citep[SMC][]{delmoral}.
+\citep[SMC; ][]{delmoral}, another family of algorithms closely related
+to Nested Sampling and annealing in the sense that they work with
+sequences of probability distributions.
+
+NS can be used for ABC by noting that Equation~\ref{eqn:abc_posterior}
+is of the same form as a constrained prior.
 

### 2016-06-02 paper/ms.tex
microtype

--- 2016-06-02 paper/ms.tex
@@
 \usepackage{enumerate}
+\usepackage{microtype}
 \usepackage{dsfont}
@@
 \prod_{i=1}^N
-p\left(x_i | \alpha\right).\label{eqn:rjobject}
+p\left(x_i | \alpha\right).
 \end{align}
@@
 distributed given the hyperparameters.
-A probabilistic graphical model (PGM) is showing this dependence
-structure is shown in Figure~\ref{fig:rjobject_pgm}. There are no observed
-data nodes here, but if such a structure forms part of a
-Bayesian model, an \code{RJObject} object within your model class
-can encapsulate this part of the model.
+A probabilistic graphical model (PGM) is shown in Figure~\ref{fig:rjobject_pgm}.
 
@@
 implement a finite mixture model using \code{RJObject}.
-This example can be found in the
-\begin{center}
-{\tt code/Examples/RJObject\_1DMixture}
-\end{center}
-directory.
 
@@
 \end{align}
-Colloquially, one might say the data were 'drawn from' a mixture of
-$N$ normal distributions, and we want to infer $N$ along with the
-properties of those normal distributions.
-
-Making the connection with Equation~\ref{eqn:rjobject} and
-Figure~\ref{fig:rjobject_pgm}, the ``components'' are the gaussians,
-the gaussian parameters are $x_j = \{\mu_j, \sigma_j, w_j\}$
-and hyperparameters $\alpha$ may be used to help specify
-a sensible joint prior for the gaussian parameters.
-To implement this model for \pkg{DNest4}, the model
-class only needs to contain an instance of an
-\code{RJObject}, like so:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-class MyModel
-{
-	private:
-        DNest4::RJObject<MyConditionalPrior> gaussians;
-....
-\end{CodeInput}
-\end{CodeChunk}
-
-The specific prior for $N$ was
+
+The prior for $N$ was
 \begin{align}
@@
 
-\section{Python Bindings}\label{sec:python_bindings}
+\subsection{Python Bindings}\label{sec:python_bindings}
 

### 2016-06-03 paper/ms.tex
New figure
I --> M
Joint prior

--- 2016-06-02 paper/ms.tex
@@
 \begin{align}
-\mathcal{H} &= \int p(\theta|D, I)
-\ln\left[\frac{p(\theta | D, I)}{p(\theta | I)}\right] \, d\theta,
+\mathcal{H} &= \int p(\theta|D, M)
+\ln\left[\frac{p(\theta | D, M)}{p(\theta | M)}\right] \, d\theta,
 \end{align}
@@
 
+\begin{figure}[ht!]
+\centering
+\includegraphics[width=0.7\textwidth]{figures/galaxies_fig3.pdf}
+\caption{The log-likelihood and posterior weights as a function
+of compression $X$ for the galaxy data. There are phase changes
+\citep{skilling} for which Nested Sampling is the best known
+solution.
+\label{fig:galaxies_fig3}}
+\end{figure}
+
 \section{Approximate Bayesian Computation}
@@
 NS can be used for ABC by noting that Equation~\ref{eqn:abc_posterior}
-is of the same form as a constrained prior.
+is of the same form as a constrained prior, only instead of being the prior
+for the parameters $p(\theta | M)$ defined on the parameter space, it is
+the {\em joint prior} for the parameters and the data,
+$p(\theta, D |M) = p(\theta | M)p(D | \theta, M)$, defined on the product
+space of possible parameters and datasets.
 

### 2016-06-03 paper/ms.tex
RJObject writing

@@
 
+To help with implementing the member functions for the \code{MyModel} class,
+the \code{gaussians} object can be generated from its prior, proposed,
+and printed trivially as follows:
+\begin{CodeInput}
+\begin{CodeChunk}
+    // Generate the components from the prior
+    gaussians.from_prior(rng);
+
+    // Do a Metropolis proposal
+    double logH = gaussians.perturb(rng);
+
+    // Print to output stream 'out'
+    gaussians.print(out);
+\end{CodeChunk}
+\end{CodeInput}
+
 The specific prior for $N$ was
@@
 distribution with location parameter $a$ and scale parameter
-$b$ has density $p(x | a, b) = \frac{1}{2}\exp\left(-\frac{1}{b}|x - a|\right)$.}
+$b$ has density $p(x | a, b) = \frac{1}{2}\exp\left(-\frac{1}{b}|x - a|\right)$.}.
+Normal distributions would be more convenional, but the analytic cumulative
+distribution function (CDF) of the Laplace distribution makes it easier to
+implement.
 

### 2016-06-03 paper/ms.tex
Simpler figure

@@
 
-Some illustrative posterior samples are plotted
+The data and the posterior mean fit are shown
 in Figure~\ref{fig:galaxies}.
@@
 \includegraphics[width=0.95\textwidth]{figures/galaxies.pdf}
-\caption{The `galaxy' data, with some illustrative fits drawn
-from the posterior distribution.\label{fig:galaxies}}
+\caption{The `galaxy' data, with the posterior mean fit
+(equivalent to the predictive distribution for the ``next'' data point).\label{fig:galaxies}}
 \end{minipage} \hspace{0.05\textwidth}

### 2016-06-03 paper/ms.tex
Changed plot, more writing

@@
 This example can be found in the
-\begin{center}
 {\tt code/Examples/RJObject\_1DMixture}
-\end{center}
-directory.
+directory. The ``SineWave'' and ``GalaxyField'' models
+from \citet{rjobject} are also included in the \pkg{DNest4} repository.
 
@@
 Figure~\ref{fig:rjobject_pgm}, the ``components'' are the gaussians,
-the gaussian parameters are $x_j = \{\mu_j, \sigma_j, w_j\}$
+the gaussian parameters are $\{x_j\} = \{(\mu_j, \sigma_j, w_j)\}$,
 and hyperparameters $\alpha$ may be used to help specify
@@
 and printed trivially as follows:
+\begin{CodeChunk}
 \begin{CodeInput}
-\begin{CodeChunk}
     // Generate the components from the prior
@@
     gaussians.print(out);
+\end{CodeInput}
 \end{CodeChunk}
+
+The \code{RJObject} constructor definition is:
+\begin{CodeChunk}
+\begin{CodeInput}
+RJObject(int num_dimensions, int max_num_components, bool fixed,
+         const ConditionalPrior& conditional_prior,
+         PriorType prior_type=PriorType::uniform);
 \end{CodeInput}
+\end{CodeChunk}
 

### 2016-06-03 code/Examples/RJObject_1DMixture/display.py

@@ -23,12 +23,15 @@ def mixture(x, params):
 
 
 hold(False)
-hist(data, 100, alpha=0.3, normed=True)
+hist(data, 100, alpha=0.2, normed=True)
+y_tot = zeros(len(x))
 
-for i in range(0, min([posterior_sample.shape[0], 100])):
-    hold(True)
+for i in range(0, posterior_sample.shape[0]):
     y = mixture(x, posterior_sample[i, :])
-    plot(x, y, 'k', linewidth=2, alpha=0.2)
+    y_tot += y
+
+hold(True)
+plot(x, y_tot/posterior_sample.shape[0], 'k', linewidth=2)
 
 xlabel("Velocity (1000 km/s)")
 ylabel("Density")

### 2016-06-03 paper/ms.tex
Lots of ABC writing

@@
 \section{Approximate Bayesian Computation}
+\subsection{Background}
 Nested Sampling can be used to solve Approximate Bayesian Computation (ABC)
@@
 imply it).
-
 The weaker proposition is defined using
@@
 \end{align}
-The discrepancy function should take the value zero when
+The discrepancy function should take a minimum value of zero when
 $D = D_{\rm observed}$.
-
 The analysis then proceeds by Monte Carlo sampling the
@@
 
+\subsection{ABC with Nested Sampling}
 NS can be used for ABC by noting that Equation~\ref{eqn:abc_posterior}
@@
 $p(\theta, D |M) = p(\theta | M)p(D | \theta, M)$, defined on the product
-space of possible parameters and datasets.
+space of possible parameters and datasets. Therefore, to implement
+ABC in \pkg{DNest4}, you implement a model class
+whose \code{from\_prior} member function generates parameters {\em and simulated
+data} from the joint prior, and whose \code{perturb} member function
+proposes changes to the parameters and/or the simulated data, in a
+manner consistent with the joint prior. The \code{log\_likelihood} function
+of the class, instead of actually evaluating the log likelihood, should
+evaluate $-\rho(D; D_{\rm observed})$. Then, running the sampler
+will create levels which systematically decrease $\rho$. The samples from
+a particular level then represent the ABC posterior defined in
+Equation~\ref{eqn:abc_posterior} with $\epsilon$ corresponding to minus
+the ``log likelihood'' of the level.
+A single \pkg{DNest4} run allows you to test sensitivity
+to the value of $\epsilon$ using a single run.
+
+\subsection{ABC Example}
+A simple ABC example is included in the directory
+{\tt code/Examples/ABC}. The example tries to infer the mean
+$\mu$ and standard deviation $\sigma$ of a normal distribution
+from a set of samples. The sampling distribution is
+\begin{align}
+x_i &\sim \textnormal{Normal}(\mu, \sigma^2).
+\end{align}
+and the prior is
+\begin{align}
+\mu &\sim \textnormal{Uniform}(-10, 10)\\
+\ln\sigma &= \textnormal{Uniform}(-10, 10).
+\end{align}
+However, instead of conditioning on the full dataset $\{x_i\}$, the example
+uses the minimum and maximum values in the data,
+$\min(\{x_i\})$ and $\max(\{x_i\})$, as summary statistics\footnote{This is
+not a serious application of ABC, since the likelihood function can easily
+be evaluated. It is included to demonstrate the principle of using
+Nested Sampling as an ABC method.}.
 
@@
 
-\bibitem[Giffin and Caticha(2007)]{giffin} Giffin, A., Caticha, A.\ 2007.\ Updating Probabilities with Data and Moments.\ Bayesian Inference and Maximum Entropy Methods in Science and Engineering 954, 74-84.
+\bibitem[Giffin and Caticha(2007)]{giffin} Giffin, A., Caticha, A.\ 2007.\ Updating Probabilities with Data and Moments.\ Bayesian Inference and Maximum Entropy
+Methods in Science and Engineering 954, 74-84.
 

### 2016-06-03 code/Examples/ABC/showresults.py
Efficiency

@@ -1,32 +1,86 @@
-# For ABC, the showresults needs to do different things...
 import copy
 import numpy as np
+import pandas as pd
 import matplotlib.pyplot as plt
-from dnest4.classic import logsumexp, logdiffexp
 
-def postprocess(temperature=1E300, numResampleLogX=1, plot=True, loaded=[], \
-			cut=0., save=True, zoom_in=True, compression_bias_min=1., compression_scatter=0., moreSamples=1., compression_assert=None):
+def logsumexp(values):
+	biggest = np.max(values)
+	x = values - biggest
+	result = np.log(np.sum(np.exp(x))) + biggest
+	return result
+
+def logdiffexp(x1, x2):
+	biggest = x1
+	xx1 = x1 - biggest
+	xx2 = x2 - biggest
+	result = np.log(np.exp(xx1) - np.exp(xx2)) + biggest
+	return result
+
+def my_loadtxt(filename, single_precision=False):
+    if single_precision:
+        return pd.read_csv(filename, header=None, sep=' ',\
+                           comment="#", dtype=np.float32)\
+                                    .dropna(axis=1).values
+    return pd.read_csv(filename, header=None, sep=' ', comment="#")\
+                                                 .dropna(axis=1).values
+
+def loadtxt_rows(filename, rows, single_precision=False):
+    """
+    Load only certain rows
+    """
+    # Open the file
+    f = open(filename, "r")
+
+    # Storage
+    results = {}
+
+    # Row number
+    i = 0
+
+    # Number of columns
+    ncol = None
+
+    while(True):
+        # Read the line and split by whitespace
+        line = f.readline()
+        cells = line.split()
+
+        # Quit when you see a different number of columns
+        if ncol is not None and len(cells) != ncol:
+            break
+
+        # Non-comment lines
+        if cells[0] != "#":
+            # If it's the first one, get the number of columns
+            if ncol is None:
+                ncol = len(cells)
+
+            # Otherwise, include in results
+            if i in rows:
+                if single_precision:
+                    results[i] = np.array([float(cell) for cell in cells],\
[…]

### 2016-06-03 code/Examples/ABC/showresults.py
ABC results plot

@@ -2,71 +2,7 @@ import copy
 import numpy as np
 import pandas as pd
 import matplotlib.pyplot as plt
-
-def logsumexp(values):
-	biggest = np.max(values)
-	x = values - biggest
-	result = np.log(np.sum(np.exp(x))) + biggest
-	return result
-
-def logdiffexp(x1, x2):
-	biggest = x1
-	xx1 = x1 - biggest
-	xx2 = x2 - biggest
-	result = np.log(np.exp(xx1) - np.exp(xx2)) + biggest
-	return result
-
-def my_loadtxt(filename, single_precision=False):
-    if single_precision:
-        return pd.read_csv(filename, header=None, sep=' ',\
-                           comment="#", dtype=np.float32)\
-                                    .dropna(axis=1).values
-    return pd.read_csv(filename, header=None, sep=' ', comment="#")\
-                                                 .dropna(axis=1).values
-
-def loadtxt_rows(filename, rows, single_precision=False):
-    """
-    Load only certain rows
-    """
-    # Open the file
-    f = open(filename, "r")
-
-    # Storage
-    results = {}
-
-    # Row number
-    i = 0
-
-    # Number of columns
-    ncol = None
-
-    while(True):
-        # Read the line and split by whitespace
-        line = f.readline()
-        cells = line.split()
-
-        # Quit when you see a different number of columns
-        if ncol is not None and len(cells) != ncol:
-            break
-
-        # Non-comment lines
-        if cells[0] != "#":
-            # If it's the first one, get the number of columns
-            if ncol is None:
-                ncol = len(cells)
-
-            # Otherwise, include in results
-            if i in rows:
-                if single_precision:
-                    results[i] = np.array([float(cell) for cell in cells],\
-                                                              dtype="float32")
-                else:
-                    results[i] = np.array([float(cell) for cell in cells])
-            i += 1
-
-    results["ncol"] = ncol
-    return results
-
+from dnest4.classic import logsumexp, logdiffexp, my_loadtxt, loadtxt_rows
 
 def postprocess(temperature=1., numResampleLogX=1, plot=True, loaded=[], \
[…]

### 2016-06-05 code/Examples/ABC/showresults.py
showresults_abc in deprecated.py

@@ -1,235 +1,12 @@
-import copy
 import numpy as np
-import pandas as pd
 import matplotlib.pyplot as plt
-from dnest4.classic import logsumexp, logdiffexp, my_loadtxt, loadtxt_rows
-
-def postprocess(temperature=1., numResampleLogX=1, plot=True, loaded=[], \
-			cut=0., save=True, zoom_in=True, compression_bias_min=1., compression_scatter=0., moreSamples=1., compression_assert=None, single_precision=False):
-	if len(loaded) == 0:
-		levels_orig = np.atleast_2d(my_loadtxt("levels.txt"))
-		sample_info = np.atleast_2d(my_loadtxt("sample_info.txt"))
-	else:
-		levels_orig, sample_info = loaded[0], loaded[1]
-
-	# Remove regularisation from levels_orig if we asked for it
-	if compression_assert is not None:
-		levels_orig[1:,0] = -np.cumsum(compression_assert*np.ones(levels_orig.shape[0] - 1))
-
-	sample_info = sample_info[int(cut*sample_info.shape[0]):, :]
-
-	if plot:
-		if numResampleLogX > 1:
-			plt.ion()
-
-		plt.figure(1)
-		plt.plot(sample_info[:,0])
-		plt.xlabel("Iteration")
-		plt.ylabel("Level")
-		if numResampleLogX > 1:
-			plt.draw()
-
-		plt.figure(2)
-		plt.subplot(2,1,1)
-		plt.plot(np.diff(levels_orig[:,0]))
-		plt.ylabel("Compression")
-		plt.xlabel("Level")
-		xlim = plt.gca().get_xlim()
-		plt.axhline(-1., color='r')
-		plt.axhline(-np.log(10.), color='g')
-		plt.ylim(ymax=0.05)
-		if numResampleLogX > 1:
-			plt.draw()
-
-		plt.subplot(2,1,2)
-		good = np.nonzero(levels_orig[:,4] > 0)[0]
-		plt.plot(levels_orig[good,3]/levels_orig[good,4])
-		plt.xlim(xlim)
-		plt.ylim([0., 1.])
-		plt.xlabel("Level")
-		plt.ylabel("MH Acceptance")
-		if numResampleLogX > 1:
-			plt.draw()
-
-	# Convert to lists of tuples
-	logl_levels = [(levels_orig[i,1], levels_orig[i, 2]) for i in range(0, levels_orig.shape[0])] # logl, tiebreaker
[…]

### 2016-06-05 python/dnest4/deprecated.py

@@ -280,6 +280,227 @@ def postprocess(temperature=1., numResampleLogX=1, plot=True, loaded=[], \
 
 	return [logz_estimate, H_estimate, logx_samples]
 
+def postprocess_abc(temperature=1., numResampleLogX=1, plot=True, loaded=[], \
+			cut=0., save=True, zoom_in=True, compression_bias_min=1., compression_scatter=0., moreSamples=1., compression_assert=None, single_precision=False):
+	if len(loaded) == 0:
+		levels_orig = np.atleast_2d(my_loadtxt("levels.txt"))
+		sample_info = np.atleast_2d(my_loadtxt("sample_info.txt"))
+	else:
+		levels_orig, sample_info = loaded[0], loaded[1]
+
+	# Remove regularisation from levels_orig if we asked for it
+	if compression_assert is not None:
+		levels_orig[1:,0] = -np.cumsum(compression_assert*np.ones(levels_orig.shape[0] - 1))
+
+	sample_info = sample_info[int(cut*sample_info.shape[0]):, :]
+
+	if plot:
+		if numResampleLogX > 1:
+			plt.ion()
+
+		plt.figure(1)
+		plt.plot(sample_info[:,0])
+		plt.xlabel("Iteration")
+		plt.ylabel("Level")
+		if numResampleLogX > 1:
+			plt.draw()
+
+		plt.figure(2)
+		plt.subplot(2,1,1)
+		plt.plot(np.diff(levels_orig[:,0]))
+		plt.ylabel("Compression")
+		plt.xlabel("Level")
+		xlim = plt.gca().get_xlim()
+		plt.axhline(-1., color='r')
+		plt.axhline(-np.log(10.), color='g')
+		plt.ylim(ymax=0.05)
+		if numResampleLogX > 1:
+			plt.draw()
+
+		plt.subplot(2,1,2)
+		good = np.nonzero(levels_orig[:,4] > 0)[0]
+		plt.plot(levels_orig[good,3]/levels_orig[good,4])
+		plt.xlim(xlim)
+		plt.ylim([0., 1.])
+		plt.xlabel("Level")
+		plt.ylabel("MH Acceptance")
+		if numResampleLogX > 1:
+			plt.draw()
+
+	# Convert to lists of tuples
+	logl_levels = [(levels_orig[i,1], levels_orig[i, 2]) for i in range(0, levels_orig.shape[0])] # logl, tiebreaker
[…]

### 2016-06-06 paper/ms.tex
ABC results plot
Mention postprocess_abc
Change of coordinates

@@
 
+In ABC applications the model class needs to describe a point in the joint
+(parameters, data) space, and the \code{from\_prior} and \code{perturb}
+functions need to respect the joint prior distribution.
+In Bayesian inference the joint prior tends to make the
+parameters and data highly dependent
+--- otherwise the data wouldn't ever be informative about the parameters.
+To improve efficiency, it is helpful to use a change of variables such that
+the parameters and data-controlling variables are independent. In this
+example we use
+
+\begin{align}
+\mu &\sim \textnormal{Uniform}(-10, 10)\\
+\ln\sigma &= \textnormal{Uniform}(-10, 10)\\
+n_i &= \textnormal{Normal}(0, 1)
+\end{align}
+where each of the $n_i$ enables a data point to be calculated using
+\begin{align}
+x_i &= \mu + \sigma n_i
+\end{align}
+
+We show posterior samples in Figure~\ref{fig:abc_results}.
+
+\begin{figure}[ht!]
+\centering
+\includegraphics[width=0.6\textwidth]{figures/abc_results.pdf}
+\caption{ABC results.\label{fig:abc_results}}
+\end{figure}
+
+For ABC applications, there is an alternate \code{postprocess()} function:
+\begin{CodeChunk}
+\begin{CodeInput}
+import dnest4.classic as dn4
+dn4.postprocess_abc()
+\end{CodeInput}
+\end{CodeChunk}
+
 %\begin{table}[ht!]

### 2016-06-06 paper/ms.tex
random numbers
editing paragraph

@@
 \begin{align}
-x_i &= \mu + \sigma n_i
-\end{align}
+x_i &= \mu + \sigma n_i.
+\end{align}
+A general procedure for achieving this is to imagine generating a
+simulated data set from the parameters. In this process, you would need
+to call a random number generator many times.
+The results of the random number calls
+are independent of the parameters and
+can be used to parameterize the data set. In the example,
+the $\{n_i\}$ variables play this role.
 

### 2016-06-06 code/Examples/ABC/MyModel.cpp
putting the code in the paper

@@ -7,7 +7,7 @@ using namespace std;
 using namespace DNest4;
 
 MyModel::MyModel()
-:x_fake_latent(Data::get_instance().get_x().size())
+:n(Data::get_instance().get_x().size())
 {
 
 }
@@ -16,8 +16,8 @@ void MyModel::from_prior(RNG& rng)
 {
 	mu = -10. + 20.*rng.rand();
 	log_sigma = -10. + 20.*rng.rand();
-	for(size_t i=0; i<x_fake_latent.size(); i++)
-		x_fake_latent[i] = rng.randn();
+	for(size_t i=0; i<n.size(); i++)
+		n[i] = rng.randn();
 }
 
 double MyModel::perturb(RNG& rng)
@@ -36,8 +36,8 @@ double MyModel::perturb(RNG& rng)
 	}
 	if(which == 2)
 	{
-		int i = rng.rand_int(x_fake_latent.size());
-		x_fake_latent[i] = rng.randn();
+		int i = rng.rand_int(n.size());
+		n[i] = rng.randn();
 	}
 
 	return 0.;
@@ -51,7 +51,7 @@ double MyModel::log_likelihood() const
 	double sigma = exp(log_sigma);
 
 	// Assemble fake dataset
-	vector<double> x_fake = x_fake_latent;
+	vector<double> x_fake = n;
 	for(size_t i=0; i<x_fake.size(); i++)
 		x_fake[i] = mu + sigma*x_fake[i];
 
@@ -66,8 +66,8 @@ double MyModel::log_likelihood() const
 void MyModel::print(std::ostream& out) const
 {
 	out<<mu<<' '<<log_sigma<<' ';
-	for(size_t i=0; i<x_fake_latent.size(); i++)
-		out<<x_fake_latent[i]<<' ';
+	for(size_t i=0; i<n.size(); i++)
+		out<<n[i]<<' ';
 }
 
 string MyModel::description() const

### 2016-06-06 code/Examples/ABC/MyModel.h

@@ -9,7 +9,7 @@ class MyModel
 {
 	private:
 		double mu, log_sigma;
-		std::vector<double> x_fake_latent;
+		std::vector<double> n;
 
 
 	public:

### 2016-06-06 paper/ms.tex
putting the code in the paper
threshold_fraction argument

@@
 the $\{n_i\}$ variables play this role.
+Therefore, the variables in the model class are:
+\begin{CodeChunk}
+\begin{CodeInput}
+class MyModel
+{
+    private:
+        double mu, log_sigma;
+        std::vector<double> n;
+\end{CodeInput}
+\end{CodeChunk}
 
@@
 
-For ABC applications, there is an alternate \code{postprocess()} function:
-\begin{CodeChunk}
-\begin{CodeInput}
-import dnest4.classic as dn4
-dn4.postprocess_abc()
+For ABC applications, there is an alternate \code{postprocess} function
+called \code{postprocess_abc}.
+This generates posterior samples using the $\epsilon$ value that
+compresses the joint prior by 80\% as much as the final level.
+\begin{CodeChunk}
+\begin{CodeInput}
+dnest4.classic.postprocess_abc()
+\end{CodeInput}
+\end{CodeChunk}
+To try different values of $\epsilon$, you can set the argument
+\code{threshold\_fraction} to a value other than 0.8. For example
+\begin{CodeChunk}
+\begin{CodeInput}
+dnest4.classic.postprocess_abc(threshold_fraction=0.6)
 \end{CodeInput}

### 2016-06-06 python/dnest4/deprecated.py

@@ -281,7 +281,7 @@ def postprocess(temperature=1., numResampleLogX=1, plot=True, loaded=[], \
 	return [logz_estimate, H_estimate, logx_samples]
 
 def postprocess_abc(temperature=1., numResampleLogX=1, plot=True, loaded=[], \
-			cut=0., save=True, zoom_in=True, compression_bias_min=1., compression_scatter=0., moreSamples=1., compression_assert=None, single_precision=False):
+			cut=0., save=True, zoom_in=True, compression_bias_min=1., compression_scatter=0., moreSamples=1., compression_assert=None, single_precision=False, threshold_fraction=0.8):
 	if len(loaded) == 0:
 		levels_orig = np.atleast_2d(my_loadtxt("levels.txt"))
 		sample_info = np.atleast_2d(my_loadtxt("sample_info.txt"))
@@ -398,7 +398,7 @@ def postprocess_abc(temperature=1., numResampleLogX=1, plot=True, loaded=[], \
 		logp_samples[:,z] = logp_samples[:,z] - logsumexp(logp_samples[:,z])
 
 		# Define the threshold for ABC, in terms of log(X)
-		threshold = 0.8*levels[:,0].min()
+		threshold = threshold_fraction*levels[:,0].min()
 
 		# Particles below threshold get no posterior weight
 		logp_samples[logx_samples > threshold] = -1E300

### 2016-06-06 paper/ms.tex
Random work

@@
 \begin{align}
-p(D_i | N, \{\mu_j\}, \{\sigma_j\}, \{w_j\}) &=
+p\left(D_i | N, \{\mu_j\}, \{\sigma_j\}, \{w_j\}\right) &=
 \sum_{j=1}^N \frac{w_j}{\sigma_j\sqrt{2\pi}}
@@
 \end{align}
-Colloquially, one might say the data were 'drawn from' a mixture of
+Colloquially, one might say the data were `drawn from' a mixture of
 $N$ normal distributions, and we want to infer $N$ along with the
@@
 \end{CodeChunk}
+where \code{num\_dimensions} is the number of parameters needed to
+specify a single component (three in this example),
+...
 

### 2016-06-07 python/dnest4/loading.py
Commit that

@@ -1,0 +1,58 @@
+# -*- coding: utf-8 -*-
+
+import numpy as np
+import pandas as pd
+
+__all__ = ["my_loadtxt", "loadtxt_rows"]
+
+def my_loadtxt(filename, single_precision=False, delimiter=" "):
+    if single_precision:
+        return pd.read_csv(filename, header=None, delimiter=delimiter,\
+                           comment="#", dtype=np.float32)\
+                                    .dropna(axis=1).values
+    return pd.read_csv(filename, header=None, delimiter=delimiter, comment="#")\
+                                                 .dropna(axis=1).values
+
+def loadtxt_rows(filename, rows, single_precision=False):
+    """
+    Load only certain rows
+    """
+    # Open the file
+    f = open(filename, "r")
+
+    # Storage
+    results = {}
+
+    # Row number
+    i = 0
+
+    # Number of columns
+    ncol = None
+
+    while(True):
+        # Read the line and split by whitespace
+        line = f.readline()
+        cells = line.split()
+
+        # Quit when you see a different number of columns
+        if ncol is not None and len(cells) != ncol:
+            break
+
+        # Non-comment lines
+        if cells[0] != "#":
+            # If it's the first one, get the number of columns
+            if ncol is None:
+                ncol = len(cells)
+
+            # Otherwise, include in results
+            if i in rows:
+                if single_precision:
+                    results[i] = np.array([float(cell) for cell in cells],\
+                                                              dtype="float32")
+                else:
+                    results[i] = np.array([float(cell) for cell in cells])
+            i += 1
+
+    results["ncol"] = ncol
+    return results
+

### 2016-06-08 paper/ms.tex
Normalized weights
Different colour figs :)
RJObject section work

@@
 \begin{CodeOutput}
-log(Z) = -175.528264138
-Information = 15.1944947225 nats.
-Effective sample size = 745.819862383
+log(Z) = -175.532983853
+Information = 15.2112429682 nats.
+Effective sample size = 737.639033806
 \end{CodeOutput}
@@
 a sensible joint prior for the gaussian parameters.
-To implement this model for \pkg{DNest4}, the model
-class only needs to contain an instance of an
-\code{RJObject}, like so:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-class MyModel
-{
-    private:
-        DNest4::RJObject<MyConditionalPrior> gaussians;
-....
-\end{CodeInput}
-\end{CodeChunk}
-where the template argument \code{<MyConditionalPrior>}
-is a class implementing the prior for the hyperparameters
-$\alpha$ and the form of the conditional prior $p(x_i | \alpha)$.
-
-To help with implementing the member functions for the \code{MyModel} class,
-the \code{gaussians} object can be generated from its prior, proposed,
-and printed trivially as follows:
-\begin{CodeChunk}
-\begin{CodeInput}
-    // Generate the components from the prior
-    gaussians.from_prior(rng);
-
-    // Do a Metropolis proposal
-    double logH = gaussians.perturb(rng);
-
-    // Print to output stream 'out'
-    gaussians.print(out);
-\end{CodeInput}
-\end{CodeChunk}
-
-The \code{RJObject} constructor definition is:
-\begin{CodeChunk}
-\begin{CodeInput}
-RJObject(int num_dimensions, int max_num_components, bool fixed,
-         const ConditionalPrior& conditional_prior,
-         PriorType prior_type=PriorType::uniform);
-\end{CodeInput}
-\end{CodeChunk}
-where \code{num\_dimensions} is the number of parameters needed to
-specify a single component (three in this example),
-...
-
-The specific prior for $N$ was
+
+The mixture weights $\{w_j\}$ must obey a normalization condition
+$\sum_j w_j = 1$. The easiest way of implementing this is to use
+un-normalized weights $\{W_j\}$ which do not obey
+such a condition and then normalise them by dividing by their sum.
+
+We now describe the specific prior distributions we used in this example.
+The prior for $N$ was
 \begin{align}
@@
 distribution function (CDF) of the Laplace distribution makes it easier to
-implement.
-
+implement. These were:
 \begin{align}
@@
 \end{align}
+The location parameters are denoted with $a$ and scale parameters with $b$.
+These priors express the idea that the centers, widths, and relative weights
+of the gaussian mixture components are probably clustered around some typical
+value.
 
@@
 b_{\ln\sigma} &\sim \textnormal{Uniform}(0, 5)\\
-b_{\ln W} &\sim \textnormal{Uniform}(0, 5)
-\end{align}
-Location parameters $a$, scale parameters $b$.
-
+b_{\ln W} &\sim \textnormal{Uniform}(0, 5).
+\end{align}
 The normalized weights are $w_k = W_k/\sum w_j$.
+
+To implement this model for \pkg{DNest4}, the model
+class needs to contain an instance of an
+\code{RJObject}, which contains the parameters
+$\{(\mu_j, \sigma_j, w_j)\}$:
+\begin{CodeChunk}
+\begin{CodeInput}
+class MyModel
+{
+    private:
+        DNest4::RJObject<MyConditionalPrior> gaussians;
+....
+\end{CodeInput}
+\end{CodeChunk}
+where the template argument \code{<MyConditionalPrior>}
+is a class implementing the prior for the hyperparameters
+$\alpha$ and the form of the conditional prior $p(x_i | \alpha)$.
+The main advantage of the \code{RJObject} class is that
+we do not need to implement any proposals for $\{(\mu_j, \sigma_j, w_j)\}$.
+Rather, these can be done trivially as follows:
+\begin{CodeChunk}
+\begin{CodeInput}
+    // Generate the components from the prior
+    gaussians.from_prior(rng);
+
+    // Do a Metropolis proposal
+    double logH = gaussians.perturb(rng);
+
+    // Print to output stream 'out'
+    gaussians.print(out);
+\end{CodeInput}
+\end{CodeChunk}
+
+The \code{RJObject} constructor definition is:
+\begin{CodeChunk}
+\begin{CodeInput}
+RJObject(int num_dimensions, int max_num_components, bool fixed,
+         const ConditionalPrior& conditional_prior,
+         PriorType prior_type=PriorType::uniform);
+\end{CodeInput}
+\end{CodeChunk}
+where \code{num\_dimensions} is the number of parameters needed to
+specify a single component (three in this example),
+...
 

### 2016-06-08 paper/ms.tex
More RJObject writing

@@
 properties of those normal distributions.
-
-Making the connection with Equation~\ref{eqn:rjobject} and
-Figure~\ref{fig:rjobject_pgm}, the ``components'' are the gaussians,
-the gaussian parameters are $\{x_j\} = \{(\mu_j, \sigma_j, w_j)\}$,
-and hyperparameters $\alpha$ may be used to help specify
-a sensible joint prior for the gaussian parameters.
-
 The mixture weights $\{w_j\}$ must obey a normalization condition
@@
 such a condition and then normalise them by dividing by their sum.
+
+Making the connection with Equation~\ref{eqn:rjobject} and
+Figure~\ref{fig:rjobject_pgm}, the ``components'' are the gaussians,
+the gaussian parameters are $\{x_j\} = \{(\mu_j, \sigma_j, W_j)\}$,
+and hyperparameters $\alpha$ may be used to help specify
+a sensible joint prior for the gaussian parameters.
 
@@
 The main advantage of the \code{RJObject} class is that
-we do not need to implement any proposals for $\{(\mu_j, \sigma_j, w_j)\}$.
+we do not need to implement any proposals for $\{(\mu_j, \sigma_j, W_j)\}$.
 Rather, these can be done trivially as follows:
@@
 where \code{num\_dimensions} is the number of parameters needed to
-specify a single component (three in this example),
-...
+specify a single component (three in this example), \code{max\_num\_components}
+is the maximum value of $N$ allowed, \code{fixed} determines whether
+$N$ is fixed at $N_{\rm max}$ or allowed to vary,
+\code{conditional\_prior} is an instance of a conditional prior,
+and \code{prior\_type} controls the prior for $N$ (the default is uniform
+from $\{0, 1, 2, ..., N_{\rm max}\}$). The {\tt RJObject\_1DMixture}
+\code{MyModel} initializes its {\tt RJObject} as follows:
+\begin{CodeChunk}
+\begin{CodeInput}
+MyModel::MyModel()
+:gaussians(3, 100, false, MyConditionalPrior(), PriorType::log_uniform)
+{
+}
+\end{CodeInput}
+\end{CodeChunk}
+Passing \code{PriorType::log_uniform} for the final argument
+specifies the $1/(N+1)$ prior for $N$.
 

### 2016-06-08 paper/ms.tex
RJObject stuff, continued...

@@
 
+\subsection{Using the RJObject class}
 To implement this model for \pkg{DNest4}, the model
@@
 specifies the $1/(N+1)$ prior for $N$.
-
+One complication for this model is that \code{RJObject}, by default,
+allows $N$ to be zero, which makes no sense for this particular
+problem. Therefore, we prevent $N=0$ from being generated in
+\code{MyModel::from\_prior}, and assert that it should always be
+rejected if proposed in \code{MyModel::perturb}:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+void MyModel::from_prior(RNG& rng)
+{
+    do
+    {
+        gaussians.from_prior(rng);
+    }while(gaussians.get_components().size() == 0);
+}
+
+double MyModel::perturb(RNG& rng)
+{
+    double logH = 0.0;
+
+    logH += gaussians.perturb(rng);
+    if(gaussians.get_components().size() == 0)
+        return -std::numeric_limits<double>::max();
+
+    return logH;
+}
+\end{CodeInput}
+\end{CodeChunk}
+
+\subsection{Mixture model results}
 The data and the posterior mean fit are shown
-in Figure~\ref{fig:galaxies}.
+in Figure~\ref{fig:galaxies} and the posterior distribution for
+$N$ is shown in Figure~\ref{fig:galaxies_N}.
 

### 2016-06-08 paper/ms.tex
tt
More RJObject writing

@@
 
+To access the component parameters, the \code{RJObject} member function
+\code{get_components} is used. This was used in the above functions, but more
+typically it is needed in the log likelihood.
+The member function \code{get_components} returns (by const reference) a
+\code{std::vector} of \code{std::vector}s of \code{double}s. For example,
+parameter two of component zero is:
+\begin{CodeChunk}
+\begin{CodeInput}
+gaussians.get_components()[0][2]
+\end{CodeInput}
+\end{CodeChunk}
+The order of $\{(\mu_j, \sigma_j, w_j)\}$
+(i.e. which one is parameter 0, 1, and 2)
+is determined by the conditional prior class, and is
+explained in the following subsection.
+
+\subsection{Conditional priors}
+
 \subsection{Mixture model results}
@@
 \proglang{Julia} language is provided, with an example given in the
-code/Templates/JuliaModel directory.
+{\tt code/Templates/JuliaModel} directory.
 Set an enviroment variable {\tt JULIA\_LIB\_PATH} to the directory on your

### 2016-06-08 python/dnest4/classic.py
Switched code (it's in classic.py now)

@@ -1,1 +1,457 @@
-from dnest4.deprecated import *
+import copy
+import numpy as np
+import pandas as pd
+import matplotlib.pyplot as plt
+from .loading import *
+
+def logsumexp(values):
+	biggest = np.max(values)
+	x = values - biggest
+	result = np.log(np.sum(np.exp(x))) + biggest
+	return result
+
+def logdiffexp(x1, x2):
+	biggest = x1
+	xx1 = x1 - biggest
+	xx2 = x2 - biggest
+	result = np.log(np.exp(xx1) - np.exp(xx2)) + biggest
+	return result
+
+
+def postprocess(temperature=1., numResampleLogX=1, plot=True, loaded=[], \
+			cut=0., save=True, zoom_in=True, compression_bias_min=1., compression_scatter=0., moreSamples=1., compression_assert=None, single_precision=False):
+	if len(loaded) == 0:
+		levels_orig = np.atleast_2d(my_loadtxt("levels.txt"))
+		sample_info = np.atleast_2d(my_loadtxt("sample_info.txt"))
+	else:
+		levels_orig, sample_info = loaded[0], loaded[1]
+
+	# Remove regularisation from levels_orig if we asked for it
+	if compression_assert is not None:
+		levels_orig[1:,0] = -np.cumsum(compression_assert*np.ones(levels_orig.shape[0] - 1))
+
+	sample_info = sample_info[int(cut*sample_info.shape[0]):, :]
+
+	if plot:
+		plt.figure(1)
+		plt.plot(sample_info[:,0], "k")
+		plt.xlabel("Iteration")
+		plt.ylabel("Level")
+
+		plt.figure(2)
+		plt.subplot(2,1,1)
+		plt.plot(np.diff(levels_orig[:,0]), "k")
+		plt.ylabel("Compression")
+		plt.xlabel("Level")
+		xlim = plt.gca().get_xlim()
+		plt.axhline(-1., color='g')
+		plt.axhline(-np.log(10.), color='g', linestyle="--")
+		plt.ylim(ymax=0.05)
+
+		plt.subplot(2,1,2)
+		good = np.nonzero(levels_orig[:,4] > 0)[0]
+		plt.plot(levels_orig[good,3]/levels_orig[good,4], "ko-")
+		plt.xlim(xlim)
+		plt.ylim([0., 1.])
+		plt.xlabel("Level")
+		plt.ylabel("MH Acceptance")
+
+	# Convert to lists of tuples
[…]

### 2016-06-08 python/dnest4/deprecated.py

@@ -1,457 +1,2 @@
-import copy
-import numpy as np
-import pandas as pd
-import matplotlib.pyplot as plt
-from .loading import *
-
-def logsumexp(values):
-	biggest = np.max(values)
-	x = values - biggest
-	result = np.log(np.sum(np.exp(x))) + biggest
-	return result
-
-def logdiffexp(x1, x2):
-	biggest = x1
-	xx1 = x1 - biggest
-	xx2 = x2 - biggest
-	result = np.log(np.exp(xx1) - np.exp(xx2)) + biggest
-	return result
-
-
-def postprocess(temperature=1., numResampleLogX=1, plot=True, loaded=[], \
-			cut=0., save=True, zoom_in=True, compression_bias_min=1., compression_scatter=0., moreSamples=1., compression_assert=None, single_precision=False):
-	if len(loaded) == 0:
-		levels_orig = np.atleast_2d(my_loadtxt("levels.txt"))
-		sample_info = np.atleast_2d(my_loadtxt("sample_info.txt"))
-	else:
-		levels_orig, sample_info = loaded[0], loaded[1]
-
-	# Remove regularisation from levels_orig if we asked for it
-	if compression_assert is not None:
-		levels_orig[1:,0] = -np.cumsum(compression_assert*np.ones(levels_orig.shape[0] - 1))
-
-	sample_info = sample_info[int(cut*sample_info.shape[0]):, :]
-
-	if plot:
-		plt.figure(1)
-		plt.plot(sample_info[:,0], "k")
-		plt.xlabel("Iteration")
-		plt.ylabel("Level")
-
-		plt.figure(2)
-		plt.subplot(2,1,1)
-		plt.plot(np.diff(levels_orig[:,0]), "k")
-		plt.ylabel("Compression")
-		plt.xlabel("Level")
-		xlim = plt.gca().get_xlim()
-		plt.axhline(-1., color='g')
-		plt.axhline(-np.log(10.), color='g', linestyle="--")
-		plt.ylim(ymax=0.05)
-
-		plt.subplot(2,1,2)
-		good = np.nonzero(levels_orig[:,4] > 0)[0]
-		plt.plot(levels_orig[good,3]/levels_orig[good,4], "ko-")
-		plt.xlim(xlim)
-		plt.ylim([0., 1.])
-		plt.xlabel("Level")
-		plt.ylabel("MH Acceptance")
-
-	# Convert to lists of tuples
[…]

### 2016-06-08 paper/ms.tex
Writing about ConditionalPriors

@@
 components, and each component is independent and identically
-distributed given the hyperparameters.
+distributed given the hyperparameters. The sampling distribution
+$p(D | x, \alpha)$ (not shown in the PGM)
+cannot depend on the ordering of the components.
 A probabilistic graphical model (PGM) is showing this dependence
@@
 \subsection{Conditional priors}
+The \code{RJObject} class is used to manage the components. An additional
+class is needed to define and manage the hyperparameters $\alpha$ (i.e.
+define how they are generated from the prior, how they are proposed, and
+so on). The and the iid
+conditional prior $p(x_i | \alpha)$.
+In the example, the class is called \code{MyConditionalPrior} and
+is inherited from an abstract base class \code{DNest4::ConditionalPrior}.
+
+The member functions \code{from\_prior}, \code{perturb\_hyperparameters}, and
+\code{print} do the same things as the similar functions in a standard model
+class, but for the hyperparameters $\alpha$. In addition, three functions
+\code{from\_uniform}, \code{to\_uniform}, and \code{log\_pdf}
+together define the form of the conditional prior for the
+component parameters, $p(x_i | \alpha)$.
 

### 2016-06-08 paper/ms.tex
conditional prior class stuff

@@
 \begin{align}
-\mu_j &\sim \textnormal{Laplace}(a_\mu, b_\mu)\\
-\ln\sigma_j &\sim \textnormal{Laplace}(a_{\ln\sigma}, b_{\ln\sigma})\\
-\ln W_j &\sim \textnormal{Laplace}(0, b_{\ln W}).
+\mu_j &\sim \textnormal{Laplace}(a_\mu, b_\mu)\label{eqn:cond_prior1}\\
+\ln\sigma_j &\sim \textnormal{Laplace}(a_{\ln\sigma}, b_{\ln\sigma})\label{eqn:cond_prior2}\\
+\ln W_j &\sim \textnormal{Laplace}(0, b_{\ln W}).\label{eqn:cond_prior3}
 \end{align}
@@
 define how they are generated from the prior, how they are proposed, and
-so on). The and the iid
-conditional prior $p(x_i | \alpha)$.
+so on). The iid
+conditional prior $p(x_i | \alpha)$ is also defined by this extra class.
 In the example, the class is called \code{MyConditionalPrior} and
-is inherited from an abstract base class \code{DNest4::ConditionalPrior}.
+is inherited from an abstract base class \code{DNest4::ConditionalPrior}
+which insists that certain member functions be specified.
 
@@
 \code{print} do the same things as the similar functions in a standard model
-class, but for the hyperparameters $\alpha$. In addition, three functions
+class, but for the hyperparameters $\alpha$. 
+In addition, three functions
 \code{from\_uniform}, \code{to\_uniform}, and \code{log\_pdf}
 together define the form of the conditional prior for the
-component parameters, $p(x_i | \alpha)$.
+component parameters, $p(x_i | \alpha)$. Each of these takes a
+\code{std::vector} of \code{double}s by reference as input.
+
+The \code{log\_pdf} function just evaluates $\ln p(x_i | \alpha)$
+at the current value of $\alpha$. The position $x_i$ where this is evaluated
+is passed in via the input vector. In the example, these
+were defined using Equations~\ref{eqn:cond_prior1}--\ref{eqn:cond_prior3}.
+To simplify this, we have defined a class for Laplace distributions,
+rather than explicitly writing out the densities. We decided (arbitarily)
+that parameters 0,1, and 2 are $\mu$, $\ln(\sigma)$, and $W$ respectively.
+\begin{CodeChunk}
+\begin{CodeInput}
+// vec = {mu, log_sigma, log_weight}
+double MyConditionalPrior::log_pdf(const std::vector<double>& vec) const
+{
+    // Three Laplace distributions
+    Laplace l1(location_mu, scale_mu);
+    Laplace l2(location_log_sigma, scale_log_sigma);
+    Laplace l3(0.0, scale_log_weight);
+	return l1.log_pdf(vec[0]) + l2.log_pdf(vec[1]) + l3.log_pdf(vec[2]);
+}
+\end{CodeInput}
+\end{CodeChunk}
+
+The functions \code{to\_uniform} and \code{from\_uniform} must implement
+the cumulative distribution function (CDF) of the conditional
+prior and its inverse. i.e. \code{from\_uniform}, if the input vector contains
+iid draws from Uniform(0, 1), these should be modified to become draws from
+$p(x|\alpha)$, and \code{to\_uniform} should be the inverse of
+\code{from\_uniform}.
+Both of these functions modify the vector argument in-place.
+For the example, we again used an external class to define these functions
+for the Laplace distribution:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+// vec = {mu, log_sigma, log_weight}
+void MyConditionalPrior::from_uniform(std::vector<double>& vec) const
+{
+    // Three Laplace distributions
+    Laplace l1(location_mu, scale_mu);
+    Laplace l2(location_log_sigma, scale_log_sigma);
+    Laplace l3(0.0, scale_log_weight);
+
+    vec[0] = l1.cdf_inverse(vec[0]);
+    vec[1] = l2.cdf_inverse(vec[1]);
+    vec[2] = l3.cdf_inverse(vec[2]);
+}
+
+// vec = {mu, log_sigma, log_weight}
+void MyConditionalPrior::to_uniform(std::vector<double>& vec) const
+{
+    // Three Laplace distributions
+    Laplace l1(location_mu, scale_mu);
+    Laplace l2(location_log_sigma, scale_log_sigma);
+    Laplace l3(0.0, scale_log_weight);
+
+    vec[0] = l1.cdf(vec[0]);
+    vec[1] = l2.cdf(vec[1]);
+    vec[2] = l3.cdf(vec[2]);
+}
+\end{CodeInput}
+\end{CodeChunk}
 

### 2016-06-08 paper/ms.tex
More of the ABC model

@@
 
+\code{RJObject}'s \code{print} function prints the dimensionality of each
+component and the maximum value of $N$, followed by the hyperparameters,
+then parameter zero of each component (zero padded when $N<N_{\rm max}$),
+parameter one for each component, and so on.
+
 \subsection{Conditional priors}
@@
 
+The proposal involves changing either \code{mu}, \code{log\_sigma}, or
+one of the \code{n}s:
+\begin{CodeChunk}
+\begin{CodeInput}
+double MyModel::perturb(RNG& rng)
+{
+	int which = rng.rand_int(3);
+
+	if(which == 0)
+	{
+		mu += 20.*rng.randh();
+		wrap(mu, -10., 10.);
+	}
+	if(which == 1)
+	{
+		log_sigma += 20.*rng.randh();
+		wrap(log_sigma, -10., 10.);
+	}
+	if(which == 2)
+	{
+		int i = rng.rand_int(n.size());
+		n[i] = rng.randn();
+	}
+
+	return 0.0;
+}
+\end{CodeInput}
+\end{CodeChunk}
+
+The ``log-likelihood'' is really minus the discrepancy function.
+Since we parameterized joint (parameter, data) space using the
+$n$ variables instead of the data $\{x_i\}$ itself, we must
+``assemble'' the simulated dataset in order to evaluate the
+discrepancy function:
+\begin{CodeChunk}
+\begin{CodeInput}
+double MyModel::log_likelihood() const
+{
+    double x_min = Data::get_instance().get_x_min();
+    double x_max = Data::get_instance().get_x_max();
+
+    double sigma = exp(log_sigma);
+
+    // Assemble fake dataset
+    vector<double> x_fake = n;
+    for(size_t i=0; i<x_fake.size(); i++)
+        x_fake[i] = mu + sigma*x_fake[i];
+
+    // Goodness
+    double logL = 0.;
+    logL -= pow(*min_element(x_fake.begin(), x_fake.end()) - x_min, 2);
+    logL -= pow(*max_element(x_fake.begin(), x_fake.end()) - x_max, 2);
+
+    return logL;
+}
+\end{CodeInput}
+\end{CodeChunk}
+
+\subsection{ABC Example Results}
 We show posterior samples in Figure~\ref{fig:abc_results}.
@@
 \end{CodeChunk}
-
-%\begin{table}[ht!]
-%\centering
-%\small
-%\begin{tabular}{|l|l|l|}
-%\hline
-%Bayesian computation		&		Statistical mechanics		&		ABC\\
-%\hline
-%Parameter space	$\Theta$	&		Phase space	$\Omega$ 			& Product space $\Theta \times \mathcal{D}$\\
-%Posterior $p(\params | D,I)$   &  Canonical distribution $p(\boldsymbol{x}; T)$     & \\ 
-%Marginal likelihood $P(D|I)$	&	Partition function $Z(T)$	&   Marginal likelihood $p(D\in | I)$\\
-%\hline
-%\end{tabular}
-%\caption{\it The relationship between standard Bayesian computation, statistical
-%mechanics, and ABC. In each case Monte Carlo methods are used to calculate
-%integrals over a space. In standard Bayesian computation, it is the parameter
-%space of a model that is usually of interest, whereas in ABC it is the space
-%of possible parameter values {\it and} data sets.
-%\label{tab:relation}}
-%\end{table}
-
-%\begin{figure}[ht!]
-%\centering
-%\includegraphics[scale=0.5]{figures/joint.pdf}
-%\caption{\it The joint prior for parameters $\params$ and data $\data$.
-%\label{fig:joint}}
-%\end{figure}
-
-%\section{Parameterising the data space}
-%In standard MCMC-based ABC, proposal moves are made by first proposing a
-%change to the parameters (from $\params$ to $\params'$), and then generating
-%a mock dataset $\data'$ from $p(\data | \params')$. As a Metropolis proposal
-%for exploring the product space $\Theta \times \mathcal{D}$ this is a very
-%aggressive choice, since the proposed value of the
-%dataset $\params'$ is likely to be very different from the previous one.
-%One of the most elementary pieces of advice we receive when learning the
-%Metropolis algorithm is to be careful about how ambitious our proposal
-%distributions are, since tiny and huge proposal moves are both very
-%inefficient. The performance of MCMC-based ABC can therefore be improved by
-%alternative choices of proposal distribution.
-
-%In some cases, we might be able to find efficient proposal distributions for exploring $\Theta \times \mathcal{D}$ by having insight into the particular
-%problem we're working on. However, there are also techniques that are
-%completely general in that they should work on any ABC-type problem.
-
-%Consider the process of generating a mock dataset. Whatever else is involved,
-%this process will involve calls to an underlying random number generator.
-%Perhaps a function {\tt rand()}, generating variables from a Uniform(0, 1)
-%distribution, is called $n$ times in the course of generating a dataset.
-%Usually, the simulation process has the property that, if $\theta'$ is
-%close to $\theta$, then the same sequence of random numbers returned by
-%{\tt rand()} would produce a dataset $\data'$ that is very similar to $\data$.
-%This suggests an alternative parameterisation of
-%the product space $\Theta\times\mathcal{D}$: instead of using the one that
-%naturally describes the data, we can use a coordinate system defined by
-%$\theta$ together with $\{u_1, ..., u_n\}$, the $n$ uniform random numbers
-%used to construct a mock dataset.
-
-%In this alternative coordinate system, a proposed move
-%to change $\params$ while keeping
-%$\{u_i\}$ fixed results in a proposed new dataset that is similar to the
-%previous one, assuming the change to $\params$ is small. The standard procedure
-%of generating an entirely new mock dataset each step is now viewed as a
-%proposal where $\params$ is moved {\it and all of the $u_i$ coordinates are
-%re-generated from their iid uniform prior}.
 
@@
 
-\section{Specifying models in Python with dnest4.builder}
+%\section{Specifying models in Python with dnest4.builder}
 

### 2016-06-08 python/examples/straightline/straightline.py
Straight line model

@@ -1,0 +1,82 @@
+#!/usr/bin/env python
+# -*- coding: utf-8 -*-
+"""
+Demonstration of DNest4 in Python using the "StraightLine" example
+"""
+
+import dnest4
+import numpy as np
+import numpy.random as rng
+from scipy.special import erf
+
+# Load the data
+data = np.loadtxt("../../../code/Examples/StraightLine/road.txt")
+
+class Model(object):
+    """
+    Specify the model in Python.
+    """
+    def __init__(self):
+        """
+        Parameter values *are not* stored inside the class
+        """
+        pass
+
+    def from_prior(self):
+        """
+        Unlike in C++, this must *return* a numpy array of parameters.
+        """
+        m = 1E3*rng.randn()
+        b = 1E3*rng.randn()
+        sigma = np.exp(-10.0 + 20.0*rng.rand())
+        return np.array([m, b, sigma])
+
+    def perturb(self, params):
+        """
+        Unlike in C++, this takes a numpy array of parameters as input,
+        and modifies it in-place. The return value is still logH.
+        """
+        logH = 0.0
+        which = rng.randint(3)
+
+        if which == 0 or which == 1:
+            logH -= -0.5*(params[which]/1E3)**2
+            params[which] += 1E3*dnest4.randh()
+            logH += -0.5*(params[which]/1E3)**2
+        else:
+            log_sigma = np.log(params[2])
+            log_sigma += 20*dnest4.randh()
+            # Note the difference between dnest4.wrap in Python and
+            # DNest4::wrap in C++. The former *returns* the wrapped value.
+            log_sigma = dnest4.wrap(log_sigma, -10.0, 10.0)
+            params[2] = np.exp(log_sigma)
+
+        return logH
+
+    def log_likelihood(self, params):
+        """
+        Gaussian sampling distribution.
+        """
+        m, b, sigma = params
+        var = sigma**2
+        return -0.5*data.shape[0]*np.log(2*np.pi*var)\
+                - 0.5*np.sum((data[:,1] - (m*data[:,0] + b))**2)/var
+
+# Create a model object and a sampler
+model = Model()
+sampler = dnest4.DNest4Sampler(model,
+                               backend=dnest4.backends.CSVBackend(".",
+                                                                  sep=" "))
+
+# Set up the sampler. The first argument is max_num_levels
+gen = sampler.sample(max_num_levels=30, num_steps=1000, new_level_interval=10000,
[…]

### 2016-06-08 paper/ms.tex
Editing and writing about threads
Python model stuff

@@
 \section{Python Bindings}\label{sec:python_bindings}
-
-{\bf To be written by Dan FM}
+In \pkg{DNest4}, it is also possible to specify and run models in
+\proglang{Python}. Currently, the only models supported
+are ones where the parameters
+are a \pkg{NumPy} array of double precision quantities.
+Only a single thread is supported, although parallelization at
+the level of the \proglang{Python} likelihood function should work.
+
+A Model class implementing the straight line example is given below
+and included in {\tt python/examples/straightline/straightline.py}.
+The member functions of this class have similar names to those
+used in C++, but work slightly differently. Most notably,
+an instance of the class does not represent a point in parameter
+space, but represents the problem itself. The functions take
+\pkg{NumPy} arrays of parameters as input, or produce them as
+output, rather than having parameters as data members as in C++.
+
+\begin{CodeChunk}
+\begin{CodeInput}
+class Model(object):
+    """
+    Specify the model in Python.
+    """
+    def __init__(self):
+        """
+        Parameter values *are not* stored inside the class
+        """
+        pass
+
+    def from_prior(self):
+        """
+        Unlike in C++, this must *return* a numpy array of parameters.
+        """
+        m = 1E3*rng.randn()
+        b = 1E3*rng.randn()
+        sigma = np.exp(-10.0 + 20.0*rng.rand())
+        return np.array([m, b, sigma])
+
+    def perturb(self, params):
+        """
+        Unlike in C++, this takes a numpy array of parameters as input,
+        and modifies it in-place. The return value is still logH.
+        """
+        logH = 0.0
+        which = rng.randint(3)
+
+        if which == 0 or which == 1:
+            logH -= -0.5*(params[which]/1E3)**2
+            params[which] += 1E3*dnest4.randh()
+            logH += -0.5*(params[which]/1E3)**2
+        else:
+            log_sigma = np.log(params[2])
+            log_sigma += 20*dnest4.randh()
+            # Note the difference between dnest4.wrap in Python and
+            # DNest4::wrap in C++. The former *returns* the wrapped value.
+            log_sigma = dnest4.wrap(log_sigma, -10.0, 10.0)
+            params[2] = np.exp(log_sigma)
+
+        return logH
+
+    def log_likelihood(self, params):
+        """
+        Gaussian sampling distribution.
+        """
+        m, b, sigma = params
+        var = sigma**2
+        return -0.5*data.shape[0]*np.log(2*np.pi*var)\
+                - 0.5*np.sum((data[:,1] - (m*data[:,0] + b))**2)/var
+
+\end{CodeInput}
+\end{CodeChunk}
+
+A sampler can then be created and run using the following code.
+The \proglang{Python} sampler does not load an {\tt OPTIONS} file.
+Instead, the options are provided as function arguments
+when setting up the sampler.
+\begin{CodeChunk}
+\begin{CodeInput}
+# Create a model object and a sampler
+model = Model()
+sampler = dnest4.DNest4Sampler(model,
+                               backend=dnest4.backends.CSVBackend(".",
+                                                                  sep=" "))
+
+# Set up the sampler. The first argument is max_num_levels
+gen = sampler.sample(max_num_levels=30, num_steps=1000, new_level_interval=10000,
+                      num_per_step=10000, thread_steps=100,
+                      num_particles=5, lam=10, beta=100, seed=1234)
+
+# Do the sampling (one iteration here = one particle save)
+for i, sample in enumerate(gen):
+    print("# Saved {k} particles.".format(k=(i+1)))
+\end{CodeInput}
+\end{CodeChunk}
 
@@
 Run on the specified number of threads. The default is 1.
+Be aware that the number of particles specified in the {\tt OPTIONS}
+file is actually the number of particles {\em per thread}. Therefore,
+if running on 8 or more cores, we recommend setting the number of particles
+per thread to 1, so the number of particles equals the number of threads.
 

### 2016-06-09 code/RNG.cpp
Faster randh

@@ -30,8 +30,7 @@ double RNG::randn()
 double RNG::randt2()
 {
     // t-distribution with 2 degrees of freedom (less ridiculous than Cauchy)
-    return this->randn()/
-                sqrt((pow(this->randn(), 2) + pow(this->randn(), 2))/2);
+    return this->randn()/sqrt(-log(this->rand()));
 }
 
 double RNG::randh()

### 2016-06-09 python/dnest4/utils.py

@@ -4,20 +4,11 @@ __all__ = ["randh", "wrap"]
 import numpy as np
 import numpy.random as rng
 
-def randh(N=1):
+def randh():
     """
     Generate from the heavy-tailed distribution.
-    If N=1, returns a float64
-    If N>1, returns a numpy array
     """
-    assert N >= 1
-
-    t = rng.randn(N)/np.sqrt((rng.randn(N)**2 + rng.randn(N)**2)/2)
-    x = 10.0**(1.5 - 3*np.abs(t))*rng.randn(N)
-
-    if len(x) == 1:
-        return x[0]
-    return x
+    return 10.0**(1.5 - 3*np.abs(rng.randn()/np.sqrt(-np.log(rng.rand()))))*rng.randn()
 
 def wrap(x, a, b):
     assert b > a

### 2016-06-09 paper/ms.tex
stage
Faster randh
Some editing

@@
 \proglang{C++11},
-along with associated utilities including: i)
-a \pkg{Python} package allowing basic use without \proglang{C++} coding;
-ii) \pkg{RJObject}, a class template for finite mixture models;
-iii) Experimental support for models implemented in \proglang{Julia}; and
-iv) An experimental \proglang{Python} tool to
-generate \pkg{C++} code from a model specification.
+along with associated utilities including:
+i)
+\pkg{RJObject}, a class template for finite mixture models;
+(ii)
+A \pkg{Python} package allowing basic use without \proglang{C++} coding; and
+iii)
+Experimental support for models implemented in \proglang{Julia}.
 In this paper we demonstrate \pkg{DNest4} usage through examples including
@@
 \section{Introduction}
-Bayesian inference, where probability theory is used to describe degrees of
+Bayesian inference, where probability theory describes degrees of
 logical implication or subjective certainty, provides a powerful general basis
@@
 This kind of calculation is often
-called ``model selection'' or ``model averaging''.
-
+called ``model selection'' or ``model averaging'', and the results
+are often presented as ratios
+of marginal likelihoods, known as Bayes Factors.
 When discussing computational matters, the prior distribution for parameters
@@
 problems where MCMC is the only viable sampling method. DNS is based on the
-Metropolis algorithm. DNS evolves one or more particles in the parameter space, along with
+Metropolis algorithm, and evolves one or more particles in the parameter space, along with
 an integer index variable $j$ for each particle,
@@
 \end{align}
-The DNS algorithm consists of two main stages. In the first stage,
+The DNS algorithm consists of two stages. In the first stage,
 the particle(s) are initialized from the prior $\pi(\theta)$, equivalent
 to the mixture of constrained priors with a single level whose likelihood
-threshold is $-\infty$. The mixture of constrained priors evolves by adding new levels
+threshold is $-\infty$. The mixture of constrained priors evolves by adding new levels,
+each of which compresses the distribution by a factor of about $e \approx 2.71818$
 (see \citet{dnest} for details). In this stage,
@@
 performance by preventing a full re-computation of the log-likelihood when
-only a small subset of the parameters have been changed.
+only a subset of the parameters has been changed.
 
@@
 \end{align}
-where $\ell_j$ is the likelihood threshold for level $j$.
+where $\ell_j$ is the likelihood threshold for the current level $j$.
+There are also moves that propose a change to $j$ while keeping
+$\params$ fixed, but the details are less relevant to the user.
 
@@
 \section{Dependencies and installation}\label{sec:installation}
-
-\subsection{Unix-like operating systems}
 The following instructions apply to Unix-like operating systems such as
-GNU/Linux, Mac OS X, and FreeBSD.
-\pkg{DNest4} can be obtained from the \pkg{git} repository located at
+GNU/Linux, Mac OS X, and FreeBSD. Currently we have not tested
+\pkg{DNest4} on Microsoft Windows.
+
+The software can be obtained from the \pkg{git} repository located at
 \begin{center}
@@
 \end{center}
-and is licensed under the
+and is licensed under the permissive open source
 MIT license. To obtain and compile \pkg{DNest4},
@@
 and set it to the directory {\em above} the \pkg{DNest4} directory. Then,
-if the examples or templates from the \pkg{DNest4} repository are copied
+if the templates from the \pkg{DNest4} repository are copied
 to any other location on your system, their Makefiles will continue to work.
-
-\subsection{Microsoft Windows}
 
@@
 These are naive diffuse priors and do not have any special status.
-This model was implemented in \proglang{C++}
-(see Section~\ref{sec:models} for details).
 The dataset is shown in Figure~\ref{fig:regression_lines}
 and the code is included in the {\tt code/Examples/StraightLine}
-subdirectory. A simplified version of the
+subdirectory, and a slightly simplified version of the
 code is explained in Section~\ref{sec:models}.
@@
 should contain information about levels and ``saving particles to disk''.
-After 5000 particles have been saved to the disk, the run should terminate.
+After 5000 particles have been saved to the disk, the run will terminate.
 
@@
 likelihood is constant. The final column tells us which thread the particle
-belonged to: when you use \dnest~in multithreaded mode, each thread
-is responsible for evolving one or more walkers.
+belonged to: when you use \dnest~in multithreaded mode
+(see Appendix~\ref{sec:command_line_options}), each thread
+is responsible for evolving one or more particles.
 
@@
 the estimates of the level compressions (and hence the $\log(X)$ values of
-the levels in column 1), as discussed in Section 3 of the
-paper. The visits column counts the number of times a level (level $j$, say)
+the levels in column 1), as discussed in Section 3 of \citet{dnest}. The visits column counts the number of times a level (level $j$, say)
 has been visited, but only starts counting after the next level ($j+1$) has been created. The exceeds column counts the number of times a particle that was
@@
 \begin{CodeInput}
-import dnest4.classic
-dnest4.classic.postprocess()
+import dnest4
+dnest4.postprocess()
 \end{CodeInput}
@@
 \begin{enumerate}
-\item Generate $a, b, c \sim$ Normal$(0, 1)$
-\item Define $t := c/\sqrt{(a^2 + b^2)/2}$
+\item Generate $a \sim$ Normal$(0, 1)$
+\item Generate $b \sim$ Uniform$(0, 1)$
+\item Define $t := a/\sqrt{-\ln(b)}$
 \item Generate $n \sim$ Normal$(0, 1)$
@@
 \begin{CodeInput}
-dnest4.classic.postprocess_abc()
+dnest4.postprocess_abc()
 \end{CodeInput}
@@
 \begin{CodeInput}
-dnest4.classic.postprocess_abc(threshold_fraction=0.6)
+dnest4.postprocess_abc(threshold_fraction=0.6)
 \end{CodeInput}

### 2016-06-09 paper/ms.tex
Singleton pattern and stuff

@@
 
-This first chooses a random integer from $[0, 3)$ using the
+This first chooses a random integer from $\{0, 1, 2\}$ using the
 \code{rand\_int(int)} member function of the \code{DNest4::RNG} class.
@@
 it is much simpler to just use heavy-tailed proposals which have some
-probability of making a jump of appropriate size. This may seem slightly
-wasteful of CPU time, but it saves a lot of human time.
+probability of making a jump of appropriate size. This is slightly
+wasteful of CPU time, but it saves a lot of human time and is more
+fail-safe than tuned step sizes.
 In simple experiments, we have found that heavy-tailed proposals are
@@
 Alternatively, the log-uniform prior, which has density proportional
-to $1/\sigma$, could have been used directly by not using the log/exponential
-trick and adding $\ln\left[(1/\sigma')/(1/\sigma)\right]$ to the
-return value. However, this isn't recommended for a distribution like this
+to $1/\sigma$, could have been used directly by adding
+$\ln\left[(1/\sigma')/(1/\sigma)\right]$ to the
+return value instead of using the log/exp trick.
+However, this isn't recommended for a prior distribution like this
 which covers several orders of magnitude, because the appropriate scale size
@@
 accessible from anywhere) is recommended.
+The \code{Data} class has
+one static member which is itself an object of class \code{Data}, and is
+accessible using \code{Data::get_instance()} --- this is the singleton pattern,
+essentially a way of defining quasi-`global' variables:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+class Data
+{
+    private:
+        // 'static' means this exists at the level of the class
+        // instead of being truly global
+        static Data instance;
+
+    public:
+        // Getter
+        static Data& get_instance();
+};
+\end{CodeInput}
+\end{CodeChunk}
+
+Alternatively, the data could be defined using static members of your
+model class.
 
@@
 (of {\tt sample.txt}) and are separated by spaces (the
-\code{postprocess} function assumes this is the case).
+\code{postprocess} function assumes the delimeter is a space).
 
@@
 using namespace std;
-using namespace DNest4;
 
@@
     Data::get_instance().load("road.txt");
-    start<StraightLine>(argc, argv);
+    DNest4::start<StraightLine>(argc, argv);
     return 0;

### 2016-06-09 paper/ms.tex
Finished editing pass

@@
 of the $i$th component, and $\alpha$ is the vector of hyperparameters, the prior
-can be factorised via the product rule, giving
+can be factorized via the product rule, giving
 \begin{align}
@@
 $N$ normal distributions, and we want to infer $N$ along with the
-properties of those normal distributions.
+properties (positions, widths, and weights) of those normal distributions.
 The mixture weights $\{w_j\}$ must obey a normalization condition
@@
 un-normalized weights $\{W_j\}$ which do not obey
-such a condition and then normalise them by dividing by their sum.
+such a condition and then normalize them by dividing by their sum.
 
@@
 \begin{CodeInput}
-// vec = {mu, log_sigma, log_weight}
+// vec (input) = {u1, u2, u3} ~ Uniform(0, 1) in the prior
+// gets modified to {mu, log_sigma, log_weight}
 void MyConditionalPrior::from_uniform(std::vector<double>& vec) const
@@
 
-// vec = {mu, log_sigma, log_weight}
+// vec (input) = {mu, log_sigma, log_weight}
+// gets modified to {u1, u2, u3} using the CDF of the conditional prior
+// This is the inverse of from_uniform
 void MyConditionalPrior::to_uniform(std::vector<double>& vec) const
@@
 \subsection{ABC Example Results}
-We show posterior samples in Figure~\ref{fig:abc_results}.
-
-\begin{figure}[ht!]
-\centering
-\includegraphics[width=0.6\textwidth]{figures/abc_results.pdf}
-\caption{ABC results.\label{fig:abc_results}}
-\end{figure}
-
 For ABC applications, there is an alternate \code{postprocess} function
@@
 \end{CodeChunk}
+
+In the example directory there is a dataset generated from
+a standard normal distribution. The inference should
+result in a joint posterior for $\mu$ and $\sigma$ with
+significant density near $(\mu=0, \sigma=1)$.
+We show posterior samples in Figure~\ref{fig:abc_results}.
+\begin{figure}[ht!]
+\centering
+\includegraphics[width=0.6\textwidth]{figures/abc_results.pdf}
+\caption{Posterior samples for $\mu$ and $\sigma$ for the
+ABC demonstration.\label{fig:abc_results}}
+\end{figure}
 
@@
 
+The text file output can be analyzed using \code{dnest4.postprocess()}
+as usual.
+
 \section*{Acknowledgements}

### 2016-06-09 python/dnest4/classic.py
Fixed black plot

@@ -446,7 +446,7 @@ def levels_plot():
 	"""
 	levels = my_loadtxt('levels.txt')
 
-	plt.plot(np.log10(np.diff(levels[:,1])), "k.")
+	plt.plot(np.log10(np.diff(levels[:,1])), "ko-")
 	plt.ylim([-1, 4])
 	plt.axhline(0., color='g', linewidth=2)
 	plt.axhline(np.log10(np.log(10.)), color='g')

### 2016-06-10 paper/ms.tex
Minor fixes

@@
 where the $\{\ell_j\}$ are a sequence of increasing likelihood thresholds
-or {\em levels}, and
+or {\em levels},
 $\ell_0 = 0$, and $\{w_j\}$ is the marginal distribution for $j$.
@@
 the particle(s) are initialized from the prior $\pi(\theta)$, equivalent
-to the mixture of constrained priors with a single level whose likelihood
+to the mixture of constrained priors with a single level whose log likelihood
 threshold is $-\infty$. The mixture of constrained priors evolves by adding new levels,

### 2016-06-10 paper/ms.tex
Dan's logH comment

@@
 H = \frac{q(\params'|\params)}{q(\params | \params')}
-\times \frac{\pi(\params')}{\pi(\params)}
-\end{align}
-so the acceptance probability becomes
+\times \frac{\pi(\params')}{\pi(\params)}.
+\end{align}
+
+The logarithm of $H$,
+\begin{align}
+\ln(H) = \ln\left[\frac{q(\params'|\params)}{q(\params | \params')}
+\times \frac{\pi(\params')}{\pi(\params)}\right],
+\end{align}
+is the user's responsibility when implementing models,
+and will become relevant in Sections~\ref{sec:perturb}
+and~\ref{sec:proposals}.
+In terms of $\ln(H)$, the
+acceptance probability becomes
 \begin{align}
 \alpha &= \min\left[1,
-H\times
+e^{\ln(H)}\times
 \mathds{1}\left(L(\params') > \ell_j\right)
-\right]
+\right].\label{eqn:logH}
 \end{align}
@@
         change to the parameter values, and returns $\ln(H)$ as defined in
-        Equation~\ref{eqn:log_hastings};
+        Equation~\ref{eqn:logH};
 \item \code{double log\_likelihood() const}, which evaluates the log of
@@
 
-\subsection{Proposal moves}
-The \code{perturb} member function for the straight line model is:
+\subsection{Proposal moves}\label{sec:perturb}
+The \code{perturb} member function for the straight line model is given below.
+This takes a random number generator as input, makes changes to
+a subset of the parameters, and returns a \code{double} corresponding
+to $\ln(H)$ as defined in Equation~\ref{eqn:logH}. The combination of
+the choice of proposal and the $\ln(H)$ value returned must be consistent
+with the prior distribution.
 \begin{CodeChunk}
@@
 {
+    // log_H value to be returned
     double log_H = 0.0;
@@
 
-This first chooses a random integer from $\{0, 1, 2\}$ using the
+This function first chooses a random integer from $\{0, 1, 2\}$ using the
 \code{rand\_int(int)} member function of the \code{DNest4::RNG} class.
@@
 
-\subsection{Proposals for single parameters}
+\subsection{Proposals for single parameters}\label{sec:proposals}
 The proposal for $m$ involves adding a perturbation to the current

### 2016-06-10 paper/replicate.py
Replication script for JSS

@@ -1,0 +1,21 @@
+"""
+This script should replicate all the runs
+presented in the paper. Note:
+plots will be generated, which need to be manually
+closed for the script to continue. It assumes
+everything has been compiled already.
+"""
+import os
+
+# Python command on your system
+my_python = "python"
+
+def run_example(directory):
+    os.chdir(directory)
+    os.system("./main -s 0")
+    os.system(my_python + " showresults.py") 
+
+run_example("StraightLine")
+run_example("RJObject_1DMixture")
+run_example("ABC")
+

### 2016-06-10 paper/replicate.py
working

@@ -13,8 +13,10 @@ my_python = "python"
 def run_example(directory):
     os.chdir(directory)
     os.system("./main -s 0")
-    os.system(my_python + " showresults.py") 
+    os.system(my_python + " showresults.py")
+    os.chdir("..")
 
+os.chdir("../code/Examples")
 run_example("StraightLine")
 run_example("RJObject_1DMixture")
 run_example("ABC")

### 2016-06-10 paper/ms.tex
got rid of useless sentence
Update from seed 0 runs
Cite Robert

@@
 \subsection{Log-likelihood}
-The log-likelihood code is standard:
+The log-likelihood for the model is evaluated and returned by the
+\code{log\_likelihood} member function. For the straight line
+fitting example, the log likelihood is based on the
+normal density, and the code is given below.
 \begin{CodeChunk}
@@
 $N$ is shown in Figure~\ref{fig:galaxies_N}.
+The marginal likelihood was $\ln(Z) = -242.2$, and
+the information (measured in nats) was $\mathcal{H} = 32.3$.
 
@@
 to the value of $\epsilon$ using a single run.
+The marginal likelihood reported will be
+the probability that $\rho\left(D; D_{\rm observed}\right) < \epsilon$
+given the model, although caution should be applied if
+using this for model selection \citep{robert}.
 
@@
 
+\bibitem[Robert et al.(2011)]{robert}
+Robert, C.P., Cornuet, J.M., Marin, J.M.
+and Pillai, N.S., 2011.
+Lack of confidence in approximate Bayesian computation model choice.
+Proceedings of the National Academy of Sciences, 108(37), pp.15112-15117.
+
 \bibitem[\protect\citeauthoryear{Skilling}{2006}]{skilling} Skilling, J., 2006, Nested Sampling for General Bayesian Computation, Bayesian Analysis 4, pp. 833-860.

### 2016-06-10 paper/replicate.py
Note about time

@@ -4,6 +4,8 @@ presented in the paper. Note:
 plots will be generated, which need to be manually
 closed for the script to continue. It assumes
 everything has been compiled already.
+
+NOTE: the RJObject_1DMixture example is fairly time-consuming.
 """
 import os
 

### 2016-06-12 paper/ms.tex
Take out 11
title fix
Installation from a specific version, not the git repo

@@
 \Plainauthor{Brendon J. Brewer, Daniel Foreman-Mackey} %% comma-separated
-\Plaintitle{DNest4: An implementation of Diffusive Nested Sampling in
-C++11 and Python} %% without formatting
+\Plaintitle{DNest4: Diffusive Nested Sampling in
+C++ and Python} %% without formatting
 \Shorttitle{\pkg{DNest4}: Diffusive Nested Sampling} %% a short title (if necessary)
@@
 
-The software can be obtained from the \pkg{git} repository located at
+Development of \pkg{DNest4} takes place in the \pkg{git} repository located at
 \begin{center}
@@
 \end{center}
-and is licensed under the permissive open source
-MIT license. To obtain and compile \pkg{DNest4},
-you require \pkg{git} and a recent version of the GNU
+The software
+is licensed under the permissive open source
+MIT license. To compile and run \pkg{DNest4},
+you require a recent version of the GNU
 \proglang{C++} compiler, \pkg{g++}. The \proglang{Python}
@@
 \begin{CodeInput}
-git clone https://github.com/eggplantbren/DNest4
+wget https://github.com/eggplantbren/DNest4/archive/0.1.0.tar.gz
+tar xvzf 0.1.0.tar.gz
+mv DNest4-0.1.0 DNest4
 cd DNest4/code
@@
 and set it to the directory {\em above} the \pkg{DNest4} directory. Then,
-if the templates from the \pkg{DNest4} repository are copied
-to any other location on your system, their Makefiles will continue to work.
+if the model templates from the \pkg{DNest4} repository are copied
+to any other location on your system and used as the basis for
+new work, their Makefiles will continue to function.
 
@@
 \begin{CodeOutput}
-log(Z) = -175.532983853
-Information = 15.2112429682 nats.
-Effective sample size = 737.639033806
+log(Z) = -175.545067073
+Information = 15.2168377939 nats.
+Effective sample size = 658.958278958
 \end{CodeOutput}

### 2016-06-12 python/dnest4/__init__.py
release version

@@ -1,6 +1,6 @@
 # -*- coding: utf-8 -*-
 
-__version__ = "0.0.1.dev0"
+__version__ = "0.1.0"
 
 try:
     __DNEST4_SETUP__

### 2016-06-22 paper/ms.tex
nojss
Using bibtex as the journal wants

@@
-\documentclass[article]{jss}
+\documentclass[article, nojss]{jss}
 \usepackage{amsmath}
@@
 logical implication or subjective certainty, provides a powerful general basis
-for data analysis \citep{ohagan, sivia}. The result of such
+for data analysis \citep{o2004kendall, sivia2006data}. The result of such
 an analysis is typically
@@
 
-Nested Sampling \citep[NS;][]{skilling} is a Monte Carlo method whose main
+Nested Sampling \citep[NS;][]{skilling2006nested} is a Monte Carlo method whose main
 aim is to calculate $Z$. However, it can also be used to generate samples
@@
 This latter property makes NS particularly useful for statistical mechanics
-calculations \citep{2009arXiv0906.3544P, 2015arXiv150303404B}, where the
+calculations \citep{partay2010efficient, baldock2016determining},
+where the
 ``canonical'' family of distributions proportional to
@@
 is required, and Markov Chain Monte Carlo (MCMC) is a popular method of doing
-this, although alternatives exist \citep[e.g.][]{multinest, 2015MNRAS.453.4384H}.
-Diffusive Nested Sampling \citep[DNS][]{dnest} is an alternative to NS for
+this, although alternatives exist
+\citep[e.g.][]{feroz2009multinest, handley2015polychord}.
+Diffusive Nested Sampling \citep[DNS;][]{brewer2011diffusive}
+is an alternative to NS for
 problems where MCMC is the only viable sampling method. DNS is based on the
@@
 each of which compresses the distribution by a factor of about $e \approx 2.71818$
-(see \citet{dnest} for details). In this stage,
+(see \citet{brewer2011diffusive} for details). In this stage,
 the mixture weights $\{w_j\}$ are set according to
@@
 The second stage sets the mixture weights to be approximately
-uniform ($w_j \propto 1$), with some tweaks described by \citet{dnest}.
+uniform ($w_j \propto 1$), with some tweaks described by
+\citet{brewer2011diffusive}.
 
@@
 likelihood estimate can also be more accurate than standard MCMC-based NS
-\citep{dnest}, as less information is discarded.
+\citep{brewer2011diffusive}, as less information is discarded.
 DNS has been applied several times in astrophysics
-\citep[e.g.][]{2014MNRAS.445.3055P, 2015ApJ...810...66H, 2015MNRAS.448.3206B}
-and was recently used in a biological application \citep{salmonella}.
+\citep[e.g.][]{pancoast2014modelling, huppenkothen2015dissecting,
+brewer2015fast}
+and was recently used in a biological application
+\citep{dybowski2015single}.
 
@@
 the estimates of the level compressions (and hence the $\log(X)$ values of
-the levels in column 1), as discussed in Section 3 of \citet{dnest}. The visits column counts the number of times a level (level $j$, say)
+the levels in column 1), as discussed in Section 3 of
+\citet{brewer2011diffusive}.
+The visits column counts the number of times a level (level $j$, say)
 has been visited, but only starts counting after the next level ($j+1$) has been created. The exceeds column counts the number of times a particle that was
@@
 
-As described by \citet{skilling}, there is no guarantee in principle that
+As described by \citet{skilling2006nested}, there is no guarantee in principle that
 another peak might have appeared had a run continued for longer. These
@@
 in data analysis problems, but can appear in the latter
-\citep[e.g.][]{rjobject, 2015MNRAS.448.3206B}.
+\citep[e.g.][]{brewer2014inference, brewer2015fast}.
 
@@
 \subsection{Equal weight enforcement}
-This value is the parameter $\beta$ described in \citet{dnest}, and compensates
+This value is the parameter $\beta$ described in
+\citet{brewer2011diffusive}, and compensates
 for imprecision in the spacing of the levels, so that the desired mixture
@@
 In simple experiments, we have found that heavy-tailed proposals are
-about as efficient as slice sampling \citep{slice}, but much easier to
+about as efficient as slice sampling \citep{neal2003slice}, but much easier to
 implement. The following procedure generates
@@
 in Bayesian data analysis. To reduce the amount of effort needed to
-implement mixture models in DNS, \citet{rjobject} implemented
+implement mixture models in DNS, \citet{brewer2014inference} implemented
 a template class called \code{RJObject} to handle the MCMC moves
-required. The RJ in \code{RJObject} stands for Reversible Jump \citep{green}, as
+required. The RJ in \code{RJObject} stands for Reversible Jump
+\citep{green1995reversible}, as
 \code{RJObject} implements birth and death moves for mixture
@@
 
-\citet{rjobject} described the motivation for \code{RJObject} and some details
+\citet{brewer2014inference}
+described the motivation for \code{RJObject} and some details
 about the Metropolis proposals underlying it. Here, we demonstrate how to
@@
 directory. The ``SineWave'' and ``GalaxyField'' models
-from \citet{rjobject} are also included in the \pkg{DNest4} repository.
+from \citet{brewer2014inference}
+are also included in the \pkg{DNest4} repository.
 
@@
 of compression $X$ for the galaxy data. There are phase changes
-\citep{skilling} for which Nested Sampling is the best known
+\citep{skilling2006nested} for which Nested Sampling is the best known
 solution.
@@
 distribution for the parameters {\em and the data} provides a bridge
-connecting Bayesian updating to maximum entropy updating \citep{caticha,giffin}.}.
+connecting Bayesian updating to maximum entropy updating
+\citep{caticha2006updating,giffin2007updating}.}.
 
@@
 implementation. Most ABC analyses are done using Sequential Monte Carlo
-\citep[SMC; ][]{delmoral}, another family of algorithms closely related
+\citep[SMC; ][]{del2012adaptive}, another family of algorithms closely related
 to Nested Sampling and annealing in the sense that they work with
@@
 given the model, although caution should be applied if
-using this for model selection \citep{robert}.
+using this for model selection \citep{robert2011lack}.
 
@@
 
-\begin{thebibliography}{99}
-\bibitem[Baldock et al.(2015)]{2015arXiv150303404B} Baldock, R.~J.~N., 
-P{\'a}rtay, L.~v.~B., Bart{\'o}k, A.~P., Payne, M.~C., Cs{\'a}nyi, G.\ 
-2015.\ Determining pressure-temperature phase diagrams of materials.\ ArXiv 
-e-prints arXiv: 1503.03404.
-
-\bibitem[\protect\citeauthoryear{Brewer, P{\'a}rtay,
-\& Cs{\'a}nyi}{2011}]{dnest} Brewer B.~J., P{\'a}rtay L.~B., Cs{\'a}nyi G., 2011,
-Statistics and Computing, 21, 4, 649-656. arXiv:0912.2380
-
-\bibitem[Brewer(2014)]{rjobject} Brewer, B.~J.\ 2014.\ Inference for Trans-dimensional Bayesian Models with Diffusive Nested Sampling.\ ArXiv e-prints arXiv:1411.3921.
-
-\bibitem[Brewer and Donovan(2015)]{2015MNRAS.448.3206B} Brewer, B.~J., Donovan, C.~P.\ 2015.\ Fast Bayesian inference for exoplanet discovery in radial velocity data.\ Monthly Notices of the Royal Astronomical Society 448, 3206-3214.
-
-\bibitem[Caticha and Giffin(2006)]{caticha} Caticha, A., Giffin, A.\ 2006.\ Updating Probabilities.\ Bayesian Inference and Maximum Entropy Methods In Science and Engineering 872, 31-42.
-
-\bibitem[Del Moral et al(2012)]{delmoral}
-Del Moral, P., Doucet, A. and Jasra, A., 2012. An adaptive sequential Monte Carlo method for approximate Bayesian computation. Statistics and Computing, 22(5), pp.1009-1020.
-
-\bibitem[Dybowski et al(2015)]{salmonella}
-Dybowski, Richard, et al. Single passage in mouse organs enhances the survival and spread of Salmonella enterica. Journal of The Royal Society Interface 12.113 (2015): 20150702.
-
-\bibitem[\protect\citeauthoryear{Feroz, Hobson,
-\& Bridges}{2009}]{multinest} Feroz F., Hobson M.~P., Bridges M., 2009, MNRAS, 398, 1601
-
-\bibitem[Giffin and Caticha(2007)]{giffin} Giffin, A., Caticha, A.\ 2007.\ Updating Probabilities with Data and Moments.\ Bayesian Inference and Maximum Entropy
-Methods in Science and Engineering 954, 74-84.
-
-\bibitem[\protect\citeauthoryear{Green}{1995}]{green}
-Green, P.~J., 1995, Reversible Jump Markov Chain Monte Carlo Computation and Bayesian Model Determination, Biometrika 82 (4): 711–732.
-
-\bibitem[Handley et al.(2015)]{2015MNRAS.453.4384H} Handley, W.~J., Hobson, M.~P., Lasenby, A.~N.\ 2015.\ POLYCHORD: next-generation nested sampling.\ Monthly Notices of the Royal Astronomical Society 453, 4384-4398.
-
-\bibitem[Huppenkothen et al.(2015)]{2015ApJ...810...66H} Huppenkothen, D., Brewer, B.~J., Hogg, D.~W., Murray, I., Frean, M., Elenbaas, C., Watts, A.~L., Levin, Y., van der Horst, A.~J., Kouveliotou, C.\ 2015.\ Dissecting Magnetar Variability with Bayesian Hierarchical Models.\ The Astrophysical Journal 810, 66.
-
-\bibitem[Neal(2003)]{slice}
-Neal, R.M., 2003. Slice sampling. Annals of statistics, pp.705-741.
-
-\bibitem[\protect\citeauthoryear{O'Hagan and Forster}{2004}]{ohagan}
-O'Hagan, A., Forster,~J., 2004, Bayesian inference. London: Arnold.
-
-\bibitem[Pancoast et al.(2014)]{2014MNRAS.445.3055P} Pancoast, A., Brewer, B.~J., Treu, T.\ 2014.\ Modelling reverberation mapping data - I.
-Improved geometric and dynamical models and comparison with cross-correlation results.\ Monthly Notices of the Royal Astronomical Society 445, 3055-3072.
-
-\bibitem[\protect\citeauthoryear{Sivia and Skilling}{2006}]{sivia} Sivia, 
-D.~ S., Skilling, J., 2006, Data Analysis: A Bayesian Tutorial, 2nd 
-Edition, Oxford University Press
-
-\bibitem[P{\'a}rtay et al.(2009)]{2009arXiv0906.3544P} P{\'a}rtay, L.~B., 
-Bart{\'o}k, A.~P., Cs{\'a}nyi, G.\ 2009.\ Efficient sampling of atomic 
-configurational spaces.\ ArXiv e-prints arXiv: 0906.3544.
-
-\bibitem[Robert et al.(2011)]{robert}
-Robert, C.P., Cornuet, J.M., Marin, J.M.
-and Pillai, N.S., 2011.
-Lack of confidence in approximate Bayesian computation model choice.
-Proceedings of the National Academy of Sciences, 108(37), pp.15112-15117.
-
-\bibitem[\protect\citeauthoryear{Skilling}{2006}]{skilling} Skilling, J., 2006, Nested Sampling for General Bayesian Computation, Bayesian Analysis 4, pp. 833-860.
-
-\end{thebibliography}
+\bibliography{references}
 
@@
 a dataset to refer to in your {\tt log\_likelihood}. The example in
-the repository is the ``SpikeSlab'' likelihood from \citet{dnest}.
+the repository is the ``SpikeSlab'' likelihood from
+\citet{brewer2011diffusive}.
 

### 2016-06-26 paper/ms.tex
Cite python packages
Stroustrup reference
GCC citation
Julia reference

@@
 you require a recent version of the GNU
-\proglang{C++} compiler, \pkg{g++}. The \proglang{Python}
-packages \pkg{NumPy}, \pkg{matplotlib},
-\pkg{Pandas}, and \pkg{Cython} are also needed.
+\proglang{C++} compiler, \pkg{g++} \citep{gcc}.
+\pkg{DNest4} uses features from the \proglang{C++11} standard
+\citep{c++11}.
+
+The \proglang{Python}
+packages \pkg{NumPy} \citep{numpy}, \pkg{matplotlib} \citep{matplotlib},
+\pkg{Pandas} \citep{pandas}, and \pkg{Cython} \citep{cython} are also needed.
 To download and compile \pkg{DNest4}
@@
 Experimental support for model specification in the
-\proglang{Julia} language is provided, with an example given in the
+\proglang{Julia} language \citep{julia}
+is provided, with an example given in the
 {\tt code/Templates/JuliaModel} directory.

### 2016-06-27 paper/ms.tex
Comparison to other packages

@@
 compiler.
+
+\section{Relation to Other Algorithms}\label{sec:relation}
+The simple C implementation of Nested Sampling given by
+\citet{skilling2006nested}
+
+\pkg{MultiNest} \citep{feroz2009multinest} is a Nested Sampling implementation
+that dispenses with MCMC for generating new particles from constrained
+prior distributions. Instead, the new particles are generated by making
+an approximation to the constrained prior using ellipses. This tends to work
+well in low to moderate dimensional parameter spaces (up to $\sim 50$).
+In higher dimensions, \pkg{DNest4} is more useful than \pkg{MultiNest}
+since it is based on MCMC. We expect \pkg{MultiNest} to be more useful
+than
+\pkg{DNest4} only on low to moderate dimensional problems with slow
+likelihood evaluations.
+
+\pkg{emcee} \citep{emcee}
+
+The popular \pkg{JAGS} \citep{jags} and \pkg{Stan} \citep{stan} packages
+are more convenient than \pkg{DNest4} for specifying models, by virtue
+of their model-specification languages. However, being based on Gibbs
+sampling and Hamiltonian MCMC respectively, they can run into difficulty
+on multimodal posterior distributions. Also, they do not calculate
+marginal likelihoods.
 

### 2016-06-28 paper/ms.tex
Table

@@
 \usepackage{dsfont}
-
+\usepackage{pifont}
+
+\newcommand{\xmark}{\ding{55}}%
+\newcommand{\cmark}{\ding{51}}%
 \newcommand{\dnest}{\pkg{DNest4}}
@@
 
-\pkg{emcee} \citep{emcee}
+\pkg{emcee} \citep{emcee} is a popular \proglang{Python} package for MCMC
+based on affine-invariant ensemble sampling.
 
@@
 marginal likelihoods.
+
+The advantages and disadvantages of these various packages, in the opinion
+of the authors, are summarized in Table~\ref{tab:packages}.
+
+%\begin{table}
+%
+%  \begin{tabular}{cc}
+%    Knuth & Lamport
+%  \end{tabular}}
+%\end{table}
+
+\begin{table}
+\begin{center}
+\resizebox{\textwidth}{!}{%
+\begin{tabular}{|l|c|c|c|c|c|c|}
+\hline
+Package     &   Easy to     & High   &   Multimodal     & Dependent   &   Phase   &   Computes $Z$?\\
+            &   implement & dimensions? & distributions? & distributions? & changes? & \\
+& models? & & & & &\\
+\hline
+\pkg{DNest4}    & \xmark  & \cmark & \cmark & \cmark &\cmark &\cmark\\
+\pkg{emcee}    & \cmark  & \xmark & \xmark & \cmark &\xmark &\xmark\\
+\pkg{JAGS}    & \cmark  & \cmark & \xmark & \xmark & \xmark &\xmark\\
+\pkg{MultiNest}    & \cmark  & \xmark & \cmark & \cmark & \cmark &\cmark\\
+\pkg{Stan}    & \cmark  & \cmark & \xmark & \cmark & \xmark &\xmark\\
+\hline
+\end{tabular}}
+\caption{\label{tab:packages}}
+\end{center}
+\end{table}
 

### 2016-06-28 paper/ms.tex
Finished comparison section

@@
 \section{Relation to Other Algorithms}\label{sec:relation}
-The simple C implementation of Nested Sampling given by
-\citet{skilling2006nested}
-
-\pkg{MultiNest} \citep{feroz2009multinest} is a Nested Sampling implementation
+The Diffusive Nested Sampling algorithm, and its implementation in
+\pkg{DNest4}, has advantages and disadvantages compared to other Bayesian
+computation algorithms and software.
+
+In the Nested Sampling sphere,
+the simple \proglang{C} implementation of given by
+\citet{skilling2006nested} is useful for understanding the classic NS
+algorithm. Similarly to \pkg{DNest4}, it is the user's responsibility
+to implement the MCMC moves used. The DNS algorithm produces more
+accurate estimates of $Z$ and ought to outperform classic NS on multimodal
+problems.
+
+\pkg{MultiNest} \citep{feroz2009multinest} is a
+\proglang{FORTRAN}
+Nested Sampling implementation
 that dispenses with MCMC for generating new particles from constrained
 prior distributions. Instead, the new particles are generated by making
-an approximation to the constrained prior using ellipses. This tends to work
-well in low to moderate dimensional parameter spaces (up to $\sim 50$).
+an approximation to the constrained prior using ellipsoids. This tends to work
+well in low to moderate dimensional parameter spaces (up to $\sim 40$).
 In higher dimensions, \pkg{DNest4} is more useful than \pkg{MultiNest}
@@
 \pkg{DNest4} only on low to moderate dimensional problems with slow
-likelihood evaluations.
+likelihood evaluations, where \pkg{DNest4}'s reliance on MCMC becomes costly.
+The more recent \pkg{POLYCHORD} \citep{handley2015polychord} combines
+\pkg{MultiNest}-like methods with MCMC.
+
+Outside of Nested Sampling,
+the popular \pkg{JAGS} \citep{jags} and \pkg{Stan} \citep{stan} packages
+are more convenient than \pkg{DNest4} for specifying models, by virtue
+of their model-specification languages. For users who are interested only
+in the posterior distribution and do not need the marginal likelihood,
+these are very useful.
+However, being based on Gibbs
+sampling and Hamiltonian MCMC respectively, they can run into difficulty
+on multimodal posterior distributions.
 
 \pkg{emcee} \citep{emcee} is a popular \proglang{Python} package for MCMC
-based on affine-invariant ensemble sampling.
-
-The popular \pkg{JAGS} \citep{jags} and \pkg{Stan} \citep{stan} packages
-are more convenient than \pkg{DNest4} for specifying models, by virtue
-of their model-specification languages. However, being based on Gibbs
-sampling and Hamiltonian MCMC respectively, they can run into difficulty
-on multimodal posterior distributions. Also, they do not calculate
-marginal likelihoods.
-
-The advantages and disadvantages of these various packages, in the opinion
-of the authors, are summarized in Table~\ref{tab:packages}.
+based on affine-invariant ensemble sampling. The user needs only to
+specify a \proglang{Python} function evaluating the log of the posterior
+density (up to a normalizing constant). This user-friendliness is a key
+advantage of \pkg{emcee}, and its algorithm performs well on low--moderate
+dimensional (up to $\sim 50$) parameter spaces with a unimodel
+target distribution (which can be highly dependent). However, it
+can give misleading results in higher dimensions \citep{huijser}.
+
+Algorithms based on ``annealing'' or ``tempering'', such as parallel
+tempering \citep{hansmann1997parallel} and annealed importance sampling
+\citep{neal2001annealed} are
+related to Nested Sampling and are useful on multimodal posterior distributions.
+However, they require much more tuning than NS (in the form of an
+{\em annealing schedule}) and do
+not work on phase change problems, unlike Nested Sampling
+\citep{skilling2006nested}.
+
+The advantages and disadvantages of a subset of these
+packages are summarized in Table~\ref{tab:packages}.
 
@@
 \end{tabular}}
-\caption{\label{tab:packages}}
+\caption{A simplified summary of the advantages and disadvantages of some
+Bayesian computation software packages.\label{tab:packages}}
 \end{center}

### 2016-06-30 paper/replicate.py
Updated replicate so it'll work stand-alone

@@ -1,16 +1,24 @@
-"""
-This script should replicate all the runs
-presented in the paper. Note:
-plots will be generated, which need to be manually
-closed for the script to continue. It assumes
-everything has been compiled already.
-
-NOTE: the RJObject_1DMixture example is fairly time-consuming.
-"""
 import os
 
 # Python command on your system
-my_python = "python"
+my_python = "/home/brendon/local/anaconda3/bin/python"
+
+print("This script should replicate all the runs presented in the paper.\nNote:\
+plots will be generated, which need to be manually\n\
+closed for the script to continue. It assumes\n\
+everything has been compiled already.\n\
+\n\
+NOTE: the RJObject_1DMixture example is fairly time-consuming.\n\n\
+Press ENTER to begin.")
+junk = input()
+
+os.system("wget https://github.com/eggplantbren/DNest4/archive/0.1.0.tar.gz")
+os.system("tar xvzf 0.1.0.tar.gz")
+os.system("mv DNest4-0.1.0 DNest4")
+os.chdir("DNest4/code")
+os.system("make")
+os.chdir("../python")
+os.system("python setup.py install")
 
 def run_example(directory):
     os.chdir(directory)

### 2016-07-03 paper/ms.tex
Affiliations
0.1.1

@@
 %% almost as usual
-\author{Brendon J. Brewer\\The University of Auckland\And 
-        Daniel Foreman-Mackey\\University of Washington}
+\author{Brendon J. Brewer\\Department of Statistics\\The University of Auckland\And 
+        Daniel Foreman-Mackey\\Sagan Fellow\\University of Washington}
 \title{\pkg{DNest4}: Diffusive Nested Sampling in
@@
 \begin{CodeInput}
-wget https://github.com/eggplantbren/DNest4/archive/0.1.0.tar.gz
-tar xvzf 0.1.0.tar.gz
-mv DNest4-0.1.0 DNest4
+wget https://github.com/eggplantbren/DNest4/archive/0.1.1.tar.gz
+tar xvzf 0.1.1.tar.gz
+mv DNest4-0.1.1 DNest4
 cd DNest4/code

### 2016-07-03 paper/replicate.py

@@ -3,7 +3,8 @@ import os
 # Python command on your system
 my_python = "/home/brendon/local/anaconda3/bin/python"
 
-print("This script should replicate all the runs presented in the paper.\nNote:\
+print("This script should download and install DNest4 and \
+replicate all the runs presented in the paper.\nNote:\
 plots will be generated, which need to be manually\n\
 closed for the script to continue. It assumes\n\
 everything has been compiled already.\n\
@@ -12,9 +13,9 @@ NOTE: the RJObject_1DMixture example is fairly time-consuming.\n\n\
 Press ENTER to begin.")
 junk = input()
 
-os.system("wget https://github.com/eggplantbren/DNest4/archive/0.1.0.tar.gz")
-os.system("tar xvzf 0.1.0.tar.gz")
-os.system("mv DNest4-0.1.0 DNest4")
+os.system("wget https://github.com/eggplantbren/DNest4/archive/0.1.1.tar.gz")
+os.system("tar xvzf 0.1.1.tar.gz")
+os.system("mv DNest4-0.1.1 DNest4")
 os.chdir("DNest4/code")
 os.system("make")
 os.chdir("../python")

### 2016-07-03 code/Version.h
0.1.1

@@ -3,6 +3,6 @@
 
 #define DNEST4_MAJOR_VERSION 0
 #define DNEST4_MINOR_VERSION 1
-#define DNEST4_PATCH_VERSION 0
+#define DNEST4_PATCH_VERSION 1
 
 #endif  // DNest4_Version

### 2016-07-03 python/dnest4/__init__.py

@@ -1,6 +1,6 @@
 # -*- coding: utf-8 -*-
 
-__version__ = "0.1.0"
+__version__ = "0.1.1"
 
 try:
     __DNEST4_SETUP__

### 2016-07-03 paper/ms.tex
proglang

@@
 (ii)
-A \pkg{Python} package allowing basic use without \proglang{C++} coding; and
+A \proglang{Python} package allowing basic use without \proglang{C++} coding; and
 iii)

### 2016-07-11 python/dnest4/loading.py
Removed pandas dependency, my_loadtxt works well

@@ -1,17 +1,48 @@
 # -*- coding: utf-8 -*-
 
 import numpy as np
-import pandas as pd
 
 __all__ = ["my_loadtxt", "loadtxt_rows"]
 
-def my_loadtxt(filename, single_precision=False, delimiter=" "):
-    if single_precision:
-        return pd.read_csv(filename, header=None, delimiter=delimiter,\
-                           comment="#", dtype=np.float32)\
-                                    .dropna(axis=1).values
-    return pd.read_csv(filename, header=None, delimiter=delimiter, comment="#")\
-                                                 .dropna(axis=1).values
+def my_loadtxt(filename, single_precision=False):
+    """
+    Load quickly
+    """
+    # Open the file
+    f = open(filename, "r")
+
+    # Storage
+    results = []
+
+    # Rows and columns
+    nrow = 0
+    ncol = None
+
+    while(True):
+        # Read the line and split by whitespace
+        line = f.readline()
+        cells = line.split()
+
+        # Quit when you see a different number of columns
+        if ncol is not None and len(cells) != ncol:
+            break
+
+        # Non-comment lines
+        if cells[0][0] != "#":
+            # If it's the first one, get the number of columns
+            if ncol is None:
+                ncol = len(cells)
+
+            # Otherwise, include in results
+            if single_precision:
+                results.append(np.array([float(cell) for cell in cells],\
+                                                          dtype="float32"))
+            else:
+                results.append(np.array([float(cell) for cell in cells]))
+            nrow += 1
+
+    results = np.vstack(results)
+    return results
 
 def loadtxt_rows(filename, rows, single_precision=False):
     """

### 2016-07-11 paper/ms.tex
Removed pandas refs

@@
 packages \pkg{NumPy} \citep{numpy}, \pkg{matplotlib} \citep{matplotlib},
-\pkg{Pandas} \citep{pandas}, and \pkg{Cython} \citep{cython} are also needed.
+and \pkg{Cython} \citep{cython} are also needed.
 To download and compile \pkg{DNest4}

### 2016-07-26 code/Version.h
version bump

@@ -3,6 +3,6 @@
 
 #define DNEST4_MAJOR_VERSION 0
 #define DNEST4_MINOR_VERSION 1
-#define DNEST4_PATCH_VERSION 1
+#define DNEST4_PATCH_VERSION 2
 
 #endif  // DNest4_Version

### 2016-07-26 python/dnest4/__init__.py

@@ -1,6 +1,6 @@
 # -*- coding: utf-8 -*-
 
-__version__ = "0.1.1"
+__version__ = "0.1.2"
 
 try:
     __DNEST4_SETUP__

### 2016-07-26 paper/ms.tex
version bump in paper

@@
 \begin{CodeInput}
-wget https://github.com/eggplantbren/DNest4/archive/0.1.1.tar.gz
-tar xvzf 0.1.1.tar.gz
-mv DNest4-0.1.1 DNest4
+wget https://github.com/eggplantbren/DNest4/archive/0.1.2.tar.gz
+tar xvzf 0.1.2.tar.gz
+mv DNest4-0.1.2 DNest4
 cd DNest4/code

### 2016-08-10 python/dnest4/classic.py
Put header in posterior_sample.txt

@@ -195,6 +195,15 @@ def postprocess(temperature=1., numResampleLogX=1, plot=True, loaded=[], \
 				break
 		rows[i] = which + cut
 
+    # Get header row
+	f = open("sample.txt", "r")
+	line = f.readline()
+	if line[0] == "#":
+		header = line[1:]
+	else:
+		header = ""
+	f.close()
+
 	sample = loadtxt_rows("sample.txt", set(rows), single_precision)
 	posterior_sample = None
 	if single_precision:
@@ -209,9 +218,11 @@ def postprocess(temperature=1., numResampleLogX=1, plot=True, loaded=[], \
 	if save:
 		np.savetxt('weights.txt', w)
 		if single_precision:
-			np.savetxt("posterior_sample.txt", posterior_sample, fmt="%.7e")
+			np.savetxt("posterior_sample.txt", posterior_sample, fmt="%.7e",\
+													header=header)
 		else:
-			np.savetxt("posterior_sample.txt", posterior_sample)
+			np.savetxt("posterior_sample.txt", posterior_sample,\
+													header=header)
 
 	if plot:
 		plt.show()

### 2016-08-10 paper/ms.tex
Clearer figure
Sentence style

@@
 
-\section{Relation to Other Algorithms}\label{sec:relation}
+\section{Relation to other algorithms}\label{sec:relation}
 The Diffusive Nested Sampling algorithm, and its implementation in
@@
 
-\section{Markov Chain Monte Carlo}\label{sec:mcmc}
+\section{Markov chain monte carlo}\label{sec:mcmc}
 DNS is build upon the Metropolis-Hastings algorithm.
@@
 \centering
-\includegraphics[width=7cm]{figures/fig1.pdf}
+\includegraphics[width=8cm]{figures/fig1.pdf}
 \caption{The level $j$ of the saved particles over time.
@@
 
-\subsection{Log-likelihood}
+\subsection{Likelihood}
 The log-likelihood for the model is evaluated and returned by the
@@
 
-\section{Finite Mixture Models with RJObject}
+\section{Finite mixture models with the RJObject class}
 Mixture models are a useful way of representing realistic prior information
@@
 
-\section{Approximate Bayesian Computation}
+\section{Approximate Bayesian computation}
 \subsection{Background}
@@
 
-\subsection{ABC Example}
+\subsection{ABC example}
 A simple ABC example is included in the directory
@@
 
-\subsection{ABC Example Results}
+\subsection{ABC example results}
 For ABC applications, there is an alternate \code{postprocess} function
@@
 
-\section{Python Bindings}\label{sec:python_bindings}
+\section{Python bindings}\label{sec:python_bindings}
 In \pkg{DNest4}, it is also possible to specify and run models in
@@
 \appendix
-\section{Command Line Options}\label{sec:command_line_options}
+\section{Command line options}\label{sec:command_line_options}
 The following command line options are available.\\

### 2016-08-10 paper/ms.tex
comma after e.g. and i.e.

@@
 this, although alternatives exist
-\citep[e.g.][]{feroz2009multinest, handley2015polychord}.
+\citep[e.g.,][]{feroz2009multinest, handley2015polychord}.
 Diffusive Nested Sampling \citep[DNS;][]{brewer2011diffusive}
@@
 DNS has been applied several times in astrophysics
-\citep[e.g.][]{pancoast2014modelling, huppenkothen2015dissecting,
+\citep[e.g.,][]{pancoast2014modelling, huppenkothen2015dissecting,
 brewer2015fast}
@@
 to consider the Metropolis acceptance probability for fixed $j$,
-i.e. with respect to a single constrained prior like
+i.e., with respect to a single constrained prior like
 Equation~\ref{eqn:constrained_prior}.
@@
 \begin{CodeInput}
-wget https://github.com/eggplantbren/DNest4/archive/0.1.2.tar.gz
-tar xvzf 0.1.2.tar.gz
-mv DNest4-0.1.2 DNest4
-cd DNest4/code
-make
-cd ../python
-python setup.py install
+> wget https://github.com/eggplantbren/DNest4/archive/0.1.2.tar.gz
+> tar xvzf 0.1.2.tar.gz
+> mv DNest4-0.1.2 DNest4
+> cd DNest4/code
+> make
+> cd ../python
+> python setup.py install
 \end{CodeInput}
@@
 The executable {\tt main} is responsible for the exploration part of the
-DNS algorithm (i.e. running the MCMC chain, building
+DNS algorithm (i.e., running the MCMC chain, building
 levels, and then exploring all the levels). It creates three text output files,
@@
 that were built during the run. The first column has estimates of the $\log(X)$
-values of the levels, i.e. how compressed they are relative to the prior, in
+values of the levels, i.e., how compressed they are relative to the prior, in
 units of nats. For example, if a level has $\log(X) = -1.02$, its likelihood
@@
 files; what is usually called ``thinning''. Saving more frequently
-(i.e. a smaller save interval) is usually better. However, this can result
+(i.e., a smaller save interval) is usually better. However, this can result
 in
@@
 in data analysis problems, but can appear in the latter
-\citep[e.g.][]{brewer2014inference, brewer2015fast}.
+\citep[e.g.,][]{brewer2014inference, brewer2015fast}.
 
@@
         // log_H takes care of the prior ratio
-        // i.e. log_H = log(pi(theta')/pi(theta))
+        // i.e., log_H = log(pi(theta')/pi(theta))
         log_H -= -0.5*pow(m/1E3, 2);
@@
 The order of $\{(\mu_j, \sigma_j, w_j)\}$
-(i.e. which one is parameter 0, 1, and 2)
+(i.e., which one is parameter 0, 1, and 2)
 is determined by the conditional prior class, and is
@@
 The \code{RJObject} class is used to manage the components. An additional
-class is needed to define and manage the hyperparameters $\alpha$ (i.e.
+class is needed to define and manage the hyperparameters $\alpha$ (i.e.,
 define how they are generated from the prior, how they are proposed, and
@@
 the cumulative distribution function (CDF) of the conditional
-prior and its inverse. i.e. \code{from\_uniform}, if the input vector contains
+prior and its inverse. i.e., \code{from\_uniform}, if the input vector contains
 iid draws from Uniform(0, 1), these should be modified to become draws from
@@
 $D = D_{\rm observed}$ with a logically
-weaker one (i.e. one that is implied by $D = D_{\rm observed}$ but does not
+weaker one (i.e., one that is implied by $D = D_{\rm observed}$ but does not
 imply it).
@@
 \end{enumerate}
-Challenge (i) is Bayesian in nature, i.e. it relates to the very
+Challenge (i) is Bayesian in nature, i.e., it relates to the very
 definition of the posterior distribution itself.

### 2016-08-10 paper/ms.tex
Some more edits to comply with JSS

@@
 \pkg{MultiNest} \citep{feroz2009multinest} is a
-\proglang{FORTRAN}
+\proglang{Fortran}
 Nested Sampling implementation
@@
    // Naive diffuse prior
-   m = 1E3*rng.randn();
-   b = 1E3*rng.randn();
+   m = 1E3 * rng.randn();
+   b = 1E3 * rng.randn();
 
    // Log-uniform prior
-   sigma = exp(-10.0 + 20.0*rng.rand());
+   sigma = exp( - 10.0 + 20.0 * rng.rand());
 }
@@
         // i.e., log_H = log(pi(theta')/pi(theta))
-        log_H -= -0.5*pow(m/1E3, 2);
+        log_H -= -0.5 * pow(m / 1E3, 2);
 
         // Take a step
-        m += 1E3*rng.randh();
+        m += 1E3 * rng.randh();
 
         // log_H takes care of the prior ratio
-        log_H += -0.5*pow(m/1E3, 2);
+        log_H += -0.5 * pow(m / 1E3, 2);
     }
@@
     {
-        log_H -= -0.5*pow(b/1E3, 2);
-        b += 1E3*rng.randh();
-        log_H += -0.5*pow(b/1E3, 2);
+        log_H -= -0.5*pow(b / 1E3, 2);
+        b += 1E3 * rng.randh();
+        log_H += -0.5*pow(b / 1E3, 2);
     }
@@
         sigma = log(sigma);
-        sigma += 20.0*rng.randh();
+        sigma += 20.0 * rng.randh();
 
@@
 \begin{CodeInput}
-m += 1E3*rng.randh();
+m += 1E3 * rng.randh();
 \end{CodeInput}
@@
 \begin{CodeInput}
-sigma += 20.0*rng.randh();
+sigma += 20.0 * rng.randh();
 DNest4::wrap(sigma, -10.0, 10.0);
@@
     // Variance
-    double var = sigma*sigma;
+    double var = sigma * sigma;
 
@@
     double mu;
-    for(size_t i=0; i<y.size(); ++i)
-    {
-        mu = m*x[i] + b;
-        log_L += -0.5*log(2*M_PI*var) - 0.5*pow(y[i] - mu, 2)/var;
+    for(size_t i = 0; i < y.size(); ++i)
+    {
+        mu = m * x[i] + b;
+        log_L += -0.5 * log(2 * M_PI * var) - 0.5 * pow(y[i] - mu, 2) / var;
     }
@@
 {
-    out<<m<<' '<<b<<' '<<sigma;
+    out << m << ' ' << b << ' ' << sigma;
 }
@@
     {
-        mu += 20.*rng.randh();
-        wrap(mu, -10., 10.);
+        mu += 20 * rng.randh();
+        wrap(mu, -10.0, 10.0);
     }
@@
     {
-        log_sigma += 20.*rng.randh();
-        wrap(log_sigma, -10., 10.);
+        log_sigma += 20 * rng.randh();
+        wrap(log_sigma, -10.0, 10.0);
     }
@@
     vector<double> x_fake = n;
-    for(size_t i=0; i<x_fake.size(); i++)
-        x_fake[i] = mu + sigma*x_fake[i];
+    for(size_t i = 0; i < x_fake.size(); i++)
+        x_fake[i] = mu + sigma * x_fake[i];
 
@@
 \begin{CodeInput}
-dnest4.postprocess_abc(threshold_fraction=0.6)
+dnest4.postprocess_abc(threshold_fraction = 0.6)
 \end{CodeInput}
@@
         """
-        m = 1E3*rng.randn()
-        b = 1E3*rng.randn()
-        sigma = np.exp(-10.0 + 20.0*rng.rand())
+        m = 1E3 * rng.randn()
+        b = 1E3 * rng.randn()
+        sigma = np.exp(-10.0 + 20.0 * rng.rand())
         return np.array([m, b, sigma])
@@
         if which == 0 or which == 1:
-            logH -= -0.5*(params[which]/1E3)**2
-            params[which] += 1E3*dnest4.randh()
-            logH += -0.5*(params[which]/1E3)**2
+            logH -= -0.5 * (params[which] / 1E3) **2
+            params[which] += 1E3 * dnest4.randh()
+            logH += -0.5 * (params[which]/1E3) **2
         else:
             log_sigma = np.log(params[2])
-            log_sigma += 20*dnest4.randh()
+            log_sigma += 20 * dnest4.randh()
             # Note the difference between dnest4.wrap in Python and
@@
         m, b, sigma = params
-        var = sigma**2
-        return -0.5*data.shape[0]*np.log(2*np.pi*var)\
-                - 0.5*np.sum((data[:,1] - (m*data[:,0] + b))**2)/var
+        var = sigma ** 2
+        return -0.5 * data.shape[0] * np.log(2 * np.pi * var)\
+                - 0.5 * np.sum((data[:,1] - (m * data[:,0] + b)) **2) / var
 
@@
 # Set up the sampler. The first argument is max_num_levels
-gen = sampler.sample(max_num_levels=30, num_steps=1000, new_level_interval=10000,
-                      num_per_step=10000, thread_steps=100,
+gen = sampler.sample(max_num_levels=30, num_steps=1000,\
+                      new_level_interval=10000,\
+                      num_per_step=10000, thread_steps=100,\
                       num_particles=5, lam=10, beta=100, seed=1234)

### 2016-08-11 paper/ms.tex
Less caps, to be on the safe side

@@
 
-\subsection{Likelihood}
+\subsection{Log likelihood}
 The log-likelihood for the model is evaluated and returned by the
@@
 
-\subsection{ABC with Nested Sampling}
+\subsection{ABC with nested sampling}
 NS can be used for ABC by noting that Equation~\ref{eqn:abc_posterior}

### 2016-08-11 paper/ms.tex
More JSS stuff

@@
 
-\section{Running DNest4}\label{sec:running}
+\section[Running DNest4]{Running \pkg{DNest4}}\label{sec:running}
 To demonstrate \pkg{DNest4},
@@
 These are all described (using the linear regression example) in the
-following subsections.
+following sections.
 
@@
 
-\section{Finite mixture models with the RJObject class}
+\section[Finite mixture models with the RJObject class]{Finite mixture models with the \code{RJObject} class}
 Mixture models are a useful way of representing realistic prior information
@@
 is determined by the conditional prior class, and is
-explained in the following subsection.
+explained in the following section.
 
@@
 
-\section{Python bindings}\label{sec:python_bindings}
+\section[Python bindings]{\proglang{Python} bindings}\label{sec:python_bindings}
 In \pkg{DNest4}, it is also possible to specify and run models in
@@
 
-\section{Specifying models in Julia}
+\section[Specifying models in Julia]{Specifying models in \proglang{Julia}}
 Experimental support for model specification in the

### 2016-09-05 python/dnest4/builder.py
Sometimes take bolder steps

@@ -48,6 +48,11 @@ class Model:
         s = ""
         s += "double logH = 0.0;\n\n"
         s += "int which;\n"
+        s += "int reps = 1;\n"
+        s += "if(rng.rand() <= 0.5)\n"
+        s += "    reps = (int)pow(10.0, 2*rng.rand());\n"
+        s += "for(int i=0; i<reps; ++i)\n"
+        s += "{\n"
         s += "which = rng.rand_int(" + str(self.num_params) + ");\n"
 
         k = 0
@@ -59,7 +64,7 @@ class Model:
                                 .format(x=node.name);
                 k += 1
 
-        s += "\n\n"
+        s += "}\n\n"
         for node in self.nodes:
             if node.observed == False:
                 s += "" + node.distribution.from_uniform()\

### 2016-09-05 paper/ms.tex
Version bump

@@
 \begin{CodeInput}
-> wget https://github.com/eggplantbren/DNest4/archive/0.1.2.tar.gz
-> tar xvzf 0.1.2.tar.gz
-> mv DNest4-0.1.2 DNest4
+> wget https://github.com/eggplantbren/DNest4/archive/0.1.3.tar.gz
+> tar xvzf 0.1.3.tar.gz
+> mv DNest4-0.1.3 DNest4
 > cd DNest4/code

### 2016-09-05 python/dnest4/__init__.py

@@ -1,6 +1,6 @@
 # -*- coding: utf-8 -*-
 
-__version__ = "0.1.2"
+__version__ = "0.1.3"
 
 try:
     __DNEST4_SETUP__

### 2016-09-05 paper/ms.tex
Some = signs should have been \sims
murray

@@
 \mu &\sim \textnormal{Uniform}(-10, 10)\\
-\ln\sigma &= \textnormal{Uniform}(-10, 10).
+\ln\sigma &\sim \textnormal{Uniform}(-10, 10).
 \end{align}
@@
 \mu &\sim \textnormal{Uniform}(-10, 10)\\
-\ln\sigma &= \textnormal{Uniform}(-10, 10)\\
-n_i &= \textnormal{Normal}(0, 1)
+\ln\sigma &\sim \textnormal{Uniform}(-10, 10)\\
+n_i &\sim \textnormal{Normal}(0, 1)
 \end{align}
@@
 \begin{align}
-x_i &= \mu + \sigma n_i.
+x_i &:= \mu + \sigma n_i.
 \end{align}
@@
 It is a pleasure to thank Anna Pancoast (Harvard),
-Ewan Cameron (Oxford), David Hogg (NYU), and Daniela Huppenkothen (NYU)
-for valuable discussions. BJB was supported by a Marsden Fast-Start grant
+Ewan Cameron (Oxford), David Hogg (NYU), Daniela Huppenkothen (NYU),
+and Iain Murray (Edinburgh)
+for valuable discussions. This work was supported by a Marsden Fast-Start grant
 from the Royal Society of New Zealand.

### 2016-09-05 python/dnest4/classic.py
Don't need pandas

@@ -1,6 +1,5 @@
 import copy
 import numpy as np
-import pandas as pd
 import matplotlib.pyplot as plt
 from .loading import *
 

### 2016-09-07 code/RNG.cpp
A constructor that sets the seed

@@ -12,6 +12,13 @@ RNG::RNG()
 
 }
 
+RNG::RNG(unsigned int seed)
+:uniform(0., 1.)
+,normal(0., 1.)
+{
+	set_seed(seed);
+}
+
 void RNG::set_seed(unsigned int seed)
 {
 	twister.seed(seed);

### 2016-09-07 code/RNG.h

@@ -25,6 +25,9 @@ class RNG
 		// Constructor
 		RNG();
 
+		// Constructor that also sets seed
+		RNG(unsigned int seed);
+
 		// Set the seed (obviously)
 		void set_seed(unsigned int seed);
 

### 2016-09-13 paper/ms.tex
Mac OS X thing
Correction

@@
 > python setup.py install
+\end{CodeInput}
+\end{CodeChunk}
+
+In Mac OS X, the final line might need to provide information about
+your OS version, like so:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+> MACOSX_DEPLOYMENT_TARGET=10.9 python setup.py install
 \end{CodeInput}
@@
 called \code{postprocess_abc}.
-This generates posterior samples using the $\epsilon$ value that
-compresses the joint prior by 80\% as much as the final level.
+This generates posterior samples using the $\epsilon$ value corresponding
+to level $\approx$ 3/4$\times$ (maximum number of levels).
 \begin{CodeChunk}

### 2016-09-14 paper/ms.tex
Fixed borked sentence

@@
 $D = D_{\rm observed}$ could have been either
-true or false in the prior state of knowledge, but it is known to be true
+true or false, but it is known to be true
 in the posterior state of knowledge.

### 2016-10-17 paper/replicate.py
Use existing code (don't clone again)

@@ -1,25 +1,26 @@
 import os
+import matplotlib.pyplot
 
 # Best Python command on your system
-my_python = "python"
+my_python = "/home/brendon/local/anaconda3/bin/python"
 
-print("This script should download and install DNest4 and \
+print("This script should\
 replicate all the runs presented in the paper.\nNote:\
 plots will be generated, which need to be manually\n\
 closed for the script to continue. It assumes\n\
-everything has been compiled already.\n\
+everything has been compiled already as per the instructions in the paper.\n\
 \n\
 NOTE: the RJObject_1DMixture example is fairly time-consuming.\n\n\
 Press ENTER to begin.")
 junk = input()
 
-os.system("wget https://github.com/eggplantbren/DNest4/archive/0.1.3.tar.gz")
-os.system("tar xvzf 0.1.3.tar.gz")
-os.system("mv DNest4-0.1.3 DNest4")
-os.chdir("DNest4/code")
-os.system("make")
-os.chdir("../python")
-os.system("python setup.py install")
+#os.system("wget https://github.com/eggplantbren/DNest4/archive/0.1.3.tar.gz")
+#os.system("tar xvzf 0.1.3.tar.gz")
+#os.system("mv DNest4-0.1.3 DNest4")
+#os.chdir("DNest4/code")
+#os.system("make")
+#os.chdir("../python")
+#os.system("python setup.py install")
 
 def run_example(directory):
     os.chdir(directory)

### 2016-10-17 paper/ms.tex
Re-running

@@
 
-In Mac OS X, the final line might need to provide information about
-your OS version, like so:
-
+In Mac OS X, the final line (which installs the \proglang{Python} parts)
+of \pkg{DNest4})
+needs to provide information about
+your OS version. For example, if your computer runs Mac OS X 10.9,
+the installation command for the \proglang{Python} package is
 \begin{CodeChunk}
@@
 should contain information about levels and ``saving particles to disk''.
-After 5000 particles have been saved to the disk, the run will terminate.
+After 10,000 particles have been saved to the disk, the run will terminate.
 
@@
 \begin{CodeOutput}
-log(Z) = -175.545067073
-Information = 15.2168377939 nats.
-Effective sample size = 658.958278958
+log(Z) = -175.512920079
+Information = 15.1683473499 nats.
+Effective sample size = 1521.58236273
 \end{CodeOutput}
@@
 100     # Equal weight enforcement. Beta in the paper
-5000    # Maximum number of saves (0 ==> run forever)
+10000   # Maximum number of saves (0 ==> run forever)
 \end{CodeInput}

### 2016-10-17 paper/ms.tex
Removed formatting from within captions as per JSS request

@@
 \includegraphics[width=8cm]{figures/fig2.pdf}
-\caption{{\bf Top panel}: The estimated compression factor between subsequent
-levels, expressed as $\ln(X_{i+1}/X_{i})$. {\bf Bottom panel}:
+\caption{Top panel: The estimated compression factor between subsequent
+levels, expressed as $\ln(X_{i+1}/X_{i})$. Bottom panel:
 The Metropolis acceptance fraction as a function of level.
@@
 \includegraphics[width=7cm]{figures/fig3.pdf}
-\caption{{\bf Top panel}: The log-likelihood curve, showing the relationship
+\caption{Top panel: The log-likelihood curve, showing the relationship
 between log-likelihood and the enclosed prior mass.
-{\bf Bottom panel}: Posterior weights of the saved particles.
+Bottom panel: Posterior weights of the saved particles.
 For a successful run, there should be a clear peak, and saved particles

### 2016-10-18 code/Examples/RJObject_1DMixture/display.py
New figures

@@ -17,7 +17,10 @@ def mixture(x, params):
 
     y = zeros(x.shape)
     for i in range(0, N):
-        y += weights[i]/widths[i]/sqrt(2.*pi)*exp(-0.5*(x - centers[i])**2/widths[i]**2)
+        # Don't plot flukey narrow things (which ought to eventually average
+        # out, but won't in a finite sample)
+        if widths[i] >= 0.1:
+            y += weights[i]/widths[i]/sqrt(2.*pi)*exp(-0.5*(x - centers[i])**2/widths[i]**2)
 
     return y
 

### 2016-10-18 paper/replicate.py

@@ -2,7 +2,8 @@ import os
 import matplotlib.pyplot
 
 # Best Python command on your system
-my_python = "/home/brendon/local/anaconda3/bin/python"
+# (should also be the one you installed the dnest4 package for)
+my_python = "python"
 
 print("This script should\
 replicate all the runs presented in the paper.\nNote:\
@@ -17,7 +18,7 @@ junk = input()
 def run_example(directory):
     os.chdir(directory)
     os.system("./main -s 0")
-    if directory == "StraigtLine":
+    if directory == "StraightLine":
         print("Figure 1 in the paper has been zoomed in.")
     os.system(my_python + " showresults.py")
     os.chdir("..")

### 2016-10-18 paper/ms.tex
Updated a number
semicolons in itemize lists

@@
 \begin{enumerate}
-\item Generate $a \sim$ Normal$(0, 1)$
-\item Generate $b \sim$ Uniform$(0, 1)$
-\item Define $t := a/\sqrt{-\ln(b)}$
-\item Generate $n \sim$ Normal$(0, 1)$
+\item Generate $a \sim$ Normal$(0, 1)$;
+\item Generate $b \sim$ Uniform$(0, 1)$;
+\item Define $t := a/\sqrt{-\ln(b)}$;
+\item Generate $n \sim$ Normal$(0, 1)$;
 \item Set $x := 10^{1.5 - 3|t|}n$
@@
 $N$ is shown in Figure~\ref{fig:galaxies_N}.
-The marginal likelihood was $\ln(Z) = -242.2$, and
-the information (measured in nats) was $\mathcal{H} = 32.3$.
+The marginal likelihood was $\ln(Z) = -241.7$, and
+the information (measured in nats) was $\mathcal{H} = 32.0$.
 
@@
 \item The choice of discrepancy function $\rho(D; D_{\rm observed})$, which
-may involve a choice of summary statistics,
+may involve a choice of summary statistics;
 \item How to choose the value of $\epsilon$, or how to change it as a run
-progresses,
+progresses; and
 \item How to make algorithms more efficient than rejection sampling.

### 2016-10-23 paper/ms.tex
Reran RJObject_1DMixture because of previous bug in Laplace
Figure with the high posterior weight points

@@
 $N$ is shown in Figure~\ref{fig:galaxies_N}.
-The marginal likelihood was $\ln(Z) = -241.7$, and
-the information (measured in nats) was $\mathcal{H} = 32.0$.
+The marginal likelihood was $\ln(Z) = -232.2$, and
+the information (measured in nats) was $\mathcal{H} = 29.5$.
 
@@
 \citep{skilling2006nested} for which Nested Sampling is the best known
-solution.
+solution. There are some points with high posterior weight at the left
+of the plot, but there are so few of them that they make up a trivial
+fraction of the posterior mass. In practice, however, it is worth
+re-running with more levels to verify that a second peak is not forming.
 \label{fig:galaxies_fig3}}

### 2016-12-08 paper/ms.tex
url

@@
 \begin{center}
-{\tt https://github.com/eggplantbren/DNest4/}\\
+\url{https://github.com/eggplantbren/DNest4/}
 \end{center}
@@
 \begin{CodeInput}
-> wget https://github.com/eggplantbren/DNest4/archive/0.1.3.tar.gz
-> tar xvzf 0.1.3.tar.gz
-> mv DNest4-0.1.3 DNest4
+> wget https://github.com/eggplantbren/DNest4/archive/0.1.4.tar.gz
+> tar xvzf 0.1.4.tar.gz
+> mv DNest4-0.1.4 DNest4
 > cd DNest4/code
@@
 template class expresses.
-Figure created using Daft ({\tt http://daft-pgm.org}).\label{fig:rjobject_pgm}}
+Figure created using Daft (\url{http://daft-pgm.org}).\label{fig:rjobject_pgm}}
 \end{center}

### 2017-02-10 paper/ms.tex
Version bump

@@
 \begin{CodeInput}
-> wget https://github.com/eggplantbren/DNest4/archive/0.1.4.tar.gz
-> tar xvzf 0.1.4.tar.gz
-> mv DNest4-0.1.4 DNest4
+> wget https://github.com/eggplantbren/DNest4/archive/0.1.5.tar.gz
+> tar xvzf 0.1.5.tar.gz
+> mv DNest4-0.1.5 DNest4
 > cd DNest4/code

### 2017-02-10 python/dnest4/__init__.py

@@ -1,6 +1,6 @@
 # -*- coding: utf-8 -*-
 
-__version__ = "0.1.4"
+__version__ = "0.1.5"
 
 try:
     __DNEST4_SETUP__

### 2018-08-07 paper/ms.tex
Paper from JSS

@@
-\documentclass[article, nojss]{jss}
+\documentclass[article]{jss}
+\usepackage{thumbpdf,lmodern} 
+\graphicspath{{Figures/}}
+
 \usepackage{amsmath}
-\usepackage[utf8]{inputenc}
-\usepackage{graphicx}
-\usepackage{natbib}
 \usepackage{enumerate}
-\usepackage{microtype}
 \usepackage{dsfont}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% declarations for jss.cls %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%% almost as usual
 \author{Brendon J. Brewer\\Department of Statistics\\The University of Auckland\And 
@@
 
-%% for pretty printing and a nice hypersummary also set:
-\Plainauthor{Brendon J. Brewer, Daniel Foreman-Mackey} %% comma-separated
-\Plaintitle{DNest4: Diffusive Nested Sampling in
-C++ and Python} %% without formatting
-\Shorttitle{\pkg{DNest4}: Diffusive Nested Sampling} %% a short title (if necessary)
-
-%% an abstract and keywords
-\Abstract{In probabilistic (Bayesian) inferences, we typically want to compute
-properties of the posterior distribution, describing knowledge of
-unknown quantities in the context of a particular dataset and the assumed
-prior information. The marginal likelihood, also known as the ``evidence'',
-is a key quantity in Bayesian model selection. The Diffusive Nested Sampling algorithm, a variant of Nested Sampling, is a powerful tool for generating
-posterior samples and estimating marginal likelihoods. It is effective at
-solving complex problems including many where the posterior distribution is
+\Plainauthor{Brendon J. Brewer, Daniel Foreman-Mackey}
+
+\Plaintitle{DNest4: Diffusive Nested Sampling in C++ and Python}
+
+\Shorttitle{\pkg{DNest4}: Diffusive Nested Sampling}
+
+\Abstract{In probabilistic (Bayesian) inferences, we typically want to
+  compute properties of the posterior distribution, describing
+  knowledge of unknown quantities in the context of a particular
+  dataset and the assumed prior information. The marginal likelihood,
+  also known as the ``evidence'', is a key quantity in Bayesian model
+  selection. The diffusive nested sampling algorithm, a variant of
+  nested sampling, is a powerful tool for generating posterior samples
+  and estimating marginal likelihoods. It is effective at solving
+  complex problems including many where the posterior distribution is
   multimodal or has strong dependencies between variables.
-\pkg{DNest4} is an open source (MIT licensed),
-multi-threaded implementation of this algorithm in
-\proglang{C++11},
-along with associated utilities including:
-i)
-\pkg{RJObject}, a class template for finite mixture models;
-(ii)
-A \proglang{Python} package allowing basic use without \proglang{C++} coding; and
-iii)
-Experimental support for models implemented in \proglang{Julia}.
-In this paper we demonstrate \pkg{DNest4} usage through examples including
-simple Bayesian data analysis, finite mixture models, and Approximate
-Bayesian Computation.
-}
-
-\Keywords{bayesian inference, markov chain monte carlo,
-metropolis algorithm, bayesian computation, nested sampling, \proglang{c++11},
-\proglang{python}}
-\Plainkeywords{bayesian inference, markov chain monte carlo,
-metropolis algorithm, bayesian computation, nested sampling, c++11, python} %% without formatting
+  \pkg{DNest4} is an open source (MIT licensed), multi-threaded
+  implementation of this algorithm in \proglang{C++11}, along with
+  associated utilities including: i) `\code{RJObject}', a class template
+  for finite mixture models; (ii) a \proglang{Python} package allowing
+  basic use without \proglang{C++} coding; and iii) experimental
+  support for models implemented in \proglang{Julia}.  In this paper
+  we demonstrate \pkg{DNest4} usage through examples including simple
+  Bayesian data analysis, finite mixture models, and approximate
+  Bayesian computation.  }
+
+\Keywords{Bayesian inference, Markov chain Monte Carlo, Metropolis
+  algorithm, Bayesian computation, nested sampling, \proglang{C++11},
+  \proglang{Python}}
+
+\Plainkeywords{Bayesian inference, Markov chain Monte Carlo,
+Metropolis algorithm, bayesian computation, nested sampling, C++11, Python} %% without formatting
 %% at least one keyword must be supplied
@@
 
-In compact but standard notation, the posterior
-posterior distribution for parameters $\params$ given data $\data$, within
-the context of prior information $M$, is
-\begin{align}
+In compact but standard notation, the posterior distribution for
+parameters $\params$ given data $\data$, within the context of prior
+information $M$, is
+\begin{align*}
 p(\params | \data, M) &=
 \frac{p(\params | M)p(\data | \params, M)}{p(\data | M)}
-\end{align}
+\end{align*}
 or
-\begin{align}
+\begin{align*}
 \textnormal{posterior} &=
@@
      {\textnormal{marginal likelihood}}.
-\end{align}
+\end{align*}
 
 If prior information $I$ (dropped hereafter)
-implies a set of possible `models' $\{M_i\}$,
+implies a set of possible ``models'' $\{M_i\}$,
 rather than a single one $M$, the posterior model probabilities are given by
-\begin{align}
+\begin{align*}
 P(M_i | \data) &=
-\frac{P(M_i)p(\data | M_i)}{\sum_j P(M_j)p(\data | M_j)}
-\end{align}
+\frac{P(M_i)p(\data | M_i)}{\sum_j P(M_j)p(\data | M_j)},
+\end{align*}
 where
-\begin{align}
+\begin{align*}
 p(\data | M_j) &= \int p(\theta_j | M_j)p(\data | \theta_j, M_j) \, d\theta_j
-\end{align}
+\end{align*}
 is the marginal likelihood of model $j$, equal to the expected value of the
@@
 are often presented as ratios
-of marginal likelihoods, known as Bayes Factors.
+of marginal likelihoods, known as Bayes factors.
 When discussing computational matters, the prior distribution for parameters
@@
 
-Nested Sampling \citep[NS;][]{skilling2006nested} is a Monte Carlo method whose main
-aim is to calculate $Z$. However, it can also be used to generate samples
-to represent the posterior distribution $\pi(\theta)L(\theta)/Z$, or
-any other distribution proportional to
-$\pi(\theta)\Phi\left[L(\theta)\right]$ where $\Phi$ is any monotonic function.
-This latter property makes NS particularly useful for statistical mechanics
-calculations \citep{partay2010efficient, baldock2016determining},
-where the
-``canonical'' family of distributions proportional to
-$\pi(\theta)L(\theta)^\beta$ is of interest. In such applications,
-$L(\theta)$ is usually equivalent to
-$\exp(-\textnormal{energy})$. NS is particularly efficient for this, since
-only a single run (exploring regions of higher and higher $L$) is needed, and different canonical
+Nested sampling \citep[NS;][]{skilling2006nested} is a Monte Carlo
+method whose main aim is to calculate $Z$. However, it can also be
+used to generate samples to represent the posterior distribution
+$\pi(\theta)L(\theta)/Z$, or any other distribution proportional to
+$\pi(\theta)\Phi\left[L(\theta)\right]$ where $\Phi$ is any monotonic
+function.  This latter property makes nested sampling particularly useful for
+statistical mechanics calculations \citep{partay2010efficient,
+  baldock2016determining}, where the ``canonical'' family of
+distributions proportional to $\pi(\theta)L(\theta)^\beta$ is of
+interest. In such applications, $L(\theta)$ is usually equivalent to
+$\exp(-\textnormal{energy})$. Nested sampling is particularly
+efficient for this, since only a single run (exploring regions of
+higher and higher $L$) is needed, and different canonical
 distributions can be obtained by re-weighting the output points.
 
-A defining feature of NS is that it works with a sequence of
+A defining feature of nested sampling is that it works with a sequence of
 {\em constrained prior} distributions, proportional to $\pi$ but
@@
 p(\theta; \ell) &=
-\frac{\pi(\theta)\mathds{1}\left[L(\theta) > \ell\right]}{X(\ell)}
+\frac{\pi(\theta)\mathds{1}\left[L(\theta) > \ell\right]}{X(\ell)},
 \label{eqn:constrained_prior}
@@
 where
-\begin{align}
+\begin{align*}
 X(\ell) &= \int \pi(\theta) \mathds{1}\left[L(\theta) > \ell\right] \, d\theta
-\end{align}
+\end{align*}
 is the amount of prior mass which has likelihood greater than $\ell$, and
@@
 argument is true and 0 otherwise.
-In the standard NS framework, the sequence of $\ell$ values is selected
+In the standard nested sampling framework, the sequence of $\ell$ values is selected
 so that $X(\ell)$ shrinks by a factor $\approx e^{-1/N}$ per iteration, where
 $N$ is the number of particles used. This geometric compression of the
-parameter space is a defining feature of NS.
+parameter space is a defining feature of nested sampling.
 
 Sampling from the constrained priors (Equation~\ref{eqn:constrained_prior})
-is required, and Markov Chain Monte Carlo (MCMC) is a popular method of doing
+is required, and Markov chain Monte Carlo (MCMC) is a popular method of doing
 this, although alternatives exist
 \citep[e.g.,][]{feroz2009multinest, handley2015polychord}.
-Diffusive Nested Sampling \citep[DNS;][]{brewer2011diffusive}
-is an alternative to NS for
+Diffusive nested sampling \citep[DNS;][]{brewer2011diffusive}
+is an alternative to nested sampling for
 problems where MCMC is the only viable sampling method. DNS is based on the
@@
 \begin{align}
-p(\theta, j) &= p(j)p(\theta | j)\\
+p(\theta, j) &= p(j)p(\theta | j)\nonumber\\
 &= w_j \times
-\frac{\pi(\theta)\mathds{1}\left[L(\theta) > \ell_j\right]}{X(\ell_j)}.
+\frac{\pi(\theta)\mathds{1}\left[L(\theta) > \ell_j\right]}{X(\ell_j)},
 \label{eqn:target_distribution}
@@
 the particle(s) are initialized from the prior $\pi(\theta)$, equivalent
-to the mixture of constrained priors with a single level whose log likelihood
+to the mixture of constrained priors with a single level whose log-likelihood
 threshold is $-\infty$. The mixture of constrained priors evolves by adding new levels,
@@
 MCMC chain to mix between different modes in some circumstances. The marginal
-likelihood estimate can also be more accurate than standard MCMC-based NS
+likelihood estimate can also be more accurate than standard MCMC-based nested sampling
 \citep{brewer2011diffusive}, as less information is discarded.
@@
 
-Alternatively, the \proglang{Python}
+Alternatively, the \proglang{Python} \citep{python}
 bindings (Section~\ref{sec:python_bindings}) allow you
@@
 \section{Relation to other algorithms}\label{sec:relation}
-The Diffusive Nested Sampling algorithm, and its implementation in
+The diffusive nested sampling algorithm, and its implementation in
 \pkg{DNest4}, has advantages and disadvantages compared to other Bayesian
@@
 
-In the Nested Sampling sphere,
+In the nested sampling sphere,
 the simple \proglang{C} implementation of given by
-\citet{skilling2006nested} is useful for understanding the classic NS
+\citet{skilling2006nested} is useful for understanding the classic nested sampling
 algorithm. Similarly to \pkg{DNest4}, it is the user's responsibility
 to implement the MCMC moves used. The DNS algorithm produces more
-accurate estimates of $Z$ and ought to outperform classic NS on multimodal
+accurate estimates of $Z$ and ought to outperform classic nested sampling on multimodal
 problems.
@@
 \proglang{Fortran}
-Nested Sampling implementation
+nested sampling implementation
 that dispenses with MCMC for generating new particles from constrained
@@
 
-Outside of Nested Sampling,
+Outside of nested sampling,
 the popular \pkg{JAGS} \citep{jags} and \pkg{Stan} \citep{stan} packages
@@
 density (up to a normalizing constant). This user-friendliness is a key
-advantage of \pkg{emcee}, and its algorithm performs well on low--moderate
+advantage of \pkg{emcee}, and its algorithm performs well on low to moderate
 dimensional (up to $\sim 50$) parameter spaces with a unimodel
@@
 \citep{neal2001annealed} are
-related to Nested Sampling and are useful on multimodal posterior distributions.
-However, they require much more tuning than NS (in the form of an
+related to nested sampling and are useful on multimodal posterior distributions.
+However, they require much more tuning than nested sampling (in the form of an
 {\em annealing schedule}) and do
-not work on phase change problems, unlike Nested Sampling
+not work on phase change problems, unlike nested sampling
 \citep{skilling2006nested}.
@@
 
-%\begin{table}
-%
-%  \begin{tabular}{cc}
-%    Knuth & Lamport
-%  \end{tabular}}
-%\end{table}
-
-\begin{table}
-\begin{center}
+\begin{table}[t!]
+\centering
 \resizebox{\textwidth}{!}{%
-\begin{tabular}{|l|c|c|c|c|c|c|}
+\begin{tabular}{lcccccc}
 \hline
@@
 Bayesian computation software packages.\label{tab:packages}}
-\end{center}
 \end{table}
-
-\section{Markov chain monte carlo}\label{sec:mcmc}
+\newpage
+\section{Markov chain Monte Carlo}\label{sec:mcmc}
 DNS is build upon the Metropolis-Hastings algorithm.
@@
 is given by
-\begin{align}
+\begin{align*}
 \alpha &= \min\left(1,
@@
 \frac{\pi(\params')}{\pi(\params)}\frac{L(\params')}{L(\params)}
-\right)
-\end{align}
+\right),
+\end{align*}
 where $q(\theta' | \theta)$ is the proposal distribution used to generate
 a new position $\theta'$ from the current position $\theta$. Often,
-$q$ is {\em symmetric} so the $q$ terms cancel in the acceptance probability.
+$q$ is {\em symmetric} so that the $q$ terms cancel in the acceptance probability.
 
@@
 \mathds{1}\left(L(\params') > \ell_j\right)
-\right]
+\right],
 \label{eqn:log_hastings}
@@
 the likelihood-related term, and write the former as
-\begin{align}
+\begin{align*}
 H = \frac{q(\params'|\params)}{q(\params | \params')}
 \times \frac{\pi(\params')}{\pi(\params)}.
-\end{align}
+\end{align*}
 
 The logarithm of $H$,
-\begin{align}
+\begin{align*}
 \ln(H) = \ln\left[\frac{q(\params'|\params)}{q(\params | \params')}
 \times \frac{\pi(\params')}{\pi(\params)}\right],
-\end{align}
+\end{align*}
 is the user's responsibility when implementing models,
@@
 
-%The \pkg{DNest4} \code{Sampler} class handles the likelihood check,
-%while the \code{perturb(DNest4::RNG\&)} member function handles the
-%rest of the terms. The value returned from \code{perturb(DNest4::RNG\&)}
-%must be
-%for your problem.
-
-%\subsection{Diffusive Nested Sampling}
-
 \section{Dependencies and installation}\label{sec:installation}
@@
 
-Development of \pkg{DNest4} takes place in the \pkg{git} repository located at
-\begin{center}
-\url{https://github.com/eggplantbren/DNest4/}
-\end{center}
+Development of \pkg{DNest4} takes place in the git repository located at
+\url{https://github.com/eggplantbren/DNest4/}.
 The software
@@
 the following steps are sufficient:
-\begin{CodeChunk}
-\begin{CodeInput}
-> wget https://github.com/eggplantbren/DNest4/archive/0.1.5.tar.gz
-> tar xvzf 0.1.5.tar.gz
-> mv DNest4-0.1.5 DNest4
-> cd DNest4/code
-> make
-> cd ../python
-> python setup.py install
-\end{CodeInput}
-\end{CodeChunk}
-
-In Mac OS X, the final line (which installs the \proglang{Python} parts)
+%
+\begin{CodeChunk}
+\begin{CodeInput}
+$ wget https://github.com/eggplantbren/DNest4/archive/0.1.4.tar.gz
+$ tar xvzf 0.1.4.tar.gz
+$ mv DNest4-0.1.4 DNest4
+$ cd DNest4/code
+$ make
+$ cd ../python
+$ python setup.py install
+\end{CodeInput}
+\end{CodeChunk}
+%
+In Mac OS X, the final line (which installs the \proglang{Python} parts
 of \pkg{DNest4})
@@
 your OS version. For example, if your computer runs Mac OS X 10.9,
-the installation command for the \proglang{Python} package is
-\begin{CodeChunk}
-\begin{CodeInput}
-> MACOSX_DEPLOYMENT_TARGET=10.9 python setup.py install
-\end{CodeInput}
-\end{CodeChunk}
-
-We recommend you create an environment variable called \code{DNEST4\_PATH}
+the installation command for the \proglang{Python} package is:
+%
+\begin{CodeChunk}
+\begin{CodeInput}
+$ MACOSX_DEPLOYMENT_TARGET=10.9 python setup.py install
+\end{CodeInput}
+\end{CodeChunk}
+%
+We recommend to create an environment variable called \code{DNEST4\_PATH}
 and set it to the directory {\em above} the \pkg{DNest4} directory. Then,
@@
 sampling distribution is
-\begin{align}
+\begin{align*}
 y_i | m, b, \sigma &\sim \textnormal{Normal}(mx_i + b, \sigma^2)
-\end{align}
+\end{align*}
 and the priors are
-\begin{align}
-m &\sim \textnormal{Normal}(0, 1000^2)\\
-b &\sim \textnormal{Normal}(0, 1000^2)\\
+\begin{align*}
+m &\sim \textnormal{Normal}(0, 1000^2),\\
+b &\sim \textnormal{Normal}(0, 1000^2),\\
 \ln\sigma &\sim \textnormal{Uniform}(-10, 10).
-\end{align}
-These are naive diffuse priors and do not have any special status.
-The dataset is shown in Figure~\ref{fig:regression_lines}
-and the code is included in the {\tt code/Examples/StraightLine}
+\end{align*}
+These are na{\"i}ve diffuse priors and do not have any special status.
+The dataset is shown in Figure~\ref{fig:fig3} on the right
+and the code is included in the \code{code/Examples/StraightLine}
 subdirectory, and a slightly simplified version of the
@@
 To execute \pkg{DNest4} on this problem, go to this directory and
-execute {\tt main}. The output to the screen
+execute \code{main}. The output to the screen
 should contain information about levels and ``saving particles to disk''.
@@
 \section{Output files}
-The executable {\tt main} is responsible for the exploration part of the
+The executable \code{main} is responsible for the exploration part of the
 DNS algorithm (i.e., running the MCMC chain, building
 levels, and then exploring all the levels). It creates three text output files,
-{\tt sample.txt}, {\tt sample\_info.txt}, and {\tt levels.txt}.
+\code{sample.txt}, \code{sample\_info.txt}, and \code{levels.txt}.
 
 The first output
-file, {\tt sample.txt}, contains a sampling of parameter values that
+file, \code{sample.txt}, contains a sampling of parameter values that
 represents the {\it mixture of constrained priors} (the target distribution
-used in DNS), {\bf not} the
-posterior distribution. Each line of {\tt sample.txt} represents a point in
+used in DNS), \emph{not} the
+posterior distribution. Each line of \code{sample.txt} represents a point in
 parameter space. In the linear example, there are three parameters
-($m$, $b$, and $\sigma$), so there
-are three columns in {\tt sample.txt}.
-Each time a point is saved to {\tt sample.txt}, \pkg{DNest4} prints
+($m$, $b$, and $\sigma$); so there
+are three columns in \code{sample.txt}.
+Each time a point is saved to \code{sample.txt}, \pkg{DNest4} prints
 the message ``Saving a particle to disk. N = ...''.
 
-The second output file, {\tt sample\_info.txt}, should have the same number of
-rows as {\tt sample.txt}, because it contains metadata about the samples in
-{\tt sample.txt}. The first
+The second output file, \code{sample\_info.txt}, should have the same number of
+rows as the file \code{sample.txt}, because it contains metadata about the samples in
+\code{sample.txt}. The first
 column is the index $j$, which tells us which ``level'' the particle was in
@@
 The second column is the log-likelihood value, and the third column is
-the likelihood ``tiebreaker'', which allows Nested Sampling to work when
+the likelihood ``tiebreaker'', which allows nested sampling to work when
 there is a region in parameter space with nonzero prior probability where the
@@
 
-The third output file, {\tt levels.txt}, contains information about the levels
+The third output file, \code{levels.txt}, contains information about the levels
 that were built during the run. The first column has estimates of the $\log(X)$
@@
 
-The second column contains the log likelihoods of the levels.
-The first level, with a $\log(X)$ value of 0 and a log likelihood of
+The second column contains the log-likelihoods of the levels.
+The first level, with a $\log(X)$ value of 0 and a log-likelihood of
 $-10^{308}$ (basically ``minus infinity''), is simply the prior. The third
@@
 the estimates of the level compressions (and hence the $\log(X)$ values of
-the levels in column 1), as discussed in Section 3 of
+the levels in column 1), as discussed in Section~3 of
 \citet{brewer2011diffusive}.
@@
 The goal of running
-NS is usually to obtain posterior samples and the marginal likelihood $Z$,
-whereas {\tt sample.txt} only contains samples from the mixture of constrained
+nested sampling is usually to obtain posterior samples and the marginal likelihood $Z$,
+whereas \code{sample.txt} only contains samples from the mixture of constrained
 priors. Additional
@@
 \proglang{Python} function\footnote{Alternatively, the file
-{\tt showresults.py} in the example directory runs this function, and then
-calls the code in {\tt display.py} to create a further plot shown in
-Figure~\ref{fig:regression_lines}.}:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-import dnest4
-dnest4.postprocess()
-\end{CodeInput}
-\end{CodeChunk}
-This produces the three diagnostic plots in Figures~\ref{fig:fig1},
-\ref{fig:fig2}, and~\ref{fig:fig3}, along with the following output:
+\code{showresults.py} in the example directory runs this function, and then
+calls the code in \code{display.py} to create a further plot shown in
+Figure~\ref{fig:fig3} on the right.}:
+%
+\begin{CodeChunk}
+\begin{CodeInput}
+>>> import dnest4
+>>> dnest4.postprocess()
+\end{CodeInput}
+\end{CodeChunk}
+%
+This produces the three diagnostic plots in Figures~\ref{fig:fig1}
+and~\ref{fig:fig3} on the left, along with the following output:
+%
 \begin{CodeChunk}
@@
 \end{CodeChunk}
+%
 These are the natural log of the marginal likelihood, the
 information
-\begin{align}
+\begin{align*}
 \mathcal{H} &= \int p(\theta|D, M)
 \ln\left[\frac{p(\theta | D, M)}{p(\theta | M)}\right] \, d\theta,
-\end{align}
+\end{align*}
 which quantifies the degree to which $D$ restricted the
@@
 The \code{postprocess} function also saves a file,
-{\tt posterior\_sample.txt}, containing posterior samples (one per row).
+\code{posterior\_sample.txt}, containing posterior samples (one per row).
 Unfortunately, it is harder to compute justified error bars on $\ln(Z)$
-in DNS than it is in standard NS.
+in DNS than it is in standard nested sampling.
 %The \code{postprocess} function does include some options
@@
 
-\begin{figure}[ht!]
-\begin{minipage}{7.5cm}
+\begin{figure}[t!]
 \centering
-\includegraphics[width=8cm]{figures/fig1.pdf}
-\caption{The level $j$ of the saved particles over time.
+\includegraphics[width=8cm]{fig1.pdf}%
+\includegraphics[width=8cm]{fig2.pdf}
+\caption{Left: The level $j$ of the saved particles over time.
 Typically, this will trend upwards until all the levels have
 been created, and then diffuse evenly throughout all the levels.
+Right, top panel: The estimated compression factor between subsequent
+levels, expressed as $\ln(X_{i+1}/X_{i})$. Right, bottom panel:
+The Metropolis acceptance fraction as a function of level.
 \label{fig:fig1}}
-\end{minipage}\hspace{0.5cm}
-\begin{minipage}{7.5cm}
+\end{figure}
+
+\begin{figure}[t!]
   \centering
-\includegraphics[width=8cm]{figures/fig2.pdf}
-\caption{Top panel: The estimated compression factor between subsequent
-levels, expressed as $\ln(X_{i+1}/X_{i})$. Bottom panel:
-The Metropolis acceptance fraction as a function of level.
-\label{fig:fig2}}
-\end{minipage}
-\end{figure}
-
-\begin{figure}[ht!]
-\begin{minipage}{7.5cm}
-\includegraphics[width=7cm]{figures/fig3.pdf}
-\caption{Top panel: The log-likelihood curve, showing the relationship
+\includegraphics[width=7cm]{fig3.pdf}
+\includegraphics[width=7cm]{regression_lines.pdf}
+\caption{Left, top panel: The log-likelihood curve, showing the relationship
 between log-likelihood and the enclosed prior mass.
-Bottom panel: Posterior weights of the saved particles.
+Left, bottom panel: Posterior weights of the saved particles.
 For a successful run, there should be a clear peak, and saved particles
@@
 to those in the peak.
+Right: Regression lines drawn from the posterior.
 \label{fig:fig3}}
-\end{minipage}\hspace{0.5cm}
-\begin{minipage}{7.5cm}
-\includegraphics[width=7cm]{figures/regression_lines.pdf}
-\caption{Regression lines drawn from the posterior.
-\label{fig:regression_lines}}
-\end{minipage}
 \end{figure}
@@
 \subsection{Options}\label{sec:options}
-A plain-text file called {\tt OPTIONS} resides in the directory from which you
+A plain-text file called \code{OPTIONS} resides in the directory from which you
 execute a run. This file contains numerical parameters controlling the
-DNS algorithm. Here are the contents of {\tt OPTIONS} for the linear
+DNS algorithm. Here are the contents of \code{OPTIONS} for the linear
 regression example:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-# File containing parameters for DNest4
+%
+\begin{CodeChunk}
+\begin{CodeInput}
+# File containing parameters for DNest4.
 # Put comments at the top, or at the end of the line.
 5       # Number of particles
-10000   # new level interval
-10000   # save interval
-100     # threadSteps - pooling interval
-0       # maximum number of levels (0 ==> automatic)
-10      # Backtracking scale length (lambda in the paper)
-100     # Equal weight enforcement. Beta in the paper
+10000   # New level interval
+10000   # Save interval
+100     # Thread steps - pooling interval
+0       # Maximum number of levels (0 ==> automatic)
+10      # Backtracking scale length lambda in the paper.
+100     # Equal weight enforcement. Beta in the paper.
 10000   # Maximum number of saves (0 ==> run forever)
@@
 \end{CodeChunk}
-
+%
 Additional options are available on the command line. These are
@@
 in
-big output files if your model prints a lot of parameters to {\tt sample.txt}.
+big output files if your model prints a lot of parameters to \code{sample.txt}
 causing the postprocessing to take a long time and/or a lot of RAM.
@@
 compressed. An appropriate value for this quantity depends on the specific
-model and dataset at hand --- typically, a larger numbers of parameters
+model and dataset at hand -- typically, a larger numbers of parameters
 and larger (more informative) datasets will lead to a larger value of
@@
 This controls the length of a \pkg{DNest4} run, in units of saved particles
-in {\tt sample.txt}, which represent the mixture of constrained priors.
+in \code{sample.txt}, which represent the mixture of constrained priors.
 The number of posterior samples is always less than this. In most applications
@@
 For the simple linear regression example, we will call the class
-{\tt StraightLine}. The member variables representing the
-unknown parameters are defined in the header file {\tt StraightLine.h}:
-
+`\code{StraightLine}'. The member variables representing the
+unknown parameters are defined in the header file \code{StraightLine.h}:
+%
 \begin{CodeChunk}
@@
 \end{CodeChunk}
-
+%
 The class must also define and implement the following member functions:
@@
 prior is:
+%
 \begin{CodeChunk}
@@
    // Log-uniform prior
-   sigma = exp( - 10.0 + 20.0 * rng.rand());
-}
-\end{CodeInput}
-\end{CodeChunk}
-
+   sigma = exp(-10.0 + 20.0 * rng.rand());
+}
+\end{CodeInput}
+\end{CodeChunk}
+%
 This generates $m$, $b$, and $\sigma$ from their joint prior. In this case,
@@
 each from their own prior distribution. For convenience,
-\pkg{DNest4} provides an \code{RNG} class to represent random number
-generators. The \code{RNG} class is just a convenience wrapper for the
+\pkg{DNest4} provides an `\code{RNG}' class to represent random number
+generators. The `\code{RNG}' class is just a convenience wrapper for the
 random number generators built into \proglang{C++11}.
@@
 with the prior distribution.
+%
 \begin{CodeChunk}
@@
 \end{CodeChunk}
-
+%
 This function first chooses a random integer from $\{0, 1, 2\}$ using the
-\code{rand\_int(int)} member function of the \code{DNest4::RNG} class.
+\code{rand\_int(int)} member function of the `\code{DNest4::RNG}' class.
 This determines which of the three parameters
@@
 value using the line
-
+%
 \begin{CodeChunk}
@@
 \end{CodeChunk}
-
-A challenge using MCMC for standard Nested Sampling is that the target
-distribution is not static --- it gets compressed over time. Similarly, in
+%
+A challenge using MCMC for standard nested sampling is that the target
+distribution is not static -- it gets compressed over time. Similarly, in
 the first stage of DNS the target distribution gets compressed (as levels are
@@
 Overall, this procedure generates values $x$
-with a maximum scale of tens--hundreds, down to
+with a maximum scale of tens to hundreds, down to
 a minimum scale of about $10^{-30}$ with 99\% probability, by virtue
 of the $t$-distribution's heavy tails.
-For convenience, the \code{RNG} class contains a member function \code{randh()}
+For convenience, the `\code{RNG}' class contains a member function \code{randh()}
 to generate values from this distribution.
@@
 
-Recall that for Nested Sampling, the Metropolis acceptance probability,
+Recall that for nested sampling, the Metropolis acceptance probability,
 excluding the term for the likelihood, is
-\begin{align}
+\begin{align*}
 H = \frac{q(\params'|\params)}{q(\params | \params')}
 \times \frac{\pi(\params')}{\pi(\params)}.
-\end{align}
+\end{align*}
 When implementing a model class, the
@@
 mean zero and standard deviation $1000$, $H$ is
-\begin{align}
+\begin{align*}
 H &= \frac{\exp\left[-\frac{1}{2}(m'/1000)^2\right]}
 {\exp\left[-\frac{1}{2}(m/1000)^2\right]}.
-\end{align}
+\end{align*}
 This explains the use of the \code{log\_H} variable.
@@
 with respect to a uniform distribution), and then $\sigma$ is exponentiated
-again. The step for the uniform prior between -10 and +10
+again. The step for the uniform prior between $-10$ and +10
 uses the following code:
+%
 \begin{CodeChunk}
@@
 \end{CodeChunk}
+%
 The factor of 20 accounts for the prior width, and
@@
 return value instead of using the log/exp trick.
-However, this isn't recommended for a prior distribution like this
+However, this is not recommended for a prior distribution like this
 which covers several orders of magnitude, because the appropriate scale size
@@
 (by setting the maximum number of levels to 1) and inspect
-{\tt sample.txt} to ensure that each parameter is exploring
+\code{sample.txt} to ensure that each parameter is exploring
 the prior correctly.
 
-\subsection{Log likelihood}
+\subsection{Log-likelihood}
 The log-likelihood for the model is evaluated and returned by the
 \code{log\_likelihood} member function. For the straight line
-fitting example, the log likelihood is based on the
+fitting example, the log-likelihood is based on the
 normal density, and the code is given below.
+%
 \begin{CodeChunk}
@@
 
-    // Conventional gaussian sampling distribution
+    // Conventional Gaussian sampling distribution
     double log_L = 0.0;
@@
 \end{CodeChunk}
+%
 The dataset is assumed to be accessible inside this function. In the
 regression example, this is achieved by having a
-\code{Data} class to represent
+`\code{Data}' class to represent
 datasets. Since there will usually only be one dataset,
@@
 accessible from anywhere) is recommended.
-The \code{Data} class has
-one static member which is itself an object of class \code{Data}, and is
-accessible using \code{Data::get_instance()} --- this is the singleton pattern,
-essentially a way of defining quasi-`global' variables:
-
+The `\code{Data}' class has
+one static member which is itself an object of class `\code{Data}', and is
+accessible using \code{Data::get_instance()} -- this is the singleton pattern,
+essentially a way of defining quasi-``global'' variables:
+%
 \begin{CodeChunk}
@@
 \end{CodeChunk}
-
+%
 Alternatively, the data could be defined using static members of your
@@
 The print and description functions are very simple:
+%
 \begin{CodeChunk}
@@
 \end{CodeChunk}
+%
 The print function specifies that the parameters
-$m$, $b$, and $\sigma$ are printed a single line
-(of {\tt sample.txt}) and are separated by spaces (the
-\code{postprocess} function assumes the delimeter is a space).
+$m$, $b$, and $\sigma$ are printed in a single line
+(of \code{sample.txt}) and are separated by spaces (the
+\code{postprocess} function assumes the delimiter is a space).
 
 \subsection{Running the sampler}
-The file {\tt main.cpp} contains the \code{main()} function which
-is executed after compiling and linking. The contents of {\tt main.cpp}
+The file \code{main.cpp} contains the \code{main()} function which
+is executed after compiling and linking. The contents of \code{main.cpp}
 for the linear regression example are:
-
+%
 \begin{CodeChunk}
@@
 \end{CodeChunk}
-
-The first line of {\tt main()} loads the data into its global instance
+%
+The first line of \code{main()} loads the data into its global instance
 (so it can be accessed from within the log-likelihood function)
@@
 
-\section[Finite mixture models with the RJObject class]{Finite mixture models with the \code{RJObject} class}
+\section[Finite mixture models with the `RJObject' class]{Finite mixture models with the `\code{RJObject}' class}
 Mixture models are a useful way of representing realistic prior information
@@
 implement mixture models in DNS, \citet{brewer2014inference} implemented
-a template class called \code{RJObject} to handle the MCMC moves
-required. The RJ in \code{RJObject} stands for Reversible Jump
+a template class called `\code{RJObject}' to handle the MCMC moves
+required. The RJ in `\code{RJObject}' stands for reversible jump
 \citep{green1995reversible}, as
-\code{RJObject} implements birth and death moves for mixture
+`\code{RJObject}' implements birth and death moves for mixture
 models with an unknown number of components. An updated version of
-\code{RJObject} is included in \pkg{DNest4}.
+`\code{RJObject}' is included in \pkg{DNest4}.
 
@@
 can be factorized via the product rule, giving
-\begin{align}
+\begin{align*}
 p\left(N, \alpha, \left\{x_i\right\}_{i=1}^N\right)
 &= p(N)p(\alpha | N)p\left(\left\{x_i\right\}_{i=1}^N | \alpha, N\right).
-\end{align}
-The specific assumptions of \pkg{RJObject} are that this simplifies to
+\end{align*}
+The specific assumptions of `\code{RJObject}' are that this simplifies to
 \begin{align}
@@
 distributed given the hyperparameters. The sampling distribution
-$p(D | x, \alpha)$ (not shown in the PGM)
+$p(D | x, \alpha)$ (not shown in the probabilistic graphical model)
 cannot depend on the ordering of the components.
-A probabilistic graphical model (PGM) is showing this dependence
+A probabilistic graphical model (PGM) showing this dependence
 structure is shown in Figure~\ref{fig:rjobject_pgm}. There are no observed
 data nodes here, but if such a structure forms part of a
-Bayesian model, an \code{RJObject} object within your model class
+Bayesian model, an `\code{RJObject}' object within your model class
 can encapsulate this part of the model.
 
-\begin{figure}[ht!]
-\begin{center}
-\hspace*{-2cm}\includegraphics[width=0.4\textwidth]{figures/rjobject_pgm.pdf}
-\caption{A PGM showing the kind of prior information the \code{RJObject}
+\begin{figure}[t!]
+\centering
+\includegraphics[width=0.4\textwidth]{rjobject_pgm.pdf}
+\caption{A PGM showing the kind of prior information the `\code{RJObject}'
 template class expresses.
 Figure created using Daft (\url{http://daft-pgm.org}).\label{fig:rjobject_pgm}}
-\end{center}
 \end{figure}
@@
 \citet{brewer2014inference}
-described the motivation for \code{RJObject} and some details
+described the motivation for `\code{RJObject}' and some details
 about the Metropolis proposals underlying it. Here, we demonstrate how to
-implement a finite mixture model using \code{RJObject}.
+implement a finite mixture model using `\code{RJObject}'.
 This example can be found in the
-{\tt code/Examples/RJObject\_1DMixture}
+\code{code/Examples/RJObject\_1DMixture}
 directory. The ``SineWave'' and ``GalaxyField'' models
@@
 Consider a sampling distribution
-for data $D=\{D_1, D_2, ..., D_n\}$ which is a mixture of $N$ gaussians
+for data $D=\{D_1, D_2, \ldots, D_n\}$ which is a mixture of $N$ Gaussians
 with means $\{\mu_j\}$, standard deviations $\{\sigma_j\}$, and
 mixture weights $\{w_j\}$. The likelihood for the $i$th data point is
-\begin{align}
+\begin{align*}
 p\left(D_i | N, \{\mu_j\}, \{\sigma_j\}, \{w_j\}\right) &=
@@
 \exp\left[-\frac{1}{2\sigma_j^2}\left(D_i - \mu_j\right)^2\right].
-\end{align}
-Colloquially, one might say the data were `drawn from' a mixture of
+\end{align*}
+Colloquially, one might say the data were ``drawn from'' a mixture of
 $N$ normal distributions, and we want to infer $N$ along with the
@@
 Making the connection with Equation~\ref{eqn:rjobject} and
-Figure~\ref{fig:rjobject_pgm}, the ``components'' are the gaussians,
-the gaussian parameters are $\{x_j\} = \{(\mu_j, \sigma_j, W_j)\}$,
+Figure~\ref{fig:rjobject_pgm}, the ``components'' are the Gaussians,
+the Gaussian parameters are $\{x_j\} = \{(\mu_j, \sigma_j, W_j)\}$,
 and hyperparameters $\alpha$ may be used to help specify
-a sensible joint prior for the gaussian parameters.
+a sensible joint prior for the Gaussian parameters.
 
@@
 The prior for $N$ was
-\begin{align}
+\begin{align*}
 p(N) \propto \frac{1}{N+1}
-\end{align}
-for $N \in \{1, 2, ..., 100\}$.
-For the conditional prior of the gaussian parameters,
+\end{align*}
+for $N \in \{1, 2, \ldots, 100\}$.
+For the conditional prior of the Gaussian parameters,
 we used Laplace (biexponential)
@@
 $b$ has density $p(x | a, b) = \frac{1}{2}\exp\left(-\frac{1}{b}|x - a|\right)$.}.
-Normal distributions would be more convenional, but the analytic cumulative
+Normal distributions would be more conventional, but the analytic cumulative
 distribution function (CDF) of the Laplace distribution makes it easier to
@@
 These priors express the idea that the centers, widths, and relative weights
-of the gaussian mixture components are probably clustered around some typical
+of the Gaussian mixture components are probably clustered around some typical
 value.
@@
 We used the following priors for the hyperparameters:
-\begin{align}
+\begin{align*}
 a_\mu &\sim \textnormal{Uniform}(-1000, 1000)\\
@@
 b_{\ln W} &\sim \textnormal{Uniform}(0, 5).
-\end{align}
+\end{align*}
 The normalized weights are $w_k = W_k/\sum w_j$.
 
-\subsection{Using the RJObject class}
+\subsection[Using the `RJObject' class]{Using the `\code{RJObject}' class}
 To implement this model for \pkg{DNest4}, the model
 class needs to contain an instance of an
-\code{RJObject}, which contains the parameters
+`\code{RJObject}', which contains the parameters
 $\{(\mu_j, \sigma_j, w_j)\}$:
+%
 \begin{CodeChunk}
@@
 \end{CodeChunk}
+%
 where the template argument \code{<MyConditionalPrior>}
@@
 $\alpha$ and the form of the conditional prior $p(x_i | \alpha)$.
-The main advantage of the \code{RJObject} class is that
+The main advantage of the `\code{RJObject}' class is that
 we do not need to implement any proposals for $\{(\mu_j, \sigma_j, W_j)\}$.
 Rather, these can be done trivially as follows:
+%
 \begin{CodeChunk}
@@
 \end{CodeChunk}
-
-The \code{RJObject} constructor definition is:
+%
+The `\code{RJObject}' constructor definition is:
+%
 \begin{CodeChunk}
@@
 \end{CodeChunk}
+%
 where \code{num\_dimensions} is the number of parameters needed to
@@
 and \code{prior\_type} controls the prior for $N$ (the default is uniform
-from $\{0, 1, 2, ..., N_{\rm max}\}$). The {\tt RJObject\_1DMixture}
-\code{MyModel} initializes its {\tt RJObject} as follows:
+from $\{0, 1, 2, \ldots, N_{\rm max}\}$). The \code{RJObject\_1DMixture}
+\code{MyModel} initializes its `\code{RJObject}' as follows:
+%
 \begin{CodeChunk}
@@
 \end{CodeChunk}
+%
 Passing \code{PriorType::log_uniform} for the final argument
 specifies the $1/(N+1)$ prior for $N$.
-One complication for this model is that \code{RJObject}, by default,
+One complication for this model is that `\code{RJObject}', by default,
 allows $N$ to be zero, which makes no sense for this particular
@@
 rejected if proposed in \code{MyModel::perturb}:
-
+%
 \begin{CodeChunk}
@@
 \end{CodeChunk}
-
-To access the component parameters, the \code{RJObject} member function
+%
+To access the component parameters, the `\code{RJObject}' member function
 \code{get_components} is used. This was used in the above functions, but more
-typically it is needed in the log likelihood.
+typically it is needed in the log-likelihood.
 The member function \code{get_components} returns (by const reference) a
@@
 parameter two of component zero is:
+%
 \begin{CodeChunk}
@@
 \end{CodeChunk}
+%
 The order of $\{(\mu_j, \sigma_j, w_j)\}$
@@
 
-\code{RJObject}'s \code{print} function prints the dimensionality of each
-component and the maximum value of $N$, followed by the hyperparameters,
-then parameter zero of each component (zero padded when $N<N_{\rm max}$),
-parameter one for each component, and so on.
+The \code{print} function for `\code{RJObject}' objects prints the
+dimensionality of each component and the maximum value of $N$,
+followed by the hyperparameters, then parameter zero of each component
+(zero padded when $N<N_{\rm max}$), parameter one for each component,
+and so on.
 
 \subsection{Conditional priors}
-The \code{RJObject} class is used to manage the components. An additional
-class is needed to define and manage the hyperparameters $\alpha$ (i.e.,
-define how they are generated from the prior, how they are proposed, and
-so on). The iid
-conditional prior $p(x_i | \alpha)$ is also defined by this extra class.
-In the example, the class is called \code{MyConditionalPrior} and
-is inherited from an abstract base class \code{DNest4::ConditionalPrior}
+The `\code{RJObject}' class is used to manage the components. An
+additional class is needed to define and manage the hyperparameters
+$\alpha$ (i.e., define how they are generated from the prior, how they
+are proposed, and so on). The i.i.d.~conditional prior
+$p(x_i | \alpha)$ is also defined by this extra class.  In the
+example, the class is called `\code{MyConditionalPrior}' and is
+inherited from an abstract base class `\code{DNest4::ConditionalPrior}'
 which insists that certain member functions be specified.
@@
 To simplify this, we have defined a class for Laplace distributions,
-rather than explicitly writing out the densities. We decided (arbitarily)
+rather than explicitly writing out the densities. We decided (arbitrarily)
 that parameters 0,1, and 2 are $\mu$, $\ln(\sigma)$, and $W$ respectively.
+%
 \begin{CodeChunk}
@@
 \end{CodeChunk}
-
+%
 The functions \code{to\_uniform} and \code{from\_uniform} must implement
 the cumulative distribution function (CDF) of the conditional
-prior and its inverse. i.e., \code{from\_uniform}, if the input vector contains
-iid draws from Uniform(0, 1), these should be modified to become draws from
+prior and its inverse, i.e., \code{from\_uniform}, if the input vector contains
+i.i.d.~draws from Uniform(0, 1), these should be modified to become draws from
 $p(x|\alpha)$, and \code{to\_uniform} should be the inverse of
@@
 for the Laplace distribution:
-
+%
 \begin{CodeChunk}
@@
 \end{CodeChunk}
-
+%
 \subsection{Mixture model results}
 The data and the posterior mean fit are shown
-in Figure~\ref{fig:galaxies} and the posterior distribution for
-$N$ is shown in Figure~\ref{fig:galaxies_N}.
+in Figure~\ref{fig:galaxies} on the left and the posterior distribution for
+$N$ is shown in Figure~\ref{fig:galaxies} on the right.
 The marginal likelihood was $\ln(Z) = -232.2$, and
@@
 
-\begin{figure}
-\begin{minipage}{0.45\textwidth}
+\begin{figure}[t!]
 \centering
-\includegraphics[width=0.95\textwidth]{figures/galaxies.pdf}
-\caption{The `galaxy' data, with the posterior mean fit
-(equivalent to the predictive distribution for the ``next'' data point).\label{fig:galaxies}}
-\end{minipage} \hspace{0.05\textwidth}
-\begin{minipage}{0.45\textwidth}
+\includegraphics[width=0.5\textwidth, trim = 0 5 20 15]{galaxies.pdf}%
+\includegraphics[width=0.5\textwidth, trim = 0 5 20 15]{galaxies_N.pdf}
+\caption{Left: The galaxy data, with the posterior mean fit
+  (equivalent to the predictive distribution for the ``next'' data point).
+  Right: The posterior distribution for $N$ given the
+galaxy data.
+\label{fig:galaxies}}
+\end{figure}
+
+\begin{figure}[t!]
 \centering
-\includegraphics[width=0.95\textwidth]{figures/galaxies_N.pdf}
-\caption{The posterior distribution for $N$ given the
-galaxy data.\label{fig:galaxies_N}}
-\end{minipage}
-\end{figure}
-
-\begin{figure}[ht!]
-\centering
-\includegraphics[width=0.7\textwidth]{figures/galaxies_fig3.pdf}
+\includegraphics[width=0.7\textwidth, trim = 0 20 0 25]{galaxies_fig3.pdf}
 \caption{The log-likelihood and posterior weights as a function
 of compression $X$ for the galaxy data. There are phase changes
-\citep{skilling2006nested} for which Nested Sampling is the best known
+\citep{skilling2006nested} for which nested sampling is the best known
 solution. There are some points with high posterior weight at the left
@@
 \subsection{Background}
-Nested Sampling can be used to solve Approximate Bayesian Computation (ABC)
+Nested sampling can be used to solve approximate Bayesian computation (ABC)
 problems elegantly and efficiently.
@@
 the user must be able to cheaply
-generate simulated data sets from the sampling distribution $p(D|\theta, M)$.
+generate simulated datasets from the sampling distribution $p(D|\theta, M)$.
 Since a sampling distribution must be specified, it is not that there is
 no likelihood function (indeed, a sampling distribution and
-a data set imply a particular likelihood function), rather that we cannot
+a dataset imply a particular likelihood function), rather that we cannot
 evaluate it cheaply and therefore cannot use MCMC.
@@
 All Bayesian updating conditions on the truth of a proposition. We sometimes
-speak and use notation as if we're conditioning on the value of a variable, for
+speak and use notation as if we are conditioning on the value of a variable, for
 example by writing the posterior distribution as
@@
 data
-\begin{align}
+\begin{align*}
 p(\theta, D | M) &= p(\theta | M)p(D | \theta, M)
-\end{align}
+\end{align*}
 and rejecting samples for which $D \neq D_{\rm observed}$, so that the
@@
 the real one $D_{\rm observed}$:
-\begin{align}
+\begin{align*}
 \rho\left(D; D_{\rm observed}\right).
-\end{align}
+\end{align*}
 The discrepancy function should take a minimum value of zero when
@@
 conditional on the logically weaker proposition
-\begin{align}
-\rho\left(D; D_{\rm observed}\right) < \epsilon
-\end{align}
+\begin{align*}
+\rho\left(D; D_{\rm observed}\right) < \epsilon,
+\end{align*}
 where $\epsilon$ is some small number. With this proposition, the
@@
 Typically, $\rho$ is defined by introducing a few {\em summary statistics}
-$s_1(D), s_2(D), ..., s_n(D)$, which hopefully capture most of the relevant
-information in the data, and then computing the Euclidian distance
+$s_1(D), s_2(D), \ldots, s_n(D)$, which hopefully capture most of the relevant
+information in the data, and then computing the Euclidean distance
 between the summary statistics of $D$ and $D_{\rm observed}$.
 That is,
-\begin{align}
+\begin{align*}
 \rho(D; D_{\rm observed}) &=
 \sqrt{\sum_i \left[s_i(D) - s_i(D_{\rm observed})\right]^2}.
-\end{align}
+\end{align*}
 
@@
 On the other hand, challenges (ii) and (iii) are about the computational
-implementation. Most ABC analyses are done using Sequential Monte Carlo
+implementation. Most ABC analyses are done using sequential Monte Carlo
 \citep[SMC; ][]{del2012adaptive}, another family of algorithms closely related
-to Nested Sampling and annealing in the sense that they work with
+to nested sampling and annealing in the sense that they work with
 sequences of probability distributions.
@@
 \subsection{ABC with nested sampling}
-NS can be used for ABC by noting that Equation~\ref{eqn:abc_posterior}
+Nested sampling can be used for ABC by noting that Equation~\ref{eqn:abc_posterior}
 is of the same form as a constrained prior, only instead of being the prior
@@
 manner consistent with the joint prior. The \code{log\_likelihood} function
-of the class, instead of actually evaluating the log likelihood, should
+of the class, instead of actually evaluating the log-likelihood, should
 evaluate $-\rho(D; D_{\rm observed})$. Then, running the sampler
@@
 Equation~\ref{eqn:abc_posterior} with $\epsilon$ corresponding to minus
-the ``log likelihood'' of the level.
+the ``log-likelihood'' of the level.
 A single \pkg{DNest4} run allows you to test sensitivity
@@
 A simple ABC example is included in the directory
-{\tt code/Examples/ABC}. The example tries to infer the mean
+\code{code/Examples/ABC}. The example tries to infer the mean
 $\mu$ and standard deviation $\sigma$ of a normal distribution
 from a set of samples. The sampling distribution is
-\begin{align}
+\begin{align*}
 x_i &\sim \textnormal{Normal}(\mu, \sigma^2).
-\end{align}
+\end{align*}
 and the prior is
-\begin{align}
-\mu &\sim \textnormal{Uniform}(-10, 10)\\
+\begin{align*}
+\mu &\sim \textnormal{Uniform}(-10, 10),\\
 \ln\sigma &\sim \textnormal{Uniform}(-10, 10).
-\end{align}
+\end{align*}
 However, instead of conditioning on the full dataset $\{x_i\}$, the example
@@
 be evaluated. It is included to demonstrate the principle of using
-Nested Sampling as an ABC method.}.
+nested sampling as an ABC method.}.
 
@@
 parameters and data highly dependent
---- otherwise the data wouldn't ever be informative about the parameters.
+-- otherwise the data would never be informative about the parameters.
 To improve efficiency, it is helpful to use a change of variables such that
@@
 example we use
-
-\begin{align}
-\mu &\sim \textnormal{Uniform}(-10, 10)\\
-\ln\sigma &\sim \textnormal{Uniform}(-10, 10)\\
-n_i &\sim \textnormal{Normal}(0, 1)
-\end{align}
+\begin{align*}
+\mu &\sim \textnormal{Uniform}(-10, 10),\\
+\ln\sigma &\sim \textnormal{Uniform}(-10, 10),\\
+n_i &\sim \textnormal{Normal}(0, 1),
+\end{align*}
 where each of the $n_i$ enables a data point to be calculated using
-\begin{align}
+\begin{align*}
 x_i &:= \mu + \sigma n_i.
-\end{align}
+\end{align*}
 A general procedure for achieving this is to imagine generating a
-simulated data set from the parameters. In this process, you would need
+simulated dataset from the parameters. In this process, you would need
 to call a random number generator many times.
@@
 are independent of the parameters and
-can be used to parameterize the data set. In the example,
+can be used to parameterize the dataset. In the example,
 the $\{n_i\}$ variables play this role.
 Therefore, the variables in the model class are:
+%
 \begin{CodeChunk}
@@
 \end{CodeChunk}
-
+%
 The proposal involves changing either \code{mu}, \code{log\_sigma}, or
 one of the \code{n}s:
+%
 \begin{CodeChunk}
@@
 \end{CodeChunk}
-
+%
 The ``log-likelihood'' is really minus the discrepancy function.
-Since we parameterized joint (parameter, data) space using the
+Since we parameterized the joint (parameter, data) space using the
 $n$ variables instead of the data $\{x_i\}$ itself, we must
@@
 discrepancy function:
+%
 \begin{CodeChunk}
@@
 \end{CodeChunk}
-
+%
 \subsection{ABC example results}
@@
 This generates posterior samples using the $\epsilon$ value corresponding
-to level $\approx$ 3/4$\times$ (maximum number of levels).
+to level $\approx 3/4\times$~(maximum number of levels).
+%
 \begin{CodeChunk}
@@
 \end{CodeChunk}
+%
 To try different values of $\epsilon$, you can set the argument
 \code{threshold\_fraction} to a value other than 0.8. For example
+%
 \begin{CodeChunk}
@@
 \end{CodeChunk}
-
+%
 In the example directory there is a dataset generated from
@@
 We show posterior samples in Figure~\ref{fig:abc_results}.
-\begin{figure}[ht!]
+\begin{figure}[t!]
 \centering
-\includegraphics[width=0.6\textwidth]{figures/abc_results.pdf}
+\includegraphics[width=0.6\textwidth, trim = 0 5 0 25]{abc_results.pdf}
 \caption{Posterior samples for $\mu$ and $\sigma$ for the
@@
 
-A Model class implementing the straight line example is given below
-and included in {\tt python/examples/straightline/straightline.py}.
+A model class implementing the straight line example is given below
+and included in the script \code{python/examples/straightline/straightline.py}.
 The member functions of this class have similar names to those
-used in C++, but work slightly differently. Most notably,
+used in \proglang{C++}, but work slightly differently. Most notably,
 an instance of the class does not represent a point in parameter
@@
 \pkg{NumPy} arrays of parameters as input, or produce them as
-output, rather than having parameters as data members as in C++.
-
+output, rather than having parameters as data members as in \proglang{C++}.
+%
 \begin{CodeChunk}
@@
                 - 0.5 * np.sum((data[:,1] - (m * data[:,0] + b)) **2) / var
-
-\end{CodeInput}
-\end{CodeChunk}
-
+\end{CodeInput}
+\end{CodeChunk}
+%
 A sampler can then be created and run using the following code.
-The \proglang{Python} sampler does not load an {\tt OPTIONS} file.
+The \proglang{Python} sampler does not load an \code{OPTIONS} file.
 Instead, the options are provided as function arguments
 when setting up the sampler.
+%
 \begin{CodeChunk}
@@
 \end{CodeChunk}
-
+%
 The text file output can be analyzed using \code{dnest4.postprocess()}
@@
 
-\section*{Acknowledgements}
+\section*{Acknowledgments}
 John Veitch (Birmingham)
@@
 
-\bibliography{references}
-
+\bibliography{ref}
+
+\newpage
 \appendix
@@
 Run on the specified number of threads. The default is 1.
-Be aware that the number of particles specified in the {\tt OPTIONS}
+Be aware that the number of particles specified in the \code{OPTIONS}
 file is actually the number of particles {\em per thread}. Therefore,
@@
 \proglang{Julia} language \citep{julia}
-is provided, with an example given in the
-{\tt code/Templates/JuliaModel} directory.
-Set an enviroment variable {\tt JULIA\_LIB\_PATH} to the directory on your
-system that contains the libraries {\tt libjulia.so} and
-{\tt sys.so}. Set another environment variable {\tt JULIA\_INCLUDE\_PATH}
+is provided. Directory
+\code{code/Templates/JuliaModel} contains an example.
+Set an environment variable \code{JULIA\_LIB\_PATH} to the directory on your
+system that contains the libraries \code{libjulia.so} and
+\code{sys.so}. Set another environment variable \code{JULIA\_INCLUDE\_PATH}
 to the directory where the \proglang{Julia} header files are located
-(do not include the final {\tt julia} directory in this path).
-
-The log likelihood is specified in {\tt julia\_model.jl}, as a function
-called {\tt log\_likelihood} that takes a {\tt Vector::Float64} as an argument
-and returns a {\tt Float64}. You should assume that the priors for the
+(do not include the final \code{julia} directory in this path).
+
+The log-likelihood is specified in \code{julia\_model.jl}, as a function
+called \code{log\_likelihood} that takes a \code{Vector::Float64} as an argument
+and returns a \code{Float64}. You should assume that the priors for the
 parameters are independent Uniform(0, 1) distributions (sensible default
 proposals are used). To use a different prior, transform the parameters inside
-your {\tt log\_likelihood} function before using them.
-You may add other functions to {\tt julia\_model.jl}, for example to load
-a dataset to refer to in your {\tt log\_likelihood}. The example in
+your \code{log\_likelihood} function before using them.
+You may add other functions to \code{julia\_model.jl}, for example to load
+a dataset to refer to in your \code{log\_likelihood}. The example in
 the repository is the ``SpikeSlab'' likelihood from
@@
 
-You can compile the model by invoking {\tt make}
-from the {\tt JuliaModel} directory.
+You can compile the model by invoking \code{make}
+from the \code{JuliaModel} directory.
 Only a single thread is supported when running \proglang{Julia}

### 2018-08-07 paper/ms.tex
Working on the changes

@@
   associated utilities including: i) `\code{RJObject}', a class template
-  for finite mixture models; (ii) a \proglang{Python} package allowing
-  basic use without \proglang{C++} coding; and iii) experimental
-  support for models implemented in \proglang{Julia}.  In this paper
+  for finite mixture models; and (ii) a \proglang{Python} package allowing
+  basic use without \proglang{C++} coding.  In this paper
   we demonstrate \pkg{DNest4} usage through examples including simple
@@
 In the nested sampling sphere,
-the simple \proglang{C} implementation of given by
+the simple \proglang{C} implementation given by
 \citet{skilling2006nested} is useful for understanding the classic nested sampling
@@
 
-\section[Specifying models in Julia]{Specifying models in \proglang{Julia}}
-Experimental support for model specification in the
-\proglang{Julia} language \citep{julia}
-is provided. Directory
-\code{code/Templates/JuliaModel} contains an example.
-Set an environment variable \code{JULIA\_LIB\_PATH} to the directory on your
-system that contains the libraries \code{libjulia.so} and
-\code{sys.so}. Set another environment variable \code{JULIA\_INCLUDE\_PATH}
-to the directory where the \proglang{Julia} header files are located
-(do not include the final \code{julia} directory in this path).
-
-The log-likelihood is specified in \code{julia\_model.jl}, as a function
-called \code{log\_likelihood} that takes a \code{Vector::Float64} as an argument
-and returns a \code{Float64}. You should assume that the priors for the
-parameters are independent Uniform(0, 1) distributions (sensible default
-proposals are used). To use a different prior, transform the parameters inside
-your \code{log\_likelihood} function before using them.
-You may add other functions to \code{julia\_model.jl}, for example to load
-a dataset to refer to in your \code{log\_likelihood}. The example in
-the repository is the ``SpikeSlab'' likelihood from
-\citet{brewer2011diffusive}.
-
-You can compile the model by invoking \code{make}
-from the \code{JuliaModel} directory.
-Only a single thread is supported when running \proglang{Julia}
-models.
+%\section[Specifying models in Julia]{Specifying models in \proglang{Julia}}
+%Experimental support for model specification in the
+%\proglang{Julia} language \citep{julia}
+%is provided. Directory
+%\code{code/Templates/JuliaModel} contains an example.
+%Set an environment variable \code{JULIA\_LIB\_PATH} to the directory on your
+%system that contains the libraries \code{libjulia.so} and
+%\code{sys.so}. Set another environment variable \code{JULIA\_INCLUDE\_PATH}
+%to the directory where the \proglang{Julia} header files are located
+%(do not include the final \code{julia} directory in this path).
+
+%The log-likelihood is specified in \code{julia\_model.jl}, as a function
+%called \code{log\_likelihood} that takes a \code{Vector::Float64} as an argument
+%and returns a \code{Float64}. You should assume that the priors for the
+%parameters are independent Uniform(0, 1) distributions (sensible default
+%proposals are used). To use a different prior, transform the parameters inside
+%your \code{log\_likelihood} function before using them.
+%You may add other functions to \code{julia\_model.jl}, for example to load
+%a dataset to refer to in your \code{log\_likelihood}. The example in
+%the repository is the ``SpikeSlab'' likelihood from
+%\citet{brewer2011diffusive}.
+
+%You can compile the model by invoking \code{make}
+%from the \code{JuliaModel} directory.
+%Only a single thread is supported when running \proglang{Julia}
+%models.
 

### 2018-08-07 paper/ms.tex
Mention p vs. P

@@
 a joint posterior probability distribution for the values of unknown
-parameters.
-
-In compact but standard notation, the posterior distribution for
+parameters. Throughout this paper we denote probability distributions
+using a lower case $p$ and probabilities with an upper case $P$.
+
+In a standard Bayesian inference situation, the posterior distribution for
 parameters $\params$ given data $\data$, within the context of prior

### 2018-08-07 python/dnest4/classic.py
Preparing replication stuff

@@ -18,7 +18,11 @@ def logdiffexp(x1, x2):
 
 def postprocess(temperature=1., numResampleLogX=1, plot=True, loaded=[], \
 			cut=0., save=True, zoom_in=True, compression_bias_min=1., verbose=True,\
-			compression_scatter=0., moreSamples=1., compression_assert=None, single_precision=False):
+			compression_scatter=0., moreSamples=1., compression_assert=None, single_precision=False, rng_seed=None):
+
+	if rng_seed is not None:
+		rng.seed(rng_seed)
+
 	if len(loaded) == 0:
 		levels_orig = np.atleast_2d(my_loadtxt("levels.txt"))
 		sample_info = np.atleast_2d(my_loadtxt("sample_info.txt"))

### 2018-08-07 code/Examples/ABC/display.py
Working on replication - still needs new figures and numbers for galaxyfield and abc

@@ -1,0 +1,20 @@
+import matplotlib.pyplot as plt
+import numpy as np
+import dnest4.classic as dn4
+
+plt.rcParams["font.family"] = "serif"
+plt.rcParams["font.size"] = 16
+plt.rc("text", usetex=True)
+
+posterior_sample = np.atleast_2d(dn4.my_loadtxt("posterior_sample.txt"))
+plt.plot(posterior_sample[:,0], np.exp(posterior_sample[:,1]),\
+                    "ko", alpha=0.1, markersize=5, label="Posterior samples")
+plt.plot(0, 1, "r*", markersize=20, label="Truth", alpha=0.5)
+plt.xlabel("$\\mu$")
+plt.ylabel("$\\sigma$")
+plt.legend(loc="upper left", numpoints=1)
+plt.title("ABC Results")
+plt.axis([-1.5, 1.5, 0.5, 1.5])
+plt.savefig("abc_results.pdf", bbox_inches="tight")
+plt.show()
+

### 2018-08-07 code/Examples/ABC/showresults.py

@@ -3,18 +3,3 @@ import matplotlib.pyplot as plt
 import dnest4.classic as dn4
 dn4.postprocess_abc()
 
-plt.rc("font", size=18, family="serif", serif="Computer Sans")
-plt.rc("text", usetex=True)
-
-posterior_sample = np.atleast_2d(dn4.my_loadtxt("posterior_sample.txt"))
-plt.plot(posterior_sample[:,0], np.exp(posterior_sample[:,1]),\
-                    "ko", alpha=0.1, markersize=5, label="Posterior samples")
-plt.plot(0, 1, "r*", markersize=20, label="Truth", alpha=0.5)
-plt.xlabel("$\\mu$")
-plt.ylabel("$\\sigma$")
-plt.legend(loc="upper left", numpoints=1)
-plt.title("ABC Results")
-plt.axis([-1.5, 1.5, 0.5, 1.5])
-plt.savefig("abc_results.pdf", bbox_inches="tight")
-plt.show()
-

### 2018-08-07 paper/replicate.py

@@ -1,11 +1,8 @@
+import dnest4.classic as dn4
 import os
 import matplotlib.pyplot
 
-# Best Python command on your system
-# (should also be the one you installed the dnest4 package for)
-my_python = "python"
-
-print("This script should\
+print("This script should \
 replicate all the runs presented in the paper.\nNote:\
 plots will be generated, which need to be manually\n\
 closed for the script to continue. It assumes\n\
@@ -15,16 +12,22 @@ NOTE: the RJObject_1DMixture example is fairly time-consuming.\n\n\
 Press ENTER to begin.")
 junk = input()
 
-def run_example(directory):
+def run_example(directory, abc=False):
     os.chdir(directory)
     os.system("./main -s 0")
     if directory == "StraightLine":
         print("Figure 1 in the paper has been zoomed in.")
-    os.system(my_python + " showresults.py")
+
+    if abc:
+        dn4.postprocess_abc(rng_seed=0)
+    else:
+        dn4.postprocess(rng_seed=0)
+
+    exec(open("display.py").read())
     os.chdir("..")
 
 os.chdir("../code/Examples")
 run_example("StraightLine")
 run_example("RJObject_1DMixture")
-run_example("ABC")
+run_example("ABC", abc=True)
 

### 2018-08-07 python/dnest4/classic.py

@@ -1,5 +1,6 @@
 import copy
 import numpy as np
+import numpy.random as rng
 from .loading import *
 
 def logsumexp(values):
@@ -231,7 +232,11 @@ def postprocess(temperature=1., numResampleLogX=1, plot=True, loaded=[], \
 
 def postprocess_abc(temperature=1., numResampleLogX=1, plot=True, loaded=[], \
 			cut=0., save=True, zoom_in=True, compression_bias_min=1., verbose=True,\
-compression_scatter=0., moreSamples=1., compression_assert=None, single_precision=False, threshold_fraction=0.8):
+compression_scatter=0., moreSamples=1., compression_assert=None, single_precision=False, threshold_fraction=0.8,rng_seed=None):
+
+	if rng_seed is not None:
+		rng.seed(rng_seed)
+
 	if len(loaded) == 0:
 		levels_orig = np.atleast_2d(my_loadtxt("levels.txt"))
 		sample_info = np.atleast_2d(my_loadtxt("sample_info.txt"))

### 2018-08-07 paper/ms.tex
Working on replication - still needs new figures and numbers for galaxyfield and abc
Numba

@@
 packages \pkg{NumPy} \citep{numpy}, \pkg{matplotlib} \citep{matplotlib},
+\pkg{numba} \citep{numba},
 and \pkg{Cython} \citep{cython} are also needed.
@@
 \begin{CodeOutput}
-log(Z) = -175.512920079
-Information = 15.1683473499 nats.
-Effective sample size = 1521.58236273
+log(Z) = -175.49577676640516
+Information = 15.17486407892028 nats.
+Effective sample size = 1126.5312086470944
 \end{CodeOutput}

### 2018-08-07 code/Examples/RJObject_1DMixture/display.py
Scripts working

@@ -1,8 +1,8 @@
 from pylab import *
-import dnest4.classic as dn4
 
-rc("font", size=18, family="serif", serif="Computer Sans")
-rc("text", usetex=True)
+plt.rcParams["font.family"] = "serif"
+plt.rcParams["font.size"] = 16
+plt.rc("text", usetex=True)
 
 data = loadtxt('galaxies.txt')
 posterior_sample = atleast_2d(dn4.my_loadtxt('posterior_sample.txt'))

### 2018-08-07 paper/replicate.py

@@ -1,6 +1,7 @@
 import dnest4.classic as dn4
 import os
 import matplotlib.pyplot
+from pylab import *
 
 print("This script should \
 replicate all the runs presented in the paper.\nNote:\

### 2018-08-07 paper/ms.tex
Use tag 0.2.3 (doesn't exist yet)

@@
 packages \pkg{NumPy} \citep{numpy}, \pkg{matplotlib} \citep{matplotlib},
-\pkg{numba} \citep{numba},
+\pkg{Numba} \citep{numba},
 and \pkg{Cython} \citep{cython} are also needed.
@@
 \begin{CodeInput}
-$ wget https://github.com/eggplantbren/DNest4/archive/0.1.4.tar.gz
-$ tar xvzf 0.1.4.tar.gz
-$ mv DNest4-0.1.4 DNest4
+$ wget https://github.com/eggplantbren/DNest4/archive/0.2.3.tar.gz
+$ tar xvzf 0.2.3.tar.gz
+$ mv DNest4-0.2.3 DNest4
 $ cd DNest4/code

### 2018-08-07 paper/ms.tex
Contributors

@@
 \usepackage{thumbpdf,lmodern} 
+\usepackage[utf8]{inputenc}
 \graphicspath{{Figures/}}
@@
 \section*{Acknowledgments}
+Other contributors to \pkg{DNest4} include
+João Faria (Porto), Matt Pitkin (Glasgow), and Mat Varidel (Sydney).
 John Veitch (Birmingham)
@@
 and Iain Murray (Edinburgh)
-for valuable discussions. This work was supported by a Marsden Fast-Start grant
+for valuable discussions over the years.
+This work was supported by a Marsden Fast-Start grant
 from the Royal Society of New Zealand.

### 2018-08-08 paper/ms.tex
New replicate.py and correspondingly updated results

@@
 \begin{CodeOutput}
-log(Z) = -175.49577676640516
-Information = 15.17486407892028 nats.
-Effective sample size = 1126.5312086470944
+log(Z) = -175.47872682863076
+Information = 15.166340875399044 nats.
+Effective sample size = 1251.3576752091165
 \end{CodeOutput}
@@
 $N$ is shown in Figure~\ref{fig:galaxies} on the right.
-The marginal likelihood was $\ln(Z) = -232.2$, and
-the information (measured in nats) was $\mathcal{H} = 29.5$.
+The marginal likelihood was $\ln(Z) = -232.1$, and
+the information was $\mathcal{H} = 29.5$ nats.
 
@@
 This generates posterior samples using the $\epsilon$ value corresponding
-to level $\approx 3/4\times$~(maximum number of levels).
+to level $\approx 0.8\times$~(maximum number of levels).
 %

### 2018-08-08 paper/replicate.py

@@ -15,7 +15,7 @@ junk = input()
 
 def run_example(directory, abc=False):
     os.chdir(directory)
-    os.system("./main -s 0")
+    os.system("./main -s 0 -t 8")
     if directory == "StraightLine":
         print("Figure 1 in the paper has been zoomed in.")
 

### 2018-08-08 code/Version.h
Version bump

@@ -3,7 +3,7 @@
 
 #define DNEST4_MAJOR_VERSION 0
 #define DNEST4_MINOR_VERSION 2
-#define DNEST4_PATCH_VERSION 2
+#define DNEST4_PATCH_VERSION 3
 
 #endif  // DNest4_Version
 

### 2018-08-08 python/dnest4/__init__.py

@@ -1,6 +1,6 @@
 # -*- coding: utf-8 -*-
 
-__version__ = "0.2.2"
+__version__ = "0.2.3"
 
 try:
     __DNEST4_SETUP__

### 2018-08-20 paper/ms.tex
Response

@@
 \newpage
-\section{Markov chain Monte Carlo}\label{sec:mcmc}
+\section{Markov chain monte carlo}\label{sec:mcmc}
 DNS is build upon the Metropolis-Hastings algorithm.
@@
 post-processing is required. This can be achieved by running the following
-\proglang{Python} function\footnote{Alternatively, the file
+\proglang{Python} function:
+%
+\begin{CodeChunk}
+\begin{CodeInput}
+>>> import dnest4
+>>> dnest4.postprocess()
+\end{CodeInput}
+\end{CodeChunk}
+
+Alternatively, the file
 \code{showresults.py} in the example directory runs this function, and then
 calls the code in \code{display.py} to create a further plot shown in
-Figure~\ref{fig:fig3} on the right.}:
-%
-\begin{CodeChunk}
-\begin{CodeInput}
->>> import dnest4
->>> dnest4.postprocess()
-\end{CodeInput}
-\end{CodeChunk}
+Figure~\ref{fig:fig3} on the right.
+
 %
@@
 template class expresses.
-Figure created using Daft (\url{http://daft-pgm.org}).\label{fig:rjobject_pgm}}
+Figure created using \pkg{Daft} \citep{daft}.\label{fig:rjobject_pgm}}
 \end{figure}
@@
 we used Laplace (biexponential)
-distributions\footnote{A Laplace
+distributions. A Laplace
 distribution with location parameter $a$ and scale parameter
-$b$ has density $p(x | a, b) = \frac{1}{2}\exp\left(-\frac{1}{b}|x - a|\right)$.}.
+$b$ has density $p(x | a, b) = \frac{1}{2}\exp\left(-\frac{1}{b}|x - a|\right)$.
 Normal distributions would be more conventional, but the analytic cumulative
@@
 $p(\theta | D = D_{\rm observed}, M)$, that is,
-the posterior\footnote{Incidentally, this usage of the {\em joint} posterior
+the posterior. Incidentally, this usage of the {\em joint} posterior
 distribution for the parameters {\em and the data} provides a bridge
 connecting Bayesian updating to maximum entropy updating
-\citep{caticha2006updating,giffin2007updating}.}.
+\citep{caticha2006updating,giffin2007updating}.
 
@@
 uses the minimum and maximum values in the data,
-$\min(\{x_i\})$ and $\max(\{x_i\})$, as summary statistics\footnote{This is
+$\min(\{x_i\})$ and $\max(\{x_i\})$, as summary statistics. This is
 not a serious application of ABC, since the likelihood function can easily
 be evaluated. It is included to demonstrate the principle of using
-nested sampling as an ABC method.}.
+nested sampling as an ABC method.
 

### 2025-08-29 talk/talk.tex
Some talk slides

\documentclass{beamer}
\usetheme[pageofpages=of,% String used between the current page and the
                         % total page count.
          bullet=circle,% Use circles instead of squares for bullets.
          titleline=true,% Show a line below the frame title.
          alternativetitlepage=true,% Use the fancy title page.
       %   titlepagelogo=logo-polito,% Logo for the first page.
       %   watermark=watermark-polito,% Watermark used in every page.
       %   watermarkheight=100px,% Height of the watermark.
       %   watermarkheightmult=4,% The watermark image is 4 times bigger
                                % than watermarkheight.
          ]{Torino}

\usecolortheme{dolphin} 
\usefonttheme{professionalfonts}

\setbeamertemplate{footline}{
  \begin{beamercolorbox}[wd=\paperwidth,ht=1ex,dp=1ex]{footline}
    \vspace{5pt} \hspace{1em} \insertframenumber/\inserttotalframenumber
  \end{beamercolorbox}
}

\author{Brendon J. Brewer}
\title{Diffusive Nested Sampling: Not Dead Yet}
\institute{Department of Statistics, The University of Auckland}
\date{}

\linespread{1.3}
\usepackage{minted}
\usepackage[utf8]{inputenc}
\usepackage{dsfont}

\newcommand{\given}{\,|\,}

\begin{document}

\frame{\titlepage}

\begin{frame}
\frametitle{Bayesian Inference}
Bayesian inference uses probability theory (usually probability distributions)
to describe uncertainty, not variability. Bayes' theorem is used frequently:
\pause
\begin{align}
p(\theta \given D, M) &= \frac{p(\theta \given M)p(D \given \theta, M)}{P(D \given M)}
\end{align}
\pause
where
\begin{align}
p(D \given M) &= \int p(\theta \given M)p(D \given \theta, M) \, d\theta
\end{align}

\end{frame}

\begin{frame}
\frametitle{Bayesian Inference}
The equations in words:

\begin{align}
\textnormal{posterior distribution} &= 
    \frac{\textnormal{prior distribution} \times \textnormal{likelihood function}}{\textnormal{marginal likelihood value}}
\end{align}
\pause
where
\begin{align}
\textnormal{marginal likelihood value} &= \textnormal{integral of prior times likelihood}.
\end{align}

\end{frame}

\begin{frame}
\frametitle{Bayesian Inference}
The choice of the {\bf prior distribution} and the {\bf likelihood} form the
inputs to the calculation. The former involves judgement calls, and the latter
involves judgement calls and the data.\\[0.5em]\pause

The {\bf posterior distribution} and
{\bf marginal likelihood value} are the outputs. But how can we calculate them
in general?

\end{frame}

\begin{frame}
\frametitle{Posterior Only}
Many Bayesians use Markov Chain Monte Carlo (MCMC) to generate {\bf samples}
from the posterior $p(\theta \given D, M)$, describing uncertainty about the
parameters that remains after incorporating the data.\\[0.5em]\pause

It is less common to compute the marginal likelihood, which is used for
model comparison purposes. E.g. for two competing models $M_1$ and $M_2$
the marginal likelihood acts as a likelihood:

\begin{align}
P(M_1 \given D) &= \frac{P(M_1)P(D \given M_1)}
                        {P(M_1)P(D \given M_1) + P(M_2)P(D \given M_2)}.
\end{align}

\end{frame}

\begin{frame}
\frametitle{Notation Switch}
For computational discussions:

\begin{align}
\pi(\theta) &= \textnormal{prior} \\
L(\theta) &= \textnormal{likelihood} \\
Z &= \textnormal{marginal likelihood}.
\end{align}

\end{frame}

\begin{frame}
\frametitle{Thermodynamic Integration}
An older technique to calculate $P(M\given D)$ is based on the following
result:
\begin{align}
\log Z &= \int_0^1 \left<\log L(\theta)\right>_\beta \, d\theta
\end{align}
\pause
where the expectation is taken with respect to `power posteriors'
proportional to $\pi(\theta)L(\theta)^\beta$. An `annealing' process can be done
where $\beta$ is gradually raised from 0 to 1.

\end{frame}

\end{document}

### 2025-08-29 talk/talk.tex
Edited slide

@@
 \frametitle{Thermodynamic Integration}
-An older technique to calculate $P(M\given D)$ is based on the following
+An older technique to calculate $Z$ is based on the following
 result:
@@
 where the expectation is taken with respect to `power posteriors'
-proportional to $\pi(\theta)L(\theta)^\beta$. An `annealing' process can be done
-where $\beta$ is gradually raised from 0 to 1.
+proportional to $\pi(\theta)L(\theta)^\beta$. \\[0.5em]\pause
+
+An `annealing' process can be done
+where $\beta$ is gradually raised from 0 to 1. As a side effect, difficult
+posteriors such as those with multiple peaks become a bit easier.
 

### 2025-08-29 talk/talk.tex
Phase changes

@@
 
+\begin{frame}
+\frametitle{Thermodynamic Integration}
+There are many methods based on the power posterior idea:
+
+\begin{itemize}
+\item Parallel tempering
+\item Stepping-stone sampling
+\item Sequential Monte Carlo
+\item Simulated Tempering
+\item More.
+\end{itemize}
+
+\end{frame}
+
+\begin{frame}
+\frametitle{Power Posteriors: Problems}
+There is a class of problems that defeats methods based on power posteriors.
+These are {\bf phase change problems}. In statistical physics they are common,
+but they also occur in Bayesian inference --- they involve a particular
+shape to the likelihood function.
+
+\end{frame}
+
+\begin{frame}
+\frametitle{Phase Change Problems}
+
+Let's explore how thermodynamic integration fails on phase change problems.
+
+\end{frame}
+
 \end{document}

### 2025-08-29 talk/plots.py
A figure

@@ -1,0 +1,27 @@
+import numpy as np
+import numpy.random as rng
+import matplotlib.pyplot as plt
+
+rng.seed(0)
+
+plt.rcParams.update({
+    "text.usetex": True,
+    "font.size": 14,
+})
+
+
+theta = np.linspace(-5.0, 5.0, 1001)
+prior = 1.0/(1.0 + (theta - 1.0)**2)
+prior = prior/np.trapz(prior, x=theta)
+likelihood = 0.3*np.exp(-0.5*(theta + 1.5)**2)
+posterior = prior*likelihood
+posterior /= np.trapz(posterior, x=theta)
+
+plt.plot(theta, prior, color="black", linewidth=2, label="Prior")
+plt.plot(theta, likelihood, color="blue", alpha=0.5, label="Likelihood")
+plt.plot(theta, posterior, color="red", linewidth=2, label="Posterior")
+plt.legend()
+plt.xlabel("Unknown Parameter $\\theta$")
+plt.ylim(0.0)
+plt.savefig("bayes.pdf", bbox_inches="tight")
+

### 2025-08-29 code/Hypercube.hpp
Hypercube class

@@ -1,0 +1,70 @@
+#ifndef DNest4_Hypercube_h
+#define DNest4_Hypercube_h
+
+#include "RNG.h"
+#include "Utils.h"
+#include <vector>
+
+namespace DNest4
+{
+
+// Base class for hypercube-based models.
+template<int size, class T>
+class Hypercube
+{
+    private:
+        std::vector<double> us, xs;
+
+    public:
+        Hypercube();
+        void from_prior(DNest4::RNG& rng);
+        double perturb(DNest4::RNG& rng);
+};
+
+
+/* Implementations follow */
+
+template<int size, class T>
+Hypercube<size, T>::Hypercube()
+:us(size)
+,xs(size)
+{
+
+}
+
+template<int size, class T>
+void Hypercube<size, T>::from_prior(DNest4::RNG& rng)
+{
+    for(double& u: us)
+        u = rng.rand();
+
+    static_cast<T*>(this)->prior_transform();
+}
+
+
+template<int size, class T>
+double Hypercube<size, T>::perturb(DNest4::RNG& rng)
+{
+    int reps;
+    if(rng.rand() <= 0.5)
+        reps = 1;
+    else
+        reps = (int)pow(size, rng.rand());
+
+    for(int i=0; i<reps; ++i)
+    {
+        int k = rng.rand_int(us.size());
+        us[k] += rng.randh();
+        DNest4::wrap(us[k], 0.0, 1.0);
+    }
+
+    static_cast<T*>(this)->prior_transform();
+
+    return 0.0;
+}
+
+
+} // namespace
+
+#endif
+

### 2025-08-29 code/DNest4.h
Hypercube template

@@ -7,6 +7,7 @@
 
 #include "Version.h"
 #include "Barrier.h"
+#include "Hypercube.hpp"
 #include "CommandLineOptions.h"
 #include "Level.h"
 #include "LikelihoodType.h"

### 2025-08-29 code/Hypercube.hpp

@@ -12,7 +12,7 @@ namespace DNest4
 template<int size, class T>
 class Hypercube
 {
-    private:
+    protected:
         std::vector<double> us, xs;
 
     public:

### 2025-08-29 code/Templates/HypercubeModel/MyModel.cpp

@@ -1,0 +1,29 @@
+#include "MyModel.h"
+#include "DNest4/code/DNest4.h"
+
+MyModel::MyModel()
+{
+
+}
+
+void MyModel::prior_transform()
+{
+    xs = us;
+}
+
+double MyModel::log_likelihood() const
+{
+    double logL = 0.0;
+    return logL;
+}
+
+void MyModel::print(std::ostream& out) const
+{
+
+}
+
+std::string MyModel::description() const
+{
+    return std::string("");
+}
+

### 2025-08-29 code/Templates/HypercubeModel/MyModel.h

@@ -1,0 +1,30 @@
+#ifndef DNest4_Template_MyModel
+#define DNest4_Template_MyModel
+
+#include "DNest4/code/DNest4.h"
+#include <ostream>
+
+class MyModel:public DNest4::Hypercube<5, MyModel>
+{
+    private:
+
+
+    public:
+        // Constructor only gives size of params
+        MyModel();
+
+        // Transform us to xs
+        void prior_transform();
+
+        // Likelihood function
+        double log_likelihood() const;
+
+        // Print to stream
+        void print(std::ostream& out) const;
+
+        // Return string with column information
+        std::string description() const;
+};
+
+#endif
+

### 2025-08-29 code/Templates/HypercubeModel/MyModel.cpp
Just to test it actually working

@@ -14,12 +14,17 @@ void MyModel::prior_transform()
 double MyModel::log_likelihood() const
 {
     double logL = 0.0;
+
+    for(double x: xs)
+        logL += -0.5*pow((x - 0.5)/0.01, 2);
+
     return logL;
 }
 
 void MyModel::print(std::ostream& out) const
 {
-
+    for(double x: xs)
+        out << x << ' ';
 }
 
 std::string MyModel::description() const

### 2025-11-23 talk/talk.tex
A figure
Integral is over beta
A few slides

@@
 \newcommand{\given}{\,|\,}
+\newcommand{\bx}{\boldsymbol{x}}
 
@@
 \end{align}
+
+\end{frame}
+
+\begin{frame}
+\frametitle{Bayesian Inference}
+
+\begin{center}
+\includegraphics[width=0.7\textwidth]{bayes.pdf}
+\end{center}
 
@@
 \begin{frame}
+\frametitle{Sampling Strategies}
+I divide MCMC (or similar) sampling strategies into two classes.
+The first category focuses on {\bf how to move around}:
+e.g., Metropolis, Gibbs sampling, Slice Sampling, Hamiltonian MCMC,
+Affine Invariant Ensemble Sampler.\\[0.5em]\pause
+
+The second category {\bf modifies the target distribution}.
+Examples: Parallel tempering (and other methods based on `annealing'
+or `power posteriors'), Nested Sampling, Diffusive Nested Sampling.
+
+\end{frame}
+
+\begin{frame}
 \frametitle{Thermodynamic Integration}
@@
 \begin{align}
-\log Z &= \int_0^1 \left<\log L(\theta)\right>_\beta \, d\theta
+\log Z &= \int_0^1 \left<\log L(\theta)\right>_\beta \, d\beta
 \end{align}
@@
 
+\begin{frame}
+\frametitle{Mixture of Two Gaussians}
+Taken from John Skilling's 2000s Nested Sampling papers.
+$\pi(\bx)$ is Uniform$(-1, 1)$ in 20 dimensions. $L(\bx)$ is a mixture
+of a broad and a narrow gaussian:
+
+\begin{align}
+L(\bx) &= \prod_{i=1}^{20}
+            \frac{\exp\left(-\frac{1}{2}(x_i/v)^2\right)}{v\sqrt{2\pi}}
+        + 100 \prod_{i=1}^{20}
+            \frac{\exp\left(-\frac{1}{2}(x_i/u)^2\right)}{u\sqrt{2\pi}}.
+\end{align}
+With $u=0.01$ and $v=0.1$ the narrow peak dominates the posterior (but not
+entirely).
+
+\end{frame}
+
 \end{document}
